### 2019-11-07 DoubleDriftResetLimits.Rmd
started project notebook

---
title: "Limits to Spontaneous Resets in the Double-Drift Illusion"
output: html_notebook
---

# Overview

State aims of the project. Say which experiments were done and are analyzed here.

# Setting up

Before starting any analyses, we need to make sure that we have all the code and data necessary, this will be set up by the chunks below

## Libraries

We've minimized depencies on other packages, but there are two that help - even when not strictly required.

```{r}
library('svglite')
library('ez')

# these packages are not mandatory
# - svglite generates SVG versions of the figures, but the native R figures might be enough
# - ez allows one ANOVA to be run, but this part could be skipped

# if you do want to install them, use this:
# install.packages(c('svglite','ez'))

```

## Functions

There are several files with code that will do the work for us in pre-processing, plotting and doing stats.

```{r}
# The functions here allow downloading the collected data from the OSF project:
source('R/data.R')
# This file has some common helper functions:

# The functions in here are used on the data of the first tracking experiment:
source('R/tracking_bounded.R')
# This code is for the second tracking experiment:
source('R/tracking_reset_pause.R')
# And finally some code for the re-tracin experiment...
source('R/retrace_onePass_V4.R')
# As well as models fit to that data:
source('R/models.R')
```

## Download data

A last step before starting actual analyses is to download the data from the Open Science Framework:

```{r}
# from R/data.R
downloadRawData(overwrite=FALSE, experiments=c('bounded_tracking', 'reset_pause', 'onePass_V4'))
```

If there were no errors, we're now ready to start.

# Experiment 1a: continuous tracking

In this experiment, 6 participants were instructed to look at a fixation point at the left or right side of the screen, while moving a tablet controlled cursor to the center of the screen. Subsequently, a double-drift stimulus would appear and then move on a straight, 13.5 cm path along the Y axis of the screen (not sideways, but forward/backward, or upward/downward if you like), starting in one of two directions at random. When the double-drift stimulus would reach either end of the path, both it's external and internal motion would be inverted. This way it would appear to move on a diagonal, with the angle of the diagonal depending on the direction and velocity of the internal motion. Participants were to use the stylus on the tablet to track the perceived position of the stimulus while it was moving. The stimulus would complete 3 full cycles (if one path is the 13.5 cm: a half path, five full paths, and another half path) in 12 seconds. That is: a single path takes 2 seconds. The internal motion could have one of 5 values -3, -1, 0, 1 or 3 cycles per second (the number of complete phases the gabor would cycle through in 1 second). Each of these 5 internal motion values was repeated 8 times (with random initial direction and fixation side) for a total of 40 trials.

Here we plot the trajectories recorded during this task (each subplot essentially shows 8 trials overlayed):

```{r fig.width=6, fig.height=9}
# from R/tracking_bounded.R:
plotBoundedTrackingRaw()
```

We can see that the orientation of the tracking movements responds to the strength and direction of the internal motion of the stimulus. However, there are no clear spontaneous resets of the stimulus.

# Experiment 1b: continuous tracking with pauses

Therefore, we did second experiment that mimicked the first, but with three changes, all designed to increase the occurrence of resets. First, the external motion was slowed down as earlier work showed that resets often occur after around 2 seconds, which was how long one pass (from one end of the screen to the other) took in the first experiment. Second, we only used -3 and 3 cycles per second for the internal motion. Third, on 3 of every 4 trials, a ~266.66 ms "blink" (8 frames at 30 Hz) was introduced in the stimulus. During this period the stimulus disappeared from the screen and reappeared after the blink in the same location to continue as before. The blinks occured equally often in the first, second or third full pass of the stimulus, and always halfway along the path on screen. These blinks have been shown to trigger resets [@Nakayama2018], so if we can spot triggered resets in the tracking data, we can use the characteristics of that tracking data to detect spontaneous resets. However, we would primarily expect to see increases in horizontal tracking speed, perhaps accompanied by a decrease in vertical speed, when people track the "jump" in the perceived location of the stimulus.

### 2019-11-07 R/tracking_bounded.R

@@ -5,7 +5,7 @@ library('svglite')
 
 plotBoundedTrackingRaw <- function(target='inline') {
   
-  participants <- c(1,3,5,6:8)
+  participants <- c(1,3,5,6,7,8)
   
   if (target=='svg') {
     svglite(file='doc/bounded_tracking_uncluttered.svg',width=6,height=9)

### 2019-11-20 R/common.R
added color scheme and CI function

@@ -1,0 +1,58 @@
+
+
+
+getColors <- function() {
+  
+  colors <- list()
+  
+  colors[['blue']]      <- list('s'='#005de4ff', 't'='#005de42f')
+  
+  colors[['lightblue']] <- list('s'='#0fd2e2ff', 't'='#0fd2e22f')
+  
+  colors[['yorkred']]   <- list('s'='#e51636ff', 't'='#e516362f')
+  
+  colors[['orange']]    <- list('s'='#ff8200ff', 't'='#ff82002f')
+  
+  colors[['purple']]    <- list('s'='#b400e4ff', 't'='#b400e42f')
+  
+  # colorset[['onlPasS']] <- '#8266f4ff' # violet
+  # colorset[['onlPasT']] <- '#8266ff2f'
+  
+  # colorset[['onlPasS']] <- '#ff6ec7ff' # pink
+  # colorset[['onlPasT']] <- '#ff6ec72f'
+  
+  return(colors)
+  
+}
+
+
+getConfidenceInterval <- function(data, variance = var(data), conf.level = 0.95, method='t-distr', resamples=1000, FUN=mean) {
+  
+  if (method %in% c('t-distr','t')) {
+    
+    z = qt((1 - conf.level)/2, df = length(data) - 1, lower.tail = FALSE)
+    
+    xbar = mean(data)
+    sdx = sqrt(variance/length(data))
+    
+    return(c(xbar - (z * sdx), xbar + (z * sdx)))
+    
+  }
+  
+  # add sample z-distribution?
+  
+  if (method %in% c('bootstrap','b')) {
+    
+    data <- data[which(is.finite(data))] #need is.finite due to NA values
+    
+    samplematrix <- matrix(sample(data, size = resamples*length(data), replace = TRUE), nrow = resamples)
+    BS <- apply(samplematrix, c(1), FUN=FUN) 
+    
+    lo <- (1-conf.level)/2.
+    hi <- 1 - lo
+    
+    return(quantile(BS, probs = c(lo,hi)))
+    
+  }
+  
+}

### 2019-11-20 R/data.R
add meta-function that calls all data pre-processing functions

@@ -1,5 +1,7 @@
 # check / download data -----
 
+source('R/tracking_reset_pause.R')
+
 downloadRawData <- function(overwrite=TRUE, experiments=c('bounded_tracking', 'onePass_V4', 'reset_pause')) {
   
   
@@ -124,5 +126,17 @@ downloadFiles <- function(files, folder='./', overwrite=TRUE) {
     }
     
   }
+  
+}
+
+doPreProcessing <- function() {
+  
+  # bounded tracking: no pre-processing
+  
+  # paused tracking:
+  getPauseTrajectories()
+  getReachDirections()
+  
+  # re-tracing:
   
 }

### 2019-11-20 R/tracking_reset_pause.R
added paused tracking pre-processing and extended plot with CIs, and no-pause trial data

@@ -1,49 +1,536 @@
 library('svglite')
+source('R/common.R')
 
-plotResetTrial <- function(participant,trial) {
-  
-  library(colormap)
-  
-  df <- read.csv(sprintf('data/reset_pause/reset_p%02d.csv', participant), stringsAsFactors=F)
-  df <- df[which(df$trial==trial & df$step==2),]
-  
-  par(mfrow=c(1,1))
-  
-  # mycolors <- colormap(colormap=colormaps$velocity_blue, nshades=256)
-  mycolors <- colormap(colormap=colormaps$autumn, nshades=256)
-  mycolors <- colormap(colormap=colormaps$plasma, nshades=256)
-  
-  velocity <- sqrt(diff(df$handx_pix)^2 + diff(df$handy_pix)^2) / diff(df$time_ms)
-  
-  segment_colors <- mycolors[floor(velocity/max(velocity, na.rm=T)*256)+1]
-  
-  plot(-1000,-1000,main=sprintf('participant: %d, trial: %d',participant,trial),xlab='',ylab='',xlim=c(-500,500),ylim=c(-750,750),asp=1)
-  
-  N <- nrow(df)
-  
-  X0 <- df$handx_pix[1:N-1]
-  Y0 <- df$handy_pix[1:N-1]
-  X1 <- df$handx_pix[2:N]
-  Y1 <- df$handy_pix[2:N]
-  
-  segments(x0=X0, y0=Y0, x1=X1, y1=Y1, col=segment_colors)
-  
-}
+# Data handling -----
 
-plotResetData <- function(participants=c(1,2,9,10,11,12,13),trials=c(1:64)) {
+getPauseTrajectories <- function() {
   
-  for (participant in participants) {
+  participants <- c(1,2,9,10,11,12,13)
+  
+  trajectories <- NA
+  
+  for (ppidx in c(1:length(participants))) {
     
-    pdf(file=sprintf('reset_p%02d.pdf',participant),width=6,height=4)
+    ppno <- participants[ppidx]
     
-    for (trial in trials) {
+    df <- read.csv(sprintf('data/reset_pause/reset_p%02d.csv',ppno))
+    
+    trials <- unique(df$trial_no)
+    
+    for (trialn in trials) {
       
-      plotResetTrial(participant=participant, trial=trial)
+      trialdf <- df[which(df$trial_no == trialn & df$step == 2),]
+      
+      yd <- diff(trialdf$gabory_pix)
+      
+      pause.idx <- which(yd == 0)
+      
[…]

### 2019-11-20 exp/resetExperiment.py
changed pause offsets to what I intended... but never did (these values don't get used in the experiment though)

@@ -40,7 +40,8 @@ cfg['trial_duration']                   = (0.5 / cfg['externalSpeed']) * (Npasse
 # there will be 3 internal cursor movements: right, left and none:
 internalMovement                        = [-3, 3]
 resetPausePass                          = [ 1, 2, 3, Npasses+1]
-resetPauseOffset                        = [.3, .4]
+#resetPauseOffset                        = [.3, .4]
+resetPauseOffset                        = [.25, .75]
 externalDirection                       = [-1, 1]
 
 repetitions                             = 1

### 2019-11-20 R/tracking_bounded.R
code for plot streamlined a bit

@@ -1,4 +1,5 @@
 library('svglite')
+source('R/common.R')
 
 
 # Figures -----
@@ -8,37 +9,34 @@ plotBoundedTrackingRaw <- function(target='inline') {
   participants <- c(1,3,5,6,7,8)
   
   if (target=='svg') {
-    svglite(file='doc/bounded_tracking_uncluttered.svg',width=6,height=9)
+    svglite(file='doc/Fig03.svg',width=7,height=6)
   }
   
-  par(mfrow=c(length(participants),5),mar=c(4.1,4.1,1.0,1.0))
+  par(mfrow=c(1,1),mar=c(4.5,4.1,0.1,0.1))
   
-  for (ppno in participants) {
+  colors <- getColors()
+  
+  red <- colors$yorkred
+  
+  plot(-1000,-1000,main='',ylim=c(0.5,5.5),xlim=c(0.5,6.5),xlab='participant',ylab='internal speed [cps]',asp=1,bty='n',ax=F)
+
+  speeds <- c(-3,-1,0,1,3)
+  
+  for (ppidx in c(1:length(participants))) {
+    
+    ppno <- participants[ppidx]
     
     df <- read.csv(sprintf('data/bounded_tracking/bounded_tracking_p%02d.csv',ppno))
     
     df <- df[which(df$step == 2),]
     
-    for (speed in c(-3,-1,0,1,3)) {
+    for (speedidx in c(1:length(speeds))) {
+      
+      speed <- speeds[speedidx]
       
       sdf <- df[which(df$internalSpeed == speed),]
       
       trials <- unique(sdf$trial_no) 
-      
-      if (ppno == participants[1]) {
-        main=sprintf('speed: %0.0f',speed)
-      } else {
-        main=''
-      }
-      xlab=''
-      if (speed == -3) {
-        ylab = sprintf('participant %d',ppno)
-      } else {
-        ylab = ''
-      }
-      
-      plot(-1000,-1000,main=main,ylim=c(-450,450),xlim=c(-450,450),ylab=ylab,xlab=xlab,asp=1,bty='n',ax=F)
-      lines(c(0,0),c(0,12),col='#999999')
       
       for (trialno in trials) {
         
@@ -49,13 +47,22 @@ plotBoundedTrackingRaw <- function(target='inline') {
         x <- sdf$handx_pix[idx] * sdf[idx,]$externalDirection[1]
         y <- sdf$handy_pix[idx]
         
-        lines(x,y,col='#99000066')
[…]

### 2019-11-20 DoubleDriftResetLimits.Rmd
improved explanation of paused tracking experiment, added figure for paused tracking

@@
 # install.packages(c('svglite','ez'))
+# they seem to rely on other packages: 'data.table', 'ggplot2', 'dplyr', 'lme4', and 'car'
 
@@
 
-```{r fig.width=6, fig.height=9}
+```{r fig.width=7, fig.height=6}
 # from R/tracking_bounded.R:
@@
 
-Therefore, we did second experiment that mimicked the first, but with three changes, all designed to increase the occurrence of resets.
-First, the external motion was slowed down as earlier work showed that resets often occur after around 2 seconds, which was how long one pass (from one end of the screen to the other) took in the first experiment.
-Second, we only used -3 and 3 cycles per second for the internal motion.
-Third, on 3 of every 4 trials, a ~266.66 ms "blink" (8 frames at 30 Hz) was introduced in the stimulus.
+Therefore, we did another experiment that mimicked the first, but with three changes, all designed to increase the occurrence of resets.
+First, the external motion was halved as earlier work showed that resets often occur after around 2-4 seconds [@Lisi2015 ??], and 2 seconds was how long one pass (from one end of the screen to the other) took in the first experiment.
+So now, the double-drift stimulus would do a single pass in 4 seconds instead of 2 seconds.
+Second, we only used -3 and 3 cycles per second for the internal motion as this would increase the strength of the illusion and perhaps lead to a higher chance of reset.
+Third, on 2 of every 4 trials, a ~266.66 ms "blink" (8 frames at 30 Hz) was introduced in the stimulus.
 During this period the stimulus disappeared from the screen and reappeared after the blink in the same location to continue as before.
-The blinks occured equally often in the first, second or third full pass of the stimulus, and always halfway along the path on screen.
-These blinks have been shown to trigger resets [@Nakayama2018], so if we can spot triggered resets in the tracking data, we can use the characteristics of that tracking data to detect spontaneous resets.
+The blinks occured either in the first or second full pass of the stimulus, halfway along the path on screen.
+These blinks have been shown to trigger resets [@Lisis201X; @Nakayama2018], so if we can spot triggered resets in the tracking data, we can use the characteristics of that tracking data to detect spontaneous resets.
 However, we would primarily expect to see increases in horizontal tracking speed, perhaps accompanied by a decrease in vertical speed, when people track the "jump" in the perceived location of the stimulus.
+
+Here is a plot of trajectories just before (orange), during (red) as well as for 1500 ms after a pause (purple).
+The trajectories are normalized for the different directions (trajectories always shown going up and to the right, but could have been down and/or to the left) and the trajectories are all shown relative to the start of the pause, which is where they cross the gray vertical.
+
+```{r fig.width=7, fig.height=7}
+# from R/tracking_reset_pause.R
+plotPostPauseTrajectories()
+```
+
+Except for one trial in participant 9, it doesn't look like there are any resets to be seen in the trajectories.

### 2019-12-16 R/tracking_bounded.R
excluding authors from participant set

@@ -7,6 +7,7 @@ getBoundedTrackingTimeNormalizedSegments <- function(segmentpoints=101) {
   
   # use these participants:
   participants <- c(1,3,5,6,7,8)
+  participants <- c(3,6,7,8) # 1 and 5 are authors
   
   # read all the data into one data frame:
   # BTdata <- NA
@@ -117,6 +118,7 @@ getBoundedTrackingStandardizedSegments <- function(segmentpoints=101) {
   
   # use these participants:
   participants <- c(1,3,5,6,7,8)
+  participants <- c(3,6,7,8) # 1 and 5 are authors
   
   # read all the data into one data frame:
   # BTdata <- NA
@@ -239,9 +241,10 @@ getSegmentDirections <- function(allsegmentdata) {
 plotBoundedTrackingRaw <- function(target='inline') {
   
   participants <- c(1,3,5,6,7,8)
+  participants <- c(3,6,7,8)
   
   if (target=='svg') {
-    svglite(file='doc/Fig03.svg',width=7,height=6)
+    svglite(file='doc/Fig03.svg',width=6,height=6)
   }
   
   par(mfrow=c(1,1),mar=c(4.5,4.1,0.1,0.1))
@@ -250,7 +253,7 @@ plotBoundedTrackingRaw <- function(target='inline') {
   
   red <- colors$yorkred
   
-  plot(-1000,-1000,main='',ylim=c(0.5,5.5),xlim=c(0.5,6.5),xlab='participant',ylab='internal speed [cps]',asp=1,bty='n',ax=F)
+  plot(-1000,-1000,main='',ylim=c(0.5,5.5),xlim=c(0.5,length(participants)+1),xlab='participant',ylab='internal speed [cps]',asp=1,bty='n',ax=F)
 
   speeds <- c(-3,-1,0,1,3)
   
@@ -367,8 +370,8 @@ placeHolderOne <- function(target='inline') {
       idx <- which(allsegments$internalSpeed == internalSpeed & allsegments$participant == participant)
       
       # we put it into a normalized 2D histogram:
-      pfreq <- hist2d(x=allsegments$direction[idx], y=allsegments$sample_no[idx], nbins=NA, edges=list(seq(0,180,length.out=31), seq(1,101,4)+0.5))
-      pfreq$freq2D <- sqrt(pfreq$freq2D / sum(pfreq$freq2D))
[…]

### 2019-12-16 doc/DoubleDriftResetLimits.Rmd
moving notebooks to /doc folder

--- 2019-11-20 DoubleDriftResetLimits.Rmd
@@
 
+The MEAN would say: "the majority of trials / participants show no evidence of a reset".
+
+I'd like to be able to say: "there is no evidence of any resets in all of the data".
+That is hard to do, unless I find a way to do an appropriate equivalence test (where the key problem is in the word 'appropriate').
+
+But I would settle for: "the majority of data shows no resets, and there's no easily identifiable minority in the data with resets either".
+
+And we can do the latter if we find a measure that would be somewhat bimodal in case of a few resets.
+A search reveals that one of the most accepted tests is recently Hartigan's Dip Test for unimodality (H1: not unimodal), at least a few years ago, so this might have changed, but it should still work.
+It is in the package 'diptest'.
+
 # Experiment 1b: continuous tracking with pauses
@@
 
-```{r fig.width=7, fig.height=7}
+```{r fig.width=7, fig.height=3}
 # from R/tracking_reset_pause.R
@@
 
-Except for one trial in participant 9, it doesn't look like there are any resets to be seen in the trajectories.
+Except perhaps for one trial in participant 9, it doesn't look like there are any resets to be seen in the trajectories.

### 2019-12-16 doc/Manuscript.Rmd
moving notebooks to /doc folder

---
title: "Measuring the double-drift illusion with hand tracking: capturing illusion resets"
author: "Marius 't Hart"
author: "Denise Y.P. Henriques"
author: "Patrick Cavanagh"
date: "16/12/2019"
output:
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Abstract

If a gabor pattern drifts in one direction while its internal texture drifts in the orthogonal direction, its perceived direction deviates strongly from its true direction and is instead some combination of its real external motion and its internal motion (Tse & Hsieh, 2006). The perceived location of such a double-drift stimulus periodically resets toward the veridical position. There may indeed be a simple perdiodicity underlying these resets. However, since both the veridical and perceived location are available to the brain, these resets may also occur when these two positions are too far from each other to come from the same source stimulus. Here we test which of the two explanations is more likely. In a first experiment, we had participants continuously track the position of the perceived position of the stimulus. While the orientation of the trajectories reflect stimulus properties, there were few clear resets. That’s why, in a second experiment, we had participants re-trace the perceived trajectory of a short stimulus presentation, or indicate the initial direction of the stimulus they perceived. We found resets here, and they indicate a set spatial limit to the distance between the veridical and perceived location of the double-drift stimulus.

# Introduction

Encoding the position of objects in the world is necessary for the myriad of visually guided motor tasks that we do every day. It is no surprise then that the brain has several mechanisms to gauge and update the position of objects in the world. For example, when a target is moving, its motion can influence it perceived location, making it appear to be slightly ahead of its true location (e.g. the flash lag, Nijhawan, 1994; or the flash grab effect, Cavanagh & Anstis, 2013). These cases of position extrapolation may be functional, helping the motor system overcome inevitable neural delays in targeting a moving object (Nijhawan, 1994; Duhamel et al., 1992). A stronger and very different motion-induced position shift arises when a moving gabor with internal motion orthogonal to its path is viewed in the periphery (Tse & Hsieh, 2006; Shapiro et al, 2010; Gurnsey & Baird, 2012; Kwon et al, 2015; Lisi & Cavanagh, 2015). This double-drift stimulus (Fig. 1) generates extreme misjudgments of the moving target’s location which may deviate by as much as several degrees of visual angle from its true location. Surprisingly, this extraordinary perceptual illusion does not affect eye movements to the gabor: saccades to these targets are determined by their physical not their perceived location (Lisi & Cavanagh, 2015). This result indicates that both true and illusory locations are available in the brain. Indeed, although immediate eye-movements go towards the true position, delayed eye movements and pointing go to the perceived position [Lisi & Cavanagh, 2017; Massendari et al, 2018].

The strength of the double-drift probably arises because of the very poor location information for a gabor pattern in the periphery when it has the same mean luminance as the background. With poor positional certainty, the motion information contributes to the location estimate, generating a shift in position. If the gabor itself is not moving (De Valois & De Valois, 1991), the shift saturates after about 100 ms (Chung et al., 2007) perhaps because the accumulating evidence of actual location of the stationary gabor patch is sufficient to anchor its position. However, when the gabor itself also moves as in the double-drift stimulus, there is no stable location information to anchor the position estimate and the build-up continues away from its physical location to reach deviations of several times the size of the gabor. At some point, the illusion resets, the perceived location jumps back to its physical location and then accumulation begins again (Fig. 2). This long duration before resetting is well beyond the integration period of other visual effects and into the range of short-term memory.

Here, we first examine the double-drift illusion by asking participants to track its path with a stylus on a tablet that is optically aligned with the monitor surface but with the hand not visible. Earlier results for pointing to the double-drift target (Lisi & Cavanagh, 2017) show that the illusion persists for manual pointing. However, with online tracking, the presence of the hand near the location of the perceived gabor may interact with the illusion. Specifically, the additional proprioceptive position signal may bias the perception toward its veridical location.

The ability to manually track the perceived position will also allow us to study the resets – the adjustment of perceived location when the illusion has deviated too much or too long from the physical location. It could be that the imprecise, but true, retinal position signal limits the illusion, or it could be that – like other illusions, such as bi-stable stimuli – there is a simple periodicity underlying changes in perception. Here we capture where and when observers experience resets by having them draw the perceived trajectory of the stimulus after having seen it (a retrace procedure). This will allow us to distinguish between a spatial limit to the illusion or a temporal limit.

## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.

### 2019-12-17 R/tracking_reset_pause.R
no longer doing triggered reset analyses: experiment wasn't suitable to evoke them

@@ -1,1389 +1,0 @@
-library('svglite')
-source('R/common.R')
-
-# Data handling -----
-
-getPauseTrajectories <- function(normalizeX=TRUE) {
-  
-  participants <- c(1,2,9,10,11,12,13)
-  
-  trajectories <- NA
-  
-  for (ppidx in c(1:length(participants))) {
-    
-    ppno <- participants[ppidx]
-    
-    df <- read.csv(sprintf('data/reset_pause/reset_p%02d.csv',ppno))
-    
-    trials <- unique(df$trial_no)
-    
-    for (trialn in trials) {
-      
-      trialdf <- df[which(df$trial_no == trialn & df$step == 2),]
-      
-      yd <- diff(trialdf$gabory_pix)
-      
-      pause.idx <- which(yd == 0)
-      
-      if (length(pause.idx) > 0) {
-        
-        #cat(sprintf('pause samples: %d\n',length(pause.idx)))
-        # first we will determine which pass this was on:
-        pauseonsettime <- trialdf$time_ms[pause.idx[1]] - trialdf$time_ms[which(trialdf$step == 2)][1]
-        passn <- round(pauseonsettime/4000)
-        
-        # which samples do we care about?
-        # those just before the pause, and up to some time after
-        # half a second before = 15 samples
-        #    1.5 seconds after = 45 samples
-        
-        handX <- trialdf$handx_pix
-        handY <- trialdf$handy_pix
-        
-        handX <- handX[c(pause.idx[1]-15:pause.idx[length(pause.idx)]+30)]
-        handY <- handY[c(pause.idx[1]-15:pause.idx[length(pause.idx)]+30)]
-        
-        # we want to stop looking the trajectories at certain points, by two rules
-        # the first is implemented here:
-        # when people hit the edge of the drawing tablet
-        
-        upto.idx <- length(handX) # start with the full trajectory
-        if (any(handY == -554)) {
-          upto.idx <- min(upto.idx,which(handY == -554)[1])
-        }
-        if (any(handY == 495)) {
-          upto.idx <- min(upto.idx,which(handY == 495)[1])
-        }
-        
-        # we align everthing so that (0,0) is where the hand is when the pause starts:
-        if (normalizeX) {
-          handX <- handX - handX[24]
-        }
-        handY <- handY - handY[24]
-        
-        # we normalize trajectories so that they all go in the same direction:
-        if (trialdf$internalSpeed[1] < 0) {
[…]

### 2019-12-17 R/data.R
changed data files

@@ -1,8 +1,8 @@
 # check / download data -----
 
-source('R/tracking_reset_pause.R')
+#source('R/tracking_reset_pause.R')
 
-downloadRawData <- function(overwrite=TRUE, experiments=c('bounded_tracking', 'onePass_V4', 'reset_pause')) {
+downloadRawData <- function(overwrite=TRUE, experiments=c('bounded_tracking', 'bounded_tracking_V2', 'onePass_V4')) {
   
   
   for (experiment in experiments) {
@@ -67,9 +67,9 @@ downloadRawData <- function(overwrite=TRUE, experiments=c('bounded_tracking', 'o
       
     }
     
-    if (experiment == 'reset_pause') {
+    if (experiment == 'bounded_tracking_V2') {
       
-      folder = 'data/reset_pause'
+      folder = 'data/bounded_tracking_V2'
       
       if (!dir.exists(folder)) {
         dir.create(folder)
@@ -77,13 +77,21 @@ downloadRawData <- function(overwrite=TRUE, experiments=c('bounded_tracking', 'o
 
       resources <- c( 
         
-        'reset_p01.csv' = 'https://osf.io/739dp/download',
-        'reset_p02.csv' = 'https://osf.io/w7kht/download',
-        'reset_p09.csv' = 'https://osf.io/k29v3/download',
-        'reset_p10.csv' = 'https://osf.io/vb83k/download',
-        'reset_p11.csv' = 'https://osf.io/9rj8s/download',
-        'reset_p12.csv' = 'https://osf.io/4n36j/download',
-        'reset_p13.csv' = 'https://osf.io/vs6bw/download'
+        'bounded_tracking_V2_p01.csv' = 'https://osf.io/akyf4/download',
+        'bounded_tracking_V2_p02.csv' = 'https://osf.io/gzkqf/download',
+        'bounded_tracking_V2_p09.csv' = 'https://osf.io/wdq7p/download',
+        'bounded_tracking_V2_p10.csv' = 'https://osf.io/8g4q9/download',
+        'bounded_tracking_V2_p11.csv' = 'https://osf.io/axqcy/download',
+        'bounded_tracking_V2_p12.csv' = 'https://osf.io/fjpz6/download',
+        'bounded_tracking_V2_p13.csv' = 'https://osf.io/cyj49/download'
+        
+        # 'reset_p01.csv' = 'https://osf.io/739dp/download',
[…]

### 2019-12-17 R/tracking_bounded.R
finalized code for first bounded tracking experiment

@@ -1,4 +1,6 @@
 library('svglite')
+library('ez')
+
 source('R/common.R')
 
 # Data handling -----
@@ -106,7 +108,9 @@ getBoundedTrackingStandardizedSegments <- function(segmentpoints=101) {
   
 }
 
-getSegmentDirections <- function(allsegmentdata) {
+getSegmentDirections <- function() {
+  
+  allsegmentdata <- read.csv('data/bounded_tracking/standardized_segments.csv', stringsAsFactors = F) 
   
   allsegmentdata$direction <- NA
   
@@ -121,7 +125,7 @@ getSegmentDirections <- function(allsegmentdata) {
     allsegmentdata$direction[seg.idx] <- c(NA, ( atan2(diff(segment$handy_pix), diff(segment$handx_pix)) / pi ) * 180)
   }
   
-  return(allsegmentdata)
+  write.csv(allsegmentdata, 'data/bounded_tracking/segment_directions.csv', row.names = F, quote = F)
   
 }
 
@@ -133,10 +137,8 @@ plotBoundedTracking <- function(target='inline') {
   
   colors <- getColors()
 
-  #rawsegments <- read.csv('data/bounded_tracking/normalized_segments.csv', stringsAsFactors = F)  
-  stdsegments <- read.csv('data/bounded_tracking/standardized_segments.csv', stringsAsFactors = F)  
-  
-  stdsegments <- getSegmentDirections(stdsegments)
+  stdsegments <- read.csv('data/bounded_tracking/segment_directions.csv', stringsAsFactors = F)  
+
   stdsegments$direction[which(stdsegments$direction < 0)] <- NA
   stdsegments <- stdsegments[which(!is.na(stdsegments$direction)),]
   
@@ -145,7 +147,7 @@ plotBoundedTracking <- function(target='inline') {
     svglite(file='doc/Fig03b.svg',width=6,height=6)
   }
   
-  # 5 conditions (columns)
+  # 5 conditions or internal speeds (columns)
   # row 1: heading distribution
   # row 2-5: participants
   
@@ -283,8 +285,93 @@ plotBoundedTracking <- function(target='inline') {
   
 }
 
+# Statistics -----
 
 
+
+analyzeBoundedTracking <- function(target='inline') {
+  
+
+  stdsegments <- read.csv('data/bounded_tracking/segment_directions.csv', stringsAsFactors = F)  
+  
[…]

### 2019-12-17 DoubleDriftResetLimits.Rmd
rewrote stats part for first bounded tracking experiment

--- 2019-11-07 DoubleDriftResetLimits.Rmd
@@
 
-# The functions in here are used on the data of the first tracking experiment:
+# The functions in here are used on the data of the bounded tracking experiments:
 source('R/tracking_bounded.R')
-# This code is for the second tracking experiment:
-source('R/tracking_reset_pause.R')
-# And finally some code for the re-tracin experiment...
+# And finally some code for the re-tracing experiment...
 source('R/retrace_onePass_V4.R')
@@
 # from R/data.R
-downloadRawData(overwrite=FALSE, experiments=c('bounded_tracking', 'reset_pause', 'onePass_V4'))
+downloadRawData(overwrite=FALSE, experiments=c('bounded_tracking', 'bounded_tracking_V2', 'onePass_V4'))
 ```
@@
 
-```{r fig.width=6, fig.height=9}
+```{r fig.width=7, fig.height=6}
 # from R/tracking_bounded.R:
-plotBoundedTrackingRaw()
+plotBoundedTracking()
 ```
 
-We can see that the orientation of the tracking movements responds to the strength and direction of the internal motion of the stimulus. However, there are no clear spontaneous resets of the stimulus.
+The top row shows average instantaneous heading (as actual direction) over time (distance from origin), with a polar heatmap of all the data in the background.
+The middle 1 second of each pass of the gabor is used, so that the inversions at the far and close end of the workspace are excluded from the average heading data.
+Since headin doesn't change throughout these segments (stats?) we take the average heading across all segments for every participant and every internal speed to do statistical analyses.
 
-# Experiment 1b: continuous tracking with pauses
+```{r}
+analyzeBoundedTracking()
+```
 
-Therefore, we did second experiment that mimicked the first, but with three changes, all designed to increase the occurrence of resets.
-First, the external motion was slowed down as earlier work showed that resets often occur after around 2 seconds, which was how long one pass (from one end of the screen to the other) took in the first experiment.
-Second, we only used -3 and 3 cycles per second for the internal motion.
-Third, on 3 of every 4 trials, a ~266.66 ms "blink" (8 frames at 30 Hz) was introduced in the stimulus.
-During this period the stimulus disappeared from the screen and reappeared after the blink in the same location to continue as before.
-The blinks occured equally often in the first, second or third full pass of the stimulus, and always halfway along the path on screen.
-These blinks have been shown to trigger resets [@Nakayama2018], so if we can spot triggered resets in the tracking data, we can use the characteristics of that tracking data to detect spontaneous resets.
-However, we would primarily expect to see increases in horizontal tracking speed, perhaps accompanied by a decrease in vertical speed, when people track the "jump" in the perceived location of the stimulus.
+First we want to see if internal motion affects illusion strength as measured by instantaneous heading direction.
+A repeated measures ANOVA on a model using internal speed to predict heading direction shows that heading direction in tracking data is a good measure of illusion strength (F(4,12)=102.7, p<.001, ges=0.97, Greenhouse-Geisser corrected).
+This effect is not caused by one internal speed giving different directions, but by each internal speed resulting in different heading directions from it's direct neighbour (4 FDR-corrected, pairwise, one-sided t-tests: all p<.025).
+
+A few of the tracking paths might be reflecting a reset (four of them are plotted in red in Fig 2).
+However, there are only a few and it's not clear they represent resets, or tracking errors made when changing direction at the far and close end of the workspace.
+The heading distributions don't seem to show any systematic deflections from the average heading, for eample evoked after a specific amount time.
+
+# Experiment 1b: continuous tracking
+
+Therefore, we did another experiment that mimicked the first, but with two changes, all designed to increase the occurrence of resets.
+First, we only used -3 and 3 cycles per second for the internal motion as this would increase the strength of the illusion and perhaps lead to a higher chance of reset.
+Second, the external motion was halved as earlier work showed that resets often occur after around 2 seconds [@Lisi2015 ??], and 2 seconds was how long one pass (from one end of the screen to the other) took in the first experiment.
+So now, the double-drift stimulus would do a single pass in 4 seconds instead of 2 seconds.
+This allows more opportunity for resets, but might also make it easier to track the gabor's position, so that errors at the direction changes in the far and close end of the workspace are less likely.

### 2019-12-18 R/tracking_bounded.R
added functions for second bounded tracking experiment... might remove later

@@ -5,11 +5,20 @@ source('R/common.R')
 
 # Data handling -----
 
-getBoundedTrackingStandardizedSegments <- function(segmentpoints=101) {
+getBoundedTrackingStandardizedSegments <- function(segmentpoints=101, version=1) {
   
   # use these participants:
   participants <- c(1,3,5,6,7,8)
   participants <- c(3,6,7,8) # 1 and 5 are authors
+  folderstr <- 'data/bounded_tracking/'
+  filestr <- 'bounded_tracking_'
+  
+  if (version == 2) {
+    participants <- c(1,2,9,10,11,12,13)
+    participants <- c(2,9,10,11,12,13) # 1 is author
+    folderstr <- 'data/bounded_tracking_V2/'
+    filestr <- 'bounded_tracking_V2_'
+  }
   
   # read all the data into one data frame:
   # BTdata <- NA
@@ -17,7 +26,7 @@ getBoundedTrackingStandardizedSegments <- function(segmentpoints=101) {
   
   for (ppno in participants) {
     
-    ppdf <- read.csv(sprintf('data/bounded_tracking/bounded_tracking_p%02d.csv',ppno), stringsAsFactors = F)
+    ppdf <- read.csv(sprintf('%s%sp%02d.csv',folderstr,filestr,ppno), stringsAsFactors = F)
     ppdf$participant <- ppno
     
     # # get all raw data in one frame?
@@ -69,6 +78,10 @@ getBoundedTrackingStandardizedSegments <- function(segmentpoints=101) {
         segmentdf$handx_pix <- segmentdf$handx_pix - segmentdf$handx_pix[1]
         segmentdf$handy_pix <- segmentdf$handy_pix - segmentdf$handy_pix[1]
         
+        if (version == 2) {
+          segmentdf$handx_pix <- segmentdf$handx_pix * sign(segmentdf$internalSpeed[1])
+        }
+        
         # make new vectors for new data frame:
         participant <-       rep(ppno,                           segmentpoints)
         trial_no <-          rep(segmentdf$trial_no[1],          segmentpoints)
@@ -104,13 +117,18 @@ getBoundedTrackingStandardizedSegments <- function(segmentpoints=101) {
     }
   }
   
[…]

### 2019-12-18 R/plotTracking.R
added note about purpose of file

@@ -1,3 +1,8 @@
+
+# This is the first, organically grown set of functions: one big mess.
+# Kept here as source material.
+
+
 require('ez')
 require('svglite')
 

### 2019-12-18 DoubleDriftResetLimits.Rmd
updated notebook with current bounded tracking plots and statistics

@@
 title: "Limits to Spontaneous Resets in the Double-Drift Illusion"
-output: html_notebook
+output:
+  word_document: default
+  html_notebook: default
 ---
@@
 # This file has some common helper functions:
-
+source('R/common.R')
 # The functions in here are used on the data of the bounded tracking experiments:
@@
 
-Here we plot the trajectories recorded during this task (each subplot essentially shows 8 trials overlayed):
+First, we process the raw data into a more useful format for our analyses:
 
-```{r fig.width=7, fig.height=6}
+```{r}
+# from R/tracking_bounded.R:
+getBoundedTrackingStandardizedSegments(segmentpoints=101, version=1)
+getSegmentDirections(version=1)
+```
+
+Then we plot the trajectories recorded during this task (each subplot essentially shows 8 trials overlayed), and in the top row we show the average heading as a thicker line, as well as a 2D polar histogram showing the distribution of headings.
+
+```{r fig.width=7, fig.height=6, echo=F}
 # from R/tracking_bounded.R:
@@
 
-# Experiment 1b: continuous tracking
+# Experiment 1b: continuous tracking with 4 second segments
 
@@
 This allows more opportunity for resets, but might also make it easier to track the gabor's position, so that errors at the direction changes in the far and close end of the workspace are less likely.
+Since the passes are longer, we use only 3 full passes in each trial, but run 32 trials in total (instead of the 8 per condition).
+
+Before calculating heading, we've normalized the data, so that the direction of all segments _should_ be the same.
+
+First, here is a figure with the heading distributions across time for all participants as well as the group average:
+
+```{r fig.width=8, fig.height=4, echo=F}
+plotBoundedTracking_V2(target='inline')
+```
+
+Again, we see no time-dependent decrease of heading towards zero, except perhaps at the start of the data for participant 12.
+Since these are preceded by very large headings, and then settle into the normal range, this participant might have had more trouble responding to the direction reversals of the gabor.
+The group average reflects this tendency of participant 12 somewhat, but overall the average heading is very close to what was seen in the first trakcing experiment: 28.9 compared to 27.4 or 26.7 degrees.
+
+# Experiment 2: one-pass tracking and psychophysics
+
+Since it seemed that no clear spontaneous resets occurred in the continuous tracking paradigm, we used a re-tracing approach (c.f. @Nakayama2019).

### 2020-01-10 DoubleDriftResetLimits.Rmd
clarified most text and added text

@@
 ---
-title: "Limits to Spontaneous Resets in the Double-Drift Illusion"
+title: "Double-Drift Illusion Boundaries"
 output:
@@
   html_notebook: default
+  html_document:
+    df_print: paged
+  pdf_document: default
 ---
@@
 
-State aims of the project. Say which experiments were done and are analyzed here.
+We want to see if we can find reset points in tajectories people draw while or after seeing a double-drift stimulus, with the goal to see if we can figure out if those resets occur after a certain time interval (like perception in other illusions) or if it occurs when the percept reaches certain spatial difference with the true position.
 
-# Setting up
+![Stimuli and Setup.
+*Top left*: a gabor with internal motion (phase shifts) and external motion, that is viewed peripherally, appears to follow a path deviating from the true path.
+Eye-movements immediately following disappearance of the gabor land at the true position, but pointing movements and delayed eye-movements go to the illusory position.
+*Bottom left*: at some point, the perceived position will "reset" towards the true position.
+For some people this takes the form of a "jump" back to the true position, for others it means the illusory position doesn't deviate more from the true position, and there are some in between version.
+*Top right*: if these resets are due to a spatial limit on the size of the illusory position shift, the points of the resets should have a constant X coordinate.
+*Middle right*: if they occur after some period of time, they should have a constant Y coordinate.
+*Bottom right*: we have people track or re-trace their percept and try to isolate reset points from the drawn trajectories to figure out where or when resets occur.
+](doc/stimulus_reset.png){width=100%}
+
+# Setting up the R session
 
@@
 
-We've minimized depencies on other packages, but there are two that help - even when not strictly required.
+We've minimized depencies on other packages, but there are two that help. Of course, you can skip the chunks that use them, or just install them by uncommenting the second line of this chunk:
 
 ```{r}
+# install the dependencies using this command:
+# install.packages(c('svglite','ez'))
+# they seem to rely on other packages: 'data.table', 'ggplot2', 'dplyr', 'lme4', and 'car'
+
 library('svglite')
@@
 
-# these packages are not mandatory
-# - svglite generates SVG versions of the figures, but the native R figures might be enough
-# - ez allows one ANOVA to be run, but this part could be skipped
-
-# if you do want to install them, use this:
-# install.packages(c('svglite','ez'))
-# they seem to rely on other packages: 'data.table', 'ggplot2', 'dplyr', 'lme4', and 'car'
-
+# - svglite generates SVG versions of the figures, but you can see the inline R figures in the notebook:
+# - ez allows one F-test to be run, but it's not central to the paper
 ```
 
-## Functions
+## Source files
 
-There are several files with code that will do the work for us in pre-processing, plotting and doing stats.
+There are several files with functions that will do the work for us in pre-processing, plotting and doing stats.
 
@@
 
+This will create pre-processed data files in the directory: `data/bounded_tracking`.
+
 Then we plot the trajectories recorded during this task (each subplot essentially shows 8 trials overlayed), and in the top row we show the average heading as a thicker line, as well as a 2D polar histogram showing the distribution of headings.
 
-```{r fig.width=7, fig.height=6, echo=F}
+```{r fig.width=7, fig.height=6}
 # from R/tracking_bounded.R:
@@
 The top row shows average instantaneous heading (as actual direction) over time (distance from origin), with a polar heatmap of all the data in the background.
-The middle 1 second of each pass of the gabor is used, so that the inversions at the far and close end of the workspace are excluded from the average heading data.
-Since headin doesn't change throughout these segments (stats?) we take the average heading across all segments for every participant and every internal speed to do statistical analyses.
+The first and last 0.5 second of each pass is stripped, so that the inversions at the far and close end of the workspace are excluded from the average heading data.
+In this case the middle 1 second of each pass of the gabor is used for the heading plots.
+Since heading doesn't change throughout these segments (stats?) we take the average heading across all segments for every participant and every internal speed to do statistical analyses.
 
@@
 First we want to see if internal motion affects illusion strength as measured by instantaneous heading direction.
-A repeated measures ANOVA on a model using internal speed to predict heading direction shows that heading direction in tracking data is a good measure of illusion strength (F(4,12)=102.7, p<.001, ges=0.97, Greenhouse-Geisser corrected).
+A repeated measures ANOVA (or F-test) on a model using internal speed to predict heading direction shows that internal motion affects heading direction in tracking data (F(4,12)=102.7, p<.001, ges=0.97, Greenhouse-Geisser corrected).
 This effect is not caused by one internal speed giving different directions, but by each internal speed resulting in different heading directions from it's direct neighbour (4 FDR-corrected, pairwise, one-sided t-tests: all p<.025).
+That means heading direction is a good measure of illusion strength.
 
-A few of the tracking paths might be reflecting a reset (four of them are plotted in red in Fig 2).
+A few of the tracking paths might be reflecting a reset (four examples are plotted in red in Fig 2).
 However, there are only a few and it's not clear they represent resets, or tracking errors made when changing direction at the far and close end of the workspace.
-The heading distributions don't seem to show any systematic deflections from the average heading, for eample evoked after a specific amount time.
+The heading distributions don't seem to show any systematic (time-locked) deflections from the average heading.
 
@@
 
-```{r fig.width=8, fig.height=4, echo=F}
+```{r fig.width=8, fig.height=4}
 plotBoundedTracking_V2(target='inline')
@@
 Again, we see no time-dependent decrease of heading towards zero, except perhaps at the start of the data for participant 12.
-Since these are preceded by very large headings, and then settle into the normal range, this participant might have had more trouble responding to the direction reversals of the gabor.
+Since these are preceded by very large headings, and then settle into the normal range, this participant might have had trouble responding to the direction reversals of the gabor rather then show resets, but it's somewhat unclear.
 The group average reflects this tendency of participant 12 somewhat, but overall the average heading is very close to what was seen in the first trakcing experiment: 28.9 compared to 27.4 or 26.7 degrees.
+So depsite the slower external movement, the illusion was about equally strong, but did not lead to consistent resets.
 
@@
 
-Since it seemed that no clear spontaneous resets occurred in the continuous tracking paradigm, we used a re-tracing approach (c.f. @Nakayama2019).
+Since it seemed that no clear spontaneous resets occurred in the continuous tracking paradigm, we used a re-tracing approach (c.f.
+@Nakayama2019).
+In this next experiment, participants first were presented with a double-drift stimulus starting close, and moving away from them, for a single pass across the workspace (while fixating to the left or right, at random).
+In two kinds of blocks participants had to either indicate the initial direction by adjusting the orientation of a line originating at the sarting point of the gabor, or they had to retrace the perceived path of the gabor.
+When retracing the path, the trajectory so far was shown on the screen, and it could be deleted to start over.
+The idea was that if the simultaneous tracking from the first tasks either interfered with the percept or made it hard to quickly report resets, this paradigm would overcome either problem, or both.
+
+In this experiment we did not control the strength of the illusion.
+In order to sample diverse illusion strengths, the external motion of 13.5 cm took either 3 or 4 seconds (~5.2 or ~3.4 cm/s), and the internal motion was set to either 2, 3 or 4 cps (corresponding to about 3.4, 5.2 or 6.9 cm/s).
+That means there are 6 different stimuli.
+Fixation was to the left or right (at random) and the internal motion could either be to the left or right as well (counterbalanced, and normalized to rightward motion in pre-processing).
+The resulting trajectories in the re-tracing condition seemed to show a reset in the majority of trials, for a few examples, see Fig XXXX.
+
+In order to determine the point of reset automatically, we took the first sign change of the X coordinate.
+Since the start of the tracing often was somewhat "shaky" we disregarded sign changes in the first XXX cm.
+Since the trajectories are measured in discrete pixel values, sign changes were clearer after slight smoothing of the trajectories, using splines.
+
+We validate this measure of detecting resets as follows.
+The angle of a straight line to the reset point should have approximately the same angle as the oriented lines indicating the initial direction of the gabor.
+We take the average initial directions and the average X and Y coordinates of detected reset points, for every participant in each of the 6 conditions.
+First, we test if the initial perceived direction as participants inidicated with the oriented line, responds to the different internal and external motion speeds.
+Second, we perform a regression of the angle to the reset point on the angle of the initial direction responses.
+
+```{r fig.width=8, fig.height=4}
+plotValidity(target='inline')
+```
+
+It looks like the initial percept angle responds to internal and external motion (speed) somewhat, and like there is some relationship between both measures, although not as strong as it might be.
+We verify both observations.
+
+```{r}
+testValidity()
+```
+
+First, we perform a repeated measures ANOVA on a model predicting (averaged) line angle, as a measure of the initial percept of the direction of motion from the internal and external speed of the stimulus.
+There is an effect of internal speed (F(2,16)=12.7, p<.001, ges=0.049) as well as of external speed (F(1,8)=6.89, p=.030, ges=0.031) on the line angles, and no interaction.
+This means that stimulus speeds affected perceived paths as expected and that the line angle responses captured this to some extent.
+
+Now, we want to see if the re-tracing paths capture the strength of the illusion, similarly to how the line angles capture illusion strength - at least up to the reset point.
+We do a linear regression of average reset point angle on average line angle in the same condition for every participant (then remove two outliers and redo the regression), **but this is not significant (R2(50)=.051, p=.058)**.
+(This is not the regression shown in the figure.) When looking at the raw data in the figure, it seems like the error in each variable is larger for increasing values of the other variable.
+If true, this would mean two things.
+First, given the apparently increasing variability of measurement (measurement error?) of each variable on the magnitude of the other variable, the errors of the fit for each point need to be weighted (e.g.
+by the distance to the origin, or something similar).
+Second, simple linear regression strictly speaking doesn't allow for (much) measurement error in the independent variable: orthogonal distance regression (or maybe total least squares) would be better suited for such data.
+
+Let's first solve the weighting problem, as that is straightforward: we add `1 / (distance to origin)` as the vector of weights to the linear model.
+The line angle responses in this weighted linear model _do_ predict reset point angle (R2(50)=0.15, p=.003).
+Orthogonal distance regression with weights should be possible, but there is no ready-made implementation (at least not that I could find), and it may also be completely redundant.
+That is: I will skip this for now.
+
+We're reasonably sure that the method of distilling reset points from trajectories works, so let's see what data there is.
+First we look at the data, and plot the reset coordinates in a time-spatial representation: X coordinates in centimeters, and Y coordinates converted to seconds (or scaled normalized centimeters that account for the different durations of the 3 s and 4 s pass conditions).
+We can also plot the X and Y coordinates over the strength of the illusion, expressed as the angle of the line from the start point of the gabor to the reset point relative to straight forward.
+
+```{r fig.width=7, fig.height=5}
+plotResetPoints()
+```
+
+In the figure with the reset point coordinates, we show the true path of the gabor as a gray arrow, and add potential limits by plotting the median X and Y coordinates as dashed lines.
+These are put at ~1.6 cm for the spatial limit and ~1.0 s for the temporal limit.
+The median X coordinate seems to describe the reset point distribution somewhat better.
+
+We can also fit models that use one or the other limit, using a least-squared error algorithm (using a constrained, quasi-Newton method).
+In these single-limit models, the errors are the distance of the fitted reset point from the actual reset point in normalized centimeters.
+Perhaps unsurprisingly, the limits found by these two models are qualitatively not very different from the medians of the coordinates.
+Of course, using a single, set value for the Y coordinate should result in a worse fit.
+And this is confirmed by AIC-based relative likelihoods:
+
+```{r}
+fitSeparateModelsOnData(verbosity = 1)
+```
+
+The mean, squared error and AIC is larger for a temporal limit, than for a spatial limit, and the relative likelihood of a temporal limit is a fraction of that of a spatial limit (p<.001).
+It does seem that a spatial limit would fit the data better.
+
+Nevertheless, we now try to fit both coordinates with a single model.
+In a first model, we simply combine both, i.e.
+there is both a spatial and a temporal limit, and we fit a weight parameter that balances the two components.
+The data with the fitted model looks like this:
+
+```{r, fig.width=7, fig.height=5}
+plotInterdependentModel()
+```
+
+Of course the MSE for this model is lower than either of the single-limit models: 2.267313. But the AIC might not be:
+
+```{r}
+# MSEs <- c('Lx'=2.603413, 'Ly'=4.027378, 'wLx+wLy'=2.267313)
+# 
+# N <- (9*6)-1 # number of observations (participants? or participants times conditions?)
+# # this is then used for C:
+# C <- N*(log(2*pi)+1)
+# 
+# AICs <- (2 * c(1,1,3)) + N*log(MSEs) + C
+# cat('model AICs:\n')
+# print(AICs)
+#   
+# relativeLikelihoods <- exp((min(AICs)-AICs)/2)
+# cat('\nrelative log-likelihoods:\n')
+# print(relativeLikelihoods)
+```
+
+If we use the number of participants as the number of observations, the single limit on X is still the favourable model.
+If we use the rows of the data frame, or the number of points in the plot, it is the double limit model.
+
+The main advantage of this model is that it includes a weight that decides how much of the result depends on a spatial limit and also how much depends on a time limit.
+This weight is fitted at around 80%, which gives a rather different picture then the relative likelihoods of the single limit models: there is substantial room for an effect of time.
+
+It is possible that after crossing a spatial limit, some small amount of time has to pass before perception is reset towards the true, retinal position.
+Therefore, we fit another model that expresses this idea.
+
+```{r}
+plotSequentialModel()
+```
+
+The fit seems the same as that of the weighted model, and indeed the MSE is identical.
+That is: the two models lead to exactly the same predicted reset points.
+Let's see what the AICs and relative log-likelihoods are.
+
+```{r}
+MSEs <- c('Lx'=2.603413, 'Ly'=4.027378, 'wLx+wLy'=2.267313, 'Lx->Ly'=2.267313)
+
+N <- (9*6)-1 # number of observations (participants? or participants times conditions?)
+# this is then used for C:
+C <- N*(log(2*pi)+1)
+
+AICs <- (2 * c(1,1,3,2)) + N*log(MSEs) + C
+cat('model AICs:\n')
+print(AICs)
+
+relativeLikelihoods <- exp((min(AICs)-AICs)/2)
+cat('\nrelative log-likelihoods:\n')
+print(relativeLikelihoods)
+```
+
+Since the sequential model has two parameters instead of three, it is slightly better, but not significantly so.
+While the sequential model allows estimating the actual spatial and temporal limits, the weighted model allows estimating the relative importance of the spatial and temporal limits.

### 2020-01-10 R/models.R
added a non-weighted (sequential?) model

@@ -1,93 +1,6 @@
 library('optimx')
 
 
-# combined geometric model -----
-
-resetModel <- function(par,directions,verbose=FALSE) {
-  
-  Lx <- par['Lx']
-  Ly <- par['Ly']
-  a  <- par['a']
-  
-  slopes <- sin(directions) / cos(directions)
-  
-  Y <- (a * Lx * slopes) + ((1-a) * Ly)
-  X <- (a * Lx)          + ((1-a) * Ly / slopes)
-  
-  resets <- data.frame(X,Y)
-  
-  if (verbose) {
-    cat(sprintf('Lx: %0.3f, Ly: %0.3f, a: %0.3f\n',Lx,Ly,a))
-  }
-  
-  return(resets)
-  
-}
-
-resetModelMSE <- function(par,directions,coords) {
-  
-  # the Lx, Ly and a parameters should be between 0 and 1, but optimx checks that
-  
-  resets <- resetModel(par,directions)
-  
-  MSE <- mean( (resets$X - coords$X)^2 + (resets$Y - coords$Y)^2 )
-  
-  return(MSE)
-  
-}
-
-fitResetModel <- function(directions,X,Y) {
-  
-  # convert input to more useful variables:
-  coords <- data.frame(X,Y)
-  directions <- ((90 - directions) / 180) * pi
-  
-  # create search grid:
-  
-  # # bound to "reasonable" values:
-  # Lx=seq(0,3/13.5,.05)
-  # Ly=seq(.5,1,.05)
-  # a=seq(0,1,.05)
-  
-  # bound to data:
-  #Lx=seq(0,median(X),diff(range(X))/10)
-  #Ly=seq(median(Y),1,diff(range(Y))/10)
-  
-  # bound to data:
-  Lx=seq(0,median(X),diff(range(X))/10)
-  Ly=seq(median(Y),1,diff(range(Y))/10)
-  a=seq(0,1,.1)
-  
-  searchgrid <- expand.grid('Lx'=Lx, 'Ly'=Ly, 'a'=a) # 11^3 combinations
-  
-  # get MSE for points in search grid:
-  MSE <- apply(searchgrid,FUN=resetModelMSE,MARGIN=c(1),directions=directions,coords=coords)
-  
-  # get 5 best points in the grid:
-  topgrid <- searchgrid[order(MSE)[1:5],]
-  
-  allfits <- do.call("rbind",
-                     apply( topgrid,
-                            MARGIN=c(1),
[…]

### 2020-01-10 R/retrace_onePass_V4.R
fixed plotting functions that didn't show plots, added analyses, used new non-weighted model

@@ -1,19 +1,21 @@
 library('svglite')
 library('ez')
 
+source('R/common.R')
+source('R/models.R')
+
+# Data handling -----
 
 preProcessOnePass_V4 <- function(participants = c(2,3,4,5,6,8,9,10,11), target='pdf') {
-  
-  graphics.off()
   
   internalMovements <- c(2,3,4)
   externalMovements <- rev(c(.125, .167))
   
   if (target == 'pdf') {
-    cairo_pdf(filename='onePass_V4.pdf',onefile=TRUE,width=11,height=8)
+    cairo_pdf(filename='doc/onePass_V4_all_participants.pdf',onefile=TRUE,width=11,height=8)
   }
   if (target == 'svg') {
-    svglite(file='onePass_V4.svg',width=11,height=8)
+    svglite(file='doc/onePass_V4_all_participants.svg',width=11,height=8)
   }
   
   par(mfrow=c(2,1),mar=c(4,4,2,0.1))
@@ -212,3 +214,821 @@ preProcessOnePass_V4 <- function(participants = c(2,3,4,5,6,8,9,10,11), target='
   }
   
 }
+
+
+summarizeTraceBoundsV4 <- function() {
+  
+  df <- read.csv('data/onePass_V4/onePass_V4_re-trace.csv', stringsAsFactors = F)
+  
+  df2 <- read.csv('data/onePass_V4/onePass_V4_arrow.csv', stringsAsFactors = F)
+  
+  # participants <- unique(df$participant)
+  # 
+  # internalspeeds <- unique(df$internalspeed)
+  # externalspeeds <- unique(df$externalspeed)
+  
+  aggdf <- aggregate(boundX ~ participant + internalspeed + externalspeed, data=df, FUN=mean, na.rm=TRUE)
+  names(aggdf)[which(names(aggdf) == 'boundX')] <- 'boundX_mean'
+  
+  tempdf <- aggregate(boundX ~ participant + internalspeed + externalspeed, data=df, FUN=sd, na.rm=TRUE)
+  aggdf['boundX_sd'] <- tempdf$boundX
+  
+  tempdf <- aggregate(boundY ~ participant + internalspeed + externalspeed, data=df, FUN=mean, na.rm=TRUE)
+  aggdf['boundY_mean'] <- tempdf$boundY
+  
[…]

### 2020-02-05 Manuscript.Rmd
add manuscript file

---
title: "Measuring the double-drift illusion with hand tracking: capturing illusion resets"
author: 
- Bernard Marius 't Hart 1,*
- Denise Y.P. Henriques 1
- Patrick Cavanagh 1,2,3
output:
  html_document:
    df_print: paged
  word_document: default
  pdf_document: default
bibliography: bibliography.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
# The functions here allow downloading the collected data from the OSF project:
source('R/data.R')
# This file has some common helper functions:
source('R/common.R')
# The functions in here are used on the data of the bounded tracking experiments:
source('R/tracking_bounded.R')
# And finally some code for the re-tracing experiment...
#source('R/retrace_onePass_V4.R')
# As well as models fit to that data:
source('R/models.R')
```

_1: Centre for Vision Research, York University, Toronto, ON M3J1P3, Canada_\
_2: Department of Psychology, Glendon College, Toronto, ON M4N3M6, Canada_\
_3: Department of Psychological and Brain Sciences, Dartmouth College, Hanover, NJ 03755, USA_\

*: Corresponding author: \
Marius ’t Hart\
Centre for Vision Research\
York University\
Toronto, ON M3J1P3\
Canada\
thartbm@gmail.com\

# Abstract

If a gabor pattern drifts in one direction while its internal texture drifts in the orthogonal direction, its perceived direction deviates strongly from its true direction and is instead some combination of its real external motion and its internal motion (Tse & Hsieh, 2006). The perceived location of such a double-drift stimulus periodically resets toward the veridical position. There may indeed be a simple perdiodicity underlying these resets. However, since both the veridical and perceived location are available to the brain, these resets may also occur when these two positions are too far from each other to come from the same source stimulus. Here we test which of the two explanations is more likely. In a first experiment, we had participants continuously (for XX min) track the position of the perceived position of the stimulus. While the orientation of the trajectories reflect stimulus properties, there were no spontaneous nor any evoked resets.  That’s why, in a second experiment, we had participants re-trace the perceived trajectory of a short stimulus presentation, or indicate the initial direction of the stimulus they perceived. We found resets here, and they indicate a set spatial limit to the distance between the veridical and perceived location of the double-drift stimulus.

# Introduction

Encoding the position of objects in the world is necessary for the myriad of visually guided motor tasks that we do every day. It is no surprise then that the brain has several mechanisms to gauge and update the position of objects in the world. For example, when a target is moving, its motion can influence it perceived location, making it appear to be slightly ahead of its true location (e.g. the flash lag, Nijhawan, 1994; or the flash grab effect, Cavanagh & Anstis, 2013). These cases of position extrapolation may be functional, helping the motor system overcome inevitable neural delays in targeting a moving object (Nijhawan, 1994; Duhamel et al., 1992). A stronger and very different motion-induced position shift arises when a moving gabor with internal motion orthogonal to its path is viewed in the periphery (Tse & Hsieh, 2006; Shapiro et al, 2010; Gurnsey & Baird, 2012; Kwon et al, 2015; Lisi & Cavanagh, 2015). This double-drift stimulus (Fig. 1, top left) generates extreme misjudgments of the moving target’s location which may deviate by as much as several degrees of visual angle from its true location. Surprisingly, this extraordinary perceptual illusion does not affect eye movements to the gabor: saccades to these targets are determined by their physical not their perceived location (Lisi & Cavanagh, 2015). This result indicates that both true and illusory locations are available in the brain. Indeed, although immediate eye-movements go towards the true position, delayed eye movements and pointing go to the perceived position [Lisi & Cavanagh, 2017; Massendari et al, 2018].

```{r echo=F, fig.height=8, fig.width=8, message=FALSE, error=FALSE, warning=FALSE, fig.cap="**Figure 1: Stimuli and Setup.** _Top left:_ a gabor with internal motion (phase shifts) and external motion, that is viewed peripherally, appears to follow a path deviating from the true path. Eye-movements immediately following disappearance of the gabor land at the true position, but pointing movements and delayed eye-movements go to the illusory position. _Bottom left:_ at some point, the perceived position will 'reset' towards the true position. For some people this takes the form of a 'jump' back to the true position, for others it means the illusory position doesn't deviate more from the true position, and there are some in between version. _Top right:_ if these resets are due to a spatial limit on the size of the illusory position shift, the points of the resets should have a constant X coordinate. _Middle right:_ if they occur after some period of time, they should have a constant Y coordinate. _Bottom right:_ we have people track or re-trace their percept and try to isolate reset points from the drawn trajectories to figure out where or when resets occur."}
library(magick)
Fig1 <- image_read_svg('doc/stimulus_reset.svg')
par(mai=c(0,0,0,0))
plot(Fig1)
```

The strength of the double-drift probably arises because of the very poor location information for a gabor pattern in the periphery when it has the same mean luminance as the background. With poor positional certainty, the motion information contributes to the location estimate, generating a shift in position. If the gabor itself is not moving (De Valois & De Valois, 1991), the shift saturates after about 100 ms (Chung et al., 2007) perhaps because the accumulating evidence of actual location of the stationary gabor patch is sufficient to anchor its position. However, when the gabor itself also moves as in the double-drift stimulus, there is no stable location information to anchor the position estimate and the build-up continues away from its physical location to reach deviations of several times the size of the gabor. At some point, the illusion resets, the perceived location jumps back to its physical location and then accumulation begins again (Fig. 1, bottom left). This long duration before resetting is well beyond the integration period of other visual effects and into the range of short-term memory. Here, we first examine the double-drift illusion by asking participants to trace out its path with a stylus on a tablet that is optically aligned with the monitor surface but with the hand not visible. Earlier results for pointing to the double-drift target (Lisi & Cavanagh, 2017) show that the illusion persists for manual pointing. However, with online tracking, the presence of the hand near the location of the perceived gabor may interact with the illusion.

The ability to manually track the perceived position will also allow us to study the resets – the adjustment of perceived location when the illusion has deviated too much or too long from the physical location. It could be that the imprecise, but true, retinal position signal limits the illusion, or it could be that – like other illusions, such as bi-stable stimuli – there is a simple periodicity underlying changes in perception. Here we capture where and when observers experience resets by having them draw the perceived trajectory of the stimulus after having seen it (a retrace procedure). This will allow us to distinguish between a spatial limit to the illusion or a temporal limit.

# Experiment 1: Continuous tracking
In this experiment, two tasks were performed by two separate groups of participants. Both tasks asked participants to use a stylus on a drawing tablet to track a moving gabor with some amount of internal motion that elicited the double-drift illusion. The idea was that whenever spontaneous resets occur, this would be recorded in the tracking data.

## Participants

For the first task 4 participants were recruited from the lab (3 female; ages 21 – 31, mean: 25.75). For the second task,  6 participants were recruited from the lab (4 female; ages 19 – 26, mean: 21.6). All participants reported to be right handed and have either normal or corrected to normal vision.  Procedures were approved by York’s Human Participants Review Committee, and all participants provided prior, written, informed consent.

## Setup
Participants used a stylus on a horizontal drawing tablet (Intuos Pro) to indicate where they perceived the location of a gabor (see Fig 1e). Stimuli were displayed on a downward facing LED monitor (30 Hz, XX”, 1440x900), level with the drawing tablet. These were observed through an upward facing mirror placed exactly between the tablet and monitor, so that the stimuli appear to be in the same plane as the hand.
- size of stimuli in cm, speed in cm/s (with guesstimates of dva?)

## Tasks

Two similar tasks were used in this experiment. In the first task, 5 different amounts of internal motion were used, including no internal motion as a baseline (-3, -1, 0 , 1 and 3 cps). The stimulus started in the middle, and moved forward and backward several times including 5 complete passes of 13.5 cm (~XXX dva) and 2 seconds each. Each stimulus was repeated 8 times, with randomized position of the fixation point (left or right) of the stimulus and counterbalanced initial external motion direction (forward or backward). This task allowed testing how sensitive online tracking is to the illusion, and served to gauge what resets would look like in tracking data.

In the second task, only the highest internal motion was used (-3 or 3 cps), trials include only 3 complete passes, but these took 4 seconds each. As before, the fixation point (left or right) was random for each trial and initial external motion (forward or backward) was counterbalanced across trials.

## Analyses

All analyses were done in R 3.6.1 [@Rproject]

## Results

In a first experiment, participants were asked to continuously track the position of a gabor that moved forward and backward along the Y-axis (of the horizontal screen) for the duration of X seconds, using a stylus on a drawing tablet. Before trial onset, the participant moved the stylus to the middle of the screen (where the gabor would appear), and was then asked to fixate a point on the left or right of the screen, and given X seconds to do so. Since the screen was seen via a mirror, the stimulus appeared to move on a horizontal plane at the same depth as the drawing tablet. The internal motion of the gabor was inverted at the far ends of the external motion, and could be X, Y, 0, -Y or -X. Since each moving gabor was shown for Z seconds, it completed A full and B half sweeps of the path (or C cycles), which should allow for spontaneous resets to occur frequently.

Furthermore, since a short disappearance of the stimulus (~250 ms or more?) has been shown to evoke a reset (XXXX; XXXX), these were included in some of the stimuli (X/Y). Each stimulus had either 0 disappearances of the stimulus (X%), or one in the second (X%), third(X%) or fourth (X%) full sweep, and it would either occur at 25% or 75% of the sweep, but never close to the middle where the perceived and illusory position should be the same. This allows characterizing what a forced reset looks like in the trajectories, and then detect spontaneous resets, that still have a lot of opportunity to occur.

In Figure X, we plot the full trajectories of all participants in each of the 5 conditions (internal motion), as well as the distribution and average instantaneous heading direction during the middle second of each 2-second pass of the stimulus. In the 0 internal motion condition the physical and perceived path coincide and there is no illusion. Trajectories from other conditions with heading in the 0 internal motion confidence interval may be evidence of a reset. Trajectories where this occurs are drawn on top of the heading distribution for the group. The average angle of the tracked path is superimposed as both a line, and a numeral (in degrees deviation from straight forward). Both from the raw trajectories as well as from the averaged angles it becomes clear that continuous tracking is sensitive to the strength of the illusion, and this is confirmed by a repeated-measures F-test on the average angles, using internal motion as a factor (F(df)=XXX, p=XXX, eta=XXX). This means we can use the averaged angle for each of the internal motion conditions as a measure of the strength of the illusion for a subset of participants that also took part in further tasks.

However, only a few potential resets / no clear resets were captured by the continuous tracking data. It is possible that resets were perceived but were too quick to respond to in the continuous drawing. Alternatively, since the hand was at the positioned at the perceived location of the gabor this may have provided an additional position signal that stabilized the illusion, preventing resets from occurring. Hence, we performed a second experiment that would counter both explanations for the absence of (spontaneous) resets.

# Experiment 2: Re-tracing

In a second experiment, we used a retracing where the participants drew what they saw of a single pass trajectory after it was over. This provided more time for the participants to reproduce the resets if there were any – without having to trace them in real time. The double-drift gabor started at the bottom of the display, close to the participant, and moved forward along the horizontal display away from the participant. There were two response conditions. In one, the participant re-traced the path they saw the stimulus take, while in the other, they indicated the angle of the first part of the stimulus motion by orienting a line originating from the starting point of the gabor’s path. The latter has already been done [XXXX] so this allows comparison of the re-tracing responses with earlier work. In the meantime, Nakayama & Holcombe [XXXX] have also used a re-tracing paradigm to study triggered resets.

## Participants

- 9 (exp 1; x female; age: y-z) + 9 (exp 2; x female; age y-z)
- Informed consent, ethics board

## Setup

The same hardware was used as in the continuous tracking experiment, except that in one of the tasks, participants could now respond using the arrow keys on a keypad.

## Tasks

Participants alternated between blocks of two kinds of trials (in counterbalanced order). In both types of trial, the gabor was shown starting close to the participant and moving away in a forward direction, for a single pass along the workspace, taking 4 seconds. During presentation of the gabor, the participant was instructed to maintain fixation to the left or right of the stimulus. After the gabor disappeared, the participant could break fixation, and was to report the percept in one of two ways. In one kind of block, the participant re-traced the perceived path of the gabor using the stylus. The drawn path was shown as a line on the screen. Participants could “key in” their re-traced percept by pressing ENTER or start over by pressing the ESCAPE key. In the other kind of block, participants reported the initial direction of the gabor (before any resets occurred), by using the LEFT and RIGHT arrow keys to change the orientation of a line of XX cm, with it’s origin in the same spot as where the gabor started moving. Participants could change the line’s orientation it until satisfied and then “key in” their response by pressing ENTER. The arrow report task has been used before (ref?) and can be used as validation of this task. The re-tracing task has also been used before, but in order to test evoked resets (Nakayama & Holcombe, 2019).

## Results

In Figure X, we observe that the data appear to show many resets. We detect them using X, Y, Z algorithm.

[Actual results] For the majority of participants (X/9) we see that most trials show at least one reset (>90%?). In the remainder we still see >50% of the trials have a reset.

The points at which resets appear can be converted to angle relative to the origin of the stimulus, and this should correspond to the [arrow responses]. Do correlation.

It may be that during online tracking, proprioceptive information on the location of the hand is included in the estimate of the location of the stimulus. This may have decreased the likelihood of a reset, and may also changed the strength of the illusion. Since some participants did both experiments (X/9), we can at least test if illusion strength is different for online tracking and offline re-tracing.

# Discussion

We set out to test if spontaneous resets in the double-drift illusion reflect discrepancies between two position signals, occurring at some given size of this discrepancy, or if they simply occur after some time elapses, analogous to other illusions or bi-stable stimuli. To measure the location of resets we first ask participants to track the perceived location of the double-drift stimulus continuously. While the direction of the tracking movements scaled with the internal motion of the stimulus (was sensitive to the strength of the illusion), there were no clear reset points in the tracking data. In a second experiment we ask participants to draw the perceived path of the stimulus after a single pass. In this re-tracing experiment the drawn traces reflect the strength of the illusion [just] as well, and in most trials there is a clear reset point. The positions of these reset points do no vary much with elapsed time, but are better explained as triggered when the distance between perceived and true position of the stimulus exceeds some threshold. This indicates that while the perceived location of a moving object can be heavily biased by its well-defined, internal motion, an uncertain (probably retinal?) position signal is still used to keep the perceived position of moving objects close to reality.

§3 (one computational model):

- Explain paper with model by Kwon, Tadin & Knill, and VSS poster with fat-tailed noise.

While this model produces reasonable resets, the properties of resets, such as their spatial or temporal distributions have not been characterized yet. An alternative explanation for resets could be that the true but imprecise position that is available for eye-movements can limit or reset the perceived position under some circumstances. In order to get at the properties of resets

If the double-drift illusion is taken as a vector average between the external motion and some proportion of the internal motion, the initial path of the stimulus can be modelled as a straight line from the origin at some angle. This straight line segment should intersect the position of the first reset. We will now see if resets occur at a given distance perpendicular to the path of the real stimulus, after a given amount of time, or perhaps a combination of the two.

This means that it seems that resets occur predominantly at some given discrepancy between the true and illusory position of the stimulus, with a small contribution of time. It may that after seeing the illusion for some time, a ‘source estimation’ mechanism can no longer account for both the perceived, illusory position and the imprecise retinal position, as coming from the same stimulus. Then the visual system changes the position of the stimulus to the position derived from retinotopic information (instead of opting for two separate sources, which is highly unlikely in this case), resulting in a “reset”.

## Conclusion

We find that online tracking of the double-drift stimulus is sensitive to illusion strength, but doesn't display clear resets. We do find spontaneous resets in single-pass retracing. The illusion appears to resets predominantly at a spatial distance from the true stimulus location, suggesting that resets resolve a source attribution conflict between the true and illusory position of the stimulus.

# Acknowledgements

The research was support by funding from NSERC Canada to PC and DYPH , as well as by the Department of Psychological and Brain Science, Dartmouth College (PC).

# References {#references .unnumbered}

### 2020-02-12 DoubleDriftResetLimits.Rmd
rename project file

@@
 output:
+  html_notebook: default
   word_document: default
-  html_notebook: default
   html_document:

### 2020-02-12 R/data.R
added files with demographics of participants

@@ -22,7 +22,9 @@ downloadRawData <- function(overwrite=TRUE, experiments=c('bounded_tracking', 'b
         'bounded_tracking_p05.csv' = 'https://osf.io/trzdx/download',
         'bounded_tracking_p06.csv' = 'https://osf.io/qywdc/download',
         'bounded_tracking_p07.csv' = 'https://osf.io/59pcd/download',
-        'bounded_tracking_p08.csv' = 'https://osf.io/sfjc6/download'
+        'bounded_tracking_p08.csv' = 'https://osf.io/sfjc6/download',
+        
+        'bounded_tracking_participants.csv' = 'https://osf.io/gtycq/download'
         
       )
       
@@ -57,7 +59,9 @@ downloadRawData <- function(overwrite=TRUE, experiments=c('bounded_tracking', 'b
         'onepass_V4_p08.csv' = 'https://osf.io/vpw3n/download',
         'onepass_V4_p09.csv' = 'https://osf.io/tm3c8/download',
         'onepass_V4_p10.csv' = 'https://osf.io/8u9rb/download',
-        'onepass_V4_p11.csv' = 'https://osf.io/v4d8p/download'
+        'onepass_V4_p11.csv' = 'https://osf.io/v4d8p/download',
+        
+        'onepass_V4_participants.csv' = 'https://osf.io/e2k8c/download'
         
       )
       
@@ -83,7 +87,9 @@ downloadRawData <- function(overwrite=TRUE, experiments=c('bounded_tracking', 'b
         'bounded_tracking_V2_p10.csv' = 'https://osf.io/8g4q9/download',
         'bounded_tracking_V2_p11.csv' = 'https://osf.io/axqcy/download',
         'bounded_tracking_V2_p12.csv' = 'https://osf.io/fjpz6/download',
-        'bounded_tracking_V2_p13.csv' = 'https://osf.io/cyj49/download'
+        'bounded_tracking_V2_p13.csv' = 'https://osf.io/cyj49/download',
+        
+        'bounded_tracking_V2_participants.csv' = 'https://osf.io/jzp5m/download'
         
         # 'reset_p01.csv' = 'https://osf.io/739dp/download',
         # 'reset_p02.csv' = 'https://osf.io/w7kht/download',

### 2020-02-12 R/models.R
added some comments

@@ -6,6 +6,8 @@ library('optimx')
 
 resetXfromYlim <- function(par,slopes) {
   
+  # time limit
+  
   Ly <- par['Ly']
 
   X <- Ly / slopes
@@ -16,6 +18,8 @@ resetXfromYlim <- function(par,slopes) {
 
 
 resetYfromXlim <- function(par,slopes) {
+  
+  # space limit
   
   Lx <- par['Lx']
   

### 2020-02-12 Manuscript.Rmd
convert manuscript to r markdown format

@@
     df_print: paged
+  pdf_document: default
   word_document: default
-  pdf_document: default
 bibliography: bibliography.bib
+csl: journal-of-vision.csl
 ---
@@
 # And finally some code for the re-tracing experiment...
-#source('R/retrace_onePass_V4.R')
+source('R/retrace_onePass_V4.R')
 # As well as models fit to that data:
@@
 
-If a gabor pattern drifts in one direction while its internal texture drifts in the orthogonal direction, its perceived direction deviates strongly from its true direction and is instead some combination of its real external motion and its internal motion (Tse & Hsieh, 2006).
+If a gabor pattern drifts in one direction while its internal texture drifts in the orthogonal direction, its perceived direction deviates strongly from its true direction and is instead some combination of its real external motion and its internal motion [@Tse2006].
 The perceived location of such a double-drift stimulus periodically resets toward the veridical position.
@@
 For example, when a target is moving, its motion can influence it perceived location, making it appear to be slightly ahead of its true location (e.g.
-the flash lag, Nijhawan, 1994; or the flash grab effect, Cavanagh & Anstis, 2013).
-These cases of position extrapolation may be functional, helping the motor system overcome inevitable neural delays in targeting a moving object (Nijhawan, 1994; Duhamel et al., 1992).
-A stronger and very different motion-induced position shift arises when a moving gabor with internal motion orthogonal to its path is viewed in the periphery (Tse & Hsieh, 2006; Shapiro et al, 2010; Gurnsey & Baird, 2012; Kwon et al, 2015; Lisi & Cavanagh, 2015).
+the flash lag, @Nijhawan1994; or the flash grab effect, @Cavanagh2013).
+These cases of position extrapolation may be functional, helping the motor system overcome inevitable neural delays in targeting a moving object [@Nijhawan1994; @Duhamel1992].
+A stronger and very different motion-induced position shift arises when a moving gabor with internal motion orthogonal to its path is viewed in the periphery [@Tse2006; @Shapiro2010; @Gurnsey2012; @Kwon2015; @Lisi2015].
 This double-drift stimulus (Fig.
 1, top left) generates extreme misjudgments of the moving target’s location which may deviate by as much as several degrees of visual angle from its true location.
-Surprisingly, this extraordinary perceptual illusion does not affect eye movements to the gabor: saccades to these targets are determined by their physical not their perceived location (Lisi & Cavanagh, 2015).
+Surprisingly, this extraordinary perceptual illusion does not affect eye movements to the gabor: saccades to these targets are determined by their physical not their perceived location [@Lisi2015].
 This result indicates that both true and illusory locations are available in the brain.
-Indeed, although immediate eye-movements go towards the true position, delayed eye movements and pointing go to the perceived position [Lisi & Cavanagh, 2017; Massendari et al, 2018].
+Indeed, although immediate eye-movements go towards the true position, delayed eye movements and pointing go to the perceived position [@Lisi2017; @Massendari2018].
 
@@
 With poor positional certainty, the motion information contributes to the location estimate, generating a shift in position.
-If the gabor itself is not moving (De Valois & De Valois, 1991), the shift saturates after about 100 ms (Chung et al., 2007) perhaps because the accumulating evidence of actual location of the stationary gabor patch is sufficient to anchor its position.
+If the gabor itself is not moving [@DeValois1991], the shift saturates after about 100 ms [@Chung2007], perhaps because the accumulating evidence of actual location of the stationary gabor patch is sufficient to anchor its position.
 However, when the gabor itself also moves as in the double-drift stimulus, there is no stable location information to anchor the position estimate and the build-up continues away from its physical location to reach deviations of several times the size of the gabor.
@@
 Here, we first examine the double-drift illusion by asking participants to trace out its path with a stylus on a tablet that is optically aligned with the monitor surface but with the hand not visible.
-Earlier results for pointing to the double-drift target (Lisi & Cavanagh, 2017) show that the illusion persists for manual pointing.
+Earlier results for pointing to the double-drift target [@Lisi2017] show that the illusion persists for manual pointing.
 However, with online tracking, the presence of the hand near the location of the perceived gabor may interact with the illusion.
@@
 
+# Methods
+
+## Participants
+
+For the first task in experiment 1, 4 participants were recruited from the lab (3 female; ages 21 – 31, mean: 25.75).
+For the second task in experiment 1, 6 participants were recruited from the lab (4 female; ages 19 – 26, mean: 21.6).
+For experiment 2, 9 participants were recruited from the lab (6 female; ages 19 - 27, mean: 22.8).
+All participants reported to be right handed and have either normal or corrected to normal vision.
+Procedures were approved by York’s Human Participants Review Committee, and all participants provided prior, written, informed consent.
+
+## Setup
+
+Participants used a stylus on a horizontal drawing tablet (Intuos Pro) to indicate where they perceived the location of a gabor (see Fig 1e), as well as a keypad for additional responses.
+Stimuli were displayed on a downward facing LED monitor (30 Hz, 20”, 1680x1050), level with the drawing tablet.
+These were observed through an upward facing mirror placed exactly between the tablet and monitor, so that the stimuli appear to be in the same horizontal plane as the hand.
+The stimuli were phase-changing gabor patches
+- size of stimuli in cm, speed in cm/s (with guesstimates of dva?)
+
+## Analyses
+
+Experiments were run in Python 2.7 with PsychoPy [@Peirce2019] and analyses were done in R 3.6.1 [@Rproject].
+
 # Experiment 1: Continuous tracking
@@
 The idea was that whenever spontaneous resets occur, this would be recorded in the tracking data.
-
-## Participants
-
-For the first task 4 participants were recruited from the lab (3 female; ages 21 – 31, mean: 25.75).
-For the second task,  6 participants were recruited from the lab (4 female; ages 19 – 26, mean: 21.6).
-All participants reported to be right handed and have either normal or corrected to normal vision.
-Procedures were approved by York’s Human Participants Review Committee, and all participants provided prior, written, informed consent.
-
-## Setup
-Participants used a stylus on a horizontal drawing tablet (Intuos Pro) to indicate where they perceived the location of a gabor (see Fig 1e).
-Stimuli were displayed on a downward facing LED monitor (30 Hz, XX”, 1440x900), level with the drawing tablet.
-These were observed through an upward facing mirror placed exactly between the tablet and monitor, so that the stimuli appear to be in the same plane as the hand.
-- size of stimuli in cm, speed in cm/s (with guesstimates of dva?)
 
@@
 As before, the fixation point (left or right) was random for each trial and initial external motion (forward or backward) was counterbalanced across trials.
-
-## Analyses
-
-All analyses were done in R 3.6.1 [@Rproject]
 
@@
 
-## Participants
-
-- 9 (exp 1; x female; age: y-z) + 9 (exp 2; x female; age y-z)
-- Informed consent, ethics board
-
-## Setup
-
-The same hardware was used as in the continuous tracking experiment, except that in one of the tasks, participants could now respond using the arrow keys on a keypad.
-
 ## Tasks
@@
 The arrow report task has been used before (ref?) and can be used as validation of this task.
-The re-tracing task has also been used before, but in order to test evoked resets (Nakayama & Holcombe, 2019).
+The re-tracing task has also been used before, but in order to test evoked resets [@Nakayama2019].
 

### 2020-02-12 R/retrace_onePass_V4.R
working on 2-limit model

@@ -912,6 +912,8 @@ plotSequentialModel <- function(target='inline') {
 
   par <- fitResetModelSeq(slopes=df$slope, X=df$boundX_mean*13.5, Y=df$boundY_mean*13.5)
   
+  print(par)
+  
   #coords <- data.frame(X=df$boundX_mean,Y=df$boundY_mean)
   #directions <- ((90 - df$initialdirection_mean) / 180) * pi
   

### 2020-02-13 Manuscript.Rmd
added technical details to methods, worked on results

@@
 output:
+  word_document: default
+  pdf_document: default
   html_document:
     df_print: paged
-  pdf_document: default
-  word_document: default
 bibliography: bibliography.bib
@@
 
-Participants used a stylus on a horizontal drawing tablet (Intuos Pro) to indicate where they perceived the location of a gabor (see Fig 1e), as well as a keypad for additional responses.
+Participants used a stylus on a horizontal drawing tablet (Wacom Intuos Pro) to indicate where they perceived the location of a gabor (see Fig 1, bottom right), as well as a keypad for additional responses.
 Stimuli were displayed on a downward facing LED monitor (30 Hz, 20”, 1680x1050), level with the drawing tablet.
 These were observed through an upward facing mirror placed exactly between the tablet and monitor, so that the stimuli appear to be in the same horizontal plane as the hand.
-The stimuli were phase-changing gabor patches
+
+## Stimuli
+
+The stimuli were phase-changing gabor patches of XX _cm_ (150 _px_ or ~XX _dva_), based on 1.5 phases of a sine wave.
 - size of stimuli in cm, speed in cm/s (with guesstimates of dva?)
@@
 # Experiment 1: Continuous tracking
+
 In this experiment, two tasks were performed by two separate groups of participants.
 Both tasks asked participants to use a stylus on a drawing tablet to track a moving gabor with some amount of internal motion that elicited the double-drift illusion.
-The idea was that whenever spontaneous resets occur, this would be recorded in the tracking data.
+The idea was that whenever spontaneous resets ofthe illusion occur, these would be recorded in the tracking data as sudden changes in heading (cf.
+Fig 1, bottom left, the dashed purple lines).
 
@@
 Two similar tasks were used in this experiment.
-In the first task, 5 different amounts of internal motion were used, including no internal motion as a baseline (-3, -1, 0 , 1 and 3 cps).
-The stimulus started in the middle, and moved forward and backward several times including 5 complete passes of 13.5 cm (~XXX dva) and 2 seconds each.
+In the first task, 5 different amounts of internal motion were used, including no internal motion as a baseline (-3, -1, 0 , 1 and 3 _cps_).
+Before stimulus onset
+
+The stimulus started in the middle, and moved forward and backward several times including 5 complete passes of 13.5 _cm_ (~XXX _dva_) and 2 _s_ each.
+The initial external motion direction (forward or backward) was counterbalanced.
+When the external motion inverted at the far and close ends of the trajectory, the internal motion was also inverted.
 Each stimulus was repeated 8 times, with randomized position of the fixation point (left or right) of the stimulus and counterbalanced initial external motion direction (forward or backward).
@@
 
-In the second task, only the highest internal motion was used (-3 or 3 cps), trials include only 3 complete passes, but these took 4 seconds each.
-As before, the fixation point (left or right) was random for each trial and initial external motion (forward or backward) was counterbalanced across trials.
+**DOUBLE PARAGRAPH; CHECK WHAT IS MISSING**
+
+In a first experiment, participants were asked to continuously track the position of a gabor that moved forward and backward along the Y-axis (of the horizontal screen) for the duration of 12 _s_, using a stylus on a drawing tablet.
+Before trial onset, the participant moved the stylus to the middle of the screen (where the gabor would appear), and was subsequently asked to fixate a point on the left or right of the screen, and given 1.5 _s_ to do so.
+Since the screen was seen via a mirror, the stimulus appeared to move on a horizontal plane at the same depth as the drawing tablet.
+The internal motion of the gabor was inverted at the far and close ends of the physical path, where external motion was also inverted.
+Internal motion could be 3, 1, 0, -1 or -3 _cps_.
+Since each moving gabor was shown for 12 _s_, and it moved from one end to another in 2 _s_, it completed 5 full and 2 half passes of the path (or 3 cycles).
+
+In the second task, only the highest internal motion was used (±3 _cps_), trials include only 3 complete passes, but these took 4 _s_ each.
+As before, the fixation point (left or right) was random for each trial, while initial external motion (forward or backward) as well as internal motion were counterbalanced across trials.
 
@@
 
-In a first experiment, participants were asked to continuously track the position of a gabor that moved forward and backward along the Y-axis (of the horizontal screen) for the duration of X seconds, using a stylus on a drawing tablet.
-Before trial onset, the participant moved the stylus to the middle of the screen (where the gabor would appear), and was then asked to fixate a point on the left or right of the screen, and given X seconds to do so.
-Since the screen was seen via a mirror, the stimulus appeared to move on a horizontal plane at the same depth as the drawing tablet.
-The internal motion of the gabor was inverted at the far ends of the external motion, and could be X, Y, 0, -Y or -X.
-Since each moving gabor was shown for Z seconds, it completed A full and B half sweeps of the path (or C cycles), which should allow for spontaneous resets to occur frequently.
+In Figure 2 (top), we plot the average instantaneous heading during the middle second of each 2 _s_ pass (as tracking was noisy at the inversion points) for each of the 5 internal motion speeds (normalized for the external motion).
+The average angle of the tracked path is given in degrees deviation from straight forward.
+It seems that continuous tracking is sensitive to the strength of the illusion, and this is confirmed by a repeated-measures F-test on the average angles, using internal motion as a within-subjects factor (_F_(4,12)=102.6, _p_<.001, $\eta^2$=0.97).
+With FDR-corrected, paired t-tests, we find a difference in heading between 0 _cps_ internal motion and both 1 _cps_ ( _p_=.022) and -1 _cps_ ( _p_=.025) as well as between -3 and -1 _cps_ and 3 and 1 _cps_ (both _p_=<.001).
+This means we can use the averaged angle for each of the internal motion conditions as a measure of the strength of the illusion.
 
-Furthermore, since a short disappearance of the stimulus (~250 ms or more?) has been shown to evoke a reset (XXXX; XXXX), these were included in some of the stimuli (X/Y).
-Each stimulus had either 0 disappearances of the stimulus (X%), or one in the second (X%), third(X%) or fourth (X%) full sweep, and it would either occur at 25% or 75% of the sweep, but never close to the middle where the perceived and illusory position should be the same.
-This allows characterizing what a forced reset looks like in the trajectories, and then detect spontaneous resets, that still have a lot of opportunity to occur.
+The average headings are close to straight lines for all conditions.
+Since average straight trajectories could either result from uniform tracking paths, each with no signs of resets or from paths with many differently shaped, directed and timed resets, they do not prove or disprove neither the absence nor the presence of resets.
+Hence, we also show full trajectories of all participants in each of the 5 internal motion conditions.
 
-In Figure X, we plot the full trajectories of all participants in each of the 5 conditions (internal motion), as well as the distribution and average instantaneous heading direction during the middle second of each 2-second pass of the stimulus.
-In the 0 internal motion condition the physical and perceived path coincide and there is no illusion.
-Trajectories from other conditions with heading in the 0 internal motion confidence interval may be evidence of a reset.
-Trajectories where this occurs are drawn on top of the heading distribution for the group.
-The average angle of the tracked path is superimposed as both a line, and a numeral (in degrees deviation from straight forward).
-Both from the raw trajectories as well as from the averaged angles it becomes clear that continuous tracking is sensitive to the strength of the illusion, and this is confirmed by a repeated-measures F-test on the average angles, using internal motion as a factor (F(df)=XXX, p=XXX, eta=XXX).
-This means we can use the averaged angle for each of the internal motion conditions as a measure of the strength of the illusion for a subset of participants that also took part in further tasks.
+In order to detect potential resets, we consider that in the 0 internal motion condition there is no illusion, so that we measure the tracking of the veridical path of the stimulus.
+Since both strong, "jump" resets and weak "hit-the-wall" resets (see Fig 1, bottom left) would change percieved direction back to or crossing the veridical motion path, their smooth-spline interpolated paths would contain samples similar in heading to the 0 internal motion condition.
+We use a 95% confidence interval of heading in the 0 internal motion condition to select trajectories from conditions with non-zero heading that may show a reset.
+These trajectories are drawn in red (only 1 per condition/participant), but as can be seen, the deviations from the condition's dominant heading occur just before or after the external motion reversals, so they may simply reflect difficulty of tracking the stimulus, rather than resets.
 
-However, only a few potential resets / no clear resets were captured by the continuous tracking data.
-It is possible that resets were perceived but were too quick to respond to in the continuous drawing.
-Alternatively, since the hand was at the positioned at the perceived location of the gabor this may have provided an additional position signal that stabilized the illusion, preventing resets from occurring.
+Hence, no unambiguous resets were captured.
+It is possible that resets were perceived but were too quick to respond to in the continuous tracking task.
+Alternatively, trackgin may have stabilized the illusion, preventing resets from occurring.
 Hence, we performed a second experiment that would counter both explanations for the absence of (spontaneous) resets.

### 2020-02-13 R/plotTracking.R
found pixels per centimeter?

@@ -1439,7 +1439,7 @@ plotCVRdemo <- function(target='pdf') {
       
       trialdf <- ppdf[which(ppdf$trial_no == trialno & ppdf$step == 99),]
       
-      X <- trialdf$handx_pix / 38.814814815
+      X <- trialdf$handx_pix / 38.814814815 # is this pixels per centimeter?
       Y <- (trialdf$handy_pix + 262) / 38.814814815
       
       # if the endpoint Y-coordinate is negative (or close to 0?) we don't trust it

### 2020-02-19 R/common.R
added resolution and dva info (as a function)

@@ -1,4 +1,46 @@
+# 
+# # pixels per cm
+# 
+# # resolution is 1680 x 1050 pixels
+# # which spans ~43.5 x ~27 cm
+# # for width that is roughly 38.799 pixels per cm
+# # for depth that is roughly 38.888 pixels per cm
+# # on average that would be: 38.755 pixels per cm
+# 
+# PPC <- 38.754789272
+# 
 
+
+getPPCdva <- function(lo=3) {
+  
+  props <- list()
+  props['PPC'] <- 38.754789272 # pixels per cm
+  
+  # tan(A) = a/b
+  
+  dva <- expand.grid('eye.dist'  = seq(15, 25, length.out = lo), 
+                     'gabor.pos' = seq(-6.75, 6.75, length.out = lo))
+  
+  b <- 27
+  a <- dva$eye.dist + dva$gabor.pos
+  
+  # https://en.wikipedia.org/wiki/Trigonometry
+  A <- atan(a/b)
+  
+  a1 <- tan(A+((0.5/180)*pi))*b
+  a2 <- tan(A-((0.5/180)*pi))*b
+  
+  dva$cmpd <- a1 - a2
+  
+  ghw <- (100/3) / props[['PPC']]
+  
+  dva$gabor.dva <- ghw / dva$cmpd
+  
+  props[['dva']] <- dva
+  
+  return(props)
+  
+}
 
 
 getColors <- function() {
@@ -354,8 +396,8 @@ hist2d <- function(x, y=NA, nbins=c(25,25), edges=NA) {
     #cat('set edges from edges\n')
   }
   
-  #cat('length x.edges:\n')
-  #print(length(x.edges))
+  # cat('length x.edges:\n')
+  # print(length(x.edges))
   
   xbincount <- findInterval(df[,1], x.edges, rightmost.closed = T, left.open = F, all.inside = F)
   ybincount <- findInterval(df[,2], y.edges, rightmost.closed = T, left.open = F, all.inside = F)
@@ -466,3 +508,5 @@ polarHeatMap <- function(x,y,z,mincol=c(0.94,0.98,0.99),maxcol=c(0.06,0.82,0.88)
   
   
 }
+
+

### 2020-02-19 R/models.R
one function to compare models based on AIC

@@ -329,3 +329,126 @@ fitResetModelSeq <- function(slopes,X,Y) {
   return(winpar)
   
 }
+
+compareAllModels <- function(verbosity=0) {
+  
+  # get the data:
+  df <- getTimeNormalizedData()
+  
+  # convert data for simpler models:
+  coords <- data.frame('X'=df$boundX_mean,'Y'=df$boundY_mean)
+  directions <- ((90 - df$initialdirection_mean) / 180) * pi
+  slopes <- sin(directions) / cos(directions)
+  
+  
+  # get fits for all models:
+  Lx <- seq(0,1,length.out = 21)
+  Ly <- seq(0,1,length.out = 21)
+  
+  searchgridXlim <- expand.grid('Lx'=Lx)
+  searchgridYlim <- expand.grid('Ly'=Ly)
+  
+  searchgridXYlim <- expand.grid('Lx'=Lx,
+                                 'Ly'=Ly)
+  
+  
+  # get MSE for all points in search grid:
+  XlimMSE <- apply(searchgridXlim,FUN=resetXlimMSE,MARGIN=c(1),slopes=slopes,coords=coords)
+  YlimMSE <- apply(searchgridYlim,FUN=resetYlimMSE,MARGIN=c(1),slopes=slopes,coords=coords)
+  
+  XYlimMSE <- apply(searchgridXYlim,FUN=resetModelSeqMSE,MARGIN=c(1),slopes=slopes,coords=coords)
+  
+  # get 3 best points in the single limit "grids":
+  topgridXlim <- data.frame('Lx'=searchgridXlim[order(XlimMSE)[1:3],])
+  topgridYlim <- data.frame('Ly'=searchgridYlim[order(YlimMSE)[1:3],])
+
+  topgridXYlim <- searchgridXYlim[order(XYlimMSE)[1:5],]
+  
+  allXlimFits <- do.call("rbind",
+                         apply( topgridXlim,
+                                MARGIN=c(1),
+                                FUN=optimx,
+                                fn=resetXlimMSE,
+                                method='L-BFGS-B',
+                                lower=c(0),
+                                upper=c(1),
+                                slopes=slopes,
+                                coords=coords ) )
+  
+  allYlimFits <- do.call("rbind",
+                         apply( topgridYlim,
+                                MARGIN=c(1),
+                                FUN=optimx,
+                                fn=resetYlimMSE,
+                                method='L-BFGS-B',
[…]

### 2020-02-19 R/retrace_onePass_V4.R
load heading distributions if available, fixed edge-issue

@@ -296,7 +296,13 @@ getTimeNormalizedData <- function(convertSpeed=TRUE, normalizeTime=TRUE, illusio
 }
 
 
-getNormalizedHeading2Dhists <- function(segmentpoints=101, resetOnly=TRUE) {
+getNormalizedHeading2Dhists <- function(segmentpoints=101, resetOnly=TRUE, load=TRUE) {
+  
+  if (load) {
+    if(file.exists('data/onePass_V4/onePassHeading.Rds')) {
+      return(readRDS("data/onePass_V4/onePassHeading.Rds"))
+    }
+  }
   
   participants = c(2,3,4,5,6,8,9,10,11)
   
@@ -401,8 +407,10 @@ getNormalizedHeading2Dhists <- function(segmentpoints=101, resetOnly=TRUE) {
     
   }
   
-  freq2Ds[['x.edges']] <- pfreq$x.edges
-  freq2Ds[['y.edges']] <- pfreq$y.edges
+  freq2Ds[['edges']][['x.edges']] <- pfreq$x.edges
+  freq2Ds[['edges']][['y.edges']] <- pfreq$y.edges
+  
+  saveRDS(freq2Ds, file='data/onePass_V4/onePassHeading.Rds')
   
   return(freq2Ds)
   
@@ -413,7 +421,7 @@ getNormalizedHeading2Dhists <- function(segmentpoints=101, resetOnly=TRUE) {
 
 plotExampleData <- function(target='inline') {
   
-  graphics.off()
+  #graphics.off()
   
   if (target == 'pdf') {
     cairo_pdf(filename='doc/onePass_V4_boundaries.pdf',onefile=TRUE,width=8,height=4)
@@ -424,6 +432,7 @@ plotExampleData <- function(target='inline') {
   
   colors <- getColors()
   
+  # this should return polar heatmaps for all 6 conditions:
   f2ds <- getNormalizedHeading2Dhists()
   x.edges <- f2ds[['edges']]$x.edges
   y.edges <- f2ds[['edges']]$y.edges
@@ -530,7 +539,6 @@ plotExampleData <- function(target='inline') {
       
     }
     
-    
     conditionname <- sprintf('%d-%0.3f',internal_speed,external_speed)
     freq2D <- f2ds[['hists']][[conditionname]]
     
@@ -541,7 +549,7 @@ plotExampleData <- function(target='inline') {
     }
     
[…]

### 2020-02-19 exp/runOnePassExperiment_V2.py
resolution appears to be incorrect in the experiment code, uses my laptop's not the setup's monitor?

@@ -15,7 +15,7 @@ cfg['fullscr']      = True
 cfg['monitorIndex'] = 1
 cfg['flip']         = True
 
-cfg = createWindow(cfg, resolution=[1920, 1080])
+cfg = createWindow(cfg, resolution=[1920, 1080]) # this is not the real resolution?
 
 mouse = event.Mouse(win=cfg['win'])
 mouse.setVisible(False)

### 2020-02-19 R/tracking_bounded.R
minor fix in figure function

@@ -172,7 +172,7 @@ plotBoundedTracking <- function(target='inline') {
   internalSpeeds <- sort(unique(stdsegments$internalSpeed))
   participants <- sort(unique(stdsegments$participant))
   
-  par(mar=c(4.1,4.1,2.1,0.1))
+  par(mar=c(4.1,4.1,2.1,0.1),mfrow=c(1,1))
   
   plot(-1000,-1000,main='',ylim=c(0.5,length(participants)+2),xlim=c(0.5,5.5),ylab='participant',xlab='internal speed [cps]',asp=1,bty='n',ax=F)
   

### 2020-02-19 Manuscript.Rmd
added figures and "modeling" section

@@
 output:
-  word_document: default
   pdf_document: default
@@
     df_print: paged
+  word_document: default
 bibliography: bibliography.bib
 csl: journal-of-vision.csl
+link-citations: yes
 ---
@@
 It is no surprise then that the brain has several mechanisms to gauge and update the position of objects in the world.
-For example, when a target is moving, its motion can influence it perceived location, making it appear to be slightly ahead of its true location (e.g.
-the flash lag, @Nijhawan1994; or the flash grab effect, @Cavanagh2013).
+For example, when a target is moving, its motion can influence it perceived location, making it appear to be slightly ahead of its true location, e.g.
+the flash lag, [@Nijhawan1994]; or the flash grab effect, [@Cavanagh2013].
 These cases of position extrapolation may be functional, helping the motor system overcome inevitable neural delays in targeting a moving object [@Nijhawan1994; @Duhamel1992].
@@
 
-```{r echo=F, fig.height=8, fig.width=8, message=FALSE, error=FALSE, warning=FALSE, fig.cap="**Figure 1: Stimuli and Setup.** _Top left:_ a gabor with internal motion (phase shifts) and external motion, that is viewed peripherally, appears to follow a path deviating from the true path.
+```{r fig1, echo=F, fig.height=8, fig.width=8, message=FALSE, error=FALSE, warning=FALSE, fig.cap="**Figure 1: Stimuli and Setup.** _Top left:_ a gabor with internal motion (phase shifts) and external motion, that is viewed peripherally, appears to follow a path deviating from the true path.
 Eye-movements immediately following disappearance of the gabor land at the true position, but pointing movements and delayed eye-movements go to the illusory position.
@@
 Stimuli were displayed on a downward facing LED monitor (30 Hz, 20”, 1680x1050), level with the drawing tablet.
-These were observed through an upward facing mirror placed exactly between the tablet and monitor, so that the stimuli appear to be in the same horizontal plane as the hand.
+The stimuli were observed via an upward facing mirror placed exactly between the tablet and monitor, so that the stimuli appear to be in the same horizontal plane as the hand, or the tip of the stylus.
 
@@
 
-The stimuli were phase-changing gabor patches of XX _cm_ (150 _px_ or ~XX _dva_), based on 1.5 phases of a sine wave.
+The stimuli were phase-changing gabor patches with a gaussian mask with an sd of 0.43 _cm_ (or ~XX _dva_), using 1.5 phases of a sine wave in the full stimulus, so that each sine phase equals ~0.582 _cm_ or ~XX _dva_.
+By applying both real motion and phases changes of the gabor, it becomes a 'double-drift stimulus' where the perceived path deviates from the real path.
+The gabor changed phase at different rates, called 'internal motion' here, specified in cycles per second ( _cps_).
+The real displacements of the stimulus are along the Y axis of the monitor, keeping the X coordinate constant at the middle of the screen, so that a stimulus without internal motion would appear to move forward and backward on the horizontal display.
+This real motion was restricted to the central 13.5 _cm_ (~ XX _dva_) and was set so that stimulus would move from either extreme to the other (a motion we call a 'pass') in 2, 3 or 4 _s_, corresponding to speeds of 6.75, 4.5 and 3.75 _cm/s_ or roughly XX, XX or XX _dva/s_ in the middle of the screen.
+For simplicity, these will be reported as _seconds per pass_.
 - size of stimuli in cm, speed in cm/s (with guesstimates of dva?)
@@
 ## Analyses
+
+[
+
+For experiment 1, we first segmented the tracking trajectories according to stimulus passes (in between direction changes of the gabor), and removed the first and last half second.
+We then calculated the instantaneous heading (direction, disregarding velocity) of a smooth spline interpolated version of the tracking trajectories.
+This heading measure should depend on the illusion strength, and should change when a (spontaneous) reset is reported in the tracking data.
+
+]
+
+[
+
+For comparison, we use the same heading measure in experiment 2, but looking at individual trajectories reveals more points that are potential spontaneous resets.
+Using a smooth spline interpolated version of the trajectory we then detect direction changes (beyond ~0.39 _cm_ from the starting point, as the initial movement is often jittery).
+These direction changes are taken as reset points.
+We use the 54 average coordinates of the reset points in all 9 participants and 6 stimulus conditions for further analyses and modelling.
+The average line orientation is also calculated for each of those 54 combinations of partcipants and stimulus conditions.
+
+]
 
@@
 
+```{r fig2, echo=FALSE, message=FALSE, fig.width=4.5, fig.height=6, fig.cap="\\label{fig:fig2}**Figure 2.
+Heading and Trajectories in Experiment 1a.** **Top:** Average heading across the middle 1 _s_ of each pass.
+Distance from origin is time, angle is angle.
+In blue lines the average heading across participants, in light blue a polar heatmap giving an impression of the distribution of heading.
+**Bottom:** All individual trajectores.
+_Red_: Highlighted trajectories from conditions with non-zero internal speed, where at least some heading samples fell in the 95% confidence interval for that participants' heading distribution in zero internal speed condition."}
+plotBoundedTracking()
+```
+
 The average headings are close to straight lines for all conditions.
@@
 
-Hence, no unambiguous resets were captured.
+In a second task, we only used the higher internal speed (±3 _cps_) and increased the duration of a single pass of the stimulus from 2 _s_ to 4 _s_, both the increase the occurence of spontaneous resets.
+Average headings and heading distributions for 6 participants can be seen in Figure 3.
+
+```{r fig3, echo=FALSE, message=FALSE, fig.width=8, fig.height=4, fig.cap="\\label{fig:fig3}**Figure 3.
+Heading in Experiment 1b.** Average heading and heading distribution across the middle 3 _s_ of each pass.
+Distance from origin is time, angle is angle.
+In dark blue lines the average heading, in light blue a polar heatmap giving an impression of the distribution of heading.
+**Left:** Data for all participants.
+**Right:** Group average."}
+plotBoundedTracking_V2()
+```
+
+Except for the first part of participant 12's trajectories, all heading data matched that of the 3 _cps_ condition in experiment 1a.
+Those first few samples in participant 12 could reflect a participant specific delay in tracking the stimulus position.
+In other words, no unambiguous resets were captured in either experiment.
 It is possible that resets were perceived but were too quick to respond to in the continuous tracking task.
-Alternatively, trackgin may have stabilized the illusion, preventing resets from occurring.
+Alternatively, tracking may have stabilized the illusion, preventing resets from occurring.
 Hence, we performed a second experiment that would counter both explanations for the absence of (spontaneous) resets.
@@
 
-In a second experiment, we used a retracing where the participants drew what they saw of a single pass trajectory after it was over.
-This provided more time for the participants to reproduce the resets if there were any – without having to trace them in real time.
-The double-drift gabor started at the bottom of the display, close to the participant, and moved forward along the horizontal display away from the participant.
-There were two response conditions.
-In one, the participant re-traced the path they saw the stimulus take, while in the other, they indicated the angle of the first part of the stimulus motion by orienting a line originating from the starting point of the gabor’s path.
-The latter has already been done [XXXX] so this allows comparison of the re-tracing responses with earlier work.
-In the meantime, Nakayama & Holcombe [XXXX] have also used a re-tracing paradigm to study triggered resets.
+In a second experiment, we used a retracing task where the participants drew what they saw of a single pass stimulus immediately after presentation.
+This provided more time for the participants to reproduce any resets, i.e.
+without the need for real-time tracking.
+In this task, the double-drift gabor started at the near end of the display, and moved forward along the horizontal display away from the participant.
+The gabor could have one of two external speeds (corresponding to 3 or 4 _s_ presentation time), and one of three internal speeds (2, 3 or 4 _cps_).
+Apart from re-tracing the perceived paths, in half the blocks participants indicated the angle of the first part of the stimulus motion by orienting a line (75 _px_ = 3 _cm_) originating from the starting point of the gabor’s path.
+This task has been used before [XXXX].
+In the meantime, Nakayama & Holcombe [-@Nakayama2019] have also used a re-tracing paradigm to study triggered resets.
 
@@
 
-In Figure X, we observe that the data appear to show many resets. We detect them using X, Y, Z algorithm.
+In Figure 4, for each of the 6 combinations of internal speed (2, 3 or 4 _cps_) and external speed (3 or 4 _s_ stimulus pass) we show the data for one example participant, as well as the heading distribution for all participants.
+While the heading distributions don't show resets, for many of the participants, we can see that the re-traced paths do show resets, many of which are detected by our algorithm.
+
+```{r fig4, echo=FALSE, message=FALSE, fig.width=8, fig.height=5, fig.cap="\\label{fig:fig4}**Figure 4.
+Example Trajectories and Heading in Experiment 2.** For each of the six conditions, defined by 3 internal speeds (2, 3 and 4 _cps_) and two external speeds (corresponding to 3 and 4 _s_ passes), we show one example participant's trajectories and reset points, as well as the heading distribution."}
+plotExampleData()
+```
 
@@
 
-The points at which resets appear can be converted to angle relative to the origin of the stimulus, and this should correspond to the [arrow responses]. Do correlation.
+First, we consider the line angle responses as a measure of illusion strength.
+The strength of the illusion should depend on both the internal and external speed of the double drift stimulus.
+As can be seen in Figure 5 (left), at least the internal speed, and perhaps the external speed may predict illusion strength as measured by line angles.
+To test this, we perform a mixed-design ANOVA on line angle, using both internal and external speed as within-subject factors.
+We find main effects of both internal speed ( _F_(2,16)=12.7; _p_<.001, $\eta^2$=0.049) and external speed ( _F_(1,8)=6.89; _p_=.03; $\eta^2$=0.031), and no interaction.
+That is: we verify that illusion strength varied systematically with our manipulations of the stimulus.
 
-It may be that during online tracking, proprioceptive information on the location of the hand is included in the estimate of the location of the stimulus.
+```{r fig5, echo=FALSE, message=FALSE, fig.width=8, fig.height=4, fig.cap="\\label{fig:fig5}**Figure 5.
+Measures of Illusion Strength** Line angle responses and reset point direction should both measure illusion strength.
+**Left:** Line Angle response as a function of internal and external speed, averaged across 9 participants in each of the 6 conditions.
+Shaded areas denote 95% confidence intervals.
+**Right:** Angles of samples halfway to a reset over line angle response.
+Each dot corresponds to one participants average in one of the 6 conditions.
+Diagonal line indicates a weighted regression (weighted by one over the square of the line angle response to counter heteroscedasticity), and the shaded area shows a 95% confidence interval for the regression."}
+plotValidity()
+```
+
+We also take the angle of the sample on each re-tracing trajectory that is halfway between the origin and the detected reset, as an alternative measure of illusion strength.
+If both the angle half-way to reset and line angle responses are good measures of illusion strength, they should be related.
+As can be seen from the data in Figure 5 (right), there might be a relation between the two variables, but there is also heteroscadasticity.
+We counter this by doing a weighted regression, with the weights set to one over the squared line angle ($\frac{1}{\alpha^2}$).
+Reported line angles predicted reset point direction (slope=0.71; _t_(49)=4.54, _p_<.001; $R^2$=.282; _F_(1,49)=20.6), verifying that the two measures - while imperfect - are likely both indicators of double-drift illusion strength.
+From here on, we will only use the angle of samples half-way to detected reset points as measure of illusion strength.
+
+## Single Limits on the Double-Drift Illusion
+
+Now we look at the distribution of reset points, and whether they can be explained by a temporal or spatial limit.
+In Figure 6 (left), we can see that the reset points vary more along the Y-axis than the X-axis.
+That is, predicting reset points from a single Y coordinate (time-cutoff) would leave larger residuals than predicting reset points from a single X coordinate (or spatial cutoff).
+In fact, the standard deviation of reset point X-coordinates is about 21.9% of teh standard deviation of reset point Y-coordinates.
+This already indicates that a spatial limit explains about 4 times more of the variance in reset points than a temporal limit.
+
+```{r fig6, echo=FALSE, message=FALSE, fig.width=8, fig.height=6, fig.cap="\\label{fig:fig6}**Figure 6.
+Reset Points and Single Limits** **Left:** The average reset points for each participant and stimulus condition displayed spatially.
+Y-coordinates have been converted to time, in order to normalize for the two external speeds.
+Transparent purple lines indicate the standard deviation of coordinates.
+Dashed vertical red line indicates the median X-coordinate (1.6 _cm_).
+Dashed horizontal blue line indicates the median Y-coordinate (1.4 _s_).
+**Top right:** X-coordinates predicted from a single temporal limit.
+**Bottom right:** Y-coordinates predicted from a single spatial limit."}
+plotResetPoints()
+```
+
+Nevertheless, we test two simple models, based on straighforward trigonometry.
+First, we calculate the slope of a line through the origin and a sample halfway to the reset point.
+Then, if the perceived path of the gabor falls on a straight line, and reset points occur at a set limit, then a temporal limit would predict X coordinates:
+
+$x = L_y / s$
+
+And a spatial limit would predict Y coordinates:
+
+$y = L_x \cdot s$
+
+Where $L_y$ and $L_x$ are a temporal (y) and spatial (x) limit that are free parameters, and s corresponds to the slope.
+If we fit these simple models by minimizing the mean square error (as the distance in centimeters between predicted and actual reset points), the optimal parameters indicate a temporal limit of ~1.4 _s_ (AIC=258.4) and a spatial limit of ~1.7 _cm_ (AIC=217.7).
+This is close to the median values of the respective coordinates.
+The model with the spatial limit fits the data better than the model with a temporal limit ( _p_<.001, relative log-likelihood based on AICs).
+This again indicates that a spatial limit explains more of the data.
+
+However, it may be that after crossing a spatial limit, it takes some given amount of time before a reset of the illusion is perceived.
+This would mean that both limits explain separate parts of the data.
+We model this by finding optimal parameter values for both a spatial limit and a temporal limit simultaneously, when minimizing the distance between predicted and atcual reset points (in centimeters).
+The equations in the model are:
+
+$x = L_x + L_y / s$
+
+$y = L_x \cdot s + L_y$
+
+Figure 7 shows both the actual reset point coordinates (dots), given the illusion strength as the angle to a sample half-way to the reset point, as well as the model's predictions of reset point coordinates (lines).
+
+```{r fig7, echo=FALSE, message=FALSE, fig.width=6, fig.height=6, fig.cap="\\label{fig:fig7}**Figure 7. Reset Points and Combined Limits**"}
+plotSequentialModel()
+```
+
+The limits this model generates are a 1.42 _cm_ spatial limit and a 0.28 _s_ temporal limit.
+While the MSE and AIC of the model are lower than that of the model with the single spatial limit, it does not explain the data better ( _p_=.53).
+This means that a simple spatial limit accounts for most of the data.
+It is possible that the way we've modeled the additional temporal delay after crossing the spatial limit is not accurate, as it may follow a poisson distribution, which we do not account for here.
+However, regardless of whether or not a temporal limit adds anything, a spatial limit appears to explain the majority of spontaneous double-drift resets.
+
+```{r comment, echo=FALSE, message=FALSE}
+#It may be that during online tracking, proprioceptive information on the location of the hand is included in the estimate of the location of the stimulus.
 This may have decreased the likelihood of a reset, and may also changed the strength of the illusion.
 Since some participants did both experiments (X/9), we can at least test if illusion strength is different for online tracking and offline re-tracing.
+```
 

### 2020-05-28 DoubleDriftResetLimits.Rmd
added time-model exploration plot

@@
 
-We've minimized depencies on other packages, but there are two that help. Of course, you can skip the chunks that use them, or just install them by uncommenting the second line of this chunk:
+We've minimized depencies on other packages, but there are three that help. Of course, you can skip the chunks that use them, or just install them by uncommenting the second line of this chunk:
 
@@
 source('R/models.R')
+# Comparison with Sirui Liu's data:
+source('R/time.R')
 ```
@@
 In order to determine the point of reset automatically, we took the first sign change of the X coordinate.
-Since the start of the tracing often was somewhat "shaky" we disregarded sign changes in the first XXX cm.
+Since the start of the tracing often was somewhat "shaky" we disregarded sign changes in the first 75 px or XXXX cm.
 Since the trajectories are measured in discrete pixel values, sign changes were clearer after slight smoothing of the trajectories, using splines.
+
+```{r}
+retrace <- read.csv('data/onePass_V4/onePass_V4_re-trace.csv', stringsAsFactors = F)
+retrace$value <- 1
+retrace$value[which(is.na(retrace$boundX))] <- 0
+resetcounts <- aggregate(value ~ participant, data=retrace, FUN=sum)
+resetcounts$value <- resetcounts$value / (6*6*4)
+print(resetcounts)
+```
+
+We can also look at this, further split by condition:
+
+```{r}
+resetcountsCOND <- aggregate(value ~ participant + internalspeed + externalspeed, data=retrace, FUN=sum)
+print(resetcountsCOND)
+hist(resetcountsCOND$value)
+```
+
+This distribution appears to be bimodal.
+Since resets here depend on direction inversion, and some people may experience more "hit-the-wall resets", it seems that the lower peak could be from people with hit-the-wall resets, while the other participants perceive some return to the true position.
 
@@
 While the sequential model allows estimating the actual spatial and temporal limits, the weighted model allows estimating the relative importance of the spatial and temporal limits.
+
+# Alternative Time Model
+
+This model is based on the work by Sirui Liu and colleagues who found that when probing the location of the stimulus at a few fixed temporal offsets from stimulus onset it deviated less and less after more time.
+This is consistent with a simple model with a fixed, fairly low chance of a reset occurring at every time point.
+That is: as time goes by the likelihood of reset having occurred is larger, and as a result the average percept is closer to the true position over time.
+This model fits very well and is completely at odds with the findings here.
+So which is true?
+
+One way both may be true at the same time, is if after a reset, the illusion builds up less or not at all.
+In the current experiment people report the full path of the stimulus, so we can also look at the average deviation of the perceived position from the true position over time, continuously.
+This allows to see if it is in line with the data reported by Liu, Tse & Cavanagh [].

### 2020-05-28 Manuscript.Rmd
text edits

@@
 output:
+  word_document: default
   pdf_document: default
@@
     df_print: paged
-  word_document: default
+header-includes:
+- \usepackage{lineno}
+- \linenumbers
+- \usepackage{caption}
+- \captionsetup[figure]{labelformat=empty}
 bibliography: bibliography.bib
@@
 Here we test which of the two explanations is more likely.
-In a first experiment, we had participants continuously (for XX min) track the position of the perceived position of the stimulus.
-While the orientation of the trajectories reflect stimulus properties, there were no spontaneous nor any evoked resets.
+In a first experiment, we had participants continuously track the position of the perceived position of the stimulus, using a stylus on a drawing tablet.
+While the orientation of the recorded trajectories reflect stimulus properties, there were no spontaneous resets.
 That’s why, in a second experiment, we had participants re-trace the perceived trajectory of a short stimulus presentation, or indicate the initial direction of the stimulus they perceived.
@@
 For example, when a target is moving, its motion can influence it perceived location, making it appear to be slightly ahead of its true location, e.g.
-the flash lag, [@Nijhawan1994]; or the flash grab effect, [@Cavanagh2013].
+the flash lag [@Nijhawan1994]; or the flash grab effect [@Cavanagh2013].
 These cases of position extrapolation may be functional, helping the motor system overcome inevitable neural delays in targeting a moving object [@Nijhawan1994; @Duhamel1992].
@@
 
-The strength of the double-drift probably arises because of the very poor location information for a gabor pattern in the periphery when it has the same mean luminance as the background.
+The strength of the double-drift illusion probably arises because of the very poor location information for a gabor pattern in the periphery when it has the same mean luminance as the background.
 With poor positional certainty, the motion information contributes to the location estimate, generating a shift in position.
 If the gabor itself is not moving [@DeValois1991], the shift saturates after about 100 ms [@Chung2007], perhaps because the accumulating evidence of actual location of the stationary gabor patch is sufficient to anchor its position.
-However, when the gabor itself also moves as in the double-drift stimulus, there is no stable location information to anchor the position estimate and the build-up continues away from its physical location to reach deviations of several times the size of the gabor.
+However, when the gabor itself also moves as in the double-drift stimulus, there is no stable location information to anchor the position estimate and the build-up continues away from its physical location up to several times the size of the gabor.
 At some point, the illusion resets, the perceived location jumps back to its physical location and then accumulation begins again (Fig.
-1, bottom left).
-This long duration before resetting is well beyond the integration period of other visual effects and into the range of short-term memory.
-Here, we first examine the double-drift illusion by asking participants to trace out its path with a stylus on a tablet that is optically aligned with the monitor surface but with the hand not visible.
+1, bottom left) or it saturates.
+This long duration before resetting or saturation is well beyond the integration period of other visual effects and into the range of short-term memory.
+Given that both true and illusory positions are available, it could be that instead of some integration period, it is the interplay between the two position signals that leads to resets.
+
+Here, we examine the double-drift illusion by asking participants to trace out its path with a stylus on a tablet that is optically aligned with the monitor surface but with the hand not visible.
 Earlier results for pointing to the double-drift target [@Lisi2017] show that the illusion persists for manual pointing.
-However, with online tracking, the presence of the hand near the location of the perceived gabor may interact with the illusion.
-
 The ability to manually track the perceived position will also allow us to study the resets – the adjustment of perceived location when the illusion has deviated too much or too long from the physical location.
 It could be that the imprecise, but true, retinal position signal limits the illusion, or it could be that – like other illusions, such as bi-stable stimuli – there is a simple periodicity underlying changes in perception.
-Here we capture where and when observers experience resets by having them draw the perceived trajectory of the stimulus after having seen it (a retrace procedure).
-This will allow us to distinguish between a spatial limit to the illusion or a temporal limit.
+Here we capture where and when observers experience resets by having them draw the perceived trajectory of the stimulus while seeing it (a tracking procedure), or after having seen it (a retrace procedure).
+By investigating the time and location of resets in the recorded trajectories, we can distinguish between a spatial limit to the illusion or a temporal limit.
 
@@
 Participants used a stylus on a horizontal drawing tablet (Wacom Intuos Pro) to indicate where they perceived the location of a gabor (see Fig 1, bottom right), as well as a keypad for additional responses.
-Stimuli were displayed on a downward facing LED monitor (30 Hz, 20”, 1680x1050), level with the drawing tablet.
-The stimuli were observed via an upward facing mirror placed exactly between the tablet and monitor, so that the stimuli appear to be in the same horizontal plane as the hand, or the tip of the stylus.
+Stimuli were displayed on a downward facing LED monitor (30 Hz, 20”, 1680x1050), parallel with the drawing tablet.
+The stimuli were observed via an upward facing mirror placed exactly between the tablet and monitor, so that the stimuli appear to be in the same horizontal plane as the hand, and the tip of the stylus.
+
+Participants were to be made For reporting degrees of visual angle that stimuli spanned ( _dva_ ), we assume the participant is close to the setup, with their eyes at the height of the monitor.
+Since there was no requirement to use the forehead rest, the latter is probably only true for the tallest participants, so that reported _dva's_ are underestimating the size of stimuli.
 
@@
 
-The stimuli were phase-changing gabor patches with a gaussian mask with an sd of 0.43 _cm_ (or ~XX _dva_), using 1.5 phases of a sine wave in the full stimulus, so that each sine phase equals ~0.582 _cm_ or ~XX _dva_.
+The stimuli were phase-changing gabor patches with a gaussian mask with an sd of 0.43 _cm_ ( or ~.6 _dva_ when in the centre of the screen and viewed from a typical distance ).
+The whole stimulus extended 3 times the sd of the Guassian in all four cardinal directions.
+The stimulus used 1.5 phases of a sine wave, so that each sine phase equals ~0.58 _cm_ or ~1.5 _dva_.
 By applying both real motion and phases changes of the gabor, it becomes a 'double-drift stimulus' where the perceived path deviates from the real path.
-The gabor changed phase at different rates, called 'internal motion' here, specified in cycles per second ( _cps_).
+The gabor changed phase at different rates, called 'internal motion' here, specified in cycles per second ( _cps_ ).
 The real displacements of the stimulus are along the Y axis of the monitor, keeping the X coordinate constant at the middle of the screen, so that a stimulus without internal motion would appear to move forward and backward on the horizontal display.
-This real motion was restricted to the central 13.5 _cm_ (~ XX _dva_) and was set so that stimulus would move from either extreme to the other (a motion we call a 'pass') in 2, 3 or 4 _s_, corresponding to speeds of 6.75, 4.5 and 3.75 _cm/s_ or roughly XX, XX or XX _dva/s_ in the middle of the screen.
-For simplicity, these will be reported as _seconds per pass_.
-- size of stimuli in cm, speed in cm/s (with guesstimates of dva?)
+This real motion was restricted to the central 13.5 _cm_ ( ~14.0 _dva_ ) and was set so that stimulus would move from either extreme to the other ( a motion we call a 'pass' ) in 2, 3 or 4 _s_, corresponding to speeds of 6.75, 4.5 and 3.75 _cm/s_ ( or roughly 7.0, 4.7 or 3.5 _dva/s_ ).
+For simplicity, and since the distance moved is always the same, these speeds will be refered to by the duration of a single pass (2, 3 or 4 _s_).
+
+## Tasks
+
+In the two tasks in experiment 1, participants were to do online tracking (using their unseen hand / stylus) while viewing a double-drift stimulus.
+In experiment 2, participants first watched a double-drift stimulus and then either reproduce the perceived path of the stimulus by retracing it on the tablet, or indicate the initial movement direction by changing the orientation of a line originating in the same position as the doube-drift stimulus.
+More details are given below.
 
@@
 
-[
+For experiment 1, we first segmented the tracking trajectories according to stimulus passes (in between direction changes of the gabor), and removed the first and last half second ( leaving 1 _s_ trajectories for the first experiment, and 3 _s_ trajectories for the second experiment).
+We then calculated the instantaneous heading (direction, disregarding velocity or distance) of a smooth spline interpolated version of each trajectory segment (using 101 spline interpolated position samples, to get 100 heading samples).
+This heading measure should depend on the illusion strength.
+Moreoever, when spontaneous resets are either time-dependent or space-dependent, resets should occur at roughly the same points and in the same direction in a given condition for every participant.
+This means that spontaneous resets should result in systematic deviations of every participant's average heading over time.
 
-For experiment 1, we first segmented the tracking trajectories according to stimulus passes (in between direction changes of the gabor), and removed the first and last half second.
-We then calculated the instantaneous heading (direction, disregarding velocity) of a smooth spline interpolated version of the tracking trajectories.
-This heading measure should depend on the illusion strength, and should change when a (spontaneous) reset is reported in the tracking data.
-
-]
-
-[
-
-For comparison, we use the same heading measure in experiment 2, but looking at individual trajectories reveals more points that are potential spontaneous resets.
-Using a smooth spline interpolated version of the trajectory we then detect direction changes (beyond ~0.39 _cm_ from the starting point, as the initial movement is often jittery).
-These direction changes are taken as reset points.
-We use the 54 average coordinates of the reset points in all 9 participants and 6 stimulus conditions for further analyses and modelling.
+For comparison, we use the same heading measure in experiment 2.
+However, the individual trajectories reveal more points that are potential spontaneous resets that we try to extract.
+Using a smooth spline interpolated version of the trajectory we detect direction changes.
+Since the initial movement is often jittery, we only take direction changes further than ~0.39 _cm_ away from the starting point.
+The first (if any) of the remaining direction changes, are then taken as reset points.
+By averaging the x and y coordinates of reset points in all 9 participants and 6 stimulus conditions, we obtain 54 average reset points for further analyses and modelling.
 The average line orientation is also calculated for each of those 54 combinations of partcipants and stimulus conditions.
-
-]
 
@@
 
-Two similar tasks were used in this experiment.
-In the first task, 5 different amounts of internal motion were used, including no internal motion as a baseline (-3, -1, 0 , 1 and 3 _cps_).
-Before stimulus onset
-
-The stimulus started in the middle, and moved forward and backward several times including 5 complete passes of 13.5 _cm_ (~XXX _dva_) and 2 _s_ each.
-The initial external motion direction (forward or backward) was counterbalanced.
-When the external motion inverted at the far and close ends of the trajectory, the internal motion was also inverted.
-Each stimulus was repeated 8 times, with randomized position of the fixation point (left or right) of the stimulus and counterbalanced initial external motion direction (forward or backward).
-This task allowed testing how sensitive online tracking is to the illusion, and served to gauge what resets would look like in tracking data.
-
-**DOUBLE PARAGRAPH; CHECK WHAT IS MISSING**
-
-In a first experiment, participants were asked to continuously track the position of a gabor that moved forward and backward along the Y-axis (of the horizontal screen) for the duration of 12 _s_, using a stylus on a drawing tablet.
-Before trial onset, the participant moved the stylus to the middle of the screen (where the gabor would appear), and was subsequently asked to fixate a point on the left or right of the screen, and given 1.5 _s_ to do so.
-Since the screen was seen via a mirror, the stimulus appeared to move on a horizontal plane at the same depth as the drawing tablet.
+In the first task, participants were asked to continuously track the position of a gabor that moved forward and backward along the Y-axis (of the horizontal screen) for the duration of 12 _s_, using a stylus on a drawing tablet.
+Before trial onset, the participant moved the stylus to the middle of the screen ( where the gabor would appear ), and was subsequently asked to fixate a point on the left or right of the screen, and given 1.5 _s_ to do so.
+Since the screen was seen via a mirror, the stimulus appeared to move forward and backward on a horizontal plane at the same depth as the drawing tablet.
 The internal motion of the gabor was inverted at the far and close ends of the physical path, where external motion was also inverted.
 Internal motion could be 3, 1, 0, -1 or -3 _cps_.
-Since each moving gabor was shown for 12 _s_, and it moved from one end to another in 2 _s_, it completed 5 full and 2 half passes of the path (or 3 cycles).
+Since each moving gabor was shown for 12 _s_, and it moved from one end to another in 2 _s_, it completed 5 full and 2 half passes of the path.
 
@@
 It seems that continuous tracking is sensitive to the strength of the illusion, and this is confirmed by a repeated-measures F-test on the average angles, using internal motion as a within-subjects factor (_F_(4,12)=102.6, _p_<.001, $\eta^2$=0.97).
-With FDR-corrected, paired t-tests, we find a difference in heading between 0 _cps_ internal motion and both 1 _cps_ ( _p_=.022) and -1 _cps_ ( _p_=.025) as well as between -3 and -1 _cps_ and 3 and 1 _cps_ (both _p_=<.001).
+With FDR-corrected, paired t-tests, we find a difference in heading between 0 _cps_ internal motion and both 1 _cps_ ( _p_=.022) and -1 _cps_ ( _p_=.025) as well as between -3 and -1 _cps_ and 3 and 1 _cps_ (both _p_ <.001).
 This means we can use the averaged angle for each of the internal motion conditions as a measure of the strength of the illusion.
@@
 
-In a second experiment, we used a retracing task where the participants drew what they saw of a single pass stimulus immediately after presentation.
-This provided more time for the participants to reproduce any resets, i.e.
-without the need for real-time tracking.
-In this task, the double-drift gabor started at the near end of the display, and moved forward along the horizontal display away from the participant.
-The gabor could have one of two external speeds (corresponding to 3 or 4 _s_ presentation time), and one of three internal speeds (2, 3 or 4 _cps_).
-Apart from re-tracing the perceived paths, in half the blocks participants indicated the angle of the first part of the stimulus motion by orienting a line (75 _px_ = 3 _cm_) originating from the starting point of the gabor’s path.
-This task has been used before [XXXX].
-In the meantime, Nakayama & Holcombe [-@Nakayama2019] have also used a re-tracing paradigm to study triggered resets.
+In a second experiment, we ask participants to re-trace the perceived path of a double-drift stimulus, immediately after viewing the stimulus.
+This allows participants more time to carefully reproduce the path.
+We use the recorded trajectories to assess points where resets occured.
+The reset points can then be used to determine if resets are time-limited or space-limited.
 
@@
 
-Participants alternated between blocks of two kinds of trials (in counterbalanced order).
-In both types of trial, the gabor was shown starting close to the participant and moving away in a forward direction, for a single pass along the workspace, taking 4 seconds.
-During presentation of the gabor, the participant was instructed to maintain fixation to the left or right of the stimulus.
-After the gabor disappeared, the participant could break fixation, and was to report the percept in one of two ways.
-In one kind of block, the participant re-traced the perceived path of the gabor using the stylus.
-The drawn path was shown as a line on the screen.
-Participants could “key in” their re-traced percept by pressing ENTER or start over by pressing the ESCAPE key.
-In the other kind of block, participants reported the initial direction of the gabor (before any resets occurred), by using the LEFT and RIGHT arrow keys to change the orientation of a line of XX cm, with it’s origin in the same spot as where the gabor started moving.
-Participants could change the line’s orientation it until satisfied and then “key in” their response by pressing ENTER.
-The arrow report task has been used before (ref?) and can be used as validation of this task.
-The re-tracing task has also been used before, but in order to test evoked resets [@Nakayama2019].
+Here we used a re-tracing procedure where participants drew the perceived path of a double-drift stimulus making a single pass across the workspace.
+This provided more time for the participants to reproduce any resets, i.e.
+without the need for real-time tracking.
+Nakayama & Holcombe [-@Nakayama2019; -@Nakayama2019b] have also used a re-tracing paradigm, in their case to study triggered resets.
+The re-tracing task was done in half of 8 blocks.
+In the other half, participants oriented a line to indicate the initial perceived direction of the stimulus motion.
+Block types were alternated, and the order was counterbalanced across participants.
+Each block used the 6 combinations of two external speeds ( corresponding to 3 or 4 _s_ presentation time ), and three internal speeds ( 2, 3 or 4 _cps_ ), 6 times.
+Within each of the two tasks, each of the combinations of internal and external speed was presented 24 times in total.
+
+In both kinds of trials, participants first had to move the stylus to the start position of the gabor, and then fixate a point to the left or right of the centre of the workspace.
+The double-drift gabor started at the near end of the display, and moved along the horizontal display away from the participant, always making a single, forward pass.
+After, the gabor had disappeared, participants could respond in one of two ways.
+When re-tracing the perceived path with the stylus on the tablet, the drawn path would show up as a red line.
+Participants could "key in" their response by pressing ENTER, or start over by pressing ESCAPE.
+When indicating only the initial direction of movement, participants would orient a line ( 75 _px_ or ~3 _cm_ ) originating from the starting point of the gabor’s path.
+They could change the orientation by pressing either the LEFT ARROW key to orient the line more clockwise or the RIGHT ARROW key to orient the line more counter clockwise.
+They could "key in" their response by pressing ENTER.
+In both tasks, there was no time-limit for the response, and careful responding was encouraged.
+
+Re-tracing the perceived paths should allow capturing spontaneous resets of the illusion, and the depence on the strength of the illusion could be checked with the data from the line-orienting task.
 
@@
 
-[Actual results] For the majority of participants (X/9) we see that most trials show at least one reset (>90%?). In the remainder we still see >50% of the trials have a reset.
+On average, the algorithm detects resets in 75% of trials.
+For three participants, only about half the trials show resets (42%, 44% and 58%).
+For the remaining six participants the algorithm detects resets in 72%-99% of trials, with an average of 88%.
+The minimum number of resets in one of the six conditions for any participant is 7 (out of 24), with the average at 17.
+Since participants experiencing a "hit-the-wall reset" could display a direction change in their paths only by chance, 50% of trials with resets seems reasonable.
+Neither percept is incorrect, and the average reset positions still indicates some limit to the illusion.
+Therefore, we keep all participants and resets.
 
@@
 We set out to test if spontaneous resets in the double-drift illusion reflect discrepancies between two position signals, occurring at some given size of this discrepancy, or if they simply occur after some time elapses, analogous to other illusions or bi-stable stimuli.
-To measure the location of resets we first ask participants to track the perceived location of the double-drift stimulus continuously.
-While the direction of the tracking movements scaled with the internal motion of the stimulus (was sensitive to the strength of the illusion), there were no clear reset points in the tracking data.
+To measure the location of resets we first ask participants to track the perceived location of the double-drift stimulus online.
+While the heading of the tracking movements is sensitive to the strength of the illusion, there are no clear reset points in the tracking data.
 In a second experiment we ask participants to draw the perceived path of the stimulus after a single pass.
-In this re-tracing experiment the drawn traces reflect the strength of the illusion [just] as well, and in most trials there is a clear reset point.
-The positions of these reset points do no vary much with elapsed time, but are better explained as triggered when the distance between perceived and true position of the stimulus exceeds some threshold.
-This indicates that while the perceived location of a moving object can be heavily biased by its well-defined, internal motion, an uncertain (probably retinal?) position signal is still used to keep the perceived position of moving objects close to reality.
+In this re-tracing experiment the drawn traces reflect the strength of the illusion just as well, and in most trials there is a clear reset point, for either a jump reset or a hit-the-wall reset.
+The positions of these reset points do no vary much with elapsed time, but are better explained by the distance from the true position.
+This has implications for the way position perception relies on motion signals.
 
-§3 (one computational model):
+There is a generative model, based on Bayesian cue integration, that can explain many phenomena with motion induced position shifts.
+This model, formulated by Kwon, Tadin & Knill [-@Kwon2015], is compatbile with a good portion of our data.
+The model posits there are three signals contributing to an object's perceived position: it's global motion, it's local motion (or texture motion) and it's position.
+Especially when the directly perceived position is a noisy or weak signal, the two motion signals contribute more, as in the double-drift or curveball illusion.
+In this stimulus, the model predicts that the sensory motion signals bias perceived position away from the true location, but at some point this is balanced by the weak, noisy sensory position signal that bias perceived position back towards the true location [@Kwon2015, Fig 3].
+When this balance is reached the perceived path of the object would be parallel to but somewhat displaced from the true object path.
+This pattern of perceived paths is exhibited by some of our participants, and we have called these "hit-the-wall resets" here.
 
-- Explain paper with model by Kwon, Tadin & Knill, and VSS poster with fat-tailed noise.
+Some of our participants seem to perceive "jump resets" though, and the model as published, does not predict any such resets.
+This could be added by including a causal inference step [e.g.
+@Wei2011].
+For the subset of participants that experience jump resets, there may be a process that compares the two position signals (perceived and retinal) and when these exceed some threshold, decides that the two position signals do not share a common source.
+At that point the perceived position may be discarded, or reset to the retinal position, and the build-up of the illusion may start again.
+For the remaining participants, either the discrepancy between retinal and perceived location never exceeds their threshold, or perhaps they do not employ any causal inference process.
 
-While this model produces reasonable resets, the properties of resets, such as their spatial or temporal distributions have not been characterized yet.
-An alternative explanation for resets could be that the true but imprecise position that is available for eye-movements can limit or reset the perceived position under some circumstances.
-In order to get at the properties of resets
-
-If the double-drift illusion is taken as a vector average between the external motion and some proportion of the internal motion, the initial path of the stimulus can be modelled as a straight line from the origin at some angle.
-This straight line segment should intersect the position of the first reset.
-We will now see if resets occur at a given distance perpendicular to the path of the real stimulus, after a given amount of time, or perhaps a combination of the two.
-
-This means that it seems that resets occur predominantly at some given discrepancy between the true and illusory position of the stimulus, with a small contribution of time.
-It may that after seeing the illusion for some time, a ‘source estimation’ mechanism can no longer account for both the perceived, illusory position and the imprecise retinal position, as coming from the same stimulus.
-Then the visual system changes the position of the stimulus to the position derived from retinotopic information (instead of opting for two separate sources, which is highly unlikely in this case), resulting in a “reset”.
+Recently, it has been suggested that resets can be triggered by planned eye-movements to the stimulus [@Nakayama2019; @Nakayama2019b], particularly "jump" resets.
+In this experiment, we did not control for eye-movements with an eye-tracker, so we can't directly investigate this idea.
+Given a pre-motor theory of attention like mechanism, when spontaneously planned eye-movements trigger (jump) resets, such resets could occur at any point during the stimulus pass.
+However, it seems that for the participants that display "jump" resets in their re-tracing data, these happen at a fairly consistent location, which suggests that jump resets are instead stimulus triggered.
+This should however be tested with eye-tracking data, so that this will remain a speculation here.
 
@@
 
-We find that online tracking of the double-drift stimulus is sensitive to illusion strength, but doesn't display clear resets.
-We do find spontaneous resets in single-pass retracing.
-The illusion appears to resets predominantly at a spatial distance from the true stimulus location, suggesting that resets resolve a source attribution conflict between the true and illusory position of the stimulus.
+We find that online, manual tracking of the double-drift stimulus is sensitive to illusion strength, but doesn't display clear resets.
+We do find spontaneous resets in delayed, single-pass re-tracing.
+The illusion appears to reset, or max out, predominantly at a spatial distance from the true stimulus location, suggesting that resets resolve a source attribution conflict between the true (retinal) and illusory (perceived) position of the stimulus.
 
@@
 
-The research was support by funding from NSERC Canada to PC and DYPH , as well as by the Department of Psychological and Brain Science, Dartmouth College (PC).
+The research was support by funding from NSERC Canada (PC and DYPH), as well as by the Department of Psychological and Brain Science, Dartmouth College (PC).
 

### 2020-05-28 R/common.R
added & fixed code for calculating dva's

@@ -21,11 +21,11 @@ getPPCdva <- function(lo=3) {
   dva <- expand.grid('eye.dist'  = seq(15, 25, length.out = lo), 
                      'gabor.pos' = seq(-6.75, 6.75, length.out = lo))
   
-  b <- 27
-  a <- dva$eye.dist + dva$gabor.pos
+  b <- 27 # distance between mirror and monitor
+  a <- dva$eye.dist + dva$gabor.pos # horizontal distance (along monitor) 
   
   # https://en.wikipedia.org/wiki/Trigonometry
-  A <- atan(a/b)
+  A <- atan(a/b) # angle at centre of gabor
   
   a1 <- tan(A+((0.5/180)*pi))*b
   a2 <- tan(A-((0.5/180)*pi))*b
@@ -34,9 +34,24 @@ getPPCdva <- function(lo=3) {
   
   ghw <- (100/3) / props[['PPC']]
   
+  props[['gaborcm']] <- ghw
+  
   dva$gabor.dva <- ghw / dva$cmpd
   
   props[['dva']] <- dva
+  
+  # visual size of the "workspace" (13.5 cm track)
+  dva.track <- expand.grid('eye.dist'  = seq(15, 25, length.out = lo))
+  
+  b <- 27
+  a <- dva.track$eye.dist
+  
+  A1 <- atan(a/(b-6.75)) # the angle of the close end at each head position
+  A2 <- atan(a/(b+6.75)) # the angle of the far end at each head position
+  
+  dva.track$dva <- ((A1-A2)/pi)*180
+  
+  props[['track']] <- dva.track
   
   return(props)
   

### 2020-05-28 R/retrace_onePass_V4.R
testing functionality of code

@@ -340,14 +340,16 @@ getNormalizedHeading2Dhists <- function(segmentpoints=101, resetOnly=TRUE, load=
         
         step.idx <- which(trialdf$step == 99) # only for re-trace
         
+        if (length(step.idx) == 0) { next() }
+        
         # get out the data:
         x  <- ((trialdf$handx_pix[step.idx] / (524)) + 0.0) * 0.6
         y  <- ((trialdf$handy_pix[step.idx] / (524)) + 0.5) * 0.6
         # t  <- trialdf$time_ms[step.idx]
         
-        if (taskdf$internalMovement[1] < 0) {
-          x <- -x
-        }
+        # this seems not to make much of a difference, 
+        # even though it works for the individual trajectories
+        if (trialdf$internalMovement[1] < 0) x <- -x
         
         boundary <- which(diff(x) < 0)[1]
         
@@ -362,7 +364,7 @@ getNormalizedHeading2Dhists <- function(segmentpoints=101, resetOnly=TRUE, load=
         # t  <- t[-del.idx]
         cd <- cd[-del.idx]
         
-        if (length(cd) < 2) next()
+        if (length(cd) < 2) { next() }
         # scale from 0 to 1:
         cd <- cd / max(cd)
         
@@ -551,6 +553,12 @@ plotExampleData <- function(target='inline') {
     plot(-1000,-1000,main='all',xlab=xlab,ylab='',xlim=c(-1,1),ylim=c(0,1),bty='n',ax=FALSE,asp=1)
     #print(freq2D$x.edges)
     polarHeatMap(x=x.edges,y=y.edges+30,z=freq2D,mincol=c(1,1,1),maxcol=c(0.06,0.82,0.88),xlim=NA,ylim=NA,xunit='degrees',border=NA,bordercol='white',resolution=1,alpha=1,overlay=TRUE,origin=c(0,0),scale=1,main='')
+    
+    avgline <- f2ds[['avg']][[conditionname]][c(2:101),]
+    scale <- ((avgline$sample_no+30) / (max(avgline$sample_no)+30))
+    adX <- (cos((avgline$heading/180)*pi) * scale)
+    adY <- (sin((avgline$heading/180)*pi) * scale)
+    lines(adX,adY,col=colors[['blue']]$s,lw=2)
     
   }
   
[…]

### 2020-05-28 R/time.R
added a script file for comparing with Sirui's time-model

@@ -1,0 +1,190 @@
+
+getParticipantAverageDeviations <- function(participants = c(2,3,4,5,6,8,9,10,11)) {
+  
+  devdf <- NA
+  
+  for (ppno in participants) {
+    
+    df <- read.csv(sprintf('data/onePass_V4/onepass_V4_p%02d.csv', ppno), stringsAsFactors = F)
+    
+    internalMotions <- sort(unique(abs(df$internalMovement)))
+    
+    ppdf <- NA
+    
+    for (im in internalMotions) {
+      
+      imdf <- df[which(abs(df$internalMovement) == im),]
+      
+      # normalize direction:
+      idx <- which(imdf$internalMovement < 0)
+      imdf$handx_pix[idx] <- -1 * imdf$handx_pix[idx]
+      
+      # get only the re-tracing samples:
+      imdf <- imdf[which(imdf$step == 99),]
+      
+      # loop through all trials:
+      trialnos <- unique(imdf$trial_no)
+      
+      time_s <- c()
+      hordev <- c()
+      
+      for (trialno in trialnos) {
+        
+        # get out the part of the data for this trial:
+        trialdf <- imdf[which(imdf$trial_no == trialno),]
+        naidx <- which(!is.na(trialdf$handx_pix) | !is.na(trialdf$handy_pix))
+        trialdf <- trialdf[naidx,]
+        
+        # if there are very few samples (by accident) skip the trial:
+        if (dim(trialdf)[1] < 50) {
+          next
+        }
+        
+        # get the X and Y coordinates:
+        x <- trialdf$handx_pix
+        y <- trialdf$handy_pix
+        
+        if (trialdf$externalMovement[1] == 0.125) {
+          duration <- 4
+        } else {
+          duration <- 3
+        }
+        
+        #print(any(is.na(y)))
+        t <- y - min(y)
+        x <- x - min(x)
+        #print(any(is.na(t)))
+        #print(max(t))
+        x <- (x / max(t)) * duration # scale x to y (assuming y spans more pixels)
+        t <- (t / max(t)) * duration # scale y/t to duration
+        #print(any(is.na(t)))
+        
+        # remove anything that is beyond 3 seconds:
+        tidx <- which(t <= 3)
[…]

### 2020-05-28 exp/demo_onepass_V4b.py
added python code to run each of the three experiments (hoping I have the right version)

@@ -1,0 +1,517 @@
+import time
+import scipy as sp
+import pandas as pd
+
+from psychopy import visual, core, data, event, logging, sound, gui
+
+from pyglet.window import key
+
+
+def adjustTrial(cfg):
+  
+  # used to read out keyboard status:
+  keyboard = key.KeyStateHandler()
+  cfg['win'].winHandle.push_handlers(keyboard)
+  
+  # getting trial parameters:
+  trialno = cfg['trial']
+  
+  # stimulus parameters:
+  externalSpeed     = cfg['externalSpeed']
+  angle             = cfg['trialdefinitions']['angle'][trialno]
+  fixationSide      = cfg['trialdefinitions']['fixationSide'][trialno]
+  
+  #angle = 90 - angle
+  internalMovement  = 0  
+  theta = ((90 - angle) / 180.) * sp.pi
+  slope = sp.sin(theta) / sp.cos(theta)
+  
+  #print(angle)
+  #print(slope)
+  #print(fixationSide)
+  
+  # set location of fixation point:
+  cfg['point'].pos = [fixationSide * (cfg['width'] / 5.), 0]
+  
+  gaborPhase = sp.rand()
+  
+  # orientation should be the same as the movement angle?
+  cfg['gabor'].ori = (-1 * angle) + 90
+  
+  starttime = time.time()
+  prevtime = starttime
+  prevY = 0
+  
+  gaborNotStraight = True
+  
+  while (gaborNotStraight):
+    
+    # stimulus properties depend on time:
+    
+    now = time.time()
+    
+    # stimulus position depends on total elapsed time:
+
+    totalelapsed = now - starttime
+    
+    # Y = (cfg['height'] / 4) * (1 - (abs((((elapsed-trialduration)*(externalMovement/0.5))%2)) * 2))
+
+    Y = (1 - (abs((((totalelapsed)*(4*externalSpeed/0.5))%2) - 1) * 2)) * (cfg['height'] / 16)
+    X = slope * Y
+
+    cfg['gabor'].setPos([X,Y])
+    
+    # phase of the gabor changes depending on time elapsed since last the frame and on internal speed:
+    # (since the internal movement speed changes, we can't use total elapsed time)
+    # and it depends on whether or not the gabor is moving up or down... which is now weird to get
+    
+    frameelapsed = now - prevtime
+    #print(frameelapsed)
+    
[…]

### 2020-05-28 exp/onepass_V4.py

@@ -1,0 +1,396 @@
+import time
+import scipy as sp
+import pandas as pd
+
+from psychopy import visual, core, data, event, logging, sound, gui
+
+from pyglet.window import key
+
+def onePassTrial(cfg):
+  
+  # used to read out keyboard status:
+  keyboard = key.KeyStateHandler()
+  cfg['win'].winHandle.push_handlers(keyboard)
+  
+  # getting trial parameters:
+  trialno = cfg['trial']
+  
+  # stimulus parameters:
+  internalMovement  = cfg['trialdefinitions']['internalMovement'][trialno]
+  externalMovement  = cfg['trialdefinitions']['externalMovement'][trialno]
+  
+  # psychophysics:
+  recordTrace       = cfg['trialdefinitions']['recordTrace'][trialno]
+  recordPercept     = cfg['trialdefinitions']['recordPercept'][trialno]
+  perceptTask       = cfg['trialdefinitions']['perceptTask'][trialno]
+  arrowPosition     = cfg['trialdefinitions']['arrowPosition'][trialno]
+  traceTime         = cfg['trialdefinitions']['traceTime'][trialno]
+  taskname          = cfg['trialdefinitions']['taskname'][trialno]
+  
+  #TDrulerPercept['recordTrace']                = False
+  #TDrulerPercept['recordPercept']              = True
+  #TDrulerPercept['perceptTask']                = 'ruler'
+  #TDrulerPercept['arrowPosition']              = None
+  #TDrulerPercept['traceTime']                  = None
+  
+  
+  trialduration = (0.5 / externalMovement) # trial duration will always be exactly enough for a single pass
+  
+  # set location of fixation point:
+  cfg['point'].pos = [cfg['trialdefinitions']['fixationSide'][trialno] * (cfg['width'] / 3.), 0]
+  
+  # we need a few vectors to collect data in:
+  handx_pix = []
+  handy_pix = []
+  time_s = []
+  gaborx_pix = []
+  gabory_pix = []
+  gaborphase = []
+  gabororientation = []
+  step = []
+  
+  # random phase:
+  gaborPhaseOffset = sp.rand()
+  
+  
+  # first the mouse / pen cursor has to be brought to the starting position (to counter all that spatial drift)
+  trialstarttime = time.time()
+  mousepos = cfg['mouse'].Pos()
+  
[…]

### 2020-05-28 exp/runOnePassExperiment_V4.py

@@ -1,0 +1,159 @@
+from setup import *
+from onepass_V4 import *
+import pandas as pd
+
+# start building a config dictionary for the experiment:
+cfg = {}
+
+# get participant number and seed random number generator based on that:
+cfg = getParticipantID(cfg)
+seed(sum([ord(c) for c in 'illusory motion reset tracking']) + (cfg['id'] * 9999))
+
+
+# add a window for visual stimuli:
+cfg['fullscr']      = True
+cfg['monitorIndex'] = 1
+cfg['flip']         = True
+
+cfg = createWindow(cfg, resolution=[1920, 1080])
+
+mouse = event.Mouse(win=cfg['win'])
+mouse.setVisible(False)
+
+# add a 'mouse' object for pen positions,
+# use XLib if possible as it is more accurate:
+
+# this creates a mouse object in the cfg dictionary
+# it has a single method: Pos()
+# that returns vector with an X and Y coordinate, and a time stamp
+# it either uses X11, and if that fails it uses PsychoPy
+cfg = addMouse(cfg)
+
+cfg['frameno'] = 1
+
+# create some psychopy objects that will be used as stimuli:
+cfg = createStimuli(cfg)
+
+cfg['externalSpeed']                    = .125 # Hz: how often can the cursor go up and down the whole range (= 2 passes) in a second
+# these two should not be used:
+Npasses                                 = 1
+cfg['trial_duration']                   = (0.5 / cfg['externalSpeed']) * (Npasses + .5)
+
+
+# there will be 6 internal cursor movements, that induce various degrees of illusion:
+internalMovement                        = [-4, -2, -1, 1, 2, 4]
+externalMovement                        = [1./8, 1./4, 1./2]
+perceptAngle                            = [89,91]
+repetitions                             = 2
+
+internalMovement                        = [-4,-3,-2, 2,3,4]
+externalMovement                        = [1./8, 1./6]
+perceptAngle                            = [90]
+repetitions                             = 3
+
+# create all desired combinations:
+TD                                      = pd.concat([foldout(a=[internalMovement,externalMovement, perceptAngle])]*repetitions, ignore_index=True)
+TD.columns                              = ['internalMovement', 'externalMovement', 'perceptAngle']
+
[…]

### 2020-05-29 R/time.R
added a function to plot reset-time distributions

@@ -166,7 +166,7 @@ plotSplitDeviations <- function() {
         
         ppdf <- imdf[which(imdf$participant == ppno),]
         
-        linreg <- lm(hordev ~ time_s, data=ppdf[which(ppdf$time_s > 0.5),])
+        linreg <- lm(hordev ~ time_s, data=ppdf[which(ppdf$time_s >= .5),])
         if (direction *linreg$coefficients[['time_s']] > 0) {
           
           lines(x=ppdf$hordev, y=ppdf$time_s, col=transp[[sprintf('%d',im)]])
@@ -184,6 +184,52 @@ plotSplitDeviations <- function() {
       axis(side=2, at=c(0,1,2,3))
       
     }
+    
+  }
+  
+}
+
+
+plotResetYdistribution <- function() {
+  
+  colors <- getColors()
+  transp <- list('2'=colors[['purple']]$t, '3'=colors[['yorkred']]$t, '4'=colors[['orange']]$t)
+  solids <- list('2'=colors[['purple']]$s, '3'=colors[['yorkred']]$s, '4'=colors[['orange']]$s)
+  
+  # read file with all detected resets
+  df <- read.csv('data/onePass_V4/onePass_V4_re-trace.csv', stringsAsFactors = F)
+  
+  # remove trials without resets
+  df <- df[which(!is.na(df$boundX)),]
+  
+  # normalize for time:
+  idx3 <- which(df$externalspeed == 0.167)
+  idx4 <- which(df$externalspeed == 0.125)
+  df$boundY[idx3] <- df$boundY[idx3] * 3
+  df$boundY[idx4] <- df$boundY[idx4] * 4
+  
+  # we create one one plot per absolute internal speed
+  # create a layout matrix:
+  layout(matrix(c(1,2,3),nrow=1,ncol=3))
+  
+  for (im in c(2,3,4)) {
+    
+    plot(-1000,-1000,xlim=c(0,3),ylim=c(0,1.6),xlab='time [s]', ylab='relative density',ax=F,bty='n',main=sprintf('internal motion %d',im))
+    
+    for (ppno in unique(df$participant)) {
+      
+      subdf <- df[which(df$participant == ppno & df$internalspeed == im),]
+      kd <- density(subdf$boundY, bw=0.2, n=151, from=0, to=3)
+      lines(kd$x, kd$y, col=transp[[sprintf('%d',im)]])
+      
+    }
[…]

### 2020-05-29 DoubleDriftResetLimits.Rmd
changed percept average plot to be split into time-constrained and space-constrained participants, added plot of reset-time distribution

@@
 This allows to see if it is in line with the data reported by Liu, Tse & Cavanagh [].
+
+Here is a figure that splits participants by whether or not they have an average percept that is consistent with the data Lui et al found, and that is inconsistent with it.
+
+```{r}
+plotSplitDeviations()
+```
+
+Let's also have a look at the distribution of first reset points on Y-axis (~time).
+Sirui's model predicts a certain distribution of first reset points, so we can compare wat is measured here, with the model.
+
+```{r}
+plotResetYdistribution()
+```

### 2020-08-04 Manuscript.Rmd
adding comparison with Sirui's PSE data

@@
 source('R/models.R')
+# And finally, code to compare this data with Sirui's PSEs
+source('R/time.R')
 ```
@@
 
+## Comparison with Sirui's data set
+
+While here we find that a spatial limit is the better explanation for resets, other evidence may point at a temporal limit instead.
+In an as of yet unpublished study [Sirui] obtained the point of subjective equivalence (PSE) of a double drift stimulus, and a static point that appeared some time after the double drift stimulus disappeared.
+The double-drift stimulus was on a 2 or 4 DVA length trajectory, and disappeared after 1, 2 or 3 seconds.
+The PSE's can be modelled very accurately with a model that assumes a fixed chance of a reset per unit of time (~.7%/s), that is: it explains the data using only a temporal bound on the illusion.
+Interestingly, this fully contradicts the current findings.
+The model predicts that the PSE decreases over time as it becomes more likely that a reset has occurred.
+We will compare the two data sets here.
+
+```{r fig8, echo=FALSE, message=FALSE, fig.width=8, fig.height=3.5, fig.cap="\\label{fig:fig8}**Figure 8. PSE's of double-drift position and continuous tracking**"}
+plotDeviations()
+```
+
+If the PSE's express the average perceived deviations from the actual trajectory of the double-drift stimulus, then they should show a similar pattern as the averaged trajectories in experiment 2.
+In order to be able to compare the two data sets, we will include all trajectories where our simple algorithm did not detect a reset, which should also be included in the PSE data.
+Further, we scale the trajectories, so that the Y end-point matches the duration of stimulus.
+Since the PSE data was collected in degrees visual angle, we convert the unit to degrees visual angles, using the estimate at the centre of the workspace.
+For the 4-second stimuli, we removed the last second.
+Figure 8 shows the individual participants' and the averaged trajectories (purple, red, orange) as well as the PSE's obtained in the other study (shown separately for 2 and 4 DVA stimulus trajectories) in gray.
+
+For all internal motion speeds, we see a very similar average trajectory, that does not get closer to veridical over time.
+We verified this with a repeated measures ANOVA on perceived deviation from veridical stimulus position, using internal motion speed (2, 3, or 4 cycles/s) and time (1, 2, or 3 seconds after stimulus onset) as factors.
+If the trajectory data from experiment 2 here follows the same pattern as the PSE data, there should be an effect of time, and perhaps an interaction with internal motion speed.
+We find no effect of time ( _F_(2,16)=0.011, _p_=.957, $\eta^2$<.001), no effect of internal motion speed ( _F_(2,16)=0.127, _p_=.814, $\eta^2$=.0015) and no interaction ( _F_(4,32)=0.361, _p_=.635, $\eta^2$=.002).
+This means that we can't replicate the pattern found with the PSE approach.
+There may be several reasons for this.
+First, we notice that the trajectory data is highly dependent on participants.
+It could be that this method provides very noisy estimates, hiding any effect.
+It could also be that the effects truly vary a lot between participants.
+Since we can see that some participants do seem to follow the pattern found with the PSE approach, and some show a pattern of increasing average deviation between perceived and true position, it may that the difference between the data sets can simply be explained by random sampling of relatively small numbers of participants.
+
 # Discussion
@@
 At that point the perceived position may be discarded, or reset to the retinal position, and the build-up of the illusion may start again.
-For the remaining participants, either the discrepancy between retinal and perceived location never exceeds their threshold, or perhaps they do not employ any causal inference process.
+For the remaining participants, either the discrepancy between retinal and perceived location never exceeds their threshold, or there may be another reason they do not employ any causal inference process.
 
@@
 The illusion appears to reset, or max out, predominantly at a spatial distance from the true stimulus location, suggesting that resets resolve a source attribution conflict between the true (retinal) and illusory (perceived) position of the stimulus.
+This confirms the notion that we gauge the position of objects in the world by combining both retinal position signals as well as accumulated position shifts derived from motion signals.
 

### 2020-08-04 R/time.R

@@ -1,4 +1,7 @@
 
+source('R/common.R')
+library('ez')
+
 getParticipantAverageDeviations <- function(participants = c(2,3,4,5,6,8,9,10,11)) {
   
   devdf <- NA
@@ -55,8 +58,20 @@ getParticipantAverageDeviations <- function(participants = c(2,3,4,5,6,8,9,10,11
         x <- x - min(x)
         #print(any(is.na(t)))
         #print(max(t))
-        x <- (x / max(t)) * duration # scale x to y (assuming y spans more pixels)
-        t <- (t / max(t)) * duration # scale y/t to duration
+        
+        use_cm <- TRUE
+        
+        if (use_cm) {
+          #x <- (x / 524.) * 13.5
+          x <- (x / max(t)) * 13.5
+          x <- x / 0.7298552 # cm/d on the centre of the track
+        } else {
+          # express x in seconds? counter intuitive, but can be scaled with the Y 
+          x <- (x / max(t)) * duration # scale x to y (assuming y spans more pixels)
+        }
+        
+        # t should be scaled always
+        t <- (t / max(t)) * duration # scale y (=t) to duration too
         #print(any(is.na(t)))
         
         # remove anything that is beyond 3 seconds:
@@ -116,7 +131,20 @@ plotDeviations <- function() {
   
   for (im in c(2,3,4)) {
     
-    plot(x=c(0,0),y=c(0,3),type='l',col='#999999',main=sprintf('internal motion %d',im), xlim=c(-.5,2), ylim=c(-.5,3.5), xlab='percept deviation', ylab='time [s]', asp=1, bty='n', ax=F)
+    plot(x=c(0,0),y=c(0,3),type='l',lty=3,col='#000000',main=sprintf('internal motion %d',im), xlim=c(-1,4), ylim=c(-.5,3.5), xlab='percept deviation [dva]', ylab='time [s]', bty='n', ax=F)
+    
+    #PSEs <- c(1.00, 0.83, 0.66, 1.65, 1.50, 1.20)
+    
+    # path length: 2 DVA
+    # the unit is not centimeters, though?
+    points(c(1.00, 0.83, 0.66),c(1,2,3),col='#999999',pch=16,cex=2)
+    lines(c(1.00, 0.83, 0.66),c(1,2,3),col='#999999',lw=3,lty=2)
[…]

### 2020-09-21 R/time.R
fixed average trajectory plotting and added distributions

@@ -2,110 +2,149 @@
 source('R/common.R')
 library('ez')
 
-getParticipantAverageDeviations <- function(participants = c(2,3,4,5,6,8,9,10,11)) {
+# Comparing trajectories with PSE/time model ====
+
+getParticipantAverageDeviations <- function(participants = c(2,3,4,5,6,8,9,10,11), getDistribution=FALSE) {
   
+  # data frame: horizontal deviation ~ participant * IM * EM * time 
+  # for plotting & ANOVA
   devdf <- NA
   
-  for (ppno in participants) {
+  # interpolation time vector:
+  t_p <- seq(0,3,length.out = 151)
+  
+  if (getDistribution) {
+  # array for distributions:
+    ddistr <- array(data=NA, dim=c(3,2,length(participants),3,101))
+  }
+  
+  for (ppno in c(1:length(participants))) {
     
-    df <- read.csv(sprintf('data/onePass_V4/onepass_V4_p%02d.csv', ppno), stringsAsFactors = F)
+    ppid <- participants[ppno]
+    
+    df <- read.csv(sprintf('data/onePass_V4/onepass_V4_p%02d.csv', ppid), stringsAsFactors = F)
+    
+    df$externalMovement[which(df$externalMovement == 0.125)] <- 4
+    df$externalMovement[which(df$externalMovement == 0.167)] <- 3
     
     internalMotions <- sort(unique(abs(df$internalMovement)))
+    externalMotions <- sort(unique(abs(df$externalMovement)))
     
     ppdf <- NA
     
-    for (im in internalMotions) {
+    for (exno in c(1:length(externalMotions))) {
       
-      imdf <- df[which(abs(df$internalMovement) == im),]
+      ex <- externalMotions[exno]
       
-      # normalize direction:
-      idx <- which(imdf$internalMovement < 0)
-      imdf$handx_pix[idx] <- -1 * imdf$handx_pix[idx]
-      
-      # get only the re-tracing samples:
-      imdf <- imdf[which(imdf$step == 99),]
-      
-      # loop through all trials:
-      trialnos <- unique(imdf$trial_no)
-      
-      time_s <- c()
-      hordev <- c()
-      
-      for (trialno in trialnos) {
+      for (imno in c(1:length(internalMotions))) {
         
-        # get out the part of the data for this trial:
-        trialdf <- imdf[which(imdf$trial_no == trialno),]
-        naidx <- which(!is.na(trialdf$handx_pix) | !is.na(trialdf$handy_pix))
[…]

### 2020-09-21 R/common.R
added %notin% operator defintion

@@ -1,3 +1,9 @@
+
+
+`%notin%` <- Negate(`%in%`)
+
+
+
 # 
 # # pixels per cm
 # 

### 2020-09-21 Manuscript.Rmd
incorporated a lot of Patrick's feedback

@@
 While the orientation of the recorded trajectories reflect stimulus properties, there were no spontaneous resets.
-That’s why, in a second experiment, we had participants re-trace the perceived trajectory of a short stimulus presentation, or indicate the initial direction of the stimulus they perceived.
+In a second experiment, we had participants re-trace the perceived trajectory of a short stimulus presentation, or indicate the initial direction of the stimulus they perceived.
 We found resets here, and they indicate a set spatial limit to the distance between the veridical and perceived location of the double-drift stimulus.
@@
 It is no surprise then that the brain has several mechanisms to gauge and update the position of objects in the world.
-For example, when a target is moving, its motion can influence it perceived location, making it appear to be slightly ahead of its true location, e.g.
+For example, when a target is moving, its motion can influence its perceived location, making it appear to be slightly ahead of its true location, e.g.
 the flash lag [@Nijhawan1994]; or the flash grab effect [@Cavanagh2013].
@@
 Surprisingly, this extraordinary perceptual illusion does not affect eye movements to the gabor: saccades to these targets are determined by their physical not their perceived location [@Lisi2015].
-This result indicates that both true and illusory locations are available in the brain.
-Indeed, although immediate eye-movements go towards the true position, delayed eye movements and pointing go to the perceived position [@Lisi2017; @Massendari2018].
+And although immediate eye-movements go towards the true position, delayed eye movements and pointing go to the perceived position [@Lisi2017; @Massendari2018].
+These findings suggest that both the true and illusory positions are available in the brain.
 
@@
 The strength of the double-drift illusion probably arises because of the very poor location information for a gabor pattern in the periphery when it has the same mean luminance as the background.
-With poor positional certainty, the motion information contributes to the location estimate, generating a shift in position.
+With poor positional certainty, the motion information contributes to the location estimate, generating a shift in perceived position.
 If the gabor itself is not moving [@DeValois1991], the shift saturates after about 100 ms [@Chung2007], perhaps because the accumulating evidence of actual location of the stationary gabor patch is sufficient to anchor its position.
@@
 This long duration before resetting or saturation is well beyond the integration period of other visual effects and into the range of short-term memory.
-Given that both true and illusory positions are available, it could be that instead of some integration period, it is the interplay between the two position signals that leads to resets.
+Given that both true and illusory positions are available, it could be that instead of some integration period, it is the interplay between the two position signals that leads to spontaneous resets.
+Alternatively, there is a simple periodicity or stochastics process underlies the changes in perception.
 
@@
 Earlier results for pointing to the double-drift target [@Lisi2017] show that the illusion persists for manual pointing.
+However, recent work has suggested that movement-related attention may lead to resets [@Nakayama2020], so that continuous real-time tracking of the gabor's position, may prevent the illusory shift in position.
 The ability to manually track the perceived position will also allow us to study the resets – the adjustment of perceived location when the illusion has deviated too much or too long from the physical location.
-It could be that the imprecise, but true, retinal position signal limits the illusion, or it could be that – like other illusions, such as bi-stable stimuli – there is a simple periodicity underlying changes in perception.
-Here we capture where and when observers experience resets by having them draw the perceived trajectory of the stimulus while seeing it (a tracking procedure), or after having seen it (a retrace procedure).
-By investigating the time and location of resets in the recorded trajectories, we can distinguish between a spatial limit to the illusion or a temporal limit.
+Hence, we first test if the illusion occurs when people continunously track the position of the double drift stimulus.
+Then we try to capture where and when observers experience resets by having them draw the perceived trajectory of the stimulus while seeing it (a tracking procedure), or after having seen it (a retrace procedure).
+Investigating the time and location of resets in the recorded trajectories, allows us to distinguish between a spatial limit to the illusion or a temporal limit.
 
@@
 For experiment 2, 9 participants were recruited from the lab (6 female; ages 19 - 27, mean: 22.8).
-All participants reported to be right handed and have either normal or corrected to normal vision.
+All participants reported to be right handed and had either normal or corrected to normal vision.
 Procedures were approved by York’s Human Participants Review Committee, and all participants provided prior, written, informed consent.
@@
 
-Participants were to be made For reporting degrees of visual angle that stimuli spanned ( _dva_ ), we assume the participant is close to the setup, with their eyes at the height of the monitor.
-Since there was no requirement to use the forehead rest, the latter is probably only true for the tallest participants, so that reported _dva's_ are underestimating the size of stimuli.
+For reporting degrees of visual angle that stimuli spanned ( _dva_ ), we assumed the participants to be close to the setup, with their eyes at the height of the monitor.
+Since there was no requirement to use the forehead rest, the latter is probably only true for the tallest participants, so that reported _dva's_ are slightly underestimating the size of stimuli.
 
@@
 Both tasks asked participants to use a stylus on a drawing tablet to track a moving gabor with some amount of internal motion that elicited the double-drift illusion.
-The idea was that whenever spontaneous resets ofthe illusion occur, these would be recorded in the tracking data as sudden changes in heading (cf.
-Fig 1, bottom left, the dashed purple lines).
+The idea was to test both if tracking the illusion continuously would preserve the illusion and if the tracked paths reveal resets.
 
@@
 The average angle of the tracked path is given in degrees deviation from straight forward.
-It seems that continuous tracking is sensitive to the strength of the illusion, and this is confirmed by a repeated-measures F-test on the average angles, using internal motion as a within-subjects factor (_F_(4,12)=102.6, _p_<.001, $\eta^2$=0.97).
+It seems that continuous tracking is sensitive to the strength of the illusion, and this is confirmed by a repeated-measures F-test on the average angles, using internal motion as a within-subjects factor ( _F_(4,12)=102.6, _p_<.001, $\eta^2$=0.97 ).
 With FDR-corrected, paired t-tests, we find a difference in heading between 0 _cps_ internal motion and both 1 _cps_ ( _p_=.022) and -1 _cps_ ( _p_=.025) as well as between -3 and -1 _cps_ and 3 and 1 _cps_ (both _p_ <.001).
-This means we can use the averaged angle for each of the internal motion conditions as a measure of the strength of the illusion.
+This makes it clear that continuous tracking preserves the illusion.
+We will use the averaged angle for each of the internal motion conditions as a measure of the strength of the illusion.
 
@@
 We use a 95% confidence interval of heading in the 0 internal motion condition to select trajectories from conditions with non-zero heading that may show a reset.
-These trajectories are drawn in red (only 1 per condition/participant), but as can be seen, the deviations from the condition's dominant heading occur just before or after the external motion reversals, so they may simply reflect difficulty of tracking the stimulus, rather than resets.
+These trajectories are drawn in red (only 1 per condition/participant), but as can be seen, the deviations from the condition's dominant heading occur just before or after the external motion reversals, so they may simply reflect difficulty of tracking the stimulus at the reversal points, rather than resets.
 
@@
 without the need for real-time tracking.
-Nakayama & Holcombe [-@Nakayama2019; -@Nakayama2019b] have also used a re-tracing paradigm, in their case to study triggered resets.
+Nakayama & Holcombe [-@Nakayama2020] have also used a re-tracing paradigm, in their case to study triggered resets.
 The re-tracing task was done in half of 8 blocks.
@@
 
-## Comparison with Sirui's data set
+## Comparison with PSE data
 
@@
 In an as of yet unpublished study [Sirui] obtained the point of subjective equivalence (PSE) of a double drift stimulus, and a static point that appeared some time after the double drift stimulus disappeared.
-The double-drift stimulus was on a 2 or 4 DVA length trajectory, and disappeared after 1, 2 or 3 seconds.
-The PSE's can be modelled very accurately with a model that assumes a fixed chance of a reset per unit of time (~.7%/s), that is: it explains the data using only a temporal bound on the illusion.
+The double-drift stimulus was on a 2 or 4 DVA length trajectory, traversed in 1, 2 or 3 seconds.
+The PSE's can be modelled very accurately with a model that assumes a fixed chance of a reset per unit of time (~0.7%/s), that is: it explains the data using only a temporal bound on the illusion.
 Interestingly, this fully contradicts the current findings.
@@
 
-```{r fig8, echo=FALSE, message=FALSE, fig.width=8, fig.height=3.5, fig.cap="\\label{fig:fig8}**Figure 8. PSE's of double-drift position and continuous tracking**"}
-plotDeviations()
+```{r fig8, echo=FALSE, message=FALSE, fig.width=6, fig.height=8, fig.cap="\\label{fig:fig8}**Figure 8. PSE's of double-drift position and continuous tracking**"}
+plotAverageTrajectories()
 ```
@@
 If the PSE's express the average perceived deviations from the actual trajectory of the double-drift stimulus, then they should show a similar pattern as the averaged trajectories in experiment 2.
-In order to be able to compare the two data sets, we will include all trajectories where our simple algorithm did not detect a reset, which should also be included in the PSE data.
-Further, we scale the trajectories, so that the Y end-point matches the duration of stimulus.
+In order to be able to compare the two data sets, we will include all trajectories, including those where our simple algorithm did not detect a reset, as such trials should also be included in the PSE data.
+Further, we scale the trajectories on both axes, so that the Y end-point matches the duration of stimulus.
 Since the PSE data was collected in degrees visual angle, we convert the unit to degrees visual angles, using the estimate at the centre of the workspace.
@@
 First, we notice that the trajectory data is highly dependent on participants.
-It could be that this method provides very noisy estimates, hiding any effect.
+It could be that the re-tracing method provides very noisy estimates, hiding any effect.
 It could also be that the effects truly vary a lot between participants.
-Since we can see that some participants do seem to follow the pattern found with the PSE approach, and some show a pattern of increasing average deviation between perceived and true position, it may that the difference between the data sets can simply be explained by random sampling of relatively small numbers of participants.
+Since we can see that some participants do seem to follow the pattern found with the PSE approach, and some show a pattern of increasing average deviation between perceived and true position, it may that the differences between the data sets can simply be explained by random sampling of relatively small numbers of participants.
 
@@
 
-Recently, it has been suggested that resets can be triggered by planned eye-movements to the stimulus [@Nakayama2019; @Nakayama2019b], particularly "jump" resets.
+Recently, it has been suggested that resets can be triggered by planned eye-movements to the stimulus [@Nakayama2020], particularly "jump" resets.
 In this experiment, we did not control for eye-movements with an eye-tracker, so we can't directly investigate this idea.

### 2020-09-29 Manuscript.Rmd
added comparison of tracking data with Cavanagh & Tse 2019

@@
 output:
+  pdf_document: default
   word_document: default
-  pdf_document: default
   html_document:
@@
 source('R/tracking_bounded.R')
-# And finally some code for the re-tracing experiment...
+# Here we compare the tracking data with an earlier paper:
+source('R/compare.R')
+# And then some code for the re-tracing experiment...
 source('R/retrace_onePass_V4.R')
@@
 However, when the gabor itself also moves as in the double-drift stimulus, there is no stable location information to anchor the position estimate and the build-up continues away from its physical location up to several times the size of the gabor.
+The strength of the illusion, or how much the perceived position shifts from the real position over time, depends on the speed of both the internal and external motion [@Tse2006; @Shapiro2010; @Kwon2015].
+
 At some point, the illusion resets, the perceived location jumps back to its physical location and then accumulation begins again (Fig.
@@
 Alternatively, there is a simple periodicity or stochastics process underlies the changes in perception.
+Recent work has shown that independent of the stimulus, movement-related attention may lead to resets [@Nakayama2020].
+This means that participants' responses to the stimulus may change the illusion.
+Here we will first test if real-time as well as delayed, manual tracking of the gabor's position, prevents or decreases the illusory shift in position.
 
-Here, we examine the double-drift illusion by asking participants to trace out its path with a stylus on a tablet that is optically aligned with the monitor surface but with the hand not visible.
-Earlier results for pointing to the double-drift target [@Lisi2017] show that the illusion persists for manual pointing.
-However, recent work has suggested that movement-related attention may lead to resets [@Nakayama2020], so that continuous real-time tracking of the gabor's position, may prevent the illusory shift in position.
-The ability to manually track the perceived position will also allow us to study the resets – the adjustment of perceived location when the illusion has deviated too much or too long from the physical location.
-Hence, we first test if the illusion occurs when people continunously track the position of the double drift stimulus.
-Then we try to capture where and when observers experience resets by having them draw the perceived trajectory of the stimulus while seeing it (a tracking procedure), or after having seen it (a retrace procedure).
+The ability to record trajectories the perceived position will then allow us to study spontaneous resets – the adjustment of perceived location when the illusion has deviated too much or too long from the physical location.
+In particular, if a conflict between two location signals [@Lisi2015] leads to resets this would predict a set distance between the real and perceived position where resets occur.
+Under this hypothesis, the timing of spontaneous resets would depend on the speed of internal motion (given constant external motion).
+On the other hand, if spontaeous resets are the result of a process that integrates over time, resets would occur after a certain amount of time, but not be dependent on internal motion.
 Investigating the time and location of resets in the recorded trajectories, allows us to distinguish between a spatial limit to the illusion or a temporal limit.
+This in turn will inform us on the processes underlying the perception of positions of moving objects.
 
@@
 The stimulus used 1.5 phases of a sine wave, so that each sine phase equals ~0.58 _cm_ or ~1.5 _dva_.
-By applying both real motion and phases changes of the gabor, it becomes a 'double-drift stimulus' where the perceived path deviates from the real path.
+By applying both real motion and phases changes to the gabor, it becomes a 'double-drift stimulus' where the perceived path deviates from the real path.
 The gabor changed phase at different rates, called 'internal motion' here, specified in cycles per second ( _cps_ ).
 The real displacements of the stimulus are along the Y axis of the monitor, keeping the X coordinate constant at the middle of the screen, so that a stimulus without internal motion would appear to move forward and backward on the horizontal display.
-This real motion was restricted to the central 13.5 _cm_ ( ~14.0 _dva_ ) and was set so that stimulus would move from either extreme to the other ( a motion we call a 'pass' ) in 2, 3 or 4 _s_, corresponding to speeds of 6.75, 4.5 and 3.75 _cm/s_ ( or roughly 7.0, 4.7 or 3.5 _dva/s_ ).
+This real motion was restricted to the central 13.5 _cm_ ( ~14.0 _dva_ ) and was set so that the stimulus would move from either extreme to the other ( a motion we call a 'pass' ) in 2, 3 or 4 _s_, corresponding to speeds of 6.75, 4.5 and 3.75 _cm/s_ ( or roughly 7.0, 4.7 or 3.5 _dva/s_ ).
 For simplicity, and since the distance moved is always the same, these speeds will be refered to by the duration of a single pass (2, 3 or 4 _s_).
@@
 In the two tasks in experiment 1, participants were to do online tracking (using their unseen hand / stylus) while viewing a double-drift stimulus.
-In experiment 2, participants first watched a double-drift stimulus and then either reproduce the perceived path of the stimulus by retracing it on the tablet, or indicate the initial movement direction by changing the orientation of a line originating in the same position as the doube-drift stimulus.
+In experiment 2, participants first watched a double-drift stimulus and then either reproduced the perceived path of the stimulus by retracing it on the tablet, or indicated the initial movement direction by changing the orientation of a line originating in the same position as the doube-drift stimulus.
 More details are given below.
@@
 This heading measure should depend on the illusion strength.
-Moreoever, when spontaneous resets are either time-dependent or space-dependent, resets should occur at roughly the same points and in the same direction in a given condition for every participant.
-This means that spontaneous resets should result in systematic deviations of every participant's average heading over time.
 
@@
 However, the individual trajectories reveal more points that are potential spontaneous resets that we try to extract.
-Using a smooth spline interpolated version of the trajectory we detect direction changes.
+Using a smooth spline interpolated version of the trajectory we detect left-right direction changes.
 Since the initial movement is often jittery, we only take direction changes further than ~0.39 _cm_ away from the starting point.
@@
 Both tasks asked participants to use a stylus on a drawing tablet to track a moving gabor with some amount of internal motion that elicited the double-drift illusion.
-The idea was to test both if tracking the illusion continuously would preserve the illusion and if the tracked paths reveal resets.
+We test if tracking the illusion continuously, prevents or decreases the illusion.
 
@@
 With FDR-corrected, paired t-tests, we find a difference in heading between 0 _cps_ internal motion and both 1 _cps_ ( _p_=.022) and -1 _cps_ ( _p_=.025) as well as between -3 and -1 _cps_ and 3 and 1 _cps_ (both _p_ <.001).
-This makes it clear that continuous tracking preserves the illusion.
+This makes it clear that continuous tracking does not prevent the illusion from ocurring.
 We will use the averaged angle for each of the internal motion conditions as a measure of the strength of the illusion.
@@
 
-The average headings are close to straight lines for all conditions.
-Since average straight trajectories could either result from uniform tracking paths, each with no signs of resets or from paths with many differently shaped, directed and timed resets, they do not prove or disprove neither the absence nor the presence of resets.
-Hence, we also show full trajectories of all participants in each of the 5 internal motion conditions.
-
-In order to detect potential resets, we consider that in the 0 internal motion condition there is no illusion, so that we measure the tracking of the veridical path of the stimulus.
-Since both strong, "jump" resets and weak "hit-the-wall" resets (see Fig 1, bottom left) would change percieved direction back to or crossing the veridical motion path, their smooth-spline interpolated paths would contain samples similar in heading to the 0 internal motion condition.
-We use a 95% confidence interval of heading in the 0 internal motion condition to select trajectories from conditions with non-zero heading that may show a reset.
-These trajectories are drawn in red (only 1 per condition/participant), but as can be seen, the deviations from the condition's dominant heading occur just before or after the external motion reversals, so they may simply reflect difficulty of tracking the stimulus at the reversal points, rather than resets.
-
-In a second task, we only used the higher internal speed (±3 _cps_) and increased the duration of a single pass of the stimulus from 2 _s_ to 4 _s_, both the increase the occurence of spontaneous resets.
+In a second task, we only used the higher internal speed (±3 _cps_) and increased the duration of a single pass of the stimulus from 2 _s_ to 4 _s_, both to increase the occurence of spontaneous resets.
 Average headings and heading distributions for 6 participants can be seen in Figure 3.
+Despite lower external motion speed, the illusion is as strong as in the conditions with the same internal motion in the first task.
 
@@
 ```
+
+### Illusion strength
+
+The illusion is clearly there in both tasks, but does tracking decrease the strength of the illusion compared to simply observing it?
+We compare illusion strength to that reported previously [@Cavanagh2019] in Figure 4.
+As can be seen the averages from our data here (in red) coincide almost exactly with the illusion strength predicted by the model.
+That is; tracking the illusion while observing it does not change the strength of the illusion.
+
+```{r fig4, echo=FALSE, message=FALSE, fig.width=8, fig.height=5, fig.cap="\\label{fig:fig4}**Figure 4.
+Illusion strength.** The strength of the illusion in degrees angle deviation from the physical path is plotted over external motion speed ( _dva/s_ ).
+In red are the average illusions strengths for ±3 _cps_, ±1 _cps and 0 cps.
+The black circles and line are the data and model from Cavanagh & Tse (2019).
+The grey lines are the illusion strengths their model predicts for the internal motion speeds used here."}
+plotStrengthComparison()
+```
+
+### Resets
+
+In order to detect potential resets in the first task, we consider that in the 0 internal motion condition there is no illusion, so that we measure the tracking of the veridical path of the stimulus.
+Since both strong, "jump" resets and weak "hit-the-wall" resets (see Fig 1, bottom left) would change percieved direction back to or crossing the veridical motion path, their smooth-spline interpolated paths would contain samples similar in heading to the 0 internal motion condition.
+We use a 95% confidence interval of heading in the 0 internal motion condition to select trajectories from conditions with non-zero heading that may show a reset.
+These trajectories are drawn in red (only 1 per condition/participant), but as can be seen, the deviations from the condition's dominant heading occur just before or after the external motion reversals, so they may simply reflect difficulty of tracking the stimulus at the reversal points, rather than resets.
+
+The second task does not give any additional information.
 
@@
 
-In Figure 4, for each of the 6 combinations of internal speed (2, 3 or 4 _cps_) and external speed (3 or 4 _s_ stimulus pass) we show the data for one example participant, as well as the heading distribution for all participants.
+In Figure 5, for each of the 6 combinations of internal speed (2, 3 or 4 _cps_) and external speed (3 or 4 _s_ stimulus pass) we show the data for one example participant, as well as the heading distribution for all participants.
 While the heading distributions don't show resets, for many of the participants, we can see that the re-traced paths do show resets, many of which are detected by our algorithm.
 
-```{r fig4, echo=FALSE, message=FALSE, fig.width=8, fig.height=5, fig.cap="\\label{fig:fig4}**Figure 4.
+```{r fig5, echo=FALSE, message=FALSE, fig.width=8, fig.height=5, fig.cap="\\label{fig:fig5}**Figure 5.
 Example Trajectories and Heading in Experiment 2.** For each of the six conditions, defined by 3 internal speeds (2, 3 and 4 _cps_) and two external speeds (corresponding to 3 and 4 _s_ passes), we show one example participant's trajectories and reset points, as well as the heading distribution."}
@@
 The strength of the illusion should depend on both the internal and external speed of the double drift stimulus.
-As can be seen in Figure 5 (left), at least the internal speed, and perhaps the external speed may predict illusion strength as measured by line angles.
+As can be seen in Figure 6 (left), at least the internal speed, and perhaps the external speed may predict illusion strength as measured by line angles.
 To test this, we perform a mixed-design ANOVA on line angle, using both internal and external speed as within-subject factors.
@@
 
-```{r fig5, echo=FALSE, message=FALSE, fig.width=8, fig.height=4, fig.cap="\\label{fig:fig5}**Figure 5.
+```{r fig6, echo=FALSE, message=FALSE, fig.width=8, fig.height=4, fig.cap="\\label{fig:fig6}**Figure 6.
 Measures of Illusion Strength** Line angle responses and reset point direction should both measure illusion strength.
@@
 If both the angle half-way to reset and line angle responses are good measures of illusion strength, they should be related.
-As can be seen from the data in Figure 5 (right), there might be a relation between the two variables, but there is also heteroscadasticity.
+As can be seen from the data in Figure 6 (right), there might be a relation between the two variables, but there is also heteroscadasticity.
 We counter this by doing a weighted regression, with the weights set to one over the squared line angle ($\frac{1}{\alpha^2}$).
@@
 Now we look at the distribution of reset points, and whether they can be explained by a temporal or spatial limit.
-In Figure 6 (left), we can see that the reset points vary more along the Y-axis than the X-axis.
+In Figure 7 (left), we can see that the reset points vary more along the Y-axis than the X-axis.
 That is, predicting reset points from a single Y coordinate (time-cutoff) would leave larger residuals than predicting reset points from a single X coordinate (or spatial cutoff).
@@
 
-```{r fig6, echo=FALSE, message=FALSE, fig.width=8, fig.height=6, fig.cap="\\label{fig:fig6}**Figure 6.
+```{r fig7, echo=FALSE, message=FALSE, fig.width=8, fig.height=6, fig.cap="\\label{fig:fig7}**Figure 7.
 Reset Points and Single Limits** **Left:** The average reset points for each participant and stimulus condition displayed spatially.
@@
 
-Figure 7 shows both the actual reset point coordinates (dots), given the illusion strength as the angle to a sample half-way to the reset point, as well as the model's predictions of reset point coordinates (lines).
+Figure 8 shows both the actual reset point coordinates (dots), given the illusion strength as the angle to a sample half-way to the reset point, as well as the model's predictions of reset point coordinates (lines).
 
-```{r fig7, echo=FALSE, message=FALSE, fig.width=6, fig.height=6, fig.cap="\\label{fig:fig7}**Figure 7. Reset Points and Combined Limits**"}
+```{r fig8, echo=FALSE, message=FALSE, fig.width=6, fig.height=6, fig.cap="\\label{fig:fig8}**Figure 8. Reset Points and Combined Limits**"}
 plotSequentialModel()
@@
 While here we find that a spatial limit is the better explanation for resets, other evidence may point at a temporal limit instead.
-In an as of yet unpublished study [Sirui] obtained the point of subjective equivalence (PSE) of a double drift stimulus, and a static point that appeared some time after the double drift stimulus disappeared.
+In an as of yet unpublished study [Sirui Liu] obtained the point of subjective equivalence (PSE) of a double drift stimulus, and a static point that appeared some time after the double drift stimulus disappeared.
 The double-drift stimulus was on a 2 or 4 DVA length trajectory, traversed in 1, 2 or 3 seconds.
@@
 The model predicts that the PSE decreases over time as it becomes more likely that a reset has occurred.
-We will compare the two data sets here.
+We will compare the two data sets here (see Fig 9).
 
-```{r fig8, echo=FALSE, message=FALSE, fig.width=6, fig.height=8, fig.cap="\\label{fig:fig8}**Figure 8. PSE's of double-drift position and continuous tracking**"}
+```{r fig9, echo=FALSE, message=FALSE, fig.width=6, fig.height=8, fig.cap="\\label{fig:fig9}**Figure 9. PSE's of double-drift position and continuous tracking**"}
 plotAverageTrajectories()
@@
 For the 4-second stimuli, we removed the last second.
-Figure 8 shows the individual participants' and the averaged trajectories (purple, red, orange) as well as the PSE's obtained in the other study (shown separately for 2 and 4 DVA stimulus trajectories) in gray.
+Figure 9 shows the individual participants' and the averaged trajectories (purple, red, orange) as well as the PSE's obtained in the other study (shown separately for 2 and 4 DVA stimulus trajectories) in gray.
 
@@
 
-We set out to test if spontaneous resets in the double-drift illusion reflect discrepancies between two position signals, occurring at some given size of this discrepancy, or if they simply occur after some time elapses, analogous to other illusions or bi-stable stimuli.
+We first tested if tracking the position of a double-drift stimulus changes the illusory percept of it's position, and found that it doesn't.
+We then set out to test if spontaneous resets in the double-drift illusion reflect discrepancies between two position signals, occurring at some given size of this discrepancy, or if they simply occur after some time elapses, analogous to other illusions or bi-stable stimuli.
 To measure the location of resets we first ask participants to track the perceived location of the double-drift stimulus online.

### 2020-09-29 R/compare.R
create plot for comparison of tracking data with Cavanagh & Tse, 2019

@@ -1,0 +1,85 @@
+
+
+functionCavanaghTse2019 <- function(internal=c(), external=c(), k=0.81) {
+  
+  # speeds need to be in dva/s
+  
+  # perceived angle = arctan of ( k times internal motion over external motion )
+  illusionstrength <- atan ( k * (internal %o% (1/external)) )
+  
+  # convert to degrees deviation from veridical:
+  illusionstrength <- (illusionstrength / pi) * 180
+  
+  # add row and column names containing the speeds
+  rownames(illusionstrength) <- as.character(internal)
+  colnames(illusionstrength) <- as.character(external)
+  
+  # return the info
+  return(illusionstrength)
+  
+}
+
+plotStrengthComparison <- function() {
+  
+
+  #plot(c(5:35),IS['7.800',], type='l')
+  
+  plot(-1000,-1000,xlim=c(0,40), ylim=c(0,60), xlab='external motion [dva/s]', ylab='illusion strength [deg]', bty='n', ax=F)
+  
+  # use these external models for prediction lines:
+  ems <- seq(3,35,0.2)
+  
+  
+  # Cavanagh & Tse 2019 model:
+  external <- c(3:35)
+  internal <- c(7.8)
+  IS_CT19 <- functionCavanaghTse2019(internal, external=ems)
+  lines(ems,IS_CT19['7.8',],lty=1,col='#000000')
+  
+  text(x=36, y=IS_CT19['7.8', length(ems)], '7.8 dva/s', adj=c(0,0.5))
+  text(x=36, y=20, 'internal\nmotion:', adj=c(0,0.5))
+  
+  ex_spd <- c(31.25, 15.63,  7.81, 15.63,  7.81,  3.91)
+  illstr <- c(11.31, 21.57, 37.03, 19.41, 38.41, 60.12)
+  
+  points(ex_spd, illstr, pch=1, col='#000000')
+  
+  # predictions for experiment 1a:
+  external <- c(7)
+  internal <- c(1,3) * 1.5
+  # our conditions:
+  IS_Ex1a <- functionCavanaghTse2019(internal, external)
+  lines(c(7,7),c(0,IS_Ex1a[2,'7']),col='#999999',lty=3)
+  #text(x=7,y=-2.5,'1a',adj=c(0.5,0.5))
+  
+
[…]

### 2020-10-05 Manuscript.Rmd
edits to writing

@@
 output:
-  pdf_document: default
   word_document: default
@@
     df_print: paged
+  pdf_document: default
 header-includes:
@@
 
-In this experiment, two tasks were performed by two separate groups of participants.
-Both tasks asked participants to use a stylus on a drawing tablet to track a moving gabor with some amount of internal motion that elicited the double-drift illusion.
-We test if tracking the illusion continuously, prevents or decreases the illusion.
+In this experiment, participants were asked to use a stylus on a drawing tablet to track a moving gabor with some amount of internal motion that elicited the double-drift illusion.
+We test if tracking the illusion continuously is a good measure of illusion strength, or if it prevents or decreases the illusion.
 
-## Tasks
+## Task
 
-In the first task, participants were asked to continuously track the position of a gabor that moved forward and backward along the Y-axis (of the horizontal screen) for the duration of 12 _s_, using a stylus on a drawing tablet.
+In this task, participants were asked to continuously track the position of a gabor that moved forward and backward along the Y-axis (of the horizontal screen) for the duration of 12 _s_, using a stylus on a drawing tablet.
 Before trial onset, the participant moved the stylus to the middle of the screen ( where the gabor would appear ), and was subsequently asked to fixate a point on the left or right of the screen, and given 1.5 _s_ to do so.
@@
 Since each moving gabor was shown for 12 _s_, and it moved from one end to another in 2 _s_, it completed 5 full and 2 half passes of the path.
-
-In the second task, only the highest internal motion was used (±3 _cps_), trials include only 3 complete passes, but these took 4 _s_ each.
-As before, the fixation point (left or right) was random for each trial, while initial external motion (forward or backward) as well as internal motion were counterbalanced across trials.
 
@@
 
-In a second task, we only used the higher internal speed (±3 _cps_) and increased the duration of a single pass of the stimulus from 2 _s_ to 4 _s_, both to increase the occurence of spontaneous resets.
-Average headings and heading distributions for 6 participants can be seen in Figure 3.
-Despite lower external motion speed, the illusion is as strong as in the conditions with the same internal motion in the first task.
+## Illusion strength
 
-```{r fig3, echo=FALSE, message=FALSE, fig.width=8, fig.height=4, fig.cap="\\label{fig:fig3}**Figure 3.
-Heading in Experiment 1b.** Average heading and heading distribution across the middle 3 _s_ of each pass.
-Distance from origin is time, angle is angle.
-In dark blue lines the average heading, in light blue a polar heatmap giving an impression of the distribution of heading.
-**Left:** Data for all participants.
-**Right:** Group average."}
-plotBoundedTracking_V2()
-```
-
-### Illusion strength
-
-The illusion is clearly there in both tasks, but does tracking decrease the strength of the illusion compared to simply observing it?
-We compare illusion strength to that reported previously [@Cavanagh2019] in Figure 4.
+The data clearly show the illusion, but does tracking decrease the strength of the illusion compared to simply observing it?
+We compare illusion strength to that reported previously [@Cavanagh2019] in Figure 3.
 As can be seen the averages from our data here (in red) coincide almost exactly with the illusion strength predicted by the model.
@@
 
-```{r fig4, echo=FALSE, message=FALSE, fig.width=8, fig.height=5, fig.cap="\\label{fig:fig4}**Figure 4.
+```{r fig3, echo=FALSE, message=FALSE, fig.width=8, fig.height=5, fig.cap="\\label{fig:fig3}**Figure 3.
 Illusion strength.** The strength of the illusion in degrees angle deviation from the physical path is plotted over external motion speed ( _dva/s_ ).
@@
 
-### Resets
+## Resets
 
-In order to detect potential resets in the first task, we consider that in the 0 internal motion condition there is no illusion, so that we measure the tracking of the veridical path of the stimulus.
+In order to detect potential resets, we consider that in the 0 internal motion condition there is no illusion, so that we measure the tracking of the veridical path of the stimulus.
 Since both strong, "jump" resets and weak "hit-the-wall" resets (see Fig 1, bottom left) would change percieved direction back to or crossing the veridical motion path, their smooth-spline interpolated paths would contain samples similar in heading to the 0 internal motion condition.
 We use a 95% confidence interval of heading in the 0 internal motion condition to select trajectories from conditions with non-zero heading that may show a reset.
-These trajectories are drawn in red (only 1 per condition/participant), but as can be seen, the deviations from the condition's dominant heading occur just before or after the external motion reversals, so they may simply reflect difficulty of tracking the stimulus at the reversal points, rather than resets.
+These trajectories are drawn in red in Figure 2 (for clarity only 1 per condition/participant is highlighted), but as can be seen, the deviations from the condition's dominant heading occur just before or after the external motion reversals, so they may simply reflect difficulty of tracking the stimulus at the reversal points, rather than resets.
 
-The second task does not give any additional information.
-
-Except for the first part of participant 12's trajectories, all heading data matched that of the 3 _cps_ condition in experiment 1a.
-Those first few samples in participant 12 could reflect a participant specific delay in tracking the stimulus position.
-In other words, no unambiguous resets were captured in either experiment.
+No unambiguous resets were captured in either experiment.
 It is possible that resets were perceived but were too quick to respond to in the continuous tracking task.
 Alternatively, tracking may have stabilized the illusion, preventing resets from occurring.
-Hence, we performed a second experiment that would counter both explanations for the absence of (spontaneous) resets.
+Either way, we performed a second experiment that would counter both explanations for the absence of (spontaneous) resets.
 
@@
 
-In a second experiment, we ask participants to re-trace the perceived path of a double-drift stimulus, immediately after viewing the stimulus.
+In a second experiment, we ask participants to re-trace the perceived path of a double-drift stimulus, immediately after viewing the stimulus, as used by Nakayama & Holcombe [-@Nakayama2020].
 This allows participants more time to carefully reproduce the path.
@@
 without the need for real-time tracking.
-Nakayama & Holcombe [-@Nakayama2020] have also used a re-tracing paradigm, in their case to study triggered resets.
 The re-tracing task was done in half of 8 blocks.
@@
 The double-drift gabor started at the near end of the display, and moved along the horizontal display away from the participant, always making a single, forward pass.
-After, the gabor had disappeared, participants could respond in one of two ways.
+After the gabor had disappeared, participants could respond in one of two ways.
 When re-tracing the perceived path with the stylus on the tablet, the drawn path would show up as a red line.
@@
 They could "key in" their response by pressing ENTER.
-In both tasks, there was no time-limit for the response, and careful responding was encouraged.
+In both tasks, there was no time-limit for the response.
 
@@
 
-In Figure 5, for each of the 6 combinations of internal speed (2, 3 or 4 _cps_) and external speed (3 or 4 _s_ stimulus pass) we show the data for one example participant, as well as the heading distribution for all participants.
+In Figure 4, for each of the 6 combinations of internal speed (2, 3 or 4 _cps_) and external speed (3 or 4 _s_ stimulus pass) we show the data for one example participant, as well as the heading distribution for all participants.
 While the heading distributions don't show resets, for many of the participants, we can see that the re-traced paths do show resets, many of which are detected by our algorithm.
 
-```{r fig5, echo=FALSE, message=FALSE, fig.width=8, fig.height=5, fig.cap="\\label{fig:fig5}**Figure 5.
+```{r fig4, echo=FALSE, message=FALSE, fig.width=8, fig.height=5, fig.cap="\\label{fig:fig4}**Figure 4.
 Example Trajectories and Heading in Experiment 2.** For each of the six conditions, defined by 3 internal speeds (2, 3 and 4 _cps_) and two external speeds (corresponding to 3 and 4 _s_ passes), we show one example participant's trajectories and reset points, as well as the heading distribution."}
@@
 
-On average, the algorithm detects resets in 75% of trials.
+On average, the algorithm detects resets in ~75% of trials.
 For three participants, only about half the trials show resets (42%, 44% and 58%).
 For the remaining six participants the algorithm detects resets in 72%-99% of trials, with an average of 88%.
-The minimum number of resets in one of the six conditions for any participant is 7 (out of 24), with the average at 17.
-Since participants experiencing a "hit-the-wall reset" could display a direction change in their paths only by chance, 50% of trials with resets seems reasonable.
-Neither percept is incorrect, and the average reset positions still indicates some limit to the illusion.
+The minimum number of resets in each of the six conditions for any participant is 7 (out of 24 trials), with the average at ~17.
+Since participants experiencing a "hit-the-wall reset" (as opposed to a "jump" reset) might not always display a direction change in their paths, 50% of trials with resets seems reasonable.
+Either type of perceived resets appears valid, and the average reset positions still indicates some limit to the illusion.
 Therefore, we keep all participants and resets.
@@
 The strength of the illusion should depend on both the internal and external speed of the double drift stimulus.
-As can be seen in Figure 6 (left), at least the internal speed, and perhaps the external speed may predict illusion strength as measured by line angles.
+As can be seen in Figure 5 (left), at least the internal speed, and perhaps the external speed may predict illusion strength as measured by line angles.
 To test this, we perform a mixed-design ANOVA on line angle, using both internal and external speed as within-subject factors.
@@
 
-```{r fig6, echo=FALSE, message=FALSE, fig.width=8, fig.height=4, fig.cap="\\label{fig:fig6}**Figure 6.
+```{r fig5, echo=FALSE, message=FALSE, fig.width=8, fig.height=4, fig.cap="\\label{fig:fig5}**Figure 5.
 Measures of Illusion Strength** Line angle responses and reset point direction should both measure illusion strength.
@@
 
-We also take the angle of the sample on each re-tracing trajectory that is halfway between the origin and the detected reset, as an alternative measure of illusion strength.
+As specified above, we also take the angle of the sample on each re-tracing trajectory that is halfway between the origin and the detected reset, as an alternative measure of illusion strength.
 If both the angle half-way to reset and line angle responses are good measures of illusion strength, they should be related.
-As can be seen from the data in Figure 6 (right), there might be a relation between the two variables, but there is also heteroscadasticity.
+As can be seen from the data in Figure 5 (right), there might be a relation between the two variables, but there is also heteroscadasticity.
 We counter this by doing a weighted regression, with the weights set to one over the squared line angle ($\frac{1}{\alpha^2}$).
@@
 That is, predicting reset points from a single Y coordinate (time-cutoff) would leave larger residuals than predicting reset points from a single X coordinate (or spatial cutoff).
-In fact, the standard deviation of reset point X-coordinates is about 21.9% of teh standard deviation of reset point Y-coordinates.
+In fact, the standard deviation of reset point X-coordinates is about 21.9% of the standard deviation of reset point Y-coordinates.
 This already indicates that a spatial limit explains about 4 times more of the variance in reset points than a temporal limit.
 
-```{r fig7, echo=FALSE, message=FALSE, fig.width=8, fig.height=6, fig.cap="\\label{fig:fig7}**Figure 7.
-Reset Points and Single Limits** **Left:** The average reset points for each participant and stimulus condition displayed spatially.
-Y-coordinates have been converted to time, in order to normalize for the two external speeds.
+```{r fig6, echo=FALSE, message=FALSE, fig.width=8, fig.height=6, fig.cap="\\label{fig:fig6}**Figure 6.
+Reset Points and Single Limits** **Left:** True path of the doubel-drift stimulus as gray arrow.
+The average reset points for each participant and stimulus condition displayed spatially, with the internal motion normalized to go rightward.
+Y-coordinates have been converted to time ( _s_ ), in order to normalize for the two external speeds.
 Transparent purple lines indicate the standard deviation of coordinates.
@@
 Nevertheless, we test two simple models, based on straighforward trigonometry.
-First, we calculate the slope of a line through the origin and a sample halfway to the reset point.
-Then, if the perceived path of the gabor falls on a straight line, and reset points occur at a set limit, then a temporal limit would predict X coordinates:
+First, we calculate the slope of a line ($s$) through the origin and a sample halfway to the reset point.
+This estimates illsuion strength as angular deviation from the true, straight path.
+If the perceived path of the gabor is approximated by a straight line (up to a reset), and reset points occur at a set limit, then a temporal limit ($L_y$) would predict X coordinates:
 
@@
 
-And a spatial limit would predict Y coordinates:
+And a spatial limit ($L_x$) would predict Y coordinates:
 
@@
 
-Where $L_y$ and $L_x$ are a temporal (y) and spatial (x) limit that are free parameters, and s corresponds to the slope.
-If we fit these simple models by minimizing the mean square error (as the distance in centimeters between predicted and actual reset points), the optimal parameters indicate a temporal limit of ~1.4 _s_ (AIC=258.4) and a spatial limit of ~1.7 _cm_ (AIC=217.7).
+Where $L_y$ and $L_x$ are a temporal (y) and spatial (x) limit that are free parameters, and s corresponds to the calculated slope.
+If we fit these simple models by minimizing the mean square error (as the absolute distance in centimeters between predicted and actual reset points), the optimal parameters indicate a temporal limit of ~1.4 _s_ (AIC=258.4) and a spatial limit of ~1.7 _cm_ (AIC=217.7).
 This is close to the median values of the respective coordinates.
@@
 
-However, it may be that after crossing a spatial limit, it takes some given amount of time before a reset of the illusion is perceived.
+However, it may be that after crossing a spatial limit, it takes some given amount of time before a reset of the illusion is perceived (or reported).
 This would mean that both limits explain separate parts of the data.
-We model this by finding optimal parameter values for both a spatial limit and a temporal limit simultaneously, when minimizing the distance between predicted and atcual reset points (in centimeters).
+We model this by finding optimal parameter values for both a spatial limit and a temporal limit simultaneously, when minimizing the distance (in _cm_ ) between predicted and actual reset points.
 The equations in the model are:
@@
 
-Figure 8 shows both the actual reset point coordinates (dots), given the illusion strength as the angle to a sample half-way to the reset point, as well as the model's predictions of reset point coordinates (lines).
+Figure 7 shows both the actual reset point coordinates (dots), given the illusion strength as the angle to a sample half-way to the reset point, as well as the model's predictions of reset point coordinates (lines).
 
-```{r fig8, echo=FALSE, message=FALSE, fig.width=6, fig.height=6, fig.cap="\\label{fig:fig8}**Figure 8. Reset Points and Combined Limits**"}
+```{r fig7, echo=FALSE, message=FALSE, fig.width=6, fig.height=6, fig.cap="\\label{fig:fig7}**Figure 7. Reset Points and Combined Limits**"}
 plotSequentialModel()
@@
 
-The limits this model generates are a 1.42 _cm_ spatial limit and a 0.28 _s_ temporal limit.
+The limits that optimally fit the model to the data are a 1.42 _cm_ spatial limit and a 0.28 _s_ temporal limit.
 While the MSE and AIC of the model are lower than that of the model with the single spatial limit, it does not explain the data better ( _p_=.53).
 This means that a simple spatial limit accounts for most of the data.
-It is possible that the way we've modeled the additional temporal delay after crossing the spatial limit is not accurate, as it may follow a poisson distribution, which we do not account for here.
+It is possible that the way we've modeled the additional temporal delay after crossing the spatial limit is not accurate, as it may not follow a normal distribution, which we do not account for here.
 However, regardless of whether or not a temporal limit adds anything, a spatial limit appears to explain the majority of spontaneous double-drift resets.
-
-```{r comment, echo=FALSE, message=FALSE}
-#It may be that during online tracking, proprioceptive information on the location of the hand is included in the estimate of the location of the stimulus.
-This may have decreased the likelihood of a reset, and may also changed the strength of the illusion.
-Since some participants did both experiments (X/9), we can at least test if illusion strength is different for online tracking and offline re-tracing.
-```
 
@@
 While here we find that a spatial limit is the better explanation for resets, other evidence may point at a temporal limit instead.
-In an as of yet unpublished study [Sirui Liu] obtained the point of subjective equivalence (PSE) of a double drift stimulus, and a static point that appeared some time after the double drift stimulus disappeared.
-The double-drift stimulus was on a 2 or 4 DVA length trajectory, traversed in 1, 2 or 3 seconds.
-The PSE's can be modelled very accurately with a model that assumes a fixed chance of a reset per unit of time (~0.7%/s), that is: it explains the data using only a temporal bound on the illusion.
+In an as of yet unpublished study [Sirui Liu] obtained the point of subjective equivalence (PSE) of a double-drift stimulus, and a static point that appeared some time after the double drift stimulus disappeared.
+The double-drift stimulus was on a 2 or 4 _dva_ length trajectory, traversed in 1, 2 or 3 _s_.
+The PSE's can be modelled very accurately with a model that assumes a fixed chance of a reset per unit of time (~0.7% / _s_), that is: it explains the data using only a temporal bound on the illusion.
 Interestingly, this fully contradicts the current findings.
 The model predicts that the PSE decreases over time as it becomes more likely that a reset has occurred.
-We will compare the two data sets here (see Fig 9).
+We will compare the two data sets here (see Fig 8).
 
-```{r fig9, echo=FALSE, message=FALSE, fig.width=6, fig.height=8, fig.cap="\\label{fig:fig9}**Figure 9. PSE's of double-drift position and continuous tracking**"}
+```{r fig8, echo=FALSE, message=FALSE, fig.width=6, fig.height=8, fig.cap="\\label{fig:fig8}**Figure 8. PSE's of double-drift position and continuous tracking**"}
 plotAverageTrajectories()
@@
 If the PSE's express the average perceived deviations from the actual trajectory of the double-drift stimulus, then they should show a similar pattern as the averaged trajectories in experiment 2.
-In order to be able to compare the two data sets, we will include all trajectories, including those where our simple algorithm did not detect a reset, as such trials should also be included in the PSE data.
-Further, we scale the trajectories on both axes, so that the Y end-point matches the duration of stimulus.
+In order to be able to compare the two data sets, we will include all trajectories, including those where our simple algorithm did not detect a reset, as such trials are also included in the PSE data.
+Further, we proportionally scale the trajectories on both axes, so that the Y end-point matches the duration of stimulus.
 Since the PSE data was collected in degrees visual angle, we convert the unit to degrees visual angles, using the estimate at the centre of the workspace.
-For the 4-second stimuli, we removed the last second.
-Figure 9 shows the individual participants' and the averaged trajectories (purple, red, orange) as well as the PSE's obtained in the other study (shown separately for 2 and 4 DVA stimulus trajectories) in gray.
+Finally, for the 4 _s_ stimuli, we removed the last second.
+Figure 9 (top row) shows the PSE's obtained in the other study (shown separately for 2 and 4 DVA stimulus trajectories) as red dots, with the time-based model's average trajectories in black and the distribution of percepts as a gray histogram.
+The bottom row of Figure 9 shows the averaged trajectories across participants for each of the internal motions (purple, red, orange), and external motions (left and right), with the distribution of percepts as a gray histogram for the same 3 time points as in the top row.
 
 For all internal motion speeds, we see a very similar average trajectory, that does not get closer to veridical over time.
-We verified this with a repeated measures ANOVA on perceived deviation from veridical stimulus position, using internal motion speed (2, 3, or 4 cycles/s) and time (1, 2, or 3 seconds after stimulus onset) as factors.
+We verified this with a repeated measures ANOVA on perceived deviation from veridical stimulus position, using internal motion speed (2, 3, or 4 _cps_) and time (1, 2, or 3 _s_ after stimulus onset) as factors.
 If the trajectory data from experiment 2 here follows the same pattern as the PSE data, there should be an effect of time, and perhaps an interaction with internal motion speed.
@@
 It could be that the re-tracing method provides very noisy estimates, hiding any effect.
-It could also be that the effects truly vary a lot between participants.
-Since we can see that some participants do seem to follow the pattern found with the PSE approach, and some show a pattern of increasing average deviation between perceived and true position, it may that the differences between the data sets can simply be explained by random sampling of relatively small numbers of participants.
+It could also be that the perception of the double-drift illusion, and it's resets, truly vary a lot between participants, so that both data sets and approaches are true.
+Since some participants do seem to follow the pattern found with the PSE approach, and some show a pattern of increasing average deviation between perceived and true position (not shown in the figure), it may that the differences between the data sets can simply be explained by random sampling from the population with relatively small numbers of participants.
 
@@
 We first tested if tracking the position of a double-drift stimulus changes the illusory percept of it's position, and found that it doesn't.
-We then set out to test if spontaneous resets in the double-drift illusion reflect discrepancies between two position signals, occurring at some given size of this discrepancy, or if they simply occur after some time elapses, analogous to other illusions or bi-stable stimuli.
-To measure the location of resets we first ask participants to track the perceived location of the double-drift stimulus online.
-While the heading of the tracking movements is sensitive to the strength of the illusion, there are no clear reset points in the tracking data.
-In a second experiment we ask participants to draw the perceived path of the stimulus after a single pass.
-In this re-tracing experiment the drawn traces reflect the strength of the illusion just as well, and in most trials there is a clear reset point, for either a jump reset or a hit-the-wall reset.
-The positions of these reset points do no vary much with elapsed time, but are better explained by the distance from the true position.
+However, tracking the percept in real time did not show any unambiguous resets.
+To measure resets, e ask participants to draw the perceived path of the stimulus after a single pass.
+In this re-tracing experiment the drawn traces reflect the strength of the illusion as well, and in most trials there is a clear reset point, for either a jump reset or a hit-the-wall reset.
+We then set out to test if these spontaneous resets in the double-drift illusion reflect discrepancies between two position signals, occurring at some given size of this discrepancy, or if they simply occur after some time elapses, analogous to other illusions or bi-stable stimuli.
+The positions of these reset points do not vary much with elapsed time, but are better explained by the distance from the true position.
 This has implications for the way position perception relies on motion signals.
@@
 At that point the perceived position may be discarded, or reset to the retinal position, and the build-up of the illusion may start again.
-For the remaining participants, either the discrepancy between retinal and perceived location never exceeds their threshold, or there may be another reason they do not employ any causal inference process.
+For the remaining participants, either the discrepancy between retinal and perceived location never exceeds their threshold, or their percept is indeed bounded to the maximum difference they allow between the two position signals.
 
@@
 Given a pre-motor theory of attention like mechanism, when spontaneously planned eye-movements trigger (jump) resets, such resets could occur at any point during the stimulus pass.
-However, it seems that for the participants that display "jump" resets in their re-tracing data, these happen at a fairly consistent location, which suggests that jump resets are instead stimulus triggered.
+However, it seems that for the participants that display "jump" resets in their re-tracing data, these happen at a fairly consistent location, which suggests that jump resets are instead usually triggered by the stimulus and not by attention.
 This should however be tested with eye-tracking data, so that this will remain a speculation here.

### 2020-12-12 Manuscript.Rmd
implemented some feedback from Patrick

@@
 ---
-title: "Measuring the double-drift illusion with hand tracking: capturing illusion resets"
+title: "Measuring the double-drift illusion with hand tracking"
 author: 
@@
   word_document: default
+  pdf_document: default
   html_document:
     df_print: paged
-  pdf_document: default
 header-includes:
@@
 
-If a gabor pattern drifts in one direction while its internal texture drifts in the orthogonal direction, its perceived direction deviates strongly from its true direction and is instead some combination of its real external motion and its internal motion [@Tse2006].
-The perceived location of such a double-drift stimulus periodically resets toward the veridical position.
-There may indeed be a simple perdiodicity underlying these resets.
-However, since both the veridical and perceived location are available to the brain, these resets may also occur when these two positions are too far from each other to come from the same source stimulus.
-Here we test which of the two explanations is more likely.
-In a first experiment, we had participants continuously track the position of the perceived position of the stimulus, using a stylus on a drawing tablet.
-While the orientation of the recorded trajectories reflect stimulus properties, there were no spontaneous resets.
-In a second experiment, we had participants re-trace the perceived trajectory of a short stimulus presentation, or indicate the initial direction of the stimulus they perceived.
-We found resets here, and they indicate a set spatial limit to the distance between the veridical and perceived location of the double-drift stimulus.
+If a gabor pattern drifts in one direction while its internal texture drifts in the orthogonal direction, its perceived position deviates further and further away from  its true path.
+We first evaluated the illusion using manual tracking.
+Participants followed the gabor with a stylus on a drawing tablet that coincided optically with the horizontal monitor surface.
+Their hand and the stylus were not visible during the tracking.
+The magnitude of the tracking illusion corresponded closely to previous perceptual and pointing measures indicating that manual tracking is a valid measure for the illusion.
+This allowed us to use it to capture the behavior of the illusion as it eventually degrades and breaks down in single trials.
+Specifically, the deviation of the gabor stops accumulating at some point and either stays at a fixed offset or resets toward the veridical position.
+In a second experiment, participants re-traced the perceived trajectory after the gabor was removed from the monitor.
+On alternate trials, they indicate the initial direction of the gabor’s path.
+Resets were detected and analyzed and they suggest that there is a spatial limit to  the perceived offset beyond which the deviation saturates or resets.
 
@@
 These cases of position extrapolation may be functional, helping the motor system overcome inevitable neural delays in targeting a moving object [@Nijhawan1994; @Duhamel1992].
-A stronger and very different motion-induced position shift arises when a moving gabor with internal motion orthogonal to its path is viewed in the periphery [@Tse2006; @Shapiro2010; @Gurnsey2012; @Kwon2015; @Lisi2015].
+A stronger and very different motion-induced position shift arises when a moving gabor has internal motion orthogonal to its path and is viewed in the periphery [@Tse2006; @Shapiro2010; @Gurnsey2012; @Kwon2015; @Lisi2015].
 This double-drift stimulus (Fig.
@@
 
-```{r fig1, echo=F, fig.height=8, fig.width=8, message=FALSE, error=FALSE, warning=FALSE, fig.cap="**Figure 1: Stimuli and Setup.** _Top left:_ a gabor with internal motion (phase shifts) and external motion, that is viewed peripherally, appears to follow a path deviating from the true path.
-Eye-movements immediately following disappearance of the gabor land at the true position, but pointing movements and delayed eye-movements go to the illusory position.
-_Bottom left:_ at some point, the perceived position will 'reset' towards the true position.
-For some people this takes the form of a 'jump' back to the true position, for others it means the illusory position doesn't deviate more from the true position, and there are some in between version.
-_Top right:_ if these resets are due to a spatial limit on the size of the illusory position shift, the points of the resets should have a constant X coordinate.
-_Middle right:_ if they occur after some period of time, they should have a constant Y coordinate.
-_Bottom right:_ we have people track or re-trace their percept and try to isolate reset points from the drawn trajectories to figure out where or when resets occur."}
+```{r fig1, echo=F, fig.height=3.2, fig.width=8, message=FALSE, error=FALSE, warning=FALSE, fig.cap="**Figure 1: Stimuli and Setup.** _Left_: a gabor with internal motion (phase shifts) and external motion, that is viewed peripherally, appears to follow a path deviating from the true path.
+_Right_: participants track their as the gabor moves back and forth along its path."}
 library(magick)
-Fig1 <- image_read_svg('doc/stimulus_reset.svg')
+Fig1 <- image_read_svg('doc/methods_fig_1.svg')
 par(mai=c(0,0,0,0))
@@
 
-The strength of the double-drift illusion probably arises because of the very poor location information for a gabor pattern in the periphery when it has the same mean luminance as the background.
+The strength of the double-drift illusion probably arises because of the very poor location information for a gabor pattern in the periphery when it has the same mean luminance as the background [@Gurnsey2012; @Kwon2015; @Cavanagh2019].
 With poor positional certainty, the motion information contributes to the location estimate, generating a shift in perceived position.
-If the gabor itself is not moving [@DeValois1991], the shift saturates after about 100 ms [@Chung2007], perhaps because the accumulating evidence of actual location of the stationary gabor patch is sufficient to anchor its position.
+If the gabor itself is not moving [@DeValois1991], the shift saturates quickly [40 _ms_: @Chung2007; 100 _ms_: @Arnold2007], perhaps because the accumulating evidence of actual location of the stationary gabor patch is sufficient to anchor its position.
 However, when the gabor itself also moves as in the double-drift stimulus, there is no stable location information to anchor the position estimate and the build-up continues away from its physical location up to several times the size of the gabor.
 The strength of the illusion, or how much the perceived position shifts from the real position over time, depends on the speed of both the internal and external motion [@Tse2006; @Shapiro2010; @Kwon2015].
+In our first experiment, participants track their percept with a stylus as the gabor moves back and forth on its trajectory (Fig.
+1) with 4 internal speeds.
+We analyze the angle of the tracking to determine if manual response is susceptible to the illusion.
 
-At some point, the illusion resets, the perceived location jumps back to its physical location and then accumulation begins again (Fig.
-1, bottom left) or it saturates.
+At some point, the illusion stops accumulating.
+The position may just remain at some fixed offset following a path parallel to the true path or the position may reset back toward its physical location and then accumulation begins again (Fig.
+4, left) or it saturates.
 This long duration before resetting or saturation is well beyond the integration period of other visual effects and into the range of short-term memory.
 Given that both true and illusory positions are available, it could be that instead of some integration period, it is the interplay between the two position signals that leads to spontaneous resets.
-Alternatively, there is a simple periodicity or stochastics process underlies the changes in perception.
+Alternatively, there is a simple periodicity or a stochastic process that underlies the changes in perception.
 Recent work has shown that independent of the stimulus, movement-related attention may lead to resets [@Nakayama2020].
@@
 
-# Methods
+# Online tracking experiment
 
-## Participants
+In this experiment, participants were asked to use a stylus on a drawing tablet to track a moving gabor with some amount of internal motion that elicited the double-drift illusion.
+We test if tracking the illusion continuously is a good measure of illusion strength, or if it prevents or decreases the illusion.
 
-For the first task in experiment 1, 4 participants were recruited from the lab (3 female; ages 21 – 31, mean: 25.75).
-For the second task in experiment 1, 6 participants were recruited from the lab (4 female; ages 19 – 26, mean: 21.6).
-For experiment 2, 9 participants were recruited from the lab (6 female; ages 19 - 27, mean: 22.8).
+## Methods
+
+### Participants
+
+For this experiment, 4 participants were recruited from the lab (3 female; ages 21 – 31, mean: 25.75).
 All participants reported to be right handed and had either normal or corrected to normal vision.
@@
 
-## Setup
+### Setup
 
@@
 The stimuli were observed via an upward facing mirror placed exactly between the tablet and monitor, so that the stimuli appear to be in the same horizontal plane as the hand, and the tip of the stylus.
+Experiments were run in Python 2.7 with PsychoPy [@Peirce2019] and analyses were done in R 3.6.1 [@Rproject].
 
 For reporting degrees of visual angle that stimuli spanned ( _dva_ ), we assumed the participants to be close to the setup, with their eyes at the height of the monitor.
-Since there was no requirement to use the forehead rest, the latter is probably only true for the tallest participants, so that reported _dva's_ are slightly underestimating the size of stimuli.
+While there was no requirement to use the forehead rest, this is a reasonable assumption, since stimuli are only visible when relatively close to the setup.
+Nevertless, reported _dva's_ are estimates of the size of stimuli.
 
-## Stimuli
+### Stimuli
 
-The stimuli were phase-changing gabor patches with a gaussian mask with an sd of 0.43 _cm_ ( or ~.6 _dva_ when in the centre of the screen and viewed from a typical distance ).
-The whole stimulus extended 3 times the sd of the Guassian in all four cardinal directions.
-The stimulus used 1.5 phases of a sine wave, so that each sine phase equals ~0.58 _cm_ or ~1.5 _dva_.
-By applying both real motion and phases changes to the gabor, it becomes a 'double-drift stimulus' where the perceived path deviates from the real path.
+The stimuli were phase-changing gabor patches with a gaussian mask with a sigma (sd) of 0.43 _cm_ or ~0.59 _dva_, spatial frequency of 0.58 _c/cm_ or ~0.42 _c/dva_, vertical orientation, and contrast of 100%.
+By simultaneously moving the gabor envelope and drifting its internal sine wave, it becomes a ‘double-drift stimulus’ where the perceived path deviates from the real path.
 The gabor changed phase at different rates, called 'internal motion' here, specified in cycles per second ( _cps_ ).
-The real displacements of the stimulus are along the Y axis of the monitor, keeping the X coordinate constant at the middle of the screen, so that a stimulus without internal motion would appear to move forward and backward on the horizontal display.
-This real motion was restricted to the central 13.5 _cm_ ( ~14.0 _dva_ ) and was set so that the stimulus would move from either extreme to the other ( a motion we call a 'pass' ) in 2, 3 or 4 _s_, corresponding to speeds of 6.75, 4.5 and 3.75 _cm/s_ ( or roughly 7.0, 4.7 or 3.5 _dva/s_ ).
+The real displacements of the stimulus are along the Y axis of the monitor, keeping the X coordinate constant at the middle of the screen, so that a stimulus without internal motion would appear to move toward and away from the participant on the horizontal display.
+
+The motion of the envelope was restricted to the central 13.5 _cm_ ( ~14.0 _dva_ ) and was set so that the stimulus would move from one end to the other of its path ( a motion we call a 'pass' ) in 2, 3 or 4 _s_, corresponding to speeds of ~6.75, ~4.5 and ~3.75 _cm/s_ ( or roughly 7.0, 4.7 or 3.5 _dva/s_ ).
 For simplicity, and since the distance moved is always the same, these speeds will be refered to by the duration of a single pass (2, 3 or 4 _s_).
 
-## Tasks
+### Procedure
 
-In the two tasks in experiment 1, participants were to do online tracking (using their unseen hand / stylus) while viewing a double-drift stimulus.
-In experiment 2, participants first watched a double-drift stimulus and then either reproduced the perceived path of the stimulus by retracing it on the tablet, or indicated the initial movement direction by changing the orientation of a line originating in the same position as the doube-drift stimulus.
-More details are given below.
+In experiment 1, participants were to do online tracking (using their unseen hand / stylus) while viewing a double-drift stimulus.
+In this task, participants were asked to continuously track the position of a gabor that moved forward and backward along the Y-axis (of the horizontal screen) for the duration of 12 s, using a stylus on a drawing tablet.
+Before trial onset, the participant moved the stylus to the middle of the screen ( where the gabor would appear ), and was subsequently asked to fixate a point on the left or right of the screen, and given 1.5 s to do so.
+Since the screen was seen via a mirror, the stimulus appeared to move forward and backward on a horizontal plane at the same depth as the drawing tablet.
+The internal motion of the gabor was inverted at the far and close ends of the physical path, where external motion was also inverted.
+Internal motion could be 3, 1, 0, -1 or -3 cps.
+Since each moving gabor was shown for 12 s, and it moved from one end to another in 2 s, it completed 5 full and 2 half passes of the path.
 
-## Analyses
+### Analyses
 
-For experiment 1, we first segmented the tracking trajectories according to stimulus passes (in between direction changes of the gabor), and removed the first and last half second ( leaving 1 _s_ trajectories for the first experiment, and 3 _s_ trajectories for the second experiment).
+We first segmented the tracking trajectories according to stimulus passes (in between direction changes of the gabor), and removed the noisy tracking during the first and last half second, leaving 1 _s_ trajectories.
 We then calculated the instantaneous heading (direction, disregarding velocity or distance) of a smooth spline interpolated version of each trajectory segment (using 101 spline interpolated position samples, to get 100 heading samples).
@@
 
-For comparison, we use the same heading measure in experiment 2.
-However, the individual trajectories reveal more points that are potential spontaneous resets that we try to extract.
-Using a smooth spline interpolated version of the trajectory we detect left-right direction changes.
-Since the initial movement is often jittery, we only take direction changes further than ~0.39 _cm_ away from the starting point.
-The first (if any) of the remaining direction changes, are then taken as reset points.
-By averaging the x and y coordinates of reset points in all 9 participants and 6 stimulus conditions, we obtain 54 average reset points for further analyses and modelling.
-The average line orientation is also calculated for each of those 54 combinations of partcipants and stimulus conditions.
-
-Experiments were run in Python 2.7 with PsychoPy [@Peirce2019] and analyses were done in R 3.6.1 [@Rproject].
-
-# Experiment 1: Continuous tracking
-
-In this experiment, participants were asked to use a stylus on a drawing tablet to track a moving gabor with some amount of internal motion that elicited the double-drift illusion.
-We test if tracking the illusion continuously is a good measure of illusion strength, or if it prevents or decreases the illusion.
-
-## Task
-
-In this task, participants were asked to continuously track the position of a gabor that moved forward and backward along the Y-axis (of the horizontal screen) for the duration of 12 _s_, using a stylus on a drawing tablet.
-Before trial onset, the participant moved the stylus to the middle of the screen ( where the gabor would appear ), and was subsequently asked to fixate a point on the left or right of the screen, and given 1.5 _s_ to do so.
-Since the screen was seen via a mirror, the stimulus appeared to move forward and backward on a horizontal plane at the same depth as the drawing tablet.
-The internal motion of the gabor was inverted at the far and close ends of the physical path, where external motion was also inverted.
-Internal motion could be 3, 1, 0, -1 or -3 _cps_.
-Since each moving gabor was shown for 12 _s_, and it moved from one end to another in 2 _s_, it completed 5 full and 2 half passes of the path.
-
 ## Results
 
-In Figure 2 (top), we plot the average instantaneous heading during the middle second of each 2 _s_ pass (as tracking was noisy at the inversion points) for each of the 5 internal motion speeds (normalized for the external motion).
+In Figure 2 (top), we plot the average instantaneous heading during the middle second of each 2 _s_ pass for each of the 5 internal motion speeds (normalized for the external motion).
 The average angle of the tracked path is given in degrees deviation from straight forward.
@@
 With FDR-corrected, paired t-tests, we find a difference in heading between 0 _cps_ internal motion and both 1 _cps_ ( _p_=.022) and -1 _cps_ ( _p_=.025) as well as between -3 and -1 _cps_ and 3 and 1 _cps_ (both _p_ <.001).
-This makes it clear that continuous tracking does not prevent the illusion from ocurring.
-We will use the averaged angle for each of the internal motion conditions as a measure of the strength of the illusion.
+This makes it clear that continuous tracking does not prevent the illusion from occurring.
+We will use heading angle as a measure of the strength of the illusion.
+
+We tried detecting resets, but none could be unambiguously identified (see Fig 2).
+It is possible that resets were perceived but were too quick to respond to in the continuous tracking task.
+Alternatively, tracking may have stabilized the illusion, preventing resets from occurring.
 
 ```{r fig2, echo=FALSE, message=FALSE, fig.width=4.5, fig.height=6, fig.cap="\\label{fig:fig2}**Figure 2.
-Heading and Trajectories in Experiment 1a.** **Top:** Average heading across the middle 1 _s_ of each pass.
+Heading and Trajectories in Experiment 1.** **Top:** Average heading across the middle 1 _s_ of each pass.
 Distance from origin is time, angle is angle.
-In blue lines the average heading across participants, in light blue a polar heatmap giving an impression of the distribution of heading.
+In darker blue lines the average heading across participants, in light blue a polar heatmap giving an impression of the distribution of heading.
 **Bottom:** All individual trajectores.
-_Red_: Highlighted trajectories from conditions with non-zero internal speed, where at least some heading samples fell in the 95% confidence interval for that participants' heading distribution in zero internal speed condition."}
+_Red_: Highlighted trajectories from conditions with non-zero internal speed, where at least some heading samples fell in the 95% confidence interval for that participants' heading distribution in zero internal speed condition.
+This would occur during a reset, but only very few trajectories have this feature, and none contain clear resets."}
 plotBoundedTracking()
@@
 The data clearly show the illusion, but does tracking decrease the strength of the illusion compared to simply observing it?
-We compare illusion strength to that reported previously [@Cavanagh2019] in Figure 3.
-As can be seen the averages from our data here (in red) coincide almost exactly with the illusion strength predicted by the model.
+We compare mean illusion strength to that reported previously [@Cavanagh2019] in Figure 3.
+As can be seen the averages from our data here (in red) coincide almost exactly with the illusion strength predicted by the model (in gray).
 That is; tracking the illusion while observing it does not change the strength of the illusion.
 
-```{r fig3, echo=FALSE, message=FALSE, fig.width=8, fig.height=5, fig.cap="\\label{fig:fig3}**Figure 3.
+```{r fig3, echo=FALSE, message=FALSE, fig.width=5, fig.height=5, fig.cap="\\label{fig:fig3}**Figure 3.
 Illusion strength.** The strength of the illusion in degrees angle deviation from the physical path is plotted over external motion speed ( _dva/s_ ).
-In red are the average illusions strengths for ±3 _cps_, ±1 _cps and 0 cps.
-The black circles and line are the data and model from Cavanagh & Tse (2019).
-The grey lines are the illusion strengths their model predicts for the internal motion speeds used here."}
+In red are the average illusions strengths for 0 _cps_, ±1 _cps_ and ±3 _cps_, and in light blue the data for individual participants.
+The gray lines are the predicted illusion strengths based on the model from Cavanagh & Tse (2019) for the internal motion speeds used here.
+*the second panel only for testing*"}
 plotStrengthComparison()
@@
 
-## Resets
+# Delayed retracing experiment
 
-In order to detect potential resets, we consider that in the 0 internal motion condition there is no illusion, so that we measure the tracking of the veridical path of the stimulus.
-Since both strong, "jump" resets and weak "hit-the-wall" resets (see Fig 1, bottom left) would change percieved direction back to or crossing the veridical motion path, their smooth-spline interpolated paths would contain samples similar in heading to the 0 internal motion condition.
-We use a 95% confidence interval of heading in the 0 internal motion condition to select trajectories from conditions with non-zero heading that may show a reset.
-These trajectories are drawn in red in Figure 2 (for clarity only 1 per condition/participant is highlighted), but as can be seen, the deviations from the condition's dominant heading occur just before or after the external motion reversals, so they may simply reflect difficulty of tracking the stimulus at the reversal points, rather than resets.
-
-No unambiguous resets were captured in either experiment.
-It is possible that resets were perceived but were too quick to respond to in the continuous tracking task.
-Alternatively, tracking may have stabilized the illusion, preventing resets from occurring.
-Either way, we performed a second experiment that would counter both explanations for the absence of (spontaneous) resets.
-
-# Experiment 2: Re-tracing
-
-In a second experiment, we ask participants to re-trace the perceived path of a double-drift stimulus, immediately after viewing the stimulus, as used by Nakayama & Holcombe [-@Nakayama2020].
+In a second experiment, aimed at collecting resets, we ask participants to re-trace the perceived path of a double-drift stimulus, immediately after viewing the stimulus, a method used by Nakayama & Holcombe [-@Nakayama2020].
 This allows participants more time to carefully reproduce the path.
-We use the recorded trajectories to assess points where resets occured.
+We use the recorded trajectories to assess points where resets occurred.
 The reset points can then be used to determine if resets are time-limited or space-limited.
 
-## Tasks
+```{r fig4b, echo=F, fig.height=3.2, fig.width=5.62, message=FALSE, error=FALSE, warning=FALSE, fig.cap="**Figure 4: Resets.** _Left_: At some point, the perceived position will stop drifting further and further away from its actual path.
+For some people this takes the form of a ‘jump’ back toward the true position, for others it means the illusory position remains at a fixed offset moving parallel to the real path, and there are some in-between versions.
+_Top right_: if these resets are due to a spatial limit on the size of the illusory position shift, the points of the resets should have a constant X coordinate.
+_Bottom right_: if they occur after some period of time, they should have a constant Y coordinate."}
+library(magick)
+Fig4b <- image_read_svg('doc/methods_fig_2.svg')
+par(mai=c(0,0,0,0))
+plot(Fig4b)
+```
 
-Here we used a re-tracing procedure where participants drew the perceived path of a double-drift stimulus making a single pass across the workspace.
+## Methods
+
+### Participants
+
+For this experiment, 9 participants were recruited from the lab (6 female; ages 19 - 27, mean: 22.8).
+All participants reported to be right handed and had either normal or corrected to normal vision.
+Procedures were approved by York’s Human Participants Review Committee, and all participants provided prior, written, informed consent.
+
+### Setup and Stimuli
+
+The setup was the same as used in the online tracking experiment, except that we now also used a small keypad, for responses (see Fig 1, right).
+Stimuli were identical to those used in the online tracking experiment, but now always started at the near end of the workspace and moved away from the participants.
+The double drift stimuli would use either 3 or 4 seconds to move 13.5 _cm_ and had an internal drift of 2, 3 or 4 _cps_.
+
+## Procedure
+
+In experiment 2, participants first watched a double-drift stimulus and then either reproduced the perceived path of the stimulus by retracing it on the tablet, or indicated the initial movement direction by changing the orientation of a line originating in the same position as the doube-drift stimulus.
+More details are given below.
 This provided more time for the participants to reproduce any resets, i.e.
@@
 The re-tracing task was done in half of 8 blocks.
-In the other half, participants oriented a line to indicate the initial perceived direction of the stimulus motion.
+In the other half, participants viewed the same stimuli, and then oriented a line to indicate the initial perceived direction of the stimulus motion.
 Block types were alternated, and the order was counterbalanced across participants.
@@
 Re-tracing the perceived paths should allow capturing spontaneous resets of the illusion, and the depence on the strength of the illusion could be checked with the data from the line-orienting task.
+
+### Analyses
+
+We use the same heading measure here as in experiment 1.
+However, the individual trajectories reveal more points that are potential spontaneous resets that we try to extract.
+Using a smooth spline interpolated version of the trajectory we detect left-right direction changes.
+Since the initial movement is often jittery, we only take direction changes further than ~0.39 cm away from the starting point.
+The first (if any) of the remaining direction changes, are then taken as reset points.
+By averaging the x and y coordinates of reset points in all 9 participants and 6 stimulus conditions, we obtain 54 average reset points for further analyses and modelling.
+The average line orientation is also calculated for each of those 54 combinations of partcipants and stimulus conditions.
 
@@
 ```{r fig6, echo=FALSE, message=FALSE, fig.width=8, fig.height=6, fig.cap="\\label{fig:fig6}**Figure 6.
-Reset Points and Single Limits** **Left:** True path of the doubel-drift stimulus as gray arrow.
+Reset Points and Single Limits** **Left:** True path of the double-drift stimulus as gray arrow.
 The average reset points for each participant and stimulus condition displayed spatially, with the internal motion normalized to go rightward.
@@
 Where $L_y$ and $L_x$ are a temporal (y) and spatial (x) limit that are free parameters, and s corresponds to the calculated slope.
-If we fit these simple models by minimizing the mean square error (as the absolute distance in centimeters between predicted and actual reset points), the optimal parameters indicate a temporal limit of ~1.4 _s_ (AIC=258.4) and a spatial limit of ~1.7 _cm_ (AIC=217.7).
+If we fit these simple models by minimizing the mean square error (as the absolute distance in centimeters between predicted and actual reset points), the optimal parameters indicate a temporal limit of ~1.4 _s_ (AIC=-167.8) and a spatial limit of ~1.7 _cm_ (AIC=-208.6).
 This is close to the median values of the respective coordinates.
@@
 The limits that optimally fit the model to the data are a 1.42 _cm_ spatial limit and a 0.28 _s_ temporal limit.
-While the MSE and AIC of the model are lower than that of the model with the single spatial limit, it does not explain the data better ( _p_=.53).
+While the AIC of the model (-209.8) is marginally lower than that of the model with the single spatial limit, but it does not explain the data better ( _p_=.53).
 This means that a simple spatial limit accounts for most of the data.
@@
 While here we find that a spatial limit is the better explanation for resets, other evidence may point at a temporal limit instead.
-In an as of yet unpublished study [Sirui Liu] obtained the point of subjective equivalence (PSE) of a double-drift stimulus, and a static point that appeared some time after the double drift stimulus disappeared.
-The double-drift stimulus was on a 2 or 4 _dva_ length trajectory, traversed in 1, 2 or 3 _s_.
-The PSE's can be modelled very accurately with a model that assumes a fixed chance of a reset per unit of time (~0.7% / _s_), that is: it explains the data using only a temporal bound on the illusion.
-Interestingly, this fully contradicts the current findings.
+An as of yet unpublished study [@Liu2019] obtained data on the PSE of a double-drift stimulus at the end of trajectories of various lengths traveled at different speeds.
+These PSEs were predicted very accurately with a model that assumes a fixed chance of a reset per unit of time (~0.7% / _s_).
+Interestingly, this temporal constraint fully contradicts the current findings.
 The model predicts that the PSE decreases over time as it becomes more likely that a reset has occurred.
-We will compare the two data sets here (see Fig 8).
-
-```{r fig8, echo=FALSE, message=FALSE, fig.width=6, fig.height=8, fig.cap="\\label{fig:fig8}**Figure 8. PSE's of double-drift position and continuous tracking**"}
-plotAverageTrajectories()
-```
-
-If the PSE's express the average perceived deviations from the actual trajectory of the double-drift stimulus, then they should show a similar pattern as the averaged trajectories in experiment 2.
-In order to be able to compare the two data sets, we will include all trajectories, including those where our simple algorithm did not detect a reset, as such trials are also included in the PSE data.
-Further, we proportionally scale the trajectories on both axes, so that the Y end-point matches the duration of stimulus.
-Since the PSE data was collected in degrees visual angle, we convert the unit to degrees visual angles, using the estimate at the centre of the workspace.
-Finally, for the 4 _s_ stimuli, we removed the last second.
-Figure 9 (top row) shows the PSE's obtained in the other study (shown separately for 2 and 4 DVA stimulus trajectories) as red dots, with the time-based model's average trajectories in black and the distribution of percepts as a gray histogram.
-The bottom row of Figure 9 shows the averaged trajectories across participants for each of the internal motions (purple, red, orange), and external motions (left and right), with the distribution of percepts as a gray histogram for the same 3 time points as in the top row.
-
-For all internal motion speeds, we see a very similar average trajectory, that does not get closer to veridical over time.
-We verified this with a repeated measures ANOVA on perceived deviation from veridical stimulus position, using internal motion speed (2, 3, or 4 _cps_) and time (1, 2, or 3 _s_ after stimulus onset) as factors.
-If the trajectory data from experiment 2 here follows the same pattern as the PSE data, there should be an effect of time, and perhaps an interaction with internal motion speed.
-We find no effect of time ( _F_(2,16)=0.011, _p_=.957, $\eta^2$<.001), no effect of internal motion speed ( _F_(2,16)=0.127, _p_=.814, $\eta^2$=.0015) and no interaction ( _F_(4,32)=0.361, _p_=.635, $\eta^2$=.002).
+If the PSE's express the average perceived deviations from the actual trajectory of the double-drift stimulus, then we should see a similar pattern in the averaged trajectories in our re-tracing experiment.
+However, these averaged trajectories reach a plateau within ~1 _s_ after which resets and illusory position shift seem to balance each other.
 This means that we can't replicate the pattern found with the PSE approach.
 There may be several reasons for this.
-First, we notice that the trajectory data is highly dependent on participants.
-It could be that the re-tracing method provides very noisy estimates, hiding any effect.
-It could also be that the perception of the double-drift illusion, and it's resets, truly vary a lot between participants, so that both data sets and approaches are true.
-Since some participants do seem to follow the pattern found with the PSE approach, and some show a pattern of increasing average deviation between perceived and true position (not shown in the figure), it may that the differences between the data sets can simply be explained by random sampling from the population with relatively small numbers of participants.
+First, our data may simply be too noisy, or the effects may be highly dependent on participants so that both studies may suffer from sampling bias.
+We can't reolve this contradiction here and leave that for future work.
 
@@
 
-We first tested if tracking the position of a double-drift stimulus changes the illusory percept of it's position, and found that it doesn't.
-However, tracking the percept in real time did not show any unambiguous resets.
-To measure resets, e ask participants to draw the perceived path of the stimulus after a single pass.
+We first tested if manual tracking of a double-drift stimulus is susceptible to the illusion as reported by perceptual judgments, pointing [@Lisi2017], and memory saccades [@Massendari2018].
+The results of this online tracking experiment show a close agreement between the angle of the manual tracking and the illusion strength expected based on the model from earlier perceptual measures [@Cavanagh2019].
+However, since no unambiguous resets were captured in the traces of this experiment, we followed up with a second experiment that did record evidence of resets.
+
+To measure resets, we ask participants to draw the perceived path of the stimulus after a single pass.
 In this re-tracing experiment the drawn traces reflect the strength of the illusion as well, and in most trials there is a clear reset point, for either a jump reset or a hit-the-wall reset.
@@
 There is a generative model, based on Bayesian cue integration, that can explain many phenomena with motion induced position shifts.
-This model, formulated by Kwon, Tadin & Knill [-@Kwon2015], is compatbile with a good portion of our data.
+This model, formulated by Kwon, Tadin & Knill [-@Kwon2015], is compatible with a good portion of our data.
 The model posits there are three signals contributing to an object's perceived position: it's global motion, it's local motion (or texture motion) and it's position.
@@
 
-## Conclusion
-
-We find that online, manual tracking of the double-drift stimulus is sensitive to illusion strength, but doesn't display clear resets.
+We find that online, manual tracking of the double-drift stimulus is sensitive to illusion strength, but doesn’t display clear resets.
 We do find spontaneous resets in delayed, single-pass re-tracing.

### 2020-12-12 R/compare.R

@@ -21,65 +21,175 @@ functionCavanaghTse2019 <- function(internal=c(), external=c(), k=0.81) {
 
 plotStrengthComparison <- function() {
   
-
+  
+  colors <- getColors()
+  
+  #par(mfrow=c(1,2))
+  
+  
+  stdsegments <- read.csv('data/bounded_tracking/segment_directions.csv', stringsAsFactors = F)  
+  
+  stdsegments$direction[which(stdsegments$direction < 0)] <- NA
+  stdsegments <- stdsegments[which(!is.na(stdsegments$direction)),]
+  
+  internalSpeeds <- sort(unique(stdsegments$internalSpeed))
+  participants <- sort(unique(stdsegments$participant))
+  
+  internalspeed <- c()
+  participant <- c()
+  avgheading <- c()
+  
+  for (internalSpeed.idx in c(1:length(internalSpeeds))) {
+    
+    internalSpeed <- internalSpeeds[internalSpeed.idx]
+    
+    for (pp in participants) {
+      
+      # this is the part of the data we're dealing with now:
+      idx <- which(stdsegments$internalSpeed == internalSpeed & stdsegments$participant == pp)
+      
+      # # we put it into a normalized 2D histogram:
+      # pfreq <- hist2d(x=stdsegments$direction[idx], y=stdsegments$sample_no[idx], nbins=NA, edges=list(seq(0,180,length.out=61), seq(1,101,4)+0.5))
+      # 
+      # if (!is.matrix(freq2D)) {
+      #   freq2D <- pfreq$freq2D
+      # } else {
+      #   freq2D <- freq2D + pfreq$freq2D
+      # }
+      
+      # also get the average, to plot on top of the polar heat map:
+      PPavgDir <- aggregate(direction ~ sample_no, data = stdsegments[idx,], FUN=mean, na.rm=T)
+      
+      internalspeed <- c(internalspeed, internalSpeed)
+      participant <- c(participant, pp)
+      avgheading <- c(avgheading, 90-mean(PPavgDir$direction))
+      
+    }
+    
+  }
+  
[…]

### 2021-02-25 Manuscript.Rmd
edits to manuscript

@@
 If the gabor itself is not moving [@DeValois1991], the shift saturates quickly [40 _ms_: @Chung2007; 100 _ms_: @Arnold2007], perhaps because the accumulating evidence of actual location of the stationary gabor patch is sufficient to anchor its position.
-However, when the gabor itself also moves as in the double-drift stimulus, there is no stable location information to anchor the position estimate and the build-up continues away from its physical location up to several times the size of the gabor.
-The strength of the illusion, or how much the perceived position shifts from the real position over time, depends on the speed of both the internal and external motion [@Tse2006; @Shapiro2010; @Kwon2015].
+However, when the gabor itself also moves as in the double-drift stimulus, there is no stable location information to anchor the position estimate and the build-up continues well beyond 100 ms.
+The perceived position conintues to move away from its physical position up to several times the size of the gabor.
+The strength of the illusion, or the deviation of perceived path of the Gabor from its true path, depends on the speed of both the internal and external motion [@Tse2006; @Shapiro2010; @Kwon2015]; @Cavanagh2019.
 In our first experiment, participants track their percept with a stylus as the gabor moves back and forth on its trajectory (Fig.
-1) with 4 internal speeds.
+1) with 5 speeds of the internal motion.
 We analyze the angle of the tracking to determine if manual response is susceptible to the illusion.
 
+In our second experiment, we look at the spatial and temporal limits of the illusion.
 At some point, the illusion stops accumulating.
-The position may just remain at some fixed offset following a path parallel to the true path or the position may reset back toward its physical location and then accumulation begins again (Fig.
-4, left) or it saturates.
-This long duration before resetting or saturation is well beyond the integration period of other visual effects and into the range of short-term memory.
-Given that both true and illusory positions are available, it could be that instead of some integration period, it is the interplay between the two position signals that leads to spontaneous resets.
-Alternatively, there is a simple periodicity or a stochastic process that underlies the changes in perception.
-Recent work has shown that independent of the stimulus, movement-related attention may lead to resets [@Nakayama2020].
-This means that participants' responses to the stimulus may change the illusion.
-Here we will first test if real-time as well as delayed, manual tracking of the gabor's position, prevents or decreases the illusory shift in position.
+The accumulation can be reset by introducing a temporal break [@Lisi2015] or by distraction attention [@Nakayama2020], but it may also reset spontaneously once the accumulation has gone on too long or too far.
+Informally, some observers have reported that the Gabor’s path may saturate so that it  remains parallel to, but offset from the true path.
+Others report that the position resets back toward its physical location whereupon accumulation begins again.
+The ability to record trajectories the perceived position will then allow us to study spontaneous reset outcomes.
 
-The ability to record trajectories the perceived position will then allow us to study spontaneous resets – the adjustment of perceived location when the illusion has deviated too much or too long from the physical location.
-In particular, if a conflict between two location signals [@Lisi2015] leads to resets this would predict a set distance between the real and perceived position where resets occur.
-Under this hypothesis, the timing of spontaneous resets would depend on the speed of internal motion (given constant external motion).
-On the other hand, if spontaeous resets are the result of a process that integrates over time, resets would occur after a certain amount of time, but not be dependent on internal motion.
-Investigating the time and location of resets in the recorded trajectories, allows us to distinguish between a spatial limit to the illusion or a temporal limit.
-This in turn will inform us on the processes underlying the perception of positions of moving objects.
-
-# Online tracking experiment
+# Experiment 1: Online tracking
 
@@
 While there was no requirement to use the forehead rest, this is a reasonable assumption, since stimuli are only visible when relatively close to the setup.
-Nevertless, reported _dva's_ are estimates of the size of stimuli.
+Nevertheless, reported _dva's_ are estimates of the size of stimuli.
 
@@
 
-The motion of the envelope was restricted to the central 13.5 _cm_ ( ~14.0 _dva_ ) and was set so that the stimulus would move from one end to the other of its path ( a motion we call a 'pass' ) in 2, 3 or 4 _s_, corresponding to speeds of ~6.75, ~4.5 and ~3.75 _cm/s_ ( or roughly 7.0, 4.7 or 3.5 _dva/s_ ).
-For simplicity, and since the distance moved is always the same, these speeds will be refered to by the duration of a single pass (2, 3 or 4 _s_).
+The motion of the envelope was restricted to the central 13.5 _cm_ ( ~14.0 _dva_ ) and was set so that the stimulus would move from one end to the other of its path ( a motion we call a 'pass' ) in 2 _s_, corresponding to a speed of ~6.75 _cm/s_ ( or roughly 7.0 _dva/s_ ).
 
@@
 
-In Figure 2 (top), we plot the average instantaneous heading during the middle second of each 2 _s_ pass for each of the 5 internal motion speeds (normalized for the external motion).
+In Figure 2 (top), we plot the average instantaneous heading during the middle second of each 2 _s_ pass for each of the 5 internal motion speeds.
 The average angle of the tracked path is given in degrees deviation from straight forward.
@@
 
-We tried detecting resets, but none could be unambiguously identified (see Fig 2).
-It is possible that resets were perceived but were too quick to respond to in the continuous tracking task.
-Alternatively, tracking may have stabilized the illusion, preventing resets from occurring.
-
 ```{r fig2, echo=FALSE, message=FALSE, fig.width=4.5, fig.height=6, fig.cap="\\label{fig:fig2}**Figure 2.
@@
 **Bottom:** All individual trajectores.
-_Red_: Highlighted trajectories from conditions with non-zero internal speed, where at least some heading samples fell in the 95% confidence interval for that participants' heading distribution in zero internal speed condition.
-This would occur during a reset, but only very few trajectories have this feature, and none contain clear resets."}
+Note that there are only a few obvious discontinuities (highlighted in red).
+In general, the traced paths showed rough linearity."}
 plotBoundedTracking()
@@
 In red are the average illusions strengths for 0 _cps_, ±1 _cps_ and ±3 _cps_, and in light blue the data for individual participants.
-The gray lines are the predicted illusion strengths based on the model from Cavanagh & Tse (2019) for the internal motion speeds used here.
-*the second panel only for testing*"}
+The gray lines are the predicted illusion strengths based on the model from Cavanagh & Tse (2019) for the internal motion speeds used here."}
 plotStrengthComparison()
@@
 
-# Delayed retracing experiment
+# Experiment 2: Delayed retracing
 
-In a second experiment, aimed at collecting resets, we ask participants to re-trace the perceived path of a double-drift stimulus, immediately after viewing the stimulus, a method used by Nakayama & Holcombe [-@Nakayama2020].
-This allows participants more time to carefully reproduce the path.
-We use the recorded trajectories to assess points where resets occurred.
-The reset points can then be used to determine if resets are time-limited or space-limited.
+The durations of each pass in the online tracking experiment was 2 seconds but the manual traces mostly showed a linear trajectory without saturation or reset of the types shown in Figure 4 on the left below.
+The continuation of the accumulation may be a result of the active tracking – either the extra attention required to track increased the extent of the accumulation or possibly the temporal demands of the tracking caused the participants to average their traces over any breaks.
+To address this, we switched to an offline recording of the perceived path.
 
-```{r fig4b, echo=F, fig.height=3.2, fig.width=5.62, message=FALSE, error=FALSE, warning=FALSE, fig.cap="**Figure 4: Resets.** _Left_: At some point, the perceived position will stop drifting further and further away from its actual path.
-For some people this takes the form of a ‘jump’ back toward the true position, for others it means the illusory position remains at a fixed offset moving parallel to the real path, and there are some in-between versions.
-_Top right_: if these resets are due to a spatial limit on the size of the illusory position shift, the points of the resets should have a constant X coordinate.
-_Bottom right_: if they occur after some period of time, they should have a constant Y coordinate."}
+```{r fig4b, echo=F, fig.height=3.2, fig.width=5.62, message=FALSE, error=FALSE, warning=FALSE, fig.cap="**Figure 4: Resets.** _Left_: At some point, the perceived position will stop moving further away from its actual path.
+For some people this takes the form of a ‘jump’ back toward the true position, for others it means the illusory position remains at a fixed offset moving parallel to the real path, or some combnation of the two.
+_Top right_: if these resets are due to a spatial limit on the size of the illusory position shift away from the true vertical path, the locations of the resets should have a constant X coordinate.
+_Bottom right_: if they occur instead after some period of time, they should have a constant Y coordinate."}
 library(magick)
@@
 ```
+
+We were interested, in particular, in whether the limitation of the accumulation would be set by space or time.
+Since both the true and the illusory position appear to be available in the visual system (the true position drives saccades whereas the illusory position drives perception, Lisi & Cavanagh; -@Lisi2015), it is possible that resets occur once the distance between the real and perceived positions exceeds some limit.
+We can speculate that this spatial limit would be related to the positional uncertainty of the Gabor.
+When the position information is reliable, the resets would occur with very little illusory deviation, keeping the perceived path close to the true path.
+When the positional uncertainty is high, in the periphery, with the Gabor’s mean luminance matched to the background [@Gurnsey2012; @Cavanagh2019], the spatial offset could be quite large before exceeding the range of positional uncertainty around the gabor’s true position.
+Under this hypothesis, the timing of spontaneous resets will depend on the spatial offset from the true path which will be directly proportional to the internal speed (for a constant external motion).
+On the other hand, if spontaneous resets are the result of the temporal limitations of the integration process, resets would occur after a certain amount of time, independently of the internal speed and the spatial offset it creates.
+Investigating the time and location of resets in the recorded trajectories, allows us to distinguish between a spatial limit to the illusion and a temporal limit.
+This in turn will inform us of the processes underlying the perception of the position of moving objects.
+
+To do so, we ask participants to re-trace the perceived path of a double-drift stimulus, immediately after viewing the stimulus, a method used by Nakayama & Holcombe [-@Nakayama2020].
+This allows participants more time to carefully reproduce the path.
+We use the recorded trajectories to assess points where resets occurred, whether they were a discontinuity in orientation reflecting a saturation (hit the wall reset, Fig.
+4) or a discontinuity in position reflecting a jump back to the real location.
+These reset points can then be used to determine if resets are time-limited or space-limited.
 
@@
 Stimuli were identical to those used in the online tracking experiment, but now always started at the near end of the workspace and moved away from the participants.
-The double drift stimuli would use either 3 or 4 seconds to move 13.5 _cm_ and had an internal drift of 2, 3 or 4 _cps_.
+The double drift stimuli would use either 3 or 4 seconds to move 13.5 _cm_ ( corresponding to speeds of ~4.5 and ~3.75 _cm/s_ or ~4.7 or ~3.5 _dva/s_ ) and had an internal drift of 2, 3 or 4 _cps_ .
 
@@
 
-Re-tracing the perceived paths should allow capturing spontaneous resets of the illusion, and the depence on the strength of the illusion could be checked with the data from the line-orienting task.
+Re-tracing the perceived paths should allow capturing spontaneous resets of the illusion, and the dependence on the strength of the illusion could be checked with the data from the line-orienting task.
 
@@
 
-We use the same heading measure here as in experiment 1.
-However, the individual trajectories reveal more points that are potential spontaneous resets that we try to extract.
+The individual trajectories reveal some points that are potential spontaneous resets that we try to extract.
 Using a smooth spline interpolated version of the trajectory we detect left-right direction changes.
-Since the initial movement is often jittery, we only take direction changes further than ~0.39 cm away from the starting point.
+Since the initial movement is often jittery, we only take direction changes further than ~0.39 cm ( ~0.23 _dva_ ) away from the starting point.
 The first (if any) of the remaining direction changes, are then taken as reset points.
 By averaging the x and y coordinates of reset points in all 9 participants and 6 stimulus conditions, we obtain 54 average reset points for further analyses and modelling.
-The average line orientation is also calculated for each of those 54 combinations of partcipants and stimulus conditions.
+
+Apart from reset points, we need a measure of illusion strength.
+We take a trajectory sample halfway between the starting point and the detected reset point.
+We then use the angle of a straight line drawn through that point and the start point as a measure of illusion strength.
+The average response in the line orientation task is also calculated for each of those 54 combinations of partcipants and stimulus conditions, as a verification for the measure of illusion strength taken from the trajectories.
 
@@
 
-In Figure 4, for each of the 6 combinations of internal speed (2, 3 or 4 _cps_) and external speed (3 or 4 _s_ stimulus pass) we show the data for one example participant, as well as the heading distribution for all participants.
-While the heading distributions don't show resets, for many of the participants, we can see that the re-traced paths do show resets, many of which are detected by our algorithm.
+In Figure 4, we show the distributions of direction halfway to the reset points as well as trajectories for participant 4 in the condition with 3 _cps_ internal motion and 4 _s_ passes.
+This exemplary participant's re-traced paths show clear 'jump' resets, all of which are detected by our simple algorithm.
+Other participants are not that clear cut.
 
-```{r fig4, echo=FALSE, message=FALSE, fig.width=8, fig.height=5, fig.cap="\\label{fig:fig4}**Figure 4.
-Example Trajectories and Heading in Experiment 2.** For each of the six conditions, defined by 3 internal speeds (2, 3 and 4 _cps_) and two external speeds (corresponding to 3 and 4 _s_ passes), we show one example participant's trajectories and reset points, as well as the heading distribution."}
+```{r fig4, echo=FALSE, message=FALSE, fig.width=10, fig.height=4, fig.cap="\\label{fig:fig4}**Figure 4.
+Illusion Strength and Example Trajectoriesin Experiment 2.** _Left_: For each of the six conditions, defined by 3 internal speeds (2, 3 and 4 _cps_) and two external speeds (corresponding to 3 and 4 _s_ passes), we show the distribution of angles for points halfway between the start point and detected resets; our measure of illusion strength.
+Dots below the distribution are individual participants' average illusion strengths.
+Gray lines indicate no illusion.
+_Right_: One exemplary participant's trajectories with very clear 'jump' resets at the purple dots.
+(Purple lines: trajectories up to resets, gray lines: trajectories after resets, blue lines: trajectories without resets.) Other participants' data was not this clear."}
 plotExampleData()
@@
 
-```{r fig5, echo=FALSE, message=FALSE, fig.width=8, fig.height=4, fig.cap="\\label{fig:fig5}**Figure 5.
+```{r fig5, echo=FALSE, message=FALSE, fig.width=8, fig.height=8, fig.cap="\\label{fig:fig5}**Figure 5.
 Measures of Illusion Strength** Line angle responses and reset point direction should both measure illusion strength.
@@
 
-As specified above, we also take the angle of the sample on each re-tracing trajectory that is halfway between the origin and the detected reset, as an alternative measure of illusion strength.
+As specified above, we also take the angle of the sample on each re-tracing trajectory that is halfway between the origin and the detected reset, as our main measure of illusion strength.
 If both the angle half-way to reset and line angle responses are good measures of illusion strength, they should be related.
@@
 Reported line angles predicted reset point direction (slope=0.71; _t_(49)=4.54, _p_<.001; $R^2$=.282; _F_(1,49)=20.6), verifying that the two measures - while imperfect - are likely both indicators of double-drift illusion strength.
-From here on, we will only use the angle of samples half-way to detected reset points as measure of illusion strength.
+From here on, we will therefore only use the angle of samples half-way to detected reset points as measure of illusion strength.
 

### 2021-02-25 R/compare.R
simplify comparison with Sirui's data

@@ -19,8 +19,14 @@ functionCavanaghTse2019 <- function(internal=c(), external=c(), k=0.81) {
   
 }
 
-plotStrengthComparison <- function() {
+plotStrengthComparison <- function(target='svg') {
   
+  if (target == 'pdf') {
+    cairo_pdf(filename='doc/tracking_CTmodel.pdf',onefile=TRUE,width=5,height=5)
+  }
+  if (target == 'svg') {
+    svglite(file='doc/tracking_CTmodel.svg',width=5,height=5)
+  }
   
   colors <- getColors()
   
@@ -58,6 +64,7 @@ plotStrengthComparison <- function() {
       # }
       
       # also get the average, to plot on top of the polar heat map:
+      # or justs to plot a point for comparison with the Cavanagh & Tse 2019 model
       PPavgDir <- aggregate(direction ~ sample_no, data = stdsegments[idx,], FUN=mean, na.rm=T)
       
       internalspeed <- c(internalspeed, internalSpeed)
@@ -73,6 +80,10 @@ plotStrengthComparison <- function() {
   headingdir$avgheading[idx] <- headingdir$avgheading[idx] * -1 
   headingdir$internalspeed[idx] <- headingdir$internalspeed[idx] * -1 
   
+  #str(headingdir)
+  
+  headingdir <- aggregate(cbind(avgheading, participant) ~ internalspeed + participant, data=headingdir, FUN=mean)
+  #str(headingdir)
   #plot(c(5:35),IS['7.800',], type='l')
   
   # plot(-1000,-1000,xlim=c(0,25), ylim=c(0,60), xlab='external motion [dva/s]', ylab='illusion strength [deg]', bty='n', ax=F)
@@ -123,7 +134,7 @@ plotStrengthComparison <- function() {
   # # text(x=3.5,y=-2.5,'1b',adj=c(0.5,0.5))
   # # 
   # # # data for exp 1b:
-  # # points(x=c(3.5), y=c(28.9), col='blue', pch=19)
+  # # points(x=c(4.5), y=c(28.9), col='blue', pch=19)
   # 
   # 
   # legend(x=4, y=62, legend=c('Cavanagh & Tse (2019)', 'experiment 1'), col=c('#999999', '#FF0000'), pch=c(NA,19), lty=c(1,0), bty='n')
@@ -160,7 +171,9 @@ plotStrengthComparison <- function() {
[…]

### 2021-02-25 R/retrace_onePass_V4.R
changes to figures

@@ -421,7 +421,7 @@ getNormalizedHeading2Dhists <- function(segmentpoints=101, resetOnly=TRUE, load=
 
 # Figures -----
 
-plotExampleData <- function(target='inline') {
+plotMoreExampleData <- function(target='inline') {
   
   #graphics.off()
   
@@ -568,9 +568,7 @@ plotExampleData <- function(target='inline') {
   
 }
 
-plotValidity <- function(target='inline') {
-  
-  #graphics.off()
+plotExampleData <- function(target='inline') {
   
   if (target == 'pdf') {
     cairo_pdf(filename='doc/onePass_V4_boundaries.pdf',onefile=TRUE,width=8,height=4)
@@ -581,15 +579,241 @@ plotValidity <- function(target='inline') {
   
   colors <- getColors()
   
+  conditions <- list( c(2, 0.125), 
+                      c(2, 0.167),
+                      c(3, 0.125), 
+                      c(3, 0.167),
+                      c(4, 0.125), 
+                      c(4, 0.167) )
+  
+  
+  lom <- matrix(c(1,3,5,7,2,4,6,7), ncol = 4, nrow = 2, byrow = TRUE)
+  layout(lom, widths=c(1,1,1,1.5))
+  
+  idd <- read.csv('data/onePass_V4/onePass_V4_re-trace.csv', stringsAsFactors = F)
+  idd <- idd[-which(is.na(idd$initialdirection)),]
+  
+  participants <- unique(idd$participant)
+  
+  for (condition in conditions) {
+    
+    plot(-1000,-1000,main=sprintf('%d cps, %s s',condition[1],list('0.125'='4', '0.167'='3')[sprintf('%0.3f',condition[2])]),
+         xlab='illusion strength [deg]', ylab='relative density',
+         xlim=c(-20,80), ylim=c(-.2,1),
+         bty='n',ax=F)
+    
+    lines(c(0,0),c(0,1),col='#999999',lty=1)
+    
+    IS <- condition[1]
+    ES <- condition[2]
+    
+    distribution <- rep(0,101)
+    directions <- c()
+    
+    for (ppno in participants) {
+      
+      ppdir <- idd$initialdirection[ which( idd$internalspeed == condition[1] &
[…]

### 2021-02-25 R/tracking_bounded.R
work on figures for first experiment

@@ -353,10 +353,14 @@ plotBoundedTracking_V2 <- function(target='inline') {
     adX <- (cos((PPavgDir$direction/180)*pi) * scale)
     adY <- (sin((PPavgDir$direction/180)*pi) * scale)
     lines(adX,adY,col=colors[['blue']]$s,lw=2)
-    
+    # text(0,0,)
   }
   
   polarHeatMap(x=pfreq$x.edges, y=pfreq$y.edges+30, z=freq2D, mincol=c(1,1,1), border=NA, ylim=c(0,1), main='group average', overlay=FALSE)
+  
+  ppAvgDirs <-  90 - colMeans( matrix(avgDir, ncol=length(participants), byrow=FALSE)  ) 
+  print(ppAvgDirs)
+  print(mean(ppAvgDirs))
   
   avgDir <- (rowMeans( matrix(avgDir, ncol=length(participants), byrow=FALSE) ) / 180) * pi
   text(0, 1.2, sprintf('%0.1f°',90 - ( ( mean(avgDir) / pi) * 180 ) ) )

### 2021-02-26 Manuscript.Rmd
minor updates to the manuscript

@@
 Canada\
-thartbm@gmail.com\
+[thartbm@gmail.com](thartbm@gmail.com)\
 
@@
 However, when the gabor itself also moves as in the double-drift stimulus, there is no stable location information to anchor the position estimate and the build-up continues well beyond 100 ms.
-The perceived position conintues to move away from its physical position up to several times the size of the gabor.
-The strength of the illusion, or the deviation of perceived path of the Gabor from its true path, depends on the speed of both the internal and external motion [@Tse2006; @Shapiro2010; @Kwon2015]; @Cavanagh2019.
+The perceived location conintues to move away from its physical position by up to several times the size of the gabor.
+The strength of the illusion, the deviation of perceived path of the Gabor from its true path, depends on the speed of both the internal and external motion [@Tse2006; @Shapiro2010; @Kwon2015; @Cavanagh2019].
 In our first experiment, participants track their percept with a stylus as the gabor moves back and forth on its trajectory (Fig.
 1) with 5 speeds of the internal motion.
-We analyze the angle of the tracking to determine if manual response is susceptible to the illusion.
+We analyze the angle of the tracking to determine if the manual response is susceptible to the illusion.
 
@@
 The continuation of the accumulation may be a result of the active tracking – either the extra attention required to track increased the extent of the accumulation or possibly the temporal demands of the tracking caused the participants to average their traces over any breaks.
-To address this, we switched to an offline recording of the perceived path.
+To address this, we switched to a delayed, offline recording of the perceived path.
 
-```{r fig4b, echo=F, fig.height=3.2, fig.width=5.62, message=FALSE, error=FALSE, warning=FALSE, fig.cap="**Figure 4: Resets.** _Left_: At some point, the perceived position will stop moving further away from its actual path.
-For some people this takes the form of a ‘jump’ back toward the true position, for others it means the illusory position remains at a fixed offset moving parallel to the real path, or some combnation of the two.
-_Top right_: if these resets are due to a spatial limit on the size of the illusory position shift away from the true vertical path, the locations of the resets should have a constant X coordinate.
-_Bottom right_: if they occur instead after some period of time, they should have a constant Y coordinate."}
+```{r fig4b, echo=F, fig.height=3.2, fig.width=5.62, message=FALSE, error=FALSE, warning=FALSE, fig.cap="**Figure 4: Resets.** **Left:** At some point, the perceived position will stop moving further away from its actual path.
+For some people this takes the form of a ‘jump’ back toward the true position, for others it means the illusory position remains at a fixed offset moving parallel to the real path, or some combination of the two.
+**Top right:** if these resets are due to a spatial limit on the size of the illusory position shift away from the true vertical path, the locations of the resets should have a constant X coordinate.
+**Bottom right:** if they occur instead after some period of time, they should have a constant Y coordinate."}
 library(magick)
@@
 
-```{r fig4, echo=FALSE, message=FALSE, fig.width=10, fig.height=4, fig.cap="\\label{fig:fig4}**Figure 4.
+```{r fig5, echo=FALSE, message=FALSE, fig.width=10, fig.height=4, fig.cap="\\label{fig:fig5}**Figure 5.
 Illusion Strength and Example Trajectoriesin Experiment 2.** _Left_: For each of the six conditions, defined by 3 internal speeds (2, 3 and 4 _cps_) and two external speeds (corresponding to 3 and 4 _s_ passes), we show the distribution of angles for points halfway between the start point and detected resets; our measure of illusion strength.
@@
 Now we look at the distribution of reset points, and whether they can be explained by a temporal or spatial limit.
-In Figure 7 (left), we can see that the reset points vary more along the Y-axis than the X-axis.
+In Figure 6 (left), we can see that the reset points vary more along the Y-axis than the X-axis.
 That is, predicting reset points from a single Y coordinate (time-cutoff) would leave larger residuals than predicting reset points from a single X coordinate (or spatial cutoff).
@@
 The limits that optimally fit the model to the data are a 1.42 _cm_ spatial limit and a 0.28 _s_ temporal limit.
-While the AIC of the model (-209.8) is marginally lower than that of the model with the single spatial limit, but it does not explain the data better ( _p_=.53).
+While the AIC of the model (-209.8) is marginally lower than that of the model with the single spatial limit, it does not explain the data better ( _p_=.53).
 This means that a simple spatial limit accounts for most of the data.
@@
 There may be several reasons for this.
-First, our data may simply be too noisy, or the effects may be highly dependent on participants so that both studies may suffer from sampling bias.
-We can't reolve this contradiction here and leave that for future work.
+Our data may simply be too noisy, or the effects may be highly dependent on participants so that both studies may suffer from sampling bias.
+But primarily, the PSE-based approach relies on all participants to exhibit full jump-resets, whereas the approach presented here is also valid for weaker resets.
+Nevertheless, we can't reolve this contradiction here and leave that for future work.
 

### 2021-03-01 Manuscript.Rmd
combined graphs, improved text of the manuscript

@@
   word_document: default
-  pdf_document: default
   html_document:
     df_print: paged
+  pdf_document: default
 header-includes:
@@
 Before trial onset, the participant moved the stylus to the middle of the screen ( where the gabor would appear ), and was subsequently asked to fixate a point on the left or right of the screen, and given 1.5 s to do so.
-Since the screen was seen via a mirror, the stimulus appeared to move forward and backward on a horizontal plane at the same depth as the drawing tablet.
 The internal motion of the gabor was inverted at the far and close ends of the physical path, where external motion was also inverted.
@@
 
-In Figure 2 (top), we plot the average instantaneous heading during the middle second of each 2 _s_ pass for each of the 5 internal motion speeds.
+In Figure 2 (top right), we plot the average instantaneous heading during the middle second of each 2 _s_ pass for each of the 5 internal motion speeds.
 The average angle of the tracked path is given in degrees deviation from straight forward.
@@
 
-```{r fig2, echo=FALSE, message=FALSE, fig.width=4.5, fig.height=6, fig.cap="\\label{fig:fig2}**Figure 2.
-Heading and Trajectories in Experiment 1.** **Top:** Average heading across the middle 1 _s_ of each pass.
+```{r fig2, echo=FALSE, message=FALSE, fig.width=8, fig.height=5, fig.cap="\\label{fig:fig2}**Figure 2.
+Online Tracking Data** **Left:** Individual trajectories.
+Note that there are only a few obvious discontinuities (highlighted in red).
+In general, the traced paths showed rough linearity.
+**Top right:** Average heading in the middle 1 _s_ of each pass.
 Distance from origin is time, angle is angle.
-In darker blue lines the average heading across participants, in light blue a polar heatmap giving an impression of the distribution of heading.
-**Bottom:** All individual trajectores.
-Note that there are only a few obvious discontinuities (highlighted in red).
-In general, the traced paths showed rough linearity."}
+Dark blue lines the average heading, in light blue a polar heatmap with the distribution of heading.
+**Bottom right: The strength of the illusion in degrees angle deviation from the physical path over unity slope (internal / external velocity).
+In red are the average illusions strengths for 0 _cps_, ±1 _cps_ and ±3 _cps_, and in light blue the data for individual participants.
+Gray dashed line: unity.
+Black line: fit from Cavanagh & Tse (2019), 81% slope.
+Purple: fit to the data, 72% slope.
+**"}
 plotBoundedTracking()
@@
 The data clearly show the illusion, but does tracking decrease the strength of the illusion compared to simply observing it?
-We compare mean illusion strength to that reported previously [@Cavanagh2019] in Figure 3.
-As can be seen the averages from our data here (in red) coincide almost exactly with the illusion strength predicted by the model (in gray).
+In Figure 2 (bottom right), we compare mean illusion strength to that reported previously [@Cavanagh2019].
+As can be seen the averages from our data here (red circles, purple line) coincide fairly well with the illusion strength predicted by the model fit to previous data (in black).
 That is; tracking the illusion while observing it does not change the strength of the illusion.
-
-```{r fig3, echo=FALSE, message=FALSE, fig.width=5, fig.height=5, fig.cap="\\label{fig:fig3}**Figure 3.
-Illusion strength.** The strength of the illusion in degrees angle deviation from the physical path is plotted over external motion speed ( _dva/s_ ).
-In red are the average illusions strengths for 0 _cps_, ±1 _cps_ and ±3 _cps_, and in light blue the data for individual participants.
-The gray lines are the predicted illusion strengths based on the model from Cavanagh & Tse (2019) for the internal motion speeds used here."}
-plotStrengthComparison()
-```
 
@@
 
-The durations of each pass in the online tracking experiment was 2 seconds but the manual traces mostly showed a linear trajectory without saturation or reset of the types shown in Figure 4 on the left below.
+The durations of each pass in the online tracking experiment was 2 seconds but the manual traces (Figure 2, left) mostly showed a linear trajectory without saturation or reset of the types shown in Figure 3 on the left below.
 The continuation of the accumulation may be a result of the active tracking – either the extra attention required to track increased the extent of the accumulation or possibly the temporal demands of the tracking caused the participants to average their traces over any breaks.
@@
 
-```{r fig4b, echo=F, fig.height=3.2, fig.width=5.62, message=FALSE, error=FALSE, warning=FALSE, fig.cap="**Figure 4: Resets.** **Left:** At some point, the perceived position will stop moving further away from its actual path.
+```{r fig3, echo=F, fig.height=4.55, fig.width=8, message=FALSE, error=FALSE, warning=FALSE, fig.cap="\\label{fig:fig3}**Figure 3: Resets.** **Left:** At some point, the perceived position will stop moving further away from its actual path.
 For some people this takes the form of a ‘jump’ back toward the true position, for others it means the illusory position remains at a fixed offset moving parallel to the real path, or some combination of the two.
@@
 
-In Figure 4, we show the distributions of direction halfway to the reset points as well as trajectories for participant 4 in the condition with 3 _cps_ internal motion and 4 _s_ passes.
+In Figure 4 (left), we show the trajectories for participant 4 in the condition with 3 _cps_ internal motion and 4 _s_ passes.
 This exemplary participant's re-traced paths show clear 'jump' resets, all of which are detected by our simple algorithm.
@@
 
-```{r fig5, echo=FALSE, message=FALSE, fig.width=10, fig.height=4, fig.cap="\\label{fig:fig5}**Figure 5.
-Illusion Strength and Example Trajectoriesin Experiment 2.** _Left_: For each of the six conditions, defined by 3 internal speeds (2, 3 and 4 _cps_) and two external speeds (corresponding to 3 and 4 _s_ passes), we show the distribution of angles for points halfway between the start point and detected resets; our measure of illusion strength.
-Dots below the distribution are individual participants' average illusion strengths.
-Gray lines indicate no illusion.
-_Right_: One exemplary participant's trajectories with very clear 'jump' resets at the purple dots.
-(Purple lines: trajectories up to resets, gray lines: trajectories after resets, blue lines: trajectories without resets.) Other participants' data was not this clear."}
-plotExampleData()
+```{r fig4, echo=FALSE, message=FALSE, fig.width=8, fig.height=3.5, fig.cap="\\label{fig:fig4}**Figure 4.
+Data in Experiment 2.** **Left:*** One exemplary participant's trajectories with very clear 'jump' resets at the purple dots.
+(Purple lines: trajectories up to resets, gray lines: trajectories after resets, blue lines: trajectories without resets.) Other participants' data was not this clear.
+**Middle:** Average participant reset points with standard deviations.
+(Gray arrow: real trajectory, blue line: spatial average, red line: temporal average.) **Right:** Illusion strength as in Figure 3; the acrtangent of the slope (internal velocity over external velocity).
+Gray dahsed line: unity, black line: data/model as in Cavanagh & Tse (2019) at 81% of unity, purple line: best fit here at 42% of unity."}
+#plotExampleData()
+plotData()
 ```
@@
 Either type of perceived resets appears valid, and the average reset positions still indicates some limit to the illusion.
-Therefore, we keep all participants and resets.
+Therefore, we keep all participants and resets (reset points in Figure 4, middle).
 
-First, we consider the line angle responses as a measure of illusion strength.
-The strength of the illusion should depend on both the internal and external speed of the double drift stimulus.
-As can be seen in Figure 5 (left), at least the internal speed, and perhaps the external speed may predict illusion strength as measured by line angles.
-To test this, we perform a mixed-design ANOVA on line angle, using both internal and external speed as within-subject factors.
-We find main effects of both internal speed ( _F_(2,16)=12.7; _p_<.001, $\eta^2$=0.049) and external speed ( _F_(1,8)=6.89; _p_=.03; $\eta^2$=0.031), and no interaction.
-That is: we verify that illusion strength varied systematically with our manipulations of the stimulus.
-
-```{r fig5, echo=FALSE, message=FALSE, fig.width=8, fig.height=8, fig.cap="\\label{fig:fig5}**Figure 5.
-Measures of Illusion Strength** Line angle responses and reset point direction should both measure illusion strength.
-**Left:** Line Angle response as a function of internal and external speed, averaged across 9 participants in each of the 6 conditions.
-Shaded areas denote 95% confidence intervals.
-**Right:** Angles of samples halfway to a reset over line angle response.
-Each dot corresponds to one participants average in one of the 6 conditions.
-Diagonal line indicates a weighted regression (weighted by one over the square of the line angle response to counter heteroscedasticity), and the shaded area shows a 95% confidence interval for the regression."}
-plotValidity()
-```
-
-As specified above, we also take the angle of the sample on each re-tracing trajectory that is halfway between the origin and the detected reset, as our main measure of illusion strength.
-If both the angle half-way to reset and line angle responses are good measures of illusion strength, they should be related.
-As can be seen from the data in Figure 5 (right), there might be a relation between the two variables, but there is also heteroscadasticity.
-We counter this by doing a weighted regression, with the weights set to one over the squared line angle ($\frac{1}{\alpha^2}$).
-Reported line angles predicted reset point direction (slope=0.71; _t_(49)=4.54, _p_<.001; $R^2$=.282; _F_(1,49)=20.6), verifying that the two measures - while imperfect - are likely both indicators of double-drift illusion strength.
-From here on, we will therefore only use the angle of samples half-way to detected reset points as measure of illusion strength.
+We compare the strength of the illusion as measured with half-way points (see Fig 4, right), with an earlier study [@Cavanagh2019].
+A line through the origin with slope 0.42 predicts illusion strength fairly well in this data, which is weaker than earlier (0.81) or with online tracking (0.72).
+We can also see that illusion strength measured with re-tracing is more variable.
+While the illusion is weaker, it is nevertheless there and depends on the relationship between internal and external speed, as expected.
 
@@
 Now we look at the distribution of reset points, and whether they can be explained by a temporal or spatial limit.
-In Figure 6 (left), we can see that the reset points vary more along the Y-axis than the X-axis.
+In Figure 4 (middle), we can see that the reset points vary more along the Y-axis than the X-axis.
 That is, predicting reset points from a single Y coordinate (time-cutoff) would leave larger residuals than predicting reset points from a single X coordinate (or spatial cutoff).
@@
 
-```{r fig6, echo=FALSE, message=FALSE, fig.width=8, fig.height=6, fig.cap="\\label{fig:fig6}**Figure 6.
-Reset Points and Single Limits** **Left:** True path of the double-drift stimulus as gray arrow.
-The average reset points for each participant and stimulus condition displayed spatially, with the internal motion normalized to go rightward.
-Y-coordinates have been converted to time ( _s_ ), in order to normalize for the two external speeds.
-Transparent purple lines indicate the standard deviation of coordinates.
-Dashed vertical red line indicates the median X-coordinate (1.6 _cm_).
-Dashed horizontal blue line indicates the median Y-coordinate (1.4 _s_).
-**Top right:** X-coordinates predicted from a single temporal limit.
-**Bottom right:** Y-coordinates predicted from a single spatial limit."}
-plotResetPoints()
+```{r fig5, echo=FALSE, message=FALSE, fig.width=8, fig.height=3, fig.cap="\\label{fig:fig5}**Figure 5.
+Reset Models** **Left:** X-coordinates predicted from a single temporal limit.
+**Middle:** Y-coordinates predicted from a single spatial limit.
+**Right:** X- and Y-coordinates predicted from a temporal and a spatial limit."}
+#plotResetPoints()
+plotModels()
 ```
@@
 Where $L_y$ and $L_x$ are a temporal (y) and spatial (x) limit that are free parameters, and s corresponds to the calculated slope.
-If we fit these simple models by minimizing the mean square error (as the absolute distance in centimeters between predicted and actual reset points), the optimal parameters indicate a temporal limit of ~1.4 _s_ (AIC=-167.8) and a spatial limit of ~1.7 _cm_ (AIC=-208.6).
-This is close to the median values of the respective coordinates.
+If we fit these simple models by minimizing the mean square error (as the absolute distance in centimeters between predicted and actual reset points), the optimal parameters indicate a temporal limit of ~1.4 _s_ (AIC=-167.8) and a spatial limit of ~1.7 _cm_ (AIC=-208.6), as illustrated in Figure 6 (left and middle).
+This is close to the mean values of the respective coordinates (Figure 5, middle).
 The model with the spatial limit fits the data better than the model with a temporal limit ( _p_<.001, relative log-likelihood based on AICs).
 This again indicates that a spatial limit explains more of the data.
+
+## Combined Limits
 
@@
 
-Figure 7 shows both the actual reset point coordinates (dots), given the illusion strength as the angle to a sample half-way to the reset point, as well as the model's predictions of reset point coordinates (lines).
-
-```{r fig7, echo=FALSE, message=FALSE, fig.width=6, fig.height=6, fig.cap="\\label{fig:fig7}**Figure 7. Reset Points and Combined Limits**"}
-plotSequentialModel()
-```
+Figure 5 (right) shows both the actual reset point coordinates (dots), given the illusion strength as the angle to a sample half-way to the reset point, as well as the model's predictions of reset point coordinates (lines).
 
 The limits that optimally fit the model to the data are a 1.42 _cm_ spatial limit and a 0.28 _s_ temporal limit.
-While the AIC of the model (-209.8) is marginally lower than that of the model with the single spatial limit, it does not explain the data better ( _p_=.53).
+While the AIC of the model (-209.8) is slightly lower than that of the model with the single spatial limit, it does not explain the data better ( _p_=.53).
 This means that a simple spatial limit accounts for most of the data.
@@
 But primarily, the PSE-based approach relies on all participants to exhibit full jump-resets, whereas the approach presented here is also valid for weaker resets.
-Nevertheless, we can't reolve this contradiction here and leave that for future work.
+Nevertheless, we can't resolve this contradiction here and leave that for future work.
 
@@
 However, it seems that for the participants that display "jump" resets in their re-tracing data, these happen at a fairly consistent location, which suggests that jump resets are instead usually triggered by the stimulus and not by attention.
-This should however be tested with eye-tracking data, so that this will remain a speculation here.
+This should however be tested with eye-tracking, so that this will remain a speculation here.
 

### 2021-03-01 R/compare.R

@@ -206,3 +206,43 @@ plotStrengthComparison2 <- function() {
   axis(side=2, at=seq(0,40,10))
   
 }
+
+plotStrengthComparison3 <- function() {
+  
+  colors <- getColors()
+  
+  df <- getAverageHeading()
+  
+  internalspeed <- df$internalspeed / 0.58 # in cm/s 
+  externalspeed <- 13.5             / 2    # in cm/s
+  
+  xcoords <- atan(internalspeed / externalspeed)
+  
+
+  plot(-1000,-1000,
+       main='',xlab=expression(paste(tan^{-1}, (V[i]/V[e]))),ylab='illusion strength [°]',
+       xlim=c(pi*-((0.25)/9),pi/4),ylim=c(-5,45),
+       bty='n',ax=F)
+  
+  angles <- seq(0,pi/2,.05)
+  lines(angles,(angles/pi)*180,col='gray',lty=2)
+  lines(angles,0.81*((angles/pi)*180),col='black',lty=1)
+  
+  
+  points(xcoords, df$avgheading, col=colors$lightblue$s, pch=1, cex=1.0)
+  
+  agg.df <- aggregate(avgheading ~ internalspeed, data=df, FUN=mean)
+  
+  points(atan((agg.df$internalspeed / 0.58) / externalspeed), agg.df$avgheading, col=colors$yorkred$s, pch=1, cex=1.5)
+  
+  
+  legend(x=2.5, y=10, 
+         legend=c('participants', 'average', 'Cavanagh & Tse (2019)'), 
+         col=c(colors$lightblue$s, colors$yorkred$s, '#999999'), 
+         pch=c(1,1,NA), lty=c(0,0,1), 
+         bty='n', cex=1)
+  
+  axis(side=1,at=seq(0,pi/4,pi/8),labels=c('0',expression(pi/8),expression(pi/4)))
+  axis(side=2,at=seq(0,45,15))
+  
+}

### 2021-03-01 R/retrace_onePass_V4.R

@@ -718,7 +718,222 @@ plotExampleData <- function(target='inline') {
   
 }
 
-plotValidity <- function(target='inline') {
+plotData <- function(target='inline') {
+  
+  if (target == 'pdf') {
+    cairo_pdf(filename='doc/onePass_V4_boundaries.pdf',onefile=TRUE,width=8,height=4)
+  }
+  if (target == 'svg') {
+    svglite(file='doc/onePass_V4_boundaries.svg',width=8,height=4)
+  }
+  
+  par(mar=c(3.5, 3.5, 2.5, 0.5))
+  
+  colors <- getColors()
+  
+  # there will be three plots:
+  # 1: example participant
+  # 2: 2D overview of average reset points
+  # 3: Cavangh & Tse comparison
+  
+  layout(matrix(c(1,2,3), nrow = 1, ncol = 3, byrow = T), widths=c(1.2,1.2,1.5))
+  
+  
+  
+  # **********************************
+  # EXAMPLE PARTICIPANT PLOT:
+  
+  plot(-1000,-1000,
+       main='participant 4, 3 cps, 4 s',xlab='',ylab='',
+       xlim=c(-.25,.5), ylim=c(-.1,.9),
+       bty='n',ax=F, asp=1)
+  
+  title(xlab='cm', line=2.5)
+  title(ylab='cm', line=2.5)
+  
+  df <- read.csv(sprintf('data/onePass_V4/onepass_V4_p%02d.csv',4))
+  tasktrials <- unique(df$trial[df$taskname == 're-trace'])
+  taskdf <- df[which(df$trial_no %in% tasktrials),]
+  
+  trials <- unique(taskdf$trial_no[which(taskdf$externalMovement == 0.125 &
+                                           abs(taskdf$internalMovement) == 3)])
+  
+  # loop through those trials:
+  for (trialno in trials) {
+    
+    trialdf <- taskdf[taskdf$trial_no == trialno,]
+    
+    step.idx <- which(trialdf$step == 99) # only for re-trace
+    
+    x <- ((trialdf$handx_pix[step.idx] / (524)) + 0.0) * 0.6
+    y <- ((trialdf$handy_pix[step.idx] / (524)) + 0.5) * 0.6
+    t <- trialdf$time_ms[step.idx]
+    
+    if (trialdf$internalMovement[1] < 0) x <- -x
+    
[…]

### 2021-03-01 R/tracking_bounded.R

@@ -147,23 +147,87 @@ getSegmentDirections <- function(version=1) {
   
 }
 
-# Figures -----
 
+getAverageHeading <- function(mirror=TRUE) {
+  
+  stdsegments <- read.csv('data/bounded_tracking/segment_directions.csv', stringsAsFactors = F)  
+  
+  stdsegments$direction[which(stdsegments$direction < 0)] <- NA
+  stdsegments <- stdsegments[which(!is.na(stdsegments$direction)),]
+  
+  internalSpeeds <- sort(unique(stdsegments$internalSpeed))
+  participants <- sort(unique(stdsegments$participant))
+  
+  internalspeed <- c()
+  participant <- c()
+  avgheading <- c()
+  
+  for (internalSpeed.idx in c(1:length(internalSpeeds))) {
+    
+    internalSpeed <- internalSpeeds[internalSpeed.idx]
+    
+    for (pp in participants) {
+      
+      # this is the part of the data we're dealing with now:
+      idx <- which(stdsegments$internalSpeed == internalSpeed & stdsegments$participant == pp)
+      
+      # # we put it into a normalized 2D histogram:
+      # pfreq <- hist2d(x=stdsegments$direction[idx], y=stdsegments$sample_no[idx], nbins=NA, edges=list(seq(0,180,length.out=61), seq(1,101,4)+0.5))
+      # 
+      # if (!is.matrix(freq2D)) {
+      #   freq2D <- pfreq$freq2D
+      # } else {
+      #   freq2D <- freq2D + pfreq$freq2D
+      # }
+      
+      # also get the average, to plot on top of the polar heat map:
+      # or justs to plot a point for comparison with the Cavanagh & Tse 2019 model
+      PPavgDir <- aggregate(direction ~ sample_no, data = stdsegments[idx,], FUN=mean, na.rm=T)
+      
+      internalspeed <- c(internalspeed, internalSpeed)
+      participant <- c(participant, pp)
+      avgheading <- c(avgheading, 90-mean(PPavgDir$direction))
+      
+    }
+    
+  }
+  
+  headingdir <- data.frame(internalspeed, participant, avgheading)
+  
+  
+  if (mirror) {
+    idx <- which(headingdir$internalspeed < 0)
[…]

### 2021-03-01 Manuscript.Rmd
implemented half-way points correctly, updated re-tracing exp results / models

@@
 Dark blue lines the average heading, in light blue a polar heatmap with the distribution of heading.
-**Bottom right: The strength of the illusion in degrees angle deviation from the physical path over unity slope (internal / external velocity).
+**Bottom right:** The strength of the illusion in degrees angle deviation from the physical path over unity slope (internal / external velocity).
 In red are the average illusions strengths for 0 _cps_, ±1 _cps_ and ±3 _cps_, and in light blue the data for individual participants.
@@
 Black line: fit from Cavanagh & Tse (2019), 81% slope.
-Purple: fit to the data, 72% slope.
-**"}
+Purple: fit to the data, 72% slope."}
 plotBoundedTracking()
@@
 We compare the strength of the illusion as measured with half-way points (see Fig 4, right), with an earlier study [@Cavanagh2019].
-A line through the origin with slope 0.42 predicts illusion strength fairly well in this data, which is weaker than earlier (0.81) or with online tracking (0.72).
+A line through the origin with slope 0.49 predicts illusion strength fairly well in this data, which is weaker than earlier (0.81) or with online tracking (0.72).
 We can also see that illusion strength measured with re-tracing is more variable.
@@
 Where $L_y$ and $L_x$ are a temporal (y) and spatial (x) limit that are free parameters, and s corresponds to the calculated slope.
-If we fit these simple models by minimizing the mean square error (as the absolute distance in centimeters between predicted and actual reset points), the optimal parameters indicate a temporal limit of ~1.4 _s_ (AIC=-167.8) and a spatial limit of ~1.7 _cm_ (AIC=-208.6), as illustrated in Figure 6 (left and middle).
+If we fit these simple models by minimizing the mean square error (as the absolute distance in centimeters between predicted and actual reset points), the optimal parameters indicate a temporal limit of ~1.4 _s_ (AIC=-167.1) and a spatial limit of ~1.8 _cm_ (AIC=-220.7), as illustrated in Figure 6 (left and middle).
 This is close to the mean values of the respective coordinates (Figure 5, middle).
@@
 
-The limits that optimally fit the model to the data are a 1.42 _cm_ spatial limit and a 0.28 _s_ temporal limit.
-While the AIC of the model (-209.8) is slightly lower than that of the model with the single spatial limit, it does not explain the data better ( _p_=.53).
+The limits that optimally fit the model to the data are a 1.47 _cm_ spatial limit and a 0.33 _s_ temporal limit.
+While the AIC of the model (-226.9) is slightly lower than that of the model with the single spatial limit, it does not explain the data better ( _p_=.45).
 This means that a simple spatial limit accounts for most of the data.
-It is possible that the way we've modeled the additional temporal delay after crossing the spatial limit is not accurate, as it may not follow a normal distribution, which we do not account for here.
+It is possible that the way we've modeled the additional temporal delay after crossing the spatial limit is not accurate, e.g.
+it may not follow a normal distribution.
 However, regardless of whether or not a temporal limit adds anything, a spatial limit appears to explain the majority of spontaneous double-drift resets.

### 2021-03-01 R/retrace_onePass_V4.R

@@ -31,6 +31,7 @@ preProcessOnePass_V4 <- function(participants = c(2,3,4,5,6,8,9,10,11), target='
     externalspeed <- c()
     fixationside <- c()
     initialdirection <- c()
+    illusionstrength <- c()
     boundX <-c()
     boundY <-c()
     
@@ -88,6 +89,7 @@ preProcessOnePass_V4 <- function(participants = c(2,3,4,5,6,8,9,10,11), target='
               lines(x+(participant.idx*2)-2+EM.idx,y+IM.idx-0.9,col='#00000033')
               
               initialdirection <- c(initialdirection, 90 - percept)
+              illusionstrength <- c(illusionstrength, 90 - percept)
               boundX <- c(boundX, NA)
               boundY <- c(boundY, NA)
               
@@ -107,14 +109,17 @@ preProcessOnePass_V4 <- function(participants = c(2,3,4,5,6,8,9,10,11), target='
               #   # print(y)
               # }
               
-              
               if (trialdf$internalMovement[1] < 0) x <- -x
               
+              # this has to be redone once the reset point is found!
               point <- which(sqrt(x^2 + y^2) > 0.15)[1]
               percept <- (atan2(y[point], x[point]) / pi) * 180
               initialdirection <- c(initialdirection, 90 - percept)
+              # IDcoords <- c(x[point],y[point])
               
               # why do this and also the smoothed spline?
+              # to test if we should be doing the splines at all?
+              # and the splines find the "more correct" point... hmmm...
               boundary <- which(diff(x) < 0)[1]
               
               if (is.na(boundary)) {
@@ -122,6 +127,7 @@ preProcessOnePass_V4 <- function(participants = c(2,3,4,5,6,8,9,10,11), target='
                 
                 boundX <- c(boundX, NA)
                 boundY <- c(boundY, NA)
+                illusionstrength <- c(illusionstrength, NA)
                 
               } else {
                 
@@ -132,20 +138,36 @@ preProcessOnePass_V4 <- function(participants = c(2,3,4,5,6,8,9,10,11), target='
                 smspl <- smooth.spline(t, x, spar=.25)
[…]

### 2021-03-04 Manuscript.Rmd
integrated feedback from Patrick

@@
 output:
-  word_document: default
-  pdf_document: default
+  word_document:
+    reference_docx: template.docx
   html_document:
     df_print: paged
+  pdf_document: default
 header-includes:
@@
 - \captionsetup[figure]{labelformat=empty}
+- \usepackage{amsmath}
 bibliography: bibliography.bib
@@
 The magnitude of the tracking illusion corresponded closely to previous perceptual and pointing measures indicating that manual tracking is a valid measure for the illusion.
-This allowed us to use it to capture the behavior of the illusion as it eventually degrades and breaks down in single trials.
+This allowed us to use it in a second experiment to capture the behavior of the illusion as it eventually degrades and breaks down in single trials.
 Specifically, the deviation of the gabor stops accumulating at some point and either stays at a fixed offset or resets toward the veridical position.
-In a second experiment, participants re-traced the perceived trajectory after the gabor was removed from the monitor.
-On alternate trials, they indicate the initial direction of the gabor’s path.
-Resets were detected and analyzed and they suggest that there is a spatial limit to  the perceived offset beyond which the deviation saturates or resets.
+To report the perceived trajectory of the gabor, participants drew it after the gabor was removed from the monitor.
+Resets were detected and analyzed and they suggest that there is a spatial limit beyond which the illusion saturates or resets.
 
@@
 the flash lag [@Nijhawan1994]; or the flash grab effect [@Cavanagh2013].
-These cases of position extrapolation may be functional, helping the motor system overcome inevitable neural delays in targeting a moving object [@Nijhawan1994; @Duhamel1992].
+These cases of position extrapolation may be functional, helping the motor system overcome inevitable neural delays in targeting a moving object [@Hogendoorn2020; @Nijhawan1994; @Duhamel1992].
 A stronger and very different motion-induced position shift arises when a moving gabor has internal motion orthogonal to its path and is viewed in the periphery [@Tse2006; @Shapiro2010; @Gurnsey2012; @Kwon2015; @Lisi2015].
@@
 1, top left) generates extreme misjudgments of the moving target’s location which may deviate by as much as several degrees of visual angle from its true location.
-Surprisingly, this extraordinary perceptual illusion does not affect eye movements to the gabor: saccades to these targets are determined by their physical not their perceived location [@Lisi2015].
-And although immediate eye-movements go towards the true position, delayed eye movements and pointing go to the perceived position [@Lisi2017; @Massendari2018].
+Surprisingly, this extraordinary perceptual illusion does not affect eye movements to the gabor: saccades to these targets are determined by their physical not their perceived location [@Lisi2015], and although immediate eye-movements go towards the true position, delayed eye movements and pointing go to the perceived position [@Lisi2017; @Massendari2018].
 These findings suggest that both the true and illusory positions are available in the brain.
 
-```{r fig1, echo=F, fig.height=3.2, fig.width=8, message=FALSE, error=FALSE, warning=FALSE, fig.cap="**Figure 1: Stimuli and Setup.** _Left_: a gabor with internal motion (phase shifts) and external motion, that is viewed peripherally, appears to follow a path deviating from the true path.
-_Right_: participants track their as the gabor moves back and forth along its path."}
+```{r fig1, echo=F, fig.height=3.2, fig.width=8, message=FALSE, error=FALSE, warning=FALSE, fig.cap="**Figure 1:** Stimuli and Setup **Left:** a gabor with internal motion (phase shifts) and external motion, that is viewed peripherally, appears to follow a path deviating from the true path.
+**Right:** participants track the perceived trajectory as the gabor moves back and forth along its path."}
 library(magick)
@@
 The perceived location conintues to move away from its physical position by up to several times the size of the gabor.
-The strength of the illusion, the deviation of perceived path of the Gabor from its true path, depends on the speed of both the internal and external motion [@Tse2006; @Shapiro2010; @Kwon2015; @Cavanagh2019].
+The strength of the illusion, the deviation of perceived path of the gabor from its true path, depends on the speed of both the internal and external motion [@Tse2006; @Shapiro2010; @Kwon2015; @Cavanagh2019].
 In our first experiment, participants track their percept with a stylus as the gabor moves back and forth on its trajectory (Fig.
@@
 The accumulation can be reset by introducing a temporal break [@Lisi2015] or by distraction attention [@Nakayama2020], but it may also reset spontaneously once the accumulation has gone on too long or too far.
-Informally, some observers have reported that the Gabor’s path may saturate so that it  remains parallel to, but offset from the true path.
+Informally, some observers have reported that the gabor’s path may saturate so that it  remains parallel to, but offset from the true path.
 Others report that the position resets back toward its physical location whereupon accumulation begins again.
-The ability to record trajectories the perceived position will then allow us to study spontaneous reset outcomes.
+The ability to record trajectories of the perceived position will then allow us to study spontaneous reset outcomes.
 
@@
 For this experiment, 4 participants were recruited from the lab (3 female; ages 21 – 31, mean: 25.75).
-All participants reported to be right handed and had either normal or corrected to normal vision.
+All participants reported being right handed and had either normal or corrected to normal vision.
 Procedures were approved by York’s Human Participants Review Committee, and all participants provided prior, written, informed consent.
@@
 
-In experiment 1, participants were to do online tracking (using their unseen hand / stylus) while viewing a double-drift stimulus.
-In this task, participants were asked to continuously track the position of a gabor that moved forward and backward along the Y-axis (of the horizontal screen) for the duration of 12 s, using a stylus on a drawing tablet.
-Before trial onset, the participant moved the stylus to the middle of the screen ( where the gabor would appear ), and was subsequently asked to fixate a point on the left or right of the screen, and given 1.5 s to do so.
+In this experiment, participants used their unseen hand and stylus to do continuous online tracking of the double-drift stimulus that moved back and forth along the Y-axis of the horizontal screen for 12 _s_.
+Before trial onset, the participant moved the stylus to the middle of the screen, and were given 1.5 _s_ to fixate a point on the left or right of the screen.
+The gabor would then appear at the middle of the screen and start moving.
+Internal motion could be 3, 1, 0, -1 or -3 _cps_.
+Since each moving gabor was shown for 12 _s_, and it moved from one end to another in _2_ s, it completed 5 full and 2 half passes of the path.
 The internal motion of the gabor was inverted at the far and close ends of the physical path, where external motion was also inverted.
-Internal motion could be 3, 1, 0, -1 or -3 cps.
-Since each moving gabor was shown for 12 s, and it moved from one end to another in 2 s, it completed 5 full and 2 half passes of the path.
 
@@
 
-We first segmented the tracking trajectories according to stimulus passes (in between direction changes of the gabor), and removed the noisy tracking during the first and last half second, leaving 1 _s_ trajectories.
+Because there is no set reference point to gauge movement direction against after each direction change, we use instantaneous heading along the trajectory instead.
+We first segmented the tracking trajectories according to stimulus passes (in between direction changes of the gabor), and removed the noisy tracking during the first and last half second, leaving 1 _s_ trajectory segments.
 We then calculated the instantaneous heading (direction, disregarding velocity or distance) of a smooth spline interpolated version of each trajectory segment (using 101 spline interpolated position samples, to get 100 heading samples).
 This heading measure should depend on the illusion strength.
+
+We will compare the illusory direction of the heading against the predictions of a simple vector combination model of perceived direction from Tse and Hsieh [-@Tse2006], Cavanagh and Tse [-@Cavanagh2019] and Heller, Patel, Faustin, Cavanagh, and Tse [-@Heller2021].
+In this model, the perceived direction is a simple vector combination of the external ($V_e$) and internal ($V_i$) motions and a constant $K$.
+That is, the deviation from the physical direction is given by:
+
+[1]  &nbsp;&nbsp;&nbsp;&nbsp;&nbsp;  $\theta=tan^{-1}K\frac{V_i}{V_e}$
+
+If $K$ is 0, then there is no illusion and the perceived direction matches the external motion direction.
+However, if $K$ is 1, then the internal motion has equal contribution with the external motion in determining the perceived direction.
+For comparison with other work, we will find the value of $K$ that best describes the strength of the illusion in each of our two experiments.
 
@@
 It seems that continuous tracking is sensitive to the strength of the illusion, and this is confirmed by a repeated-measures F-test on the average angles, using internal motion as a within-subjects factor ( _F_(4,12)=102.6, _p_<.001, $\eta^2$=0.97 ).
-With FDR-corrected, paired t-tests, we find a difference in heading between 0 _cps_ internal motion and both 1 _cps_ ( _p_=.022) and -1 _cps_ ( _p_=.025) as well as between -3 and -1 _cps_ and 3 and 1 _cps_ (both _p_ <.001).
+With FDR-corrected, paired t-tests, we find a difference in heading between internal motion speeds of 0 _cps_ and both 1 _cps_ ( _p_=.022) and -1 _cps_ ( _p_=.025) as well as between -3 and -1 _cps_ and 3 and 1 _cps_ (both _p_ <.001).
 This makes it clear that continuous tracking does not prevent the illusion from occurring.
-We will use heading angle as a measure of the strength of the illusion.
+We will use the heading angle as a measure of the strength of the illusion.
 
-```{r fig2, echo=FALSE, message=FALSE, fig.width=8, fig.height=5, fig.cap="\\label{fig:fig2}**Figure 2.
-Online Tracking Data** **Left:** Individual trajectories.
+```{r fig2, echo=FALSE, message=FALSE, fig.width=8, fig.height=5, fig.cap="\\label{fig:fig2}**Figure 2:** Online Tracking Data.
+**Left:** Individual trajectories.
 Note that there are only a few obvious discontinuities (highlighted in red).
 In general, the traced paths showed rough linearity.
-**Top right:** Average heading in the middle 1 _s_ of each pass.
-Distance from origin is time, angle is angle.
+**Top right:** Average heading of the middle 1 _s_ of each pass.
+Distance from origin is time, angle is the heading angle.
 Dark blue lines the average heading, in light blue a polar heatmap with the distribution of heading.
 **Bottom right:** The strength of the illusion in degrees angle deviation from the physical path over unity slope (internal / external velocity).
-In red are the average illusions strengths for 0 _cps_, ±1 _cps_ and ±3 _cps_, and in light blue the data for individual participants.
-Gray dashed line: unity.
-Black line: fit from Cavanagh & Tse (2019), 81% slope.
-Purple: fit to the data, 72% slope."}
+In red are the average illusions strengths for 0 _cps_, ±1 _cps_ and ±3 _cps_, and in light blue the data for the 4 individual participants.
+Gray dashed line: rediction of vector combination model of Equation 1 with $K=1$.
+Black line: fit with $K=0.81$ as in Cavanagh & Tse (2019), $K=0.81$.
+Purple: best fit to our data, $K=0.72$."}
 plotBoundedTracking()
@@
 In Figure 2 (bottom right), we compare mean illusion strength to that reported previously [@Cavanagh2019].
-As can be seen the averages from our data here (red circles, purple line) coincide fairly well with the illusion strength predicted by the model fit to previous data (in black).
+As can be seen the averages from our data here (red circles, purple line) coincide fairly well with the illusion strength predicted by the model (Eq.
+1, in black).
 That is; tracking the illusion while observing it does not change the strength of the illusion.
@@
 
-The durations of each pass in the online tracking experiment was 2 seconds but the manual traces (Figure 2, left) mostly showed a linear trajectory without saturation or reset of the types shown in Figure 3 on the left below.
+The illusion does drift continuously away from the physical path forever.
+It can be disrupted by a temporal break [@Lisi2015] or by distracting attention [@Nakayama2020].
+These disruptions may cause the perceived path to stay at a fixed offset, travelling parallel to the physical path, or it may return toward the veridical position before resuming the illusory direction (Fig.
+3).
+These resets may also occur spontaneously once the accumulation has gone on too long or too far.
+The purpose of this experiment is to capture this spontaneous resetting and determine its source.
+
+In the previous experiment with online tracking, the duration of each pass in the online tracking experiment was 2 seconds and the manual traces (Fig.
+2, left) mostly showed a linear trajectory without saturation or reset of the types shown in Figure 3 (left).
 The continuation of the accumulation may be a result of the active tracking – either the extra attention required to track increased the extent of the accumulation or possibly the temporal demands of the tracking caused the participants to average their traces over any breaks.
@@
 
-```{r fig3, echo=F, fig.height=4.55, fig.width=8, message=FALSE, error=FALSE, warning=FALSE, fig.cap="\\label{fig:fig3}**Figure 3: Resets.** **Left:** At some point, the perceived position will stop moving further away from its actual path.
-For some people this takes the form of a ‘jump’ back toward the true position, for others it means the illusory position remains at a fixed offset moving parallel to the real path, or some combination of the two.
+```{r fig3, echo=F, fig.height=4.55, fig.width=8, message=FALSE, error=FALSE, warning=FALSE, fig.cap="\\label{fig:fig3}**Figure 3:** Double-Drift Resets.
+**Left:** At some point, the perceived position may stop moving further away from its actual path.
+For some people this takes the form of a ‘return’ reset, back toward the true position, for others it means the illusory position remains at a fixed offset moving parallel to the real path, or some combination of the two.
 **Top right:** if these resets are due to a spatial limit on the size of the illusory position shift away from the true vertical path, the locations of the resets should have a constant X coordinate.
-**Bottom right:** if they occur instead after some period of time, they should have a constant Y coordinate."}
+**Bottom right:** if they occur instead after some period of time, resets should have a constant Y coordinate."}
 library(magick)
@@
 We were interested, in particular, in whether the limitation of the accumulation would be set by space or time.
-Since both the true and the illusory position appear to be available in the visual system (the true position drives saccades whereas the illusory position drives perception, Lisi & Cavanagh; -@Lisi2015), it is possible that resets occur once the distance between the real and perceived positions exceeds some limit.
-We can speculate that this spatial limit would be related to the positional uncertainty of the Gabor.
+Since both the true and the illusory position appear to be available in the visual system (the true position drives saccades whereas the illusory position drives perception, Lisi & Cavanagh, -@Lisi2015, it is possible that resets occur once the distance between the real and perceived positions exceeds some limit.
+We can speculate that this spatial limit would be related to the positional uncertainty of the gabor.
 When the position information is reliable, the resets would occur with very little illusory deviation, keeping the perceived path close to the true path.
-When the positional uncertainty is high, in the periphery, with the Gabor’s mean luminance matched to the background [@Gurnsey2012; @Cavanagh2019], the spatial offset could be quite large before exceeding the range of positional uncertainty around the gabor’s true position.
+When the positional uncertainty is high, in the periphery, with the gabor’s mean luminance matched to the background [@Gurnsey2012; @Cavanagh2019], the spatial offset could be quite large before exceeding the range of positional uncertainty around the gabor’s true position.
 Under this hypothesis, the timing of spontaneous resets will depend on the spatial offset from the true path which will be directly proportional to the internal speed (for a constant external motion).
@@
 Stimuli were identical to those used in the online tracking experiment, but now always started at the near end of the workspace and moved away from the participants.
-The double drift stimuli would use either 3 or 4 seconds to move 13.5 _cm_ ( corresponding to speeds of ~4.5 and ~3.75 _cm/s_ or ~4.7 or ~3.5 _dva/s_ ) and had an internal drift of 2, 3 or 4 _cps_ .
+The double drift stimuli would use either 3 or 4 seconds to move 13.5 _cm_, corresponding to speeds of ~4.5 and ~3.75 _cm/s_ or ~4.7 or ~3.5 _dva/s_, and had an internal drift of 2, 3 or 4 _cps_ .
 
@@
 
-In experiment 2, participants first watched a double-drift stimulus and then either reproduced the perceived path of the stimulus by retracing it on the tablet, or indicated the initial movement direction by changing the orientation of a line originating in the same position as the doube-drift stimulus.
+In experiment 2, participants first watched a double-drift stimulus and then either reproduced the perceived path of the stimulus by retracing it on the tablet, or indicated the initial movement direction by changing the orientation of a line originating in the same position as the doube-drift stimulus (data not used, but corresponds to illusion strength determined from trajectories).
 More details are given below.
@@
 without the need for real-time tracking.
-The re-tracing task was done in half of 8 blocks.
-In the other half, participants viewed the same stimuli, and then oriented a line to indicate the initial perceived direction of the stimulus motion.
+The re-tracing task was done in half of 8 blocks (the other half of the blocks are not used).
 Block types were alternated, and the order was counterbalanced across participants.
-Each block used the 6 combinations of two external speeds ( corresponding to 3 or 4 _s_ presentation time ), and three internal speeds ( 2, 3 or 4 _cps_ ), 6 times.
-Within each of the two tasks, each of the combinations of internal and external speed was presented 24 times in total.
+Each block used the 6 combinations of two external speeds ( corresponding to 3 or 4 _s_ presentation time ), and three internal speeds ( 2, 3 or 4 _cps_ ), 6 times, for a total of 36 trials per block.
+Each of the combinations of internal and external speed was presented 24 times in total.
 
@@
 Participants could "key in" their response by pressing ENTER, or start over by pressing ESCAPE.
-When indicating only the initial direction of movement, participants would orient a line ( 75 _px_ or ~3 _cm_ ) originating from the starting point of the gabor’s path.
-They could change the orientation by pressing either the LEFT ARROW key to orient the line more clockwise or the RIGHT ARROW key to orient the line more counter clockwise.
-They could "key in" their response by pressing ENTER.
-In both tasks, there was no time-limit for the response.
-
-Re-tracing the perceived paths should allow capturing spontaneous resets of the illusion, and the dependence on the strength of the illusion could be checked with the data from the line-orienting task.
+There was no time-limit for the response.
+Re-tracing the perceived paths should allow capturing spontaneous resets of the illusion.
 
@@
 The individual trajectories reveal some points that are potential spontaneous resets that we try to extract.
-Using a smooth spline interpolated version of the trajectory we detect left-right direction changes.
+Using a smooth spline interpolated version of the trajectory we detect right-left direction changes, based on sign changes.
 Since the initial movement is often jittery, we only take direction changes further than ~0.39 cm ( ~0.23 _dva_ ) away from the starting point.
 The first (if any) of the remaining direction changes, are then taken as reset points.
-By averaging the x and y coordinates of reset points in all 9 participants and 6 stimulus conditions, we obtain 54 average reset points for further analyses and modelling.
+By first averaging the x and y coordinates (in _cm_) of reset points in all 9 participants and 6 stimulus conditions, we obtain 54 average reset points for further analyses and modelling.
+Both the horizontal and vertical distance of the reset point from the origin of the trajectory are converted to centimeters for purposes of model fitting (to keep them in the same unit).
+The vertical component can then be converted to time as 1 _cm_ = 3.375 _s_ in the 4 _s_ condition and 1 _cm_ = 4.5 _s_ in the 3 _s_ condition.
 
@@
 We take a trajectory sample halfway between the starting point and the detected reset point.
-We then use the angle of a straight line drawn through that point and the start point as a measure of illusion strength.
-The average response in the line orientation task is also calculated for each of those 54 combinations of partcipants and stimulus conditions, as a verification for the measure of illusion strength taken from the trajectories.
+We then use the angular difference between the gabor's real trajectory and a straight line drawn through that point and the start of the trajectory as a measure of illusion strength.
+This is comparable to $K \cdot V_i/V_e$ in Eq.
+1, even when expressed in degrees angle.
 
@@
 In Figure 4 (left), we show the trajectories for participant 4 in the condition with 3 _cps_ internal motion and 4 _s_ passes.
-This exemplary participant's re-traced paths show clear 'jump' resets, all of which are detected by our simple algorithm.
+This particular participant's re-traced paths show clear 'return' resets, all of which are detected by our simple algorithm.
 Other participants are not that clear cut.
 
-```{r fig4, echo=FALSE, message=FALSE, fig.width=8, fig.height=3.5, fig.cap="\\label{fig:fig4}**Figure 4.
-Data in Experiment 2.** **Left:*** One exemplary participant's trajectories with very clear 'jump' resets at the purple dots.
-(Purple lines: trajectories up to resets, gray lines: trajectories after resets, blue lines: trajectories without resets.) Other participants' data was not this clear.
-**Middle:** Average participant reset points with standard deviations.
-(Gray arrow: real trajectory, blue line: spatial average, red line: temporal average.) **Right:** Illusion strength as in Figure 3; the acrtangent of the slope (internal velocity over external velocity).
-Gray dahsed line: unity, black line: data/model as in Cavanagh & Tse (2019) at 81% of unity, purple line: best fit here at 42% of unity."}
+```{r fig4, echo=FALSE, message=FALSE, fig.width=8, fig.height=3.5, fig.cap="\\label{fig:fig4}**Figure 4:** Re-tracing Data.
+**Left:** One participant's trajectories with very clear 'return' resets at the purple dots.
+(Purple lines: trajectories up to resets, gray lines: trajectories after resets, blue lines: trajectories that had no resets in the re-tracing.) Other participants' data were not this clear.
+**Middle:** Average reset point locations for all 9 participants in all 6 conditions (54 points) with standard deviations.
+(Gray arrow: real trajectory, blue line: spatial average, red line: temporal average.) **Right:** Illusion strength as in Figure 2 (bottom right); the arctangent of the slope (internal over external velocity).
+Predictions from Eq.
+1: Gray dashed line; $K=1$, black line; $K=0.81$, purple line; best fit here at $K=0.49$."}
 #plotExampleData()
@@
 On average, the algorithm detects resets in ~75% of trials.
-For three participants, only about half the trials show resets (42%, 44% and 58%).
-For the remaining six participants the algorithm detects resets in 72%-99% of trials, with an average of 88%.
-The minimum number of resets in each of the six conditions for any participant is 7 (out of 24 trials), with the average at ~17.
-Since participants experiencing a "hit-the-wall reset" (as opposed to a "jump" reset) might not always display a direction change in their paths, 50% of trials with resets seems reasonable.
+For three participants, only about half the trials showed resets (42%, 44% and 58%).
+For the remaining six participants the algorithm detected resets in 72%-99% of trials, with an average of 88%.
+The minimum number of resets in each of the six conditions for any participant is 7 (out of 24 trials), with an average of ~17.
+Since participants experiencing a "hit-the-wall reset" (Fig.
+3) as opposed to a "jump" reset, might not always display a detectable direction change in their paths, 50% of trials with resets seems reasonable.
 Either type of perceived resets appears valid, and the average reset positions still indicates some limit to the illusion.
-Therefore, we keep all participants and resets (reset points in Figure 4, middle).
+Therefore, we kept all participants and resets in our analysis (reset points in Fig.
+4, middle).
 
 We compare the strength of the illusion as measured with half-way points (see Fig 4, right), with an earlier study [@Cavanagh2019].
-A line through the origin with slope 0.49 predicts illusion strength fairly well in this data, which is weaker than earlier (0.81) or with online tracking (0.72).
+A line through the origin with slope 0.49 ($K$ from Eq.
+1) predicts illusion strength fairly well in this data, which is weaker than earlier ($K$=0.81) or with online tracking ($K$=0.72).
 We can also see that illusion strength measured with re-tracing is more variable.
-While the illusion is weaker, it is nevertheless there and depends on the relationship between internal and external speed, as expected.
+While the illusion is less strong and more variable, it is nevertheless there and depends on the relationship between internal and external speed as in previous studies.
 
-## Single Limits on the Double-Drift Illusion
+## Limits on the Double-Drift Illusion
 
 Now we look at the distribution of reset points, and whether they can be explained by a temporal or spatial limit.
-In Figure 4 (middle), we can see that the reset points vary more along the Y-axis than the X-axis.
-That is, predicting reset points from a single Y coordinate (time-cutoff) would leave larger residuals than predicting reset points from a single X coordinate (or spatial cutoff).
-In fact, the standard deviation of reset point X-coordinates is about 21.9% of the standard deviation of reset point Y-coordinates.
-This already indicates that a spatial limit explains about 4 times more of the variance in reset points than a temporal limit.
+In Figure 4 (middle), we can see that the reset points vary more in when they occur, than how far they are from the real path of the stimulus.
+That is, predicting reset points from a single Y coordinate (time limit) would leave larger residuals than predicting reset points from a single X coordinate (or spatial limit).
+This already indicates that a spatial limit explains more of the variance in reset points than a temporal limit.
 
-```{r fig5, echo=FALSE, message=FALSE, fig.width=8, fig.height=3, fig.cap="\\label{fig:fig5}**Figure 5.
-Reset Models** **Left:** X-coordinates predicted from a single temporal limit.
-**Middle:** Y-coordinates predicted from a single spatial limit.
-**Right:** X- and Y-coordinates predicted from a temporal and a spatial limit."}
+```{r fig5, echo=FALSE, message=FALSE, fig.width=8, fig.height=3, fig.cap="\\label{fig:fig5}**Figure 5:** Reset Models.
+**Left:** Distance of reset from true path predicted from a single temporal limit.
+**Middle:** Time of reset predicted from a single spatial limit.
+**Right:** Both spatial offset and time of resets predicted from a temporal and a spatial limit."}
 #plotResetPoints()
@@
 
-Nevertheless, we test two simple models, based on straighforward trigonometry.
-First, we calculate the slope of a line ($s$) through the origin and a sample halfway to the reset point.
-This estimates illsuion strength as angular deviation from the true, straight path.
-If the perceived path of the gabor is approximated by a straight line (up to a reset), and reset points occur at a set limit, then a temporal limit ($L_y$) would predict X coordinates:
+We test this with three simple models, based on straighforward trigonometry.
+First, we calculate the slope of a line ($s$) through the origin and a sample halfway to the reset point, indicating the strength of the illusion.
+If the perceived path of the gabor is approximated by a straight line (up to a reset), and reset points occur at a set limit, then a temporal limit ($L_y$) would predict spatial offset from the physical path at the time of the reset, $x$:
 
-$x = L_y / s$
+[2]  &nbsp;&nbsp;&nbsp;&nbsp;&nbsp;  $x = L_y / s$
 
-And a spatial limit ($L_x$) would predict Y coordinates:
+And a spatial limit ($L_x$) would predict time until the first reset, $y$:
 
-$y = L_x \cdot s$
+[3]  &nbsp;&nbsp;&nbsp;&nbsp;&nbsp;  $y = L_x \cdot s$
 
-Where $L_y$ and $L_x$ are a temporal (y) and spatial (x) limit that are free parameters, and s corresponds to the calculated slope.
-If we fit these simple models by minimizing the mean square error (as the absolute distance in centimeters between predicted and actual reset points), the optimal parameters indicate a temporal limit of ~1.4 _s_ (AIC=-167.1) and a spatial limit of ~1.8 _cm_ (AIC=-220.7), as illustrated in Figure 6 (left and middle).
-This is close to the mean values of the respective coordinates (Figure 5, middle).
-The model with the spatial limit fits the data better than the model with a temporal limit ( _p_<.001, relative log-likelihood based on AICs).
-This again indicates that a spatial limit explains more of the data.
+While resets relying on both limits would use these two equations:
 
-## Combined Limits
+[4a]  &nbsp;&nbsp;&nbsp;&nbsp;&nbsp;  $x = L_x + L_y / s$
 
-However, it may be that after crossing a spatial limit, it takes some given amount of time before a reset of the illusion is perceived (or reported).
-This would mean that both limits explain separate parts of the data.
-We model this by finding optimal parameter values for both a spatial limit and a temporal limit simultaneously, when minimizing the distance (in _cm_ ) between predicted and actual reset points.
-The equations in the model are:
+[4b]  &nbsp;&nbsp;&nbsp;&nbsp;&nbsp;  $y = L_x \cdot s + L_y$
 
-$x = L_x + L_y / s$
-
-$y = L_x \cdot s + L_y$
-
-Figure 5 (right) shows both the actual reset point coordinates (dots), given the illusion strength as the angle to a sample half-way to the reset point, as well as the model's predictions of reset point coordinates (lines).
-
-The limits that optimally fit the model to the data are a 1.47 _cm_ spatial limit and a 0.33 _s_ temporal limit.
-While the AIC of the model (-226.9) is slightly lower than that of the model with the single spatial limit, it does not explain the data better ( _p_=.45).
-This means that a simple spatial limit accounts for most of the data.
-It is possible that the way we've modeled the additional temporal delay after crossing the spatial limit is not accurate, e.g.
-it may not follow a normal distribution.
-However, regardless of whether or not a temporal limit adds anything, a spatial limit appears to explain the majority of spontaneous double-drift resets.
-
-## Comparison with PSE data
-
-While here we find that a spatial limit is the better explanation for resets, other evidence may point at a temporal limit instead.
-An as of yet unpublished study [@Liu2019] obtained data on the PSE of a double-drift stimulus at the end of trajectories of various lengths traveled at different speeds.
-These PSEs were predicted very accurately with a model that assumes a fixed chance of a reset per unit of time (~0.7% / _s_).
-Interestingly, this temporal constraint fully contradicts the current findings.
-The model predicts that the PSE decreases over time as it becomes more likely that a reset has occurred.
-If the PSE's express the average perceived deviations from the actual trajectory of the double-drift stimulus, then we should see a similar pattern in the averaged trajectories in our re-tracing experiment.
-However, these averaged trajectories reach a plateau within ~1 _s_ after which resets and illusory position shift seem to balance each other.
-This means that we can't replicate the pattern found with the PSE approach.
-There may be several reasons for this.
-Our data may simply be too noisy, or the effects may be highly dependent on participants so that both studies may suffer from sampling bias.
-But primarily, the PSE-based approach relies on all participants to exhibit full jump-resets, whereas the approach presented here is also valid for weaker resets.
-Nevertheless, we can't resolve this contradiction here and leave that for future work.
+Where $L_y$ and $L_x$ are a temporal and spatial limit that are free parameters, and $s$ corresponds to the calculated slope.
+If we fit these simple models by minimizing the mean square error (as the absolute distance in centimeters between predicted and actual reset points), the optimal parameters for the single limit models indicate a temporal limit of ~1.4 _s_ (AIC=-26.2) and a spatial limit of ~1.8 _cm_ (AIC=-35.1), as illustrated in Figure 6 (left and middle).
+This is close to the mean values of the respective coordinates of reset points (Fig.
+5, middle).
+The model with two limits uses a temporal limit of 0.33 _s_ and a spatial limit of 1.47 _cm_ (AIC=-34.5).
+The model with a single spatial limit fits the data better than the model with a temporal limit ( _p_=.01, relative log-likelihood based on AICs) but is not better than the model with combined limits ( _p_=.73).
+This indicates that a spatial limit explains most of the data.
 
@@
 
-We first tested if manual tracking of a double-drift stimulus is susceptible to the illusion as reported by perceptual judgments, pointing [@Lisi2017], and memory saccades [@Massendari2018].
+We first tested whether manual tracking of a double-drift stimulus is susceptible to the illusion as reported by perceptual judgments, pointing [@Lisi2017], and memory saccades [@Massendari2018].
 The results of this online tracking experiment show a close agreement between the angle of the manual tracking and the illusion strength expected based on the model from earlier perceptual measures [@Cavanagh2019].
@@
 
-To measure resets, we ask participants to draw the perceived path of the stimulus after a single pass.
-In this re-tracing experiment the drawn traces reflect the strength of the illusion as well, and in most trials there is a clear reset point, for either a jump reset or a hit-the-wall reset.
-We then set out to test if these spontaneous resets in the double-drift illusion reflect discrepancies between two position signals, occurring at some given size of this discrepancy, or if they simply occur after some time elapses, analogous to other illusions or bi-stable stimuli.
-The positions of these reset points do not vary much with elapsed time, but are better explained by the distance from the true position.
-This has implications for the way position perception relies on motion signals.
+To measure resets, we asked participants to draw the perceived path of the stimulus after a single pass.
+The orientation of the drawn traces reflect the strength of the illusion, and in most trials there was a clear reset point, for either a "return" reset or a "hit-the-wall" reset.
+We then set out to test if these spontaneous resets are triggered by a limiting distance of drift away from the physical location, or by a specific limit of time.
+The results showed that the resets are better explained by the distance from the true position, not by elapsed time.
 
-There is a generative model, based on Bayesian cue integration, that can explain many phenomena with motion induced position shifts.
-This model, formulated by Kwon, Tadin & Knill [-@Kwon2015], is compatible with a good portion of our data.
-The model posits there are three signals contributing to an object's perceived position: it's global motion, it's local motion (or texture motion) and it's position.
-Especially when the directly perceived position is a noisy or weak signal, the two motion signals contribute more, as in the double-drift or curveball illusion.
-In this stimulus, the model predicts that the sensory motion signals bias perceived position away from the true location, but at some point this is balanced by the weak, noisy sensory position signal that bias perceived position back towards the true location [@Kwon2015, Fig 3].
-When this balance is reached the perceived path of the object would be parallel to but somewhat displaced from the true object path.
-This pattern of perceived paths is exhibited by some of our participants, and we have called these "hit-the-wall resets" here.
+Recently, Kwon, Tadin, and Knill [-@Kwon2015] proposed a Bayesian object tracking model to explain the double-drift illusion.
+In their model, when the position signals are weak, the motion signals bias perceived position away from the true location.
+In their case, this deviation saturated after about 200 _ms_ and the perceived path then continued parallel to the physical path, as in the “hit the wall” resets we described above [@Kwon2015, Fig.
+3].
+In contrast to their model, our participants in Experiment 2 showed quite variable timing for the reset point where the accumulation of offset ended.
+The mean time of these reset points was not 200 _ms_ but 1.5 _s_ and, more importantly, ranged from 0.5 _s_ to almost 4 _s_ (see Fig.
+4).
+Moreover, 4/9 of our participants regularly reported “jump resets” that the Kwon et al.
+[-@Kwon2015] model cannot predict.
+The jump resets suggest that both the physical and perceived locations are available to the visual system and that when the deviation exceeds some limit, the perceived location returns toward the physical location.
+At that point, the build-up of the illusion may start again.
 
-Some of our participants seem to perceive "jump resets" though, and the model as published, does not predict any such resets.
-This could be added by including a causal inference step [e.g.
-@Wei2011].
-For the subset of participants that experience jump resets, there may be a process that compares the two position signals (perceived and retinal) and when these exceed some threshold, decides that the two position signals do not share a common source.
-At that point the perceived position may be discarded, or reset to the retinal position, and the build-up of the illusion may start again.
-For the remaining participants, either the discrepancy between retinal and perceived location never exceeds their threshold, or their percept is indeed bounded to the maximum difference they allow between the two position signals.
+Recently, it has been suggested that resets can be triggered by a distraction of attention [@Nakayama2020].
+The participants in this experiment also retraced the perceived path and reported "jump" resets.
+Nakayama and Holcombe also suggest that planned eye movements can trigger a “jump” reset back to the physical path.
+In our experiment, we did not control for eye-movements, so we can't directly investigate whether eye movements accompanied the jump resets that participants reported.
+However, "jump" resets occured at a fairly consistent location, which suggests that they are triggered by the stimulus and not by attentional distraction or eye movements.
 
-Recently, it has been suggested that resets can be triggered by planned eye-movements to the stimulus [@Nakayama2020], particularly "jump" resets.
-In this experiment, we did not control for eye-movements with an eye-tracker, so we can't directly investigate this idea.
-Given a pre-motor theory of attention like mechanism, when spontaneously planned eye-movements trigger (jump) resets, such resets could occur at any point during the stimulus pass.
-However, it seems that for the participants that display "jump" resets in their re-tracing data, these happen at a fairly consistent location, which suggests that jump resets are instead usually triggered by the stimulus and not by attention.
-This should however be tested with eye-tracking, so that this will remain a speculation here.
-
-We find that online, manual tracking of the double-drift stimulus is sensitive to illusion strength, but doesn’t display clear resets.
-We do find spontaneous resets in delayed, single-pass re-tracing.
-The illusion appears to reset, or max out, predominantly at a spatial distance from the true stimulus location, suggesting that resets resolve a source attribution conflict between the true (retinal) and illusory (perceived) position of the stimulus.
-This confirms the notion that we gauge the position of objects in the world by combining both retinal position signals as well as accumulated position shifts derived from motion signals.
+We find that manual tracking of the double-drift stimulus during its motion did show the expected illusion.
+Few if any resets were detected during the online tracking, either because of the additional attentional requirements or because of the short duration (2 s), Participants did report spontaneous resets in delayed, single-pass re-tracing of Experiment 2.
+The resets were best explained by a limiting spatial offset from the true stimulus location.
+This finding suggested that resets resolve a conflict between the physical and illusory position of the stimulus and that both are available to the visual system.
+This confirms the notion that we gauge the position of objects in the world by combining both retinal position signals and when the retinal information has high uncertainty, motion-based predictions.
 

### 2021-03-04 R/models.R
very conservative number of independent observations

@@ -424,9 +424,13 @@ compareAllModels <- function(verbosity=0) {
     print(MSEs)
     
     # the AIC parameters are the same for both models, except the MSEs
+    # Gunnar says number of independent observations can be 
     # N <- (9*6)-1
     N <- length(df$boundX_mean)
     # this is then used for C:
+    
+    # Patrick says N should be number of participants:
+    N <- 9
     
     AICs <- AIC(MSE=MSEs, k=c(1,1,2), N=N)
     cat('\nmodel AICs:\n')

### 2021-03-04 R/retrace_onePass_V4.R
changed exp 2 data figure: beter y-axis label, added a legend to the illusion strength panel

@@ -920,7 +920,7 @@ plotData <- function(target='inline') {
   avg_idxE4 <- which(avg_df$externalspeed == 4)
   
   plot(-1000,-1000,
-       main='re-tracing halfway points',xlab='',ylab='',
+       main='illusion strength',xlab='',ylab='',
        xlim=c(0,3*(pi/8)),ylim=c(0,45),
        bty='n',ax=F)
   
@@ -948,6 +948,20 @@ plotData <- function(target='inline') {
   
   points(avg_xcoords[avg_idxE3], avg_df$initialdirection_mean[avg_idxE3], col=colors$blue$s, pch=1)
   points(avg_xcoords[avg_idxE4], avg_df$initialdirection_mean[avg_idxE4], col=colors$yorkred$s, pch=1)
+  
+  legend(x=0, y=45, 
+         legend = c('participant (3 s)', 'participant (4 s)', 'group averages', 'K=1', 'K=0.81', 'K=0.49'), 
+         pch=c(16,16,1,NA,NA,NA), col=c(colors$blue$t, colors$yorkred$t, 'black', 'gray', 'black', colors$purple$s), 
+         lty = c(0,0,0,2,1,1),
+         bty='n')
+  
+  #  legend(15,10,c('reset distance','reset time'),col=c(colors[['blue']]$s, colors[['yorkred']]$s), lty=c(1,1), title='reset coordinate:',bty='n')
+  
+  # legend(x=0, y=45, 
+  #        legend=c('Cavanagh & Tse (2019)', 'participants', 'average'), 
+  #        col=c('black', colors$lightblue$s, colors$yorkred$s), 
+  #        pch=c(NA,1,1), lty=c(1,0,0), 
+  #        bty='n', cex=1)
   
   axis(side=1,at=seq(0,3*(pi/8),pi/8),labels=c('0',expression(pi/8),expression(pi/4),expression(3*pi/8)))
   axis(side=2,at=seq(0,45,15))
@@ -1611,7 +1625,6 @@ plotModels <- function(target='inline') {
   
   
   par <- fitResetModelSeq(slopes=df$slope, X=df$boundX_mean*13.5, Y=df$boundY_mean*13.5)
-  print(par)
   
   model_angles <- seq(10,40)
   model_directions <- ((90-model_angles)/180)*pi

### 2021-03-11 Manuscript.Rmd
small corrections

@@
 However, if $K$ is 1, then the internal motion has equal contribution with the external motion in determining the perceived direction.
-For comparison with other work, we will find the value of $K$ that best describes the strength of the illusion in each of our two experiments.
+For comparison with other work, we will find the value of $K$ that best describes the strength of the illusion in both of our two experiments.
 
@@
 Dark blue lines the average heading, in light blue a polar heatmap with the distribution of heading.
-**Bottom right:** The strength of the illusion in degrees angle deviation from the physical path over unity slope (internal / external velocity).
+**Bottom right:** The strength of the illusion in degrees angle deviation from the physical path.
 In red are the average illusions strengths for 0 _cps_, ±1 _cps_ and ±3 _cps_, and in light blue the data for the 4 individual participants.
-Gray dashed line: rediction of vector combination model of Equation 1 with $K=1$.
+Gray dashed line: prediction of the vector combination model of Equation 1 with $K=1$.
 Black line: fit with $K=0.81$ as in Cavanagh & Tse (2019), $K=0.81$.
@@
 
-The illusion does drift continuously away from the physical path forever.
+The illusion cannot drift continuously away from the physical path forever.
 It can be disrupted by a temporal break [@Lisi2015] or by distracting attention [@Nakayama2020].
@@
 For some people this takes the form of a ‘return’ reset, back toward the true position, for others it means the illusory position remains at a fixed offset moving parallel to the real path, or some combination of the two.
-**Top right:** if these resets are due to a spatial limit on the size of the illusory position shift away from the true vertical path, the locations of the resets should have a constant X coordinate.
-**Bottom right:** if they occur instead after some period of time, resets should have a constant Y coordinate."}
+**Top right:** if these resets are due to a spatial limit on the size of the illusory position shift away from the true vertical path, the locations of the resets should have a constant X coordinatee, the spatial offset from the vertical path.
+**Bottom right:** if they occur instead after some period of time, resets should have a constant Y coordinate, the time since the gabor’s motion began."}
 library(magick)
@@
 There was no time-limit for the response.
-Re-tracing the perceived paths should allow capturing spontaneous resets of the illusion.
+Re-tracing the perceived paths should allow capturing spontaneous resets of the illusion, if there are any.
 
@@
 By first averaging the x and y coordinates (in _cm_) of reset points in all 9 participants and 6 stimulus conditions, we obtain 54 average reset points for further analyses and modelling.
-Both the horizontal and vertical distance of the reset point from the origin of the trajectory are converted to centimeters for purposes of model fitting (to keep them in the same unit).
+Both the horizontal and vertical (straight ahead) distance of the reset point from the origin of the trajectory are converted to centimeters, and trajectories are scaled in both dimensions so that vertical distance from the origin is 13.5 cm.
 The vertical component can then be converted to time as 1 _cm_ = 3.375 _s_ in the 4 _s_ condition and 1 _cm_ = 4.5 _s_ in the 3 _s_ condition.
@@
 We then use the angular difference between the gabor's real trajectory and a straight line drawn through that point and the start of the trajectory as a measure of illusion strength.
-This is comparable to $K \cdot V_i/V_e$ in Eq.
-1, even when expressed in degrees angle.
 
@@
 **Middle:** Average reset point locations for all 9 participants in all 6 conditions (54 points) with standard deviations.
-(Gray arrow: real trajectory, blue line: spatial average, red line: temporal average.) **Right:** Illusion strength as in Figure 2 (bottom right); the arctangent of the slope (internal over external velocity).
+(Gray arrow: real trajectory, blue line: spatial average, red line: temporal average.) **Right:** Illusion strength as in Figure 2 (bottom right); plotted against predicted strength from Eq.1: the arctangent of the ratio of internal to external velocities.
+Blue dots from 3 s passes, red from 4 s passes, open circles are group averages.
 Predictions from Eq.
-1: Gray dashed line; $K=1$, black line; $K=0.81$, purple line; best fit here at $K=0.49$."}
+1: Gray dashed line for $K=1$, black line; $K=0.81$, purple line; best fit here at $K=0.51$."}
 #plotExampleData()
@@
 We compare the strength of the illusion as measured with half-way points (see Fig 4, right), with an earlier study [@Cavanagh2019].
-A line through the origin with slope 0.49 ($K$ from Eq.
-1) predicts illusion strength fairly well in this data, which is weaker than earlier ($K$=0.81) or with online tracking ($K$=0.72).
+A line through the origin with slope 0.51 ($K$ from Eq.
+1) predicts illusion strength fairly well in these data, which is weaker than the $K$=0.81 found by Cavanagh & Tse [-@Cavanagh2019] and lower than the $K$=0.72 found in Experiment 1 with online tracking.
 We can also see that illusion strength measured with re-tracing is more variable.
@@
 We test this with three simple models, based on straighforward trigonometry.
-First, we calculate the slope of a line ($s$) through the origin and a sample halfway to the reset point, indicating the strength of the illusion.
+First, we calculate the slope of a line ($s$, time/illusory offset) through the origin and a sample halfway to the reset point, indicating the strength of the illusion.
 If the perceived path of the gabor is approximated by a straight line (up to a reset), and reset points occur at a set limit, then a temporal limit ($L_y$) would predict spatial offset from the physical path at the time of the reset, $x$:
@@
 
-Where $L_y$ and $L_x$ are a temporal and spatial limit that are free parameters, and $s$ corresponds to the calculated slope.
-If we fit these simple models by minimizing the mean square error (as the absolute distance in centimeters between predicted and actual reset points), the optimal parameters for the single limit models indicate a temporal limit of ~1.4 _s_ (AIC=-26.2) and a spatial limit of ~1.8 _cm_ (AIC=-35.1), as illustrated in Figure 6 (left and middle).
-This is close to the mean values of the respective coordinates of reset points (Fig.
+Where $L_y$ and $L_x$ are a temporal and spatial limit that are free parameters, and $s$ corresponds to the calculated slope from the start of the trajectory to the halfway point.
+If we fit these simple models by minimizing the mean square error (errors are the absolute distance in centimeters between predicted and actual reset points), the optimal parameters for the single limit models indicate a temporal limit of ~1.82 _s_ (Eq.
+2; AIC=-20.3) and a spatial limit of ~2.46 _cm_ (Eq.
+3; AIC=-28.4), as illustrated in Figure 6 (left and middle).
+This is close to, but not the same as, the mean values of the respective coordinates of reset points (Fig.
 5, middle).
-The model with two limits uses a temporal limit of 0.33 _s_ and a spatial limit of 1.47 _cm_ (AIC=-34.5).
-The model with a single spatial limit fits the data better than the model with a temporal limit ( _p_=.01, relative log-likelihood based on AICs) but is not better than the model with combined limits ( _p_=.73).
+The model with two limits uses a temporal limit of 0.55 _s_ and a spatial limit of 1.89 _cm_ (Eq.
+4a,b; AIC=-28.8).
+The model with a single spatial limit fits the data better than the model with a temporal limit ( _p_=.016, relative log-likelihood based on AICs) but is not better than the model with combined limits ( _p_=.87).
 This indicates that a spatial limit explains most of the data.
@@
 In contrast to their model, our participants in Experiment 2 showed quite variable timing for the reset point where the accumulation of offset ended.
-The mean time of these reset points was not 200 _ms_ but 1.5 _s_ and, more importantly, ranged from 0.5 _s_ to almost 4 _s_ (see Fig.
+The resets were more stable in terms of spatial offset (Fig.
+4, middle) but much more variable in terms of time from the start of the gabor's motion.
+In any case, the temporal mean of these reset points was not 200 _ms_ but 1.5 _s_ and ranged from 0.5 _s_ to almost 4 _s_ (see Fig.
 4).
-Moreover, 4/9 of our participants regularly reported “jump resets” that the Kwon et al.
-[-@Kwon2015] model cannot predict.
-The jump resets suggest that both the physical and perceived locations are available to the visual system and that when the deviation exceeds some limit, the perceived location returns toward the physical location.
+Moreover, 4/9 of our participants regularly reported “return” resets that the Kwon et al.
+[-@Kwon2015] model does not predict.
+The “return” resets suggest that both the physical and perceived locations are available to the visual system and that when the deviation exceeds some limit, the perceived location returns toward the physical location.
 At that point, the build-up of the illusion may start again.
@@
 Recently, it has been suggested that resets can be triggered by a distraction of attention [@Nakayama2020].
-The participants in this experiment also retraced the perceived path and reported "jump" resets.
-Nakayama and Holcombe also suggest that planned eye movements can trigger a “jump” reset back to the physical path.
-In our experiment, we did not control for eye-movements, so we can't directly investigate whether eye movements accompanied the jump resets that participants reported.
-However, "jump" resets occured at a fairly consistent location, which suggests that they are triggered by the stimulus and not by attentional distraction or eye movements.
+The participants in this experiment also retraced the perceived path and reported “return” resets.
+Nakayama and Holcombe also suggest that planned eye movements can trigger a “return” reset back to the physical path.
+In our experiment, we did not control for eye-movements, so we can't directly investigate whether eye movements accompanied the “return” resets that participants reported.
+However, “return” resets occurred at a fairly consistent location for each participant, which suggests that they are triggered by the stimulus dynamics and not by attentional distraction or eye movements.
 

### 2021-03-11 Manuscript.Rmd

@@
 source('R/time.R')
+
+# we download the data:
+OSFdata()
+# this only downloads the data for the paper (well...)
+
+# we pre-process the data:
+getBoundedTrackingStandardizedSegments()
+getSegmentDirections()
+preProcessOnePass_V4()
 ```
@@
 **Right:** participants track the perceived trajectory as the gabor moves back and forth along its path."}
+library('rsvg')
 library(magick)

### 2021-03-12 R/data.R
download data from OSF with `osfr` package

@@ -1,145 +1,198 @@
 # check / download data -----
 
+library('osfr')
+
+# this function will download and unzip the data from OSF:
+
+OSFdata <- function(overwrite=FALSE, removezip=FALSE, extradata=FALSE) {
+  
+  repo <- osfr::osf_retrieve_node('72ndu')
+  
+  dirs <- osfr::osf_ls_files(repo, type='folder')
+  
+  idx <- which(dirs$name == 'data')
+  
+  datadir <- osfr::osf_ls_files(dirs[idx,])
+  
+  zipfiles <- c('bounded_tracking.zip','onepass_V4.zip')
+  if (extradata) {
+    zipciles <- c(zipfiles, 'infinite_tracking.zip', 'pilots2.zip', 'CVRdemo.zip')
+  }
+  
+  if (overwrite) {
+    conflicts <- 'overwrite'
+  } else {
+    conflicts <- 'skip'
+  }
+  
+  for (zipfile in zipfiles) {
+    
+    idx <- which(datadir$name == zipfile)
+    
+    if (overwrite | !file.exists(sprintf('data/%s',zipfile))) {
+      osfr::osf_download(datadir[idx,], path='data/', conflicts=conflicts)
+    }
+    
+  }
+  
+  for (zipfile in zipfiles) {
+    
+    fullzipfile <- sprintf('data/%s',zipfile)
+    unzip(fullzipfile, exdir='data')
+    if (removezip) {
+      file.remove(fullzipfile)
+    }
+    
+  }
+  
+}
+
+
+# code graveyard -----
+
+# the stuff below is old-fashioned
+# superseded by osfr functionality
+
 #source('R/tracking_reset_pause.R')
 
-downloadRawData <- function(overwrite=TRUE, experiments=c('bounded_tracking', 'bounded_tracking_V2', 'onePass_V4')) {
-  
-  
-  for (experiment in experiments) {
-    
-    if (experiment == 'bounded_tracking') {
-      
-      folder = 'data/bounded_tracking'
-      
-      if (!dir.exists(folder)) {
-        dir.create(folder)
-      }
-      
-      resources <- c( 
-        
-        'bounded_tracking_p01.csv' = 'https://osf.io/xrbp7/download',
-        'bounded_tracking_p03.csv' = 'https://osf.io/2p3nc/download',
-        'bounded_tracking_p05.csv' = 'https://osf.io/trzdx/download',
[…]

### 2021-06-22 R/retrace_onePass_V4.R
updates to retrace study code... needs to be cleaned, and integrated

@@ -12,11 +12,24 @@ preProcessOnePass_V4 <- function(participants = c(2,3,4,5,6,8,9,10,11), overwrit
     return()
   }
   
+  angle_deg <- seq(5,50,0.5)
+  angle_rad <- (angle_deg / 180) * pi
+  cos.a <- cos(angle_rad)
+  sin.a <- sin(angle_rad)
+  #slope <- sin.a / cos.a
+  slope <- cos.a / sin.a
+  
+  modeldf <- data.frame(angle_deg, angle_rad, cos.a, sin.a, slope)
+  
+  Lx_MSE <- c()
+  Ly_MSE <- c()
+  LxLy_MSE <- c()
   
   internalMovements <- c(2,3,4)
   externalMovements <- rev(c(.125, .167))
   
-  for (task in c('arrow','re-trace')) {
+  #for (task in c('arrow','re-trace')) {
+  for (task in c('re-trace')) {
     
     # we will generate data files for both tasks, with these columns:
     # except that there is no bound in the arrow task
@@ -30,9 +43,24 @@ preProcessOnePass_V4 <- function(participants = c(2,3,4,5,6,8,9,10,11), overwrit
     illusionstrength <- c()
     boundX <-c()
     boundY <-c()
+    boundXraw <- c()
+    boundYraw <- c()
     
+    # ***************************
+    # 
+    
+    pdf(file='doc/individual_participants_trials.pdf',width=12,height=8)
+    
+    #layout(matrix(c(1,2,3,4,5,6),byrow=TRUE,ncol=3,nrow=2))
     
     for (participant.idx in c(1:length(participants))) {
+      
+      plot(-1000,-1000,
+           main=sprintf('participant %d',participants[participant.idx]),xlab='',ylab='',
+           xlim=c(-4.5,10),ylim=c(-.5,14),
+           bty='n',ax=F)
+      
+      lines(x=c(0,0),y=c(0,13.5),col='blue')
       
       ppno <- participants[participant.idx]
       
@@ -76,6 +104,8 @@ preProcessOnePass_V4 <- function(participants = c(2,3,4,5,6,8,9,10,11), overwrit
               illusionstrength <- c(illusionstrength, 90 - percept)
               boundX <- c(boundX, NA)
               boundY <- c(boundY, NA)
+              boundXraw <- c(boundXraw, NA)
+              boundYraw <- c(boundYraw, NA)
               
             } else {
               
[…]

### 2021-06-22 R/common.R
write a 2D density function (Gaussian kernel)

@@ -530,4 +530,79 @@ polarHeatMap <- function(x,y,z,mincol=c(0.94,0.98,0.99),maxcol=c(0.06,0.82,0.88)
   
 }
 
+density2D <- function(x, y, bw=1, weights=NULL, n=100, from=NULL, to=NULL, cut=3, na.rm=FALSE) {
+  
+  # have weights:
+  if (!is.null(weights)) {
+    if (length(x) != length(weights)) {
+      cat('denisty2d error: weights need to be the same length as x and y\n')
+      return()
+    }
+    weights <- weights / sum(weights)
+  } else {
+    weights <- rep(1/length(x), length(x))
+  }
 
+  # remove NA values, if specified:
+  if (na.rm) {
+    idx <- intersect( which(!is.na(x)), which(!is.na(y)) )
+    x <- x[idx]
+    y <- y[idx]
+    weights <- weights[idx]
+  }
+  
+  # make sure we have bandwidth in the correct shape:
+  if (length(bw) == 1) {
+    bw <- c(bw,bw)
+  }
+  
+  # set up the coordinates for the grid:
+  if (is.null(from) | is.null(to)) {
+    from <- c(min(x)-(cut*bw[1]),min(y)-(cut*bw[2]))
+    to   <- c(max(x)+(cut*bw[1]),max(y)+(cut*bw[2]))
+  }
+  if (length(n) == 1) {
+    n = c(n,n)
+  }
+  
+  # make the actual grid:
+  X <- seq(from[1], to[1], length.out=n[1])
+  Y <- seq(from[2], to[2], length.out=n[2])
+  grid <- expand.grid('x'=X,'y'=Y,'z'=NA)
+  
+  
+  # get density at the grid points:
+  for (idx in c(1:dim(grid)[1])) {
+    grid$z[idx] <- sum( weights * ( 1/(2*pi*bw[1]+bw[2]) * exp( -1 * ((x-grid$x[idx])^2/(2*(bw[1]^2)) + (y-grid$y[idx])^2/(2*(bw[2]^2)))) ) )
+  }
+  
+  return( list('x'=X,
+               'y'=Y,
+               'z'=matrix(grid$z, nrow=length(X)) ) )
+  
+}
+
+
+
+## Transparent colors
+## Mark Gardener 2015
+## www.dataanalytics.org.uk
+
+t_col <- function(color, percent = 50, name = NULL) {
+  #      color = color name
+  #    percent = % transparency
+  #       name = an optional name for the color
+  
[…]

### 2021-06-22 R/data.R
fixed typo

@@ -16,7 +16,7 @@ OSFdata <- function(overwrite=FALSE, removezip=FALSE, extradata=FALSE) {
   
   zipfiles <- c('bounded_tracking.zip','onepass_V4.zip')
   if (extradata) {
-    zipciles <- c(zipfiles, 'infinite_tracking.zip', 'pilots2.zip', 'CVRdemo.zip')
+    zipfiles <- c(zipfiles, 'infinite_tracking.zip', 'pilots2.zip', 'CVRdemo.zip')
   }
   
   if (overwrite) {

### 2021-06-22 R/individual_residual.R
stuff on residuals, no good spot for this, probably won't use it

@@ -1,0 +1,215 @@
+
+source('R/models.R')
+source('R/retrace_onePass_V4.R')
+
+# *****************
+# plot residuals for individually fit limits
+
+individualResidualsPlot <- function() {
+  
+  df <- prepData()
+  
+  df <- addPredictedCoordinates(df)
+  
+  par(mar=c(3.5,3.5,3.5,0.5))
+  
+  layout(matrix(c(1:9),ncol=3,nrow=3,byrow=TRUE))
+  
+  for (coordinate in c('X','Y','2D')) {
+    
+    for (model in c('Lx','Ly','Lx.Ly')) {
+      
+      if (coordinate %in% c('X','Y')) {
+        
+        # predicted over actual
+        Ydots <- df[,sprintf('%s_%s',model,coordinate)]
+        Xdots <- df[,coordinate]
+        
+        xlim <- c(0,13.5)
+        ylim <- c(0,13.5)
+        
+        xlab <- 'actual'
+        ylab <- 'predicted'
+        
+      } else {
+        
+        Xdots <- df[,sprintf('%s_X',model)] - df[,'X']
+        Ydots <- df[,sprintf('%s_Y',model)] - df[,'Y']
+        
+        xlim <- c(-13.5,13.5)
+        ylim <- c(-13.5,13.5)
+        
+        xlab <- 'X error'
+        ylab <- 'Y error'
+        
+      }
+      
+      plot(Xdots,Ydots, 
+           main=sprintf('%s (%s)',model,coordinate), xlab='', ylab='',
+           xlim=xlim, ylim=ylim, 
+           col='#e516362f',
+           asp=1,bty='n',ax=F)
+      
+      title(xlab=xlab, line=2.5)
+      title(ylab=ylab, line=2.5)
+      
+      axis(side=1, at=seq(min(xlim),max(xlim),4.5))
+      axis(side=2, at=seq(min(ylim),max(ylim),4.5))
+      
+    }
+    
+  }
+
+}
+
+
+prepData <- function() {
+  
+  df <- read.csv('data/onePass_V4/onePass_V4_re-trace.csv', stringsAsFactors = FALSE)
+  df <- df[which(is.finite(df$illusionstrength)),]
+  
[…]

### 2021-06-22 R/tryout.R
file with functions I'm trying out... might have to be removed in the end: not using it

@@ -1,0 +1,194 @@
+
+
+randomProcess <- function(repeats=10000, 
+                          timepoints=5000, 
+                          lambda=0.005,
+                          eventno=2) {
+  
+  # vector of times until `eventno` events have occurred
+  times <- c()
+  
+  # we repeat the random simulation `repeat` times
+  for (r in c(1:repeats)) {
+    
+    # we get a bunch of random numbers:
+    rp <- runif(timepoints) # timepoint random uniform numbers between 0 and 1
+    
+    # we check if the random numbers are lower than `lamda` at least `eventno` times
+    if (length(which(rp <= lambda)) >= eventno) {
+      # in this case we store that time that this happens:
+      times <- c(times, which(rp <= lambda)[eventno])
+    }
+    
+  }
+  
+  # plot histogram of the times:
+  hist(times, breaks=40)
+  
+}
+
+
+
+XlikelihoodTry <- function() {
+  
+  source('R/models.R')
+  source('R/retrace_onePass_V4.R')
+  
+  df <- getData()
+  
+  singleFits <- fitSingleLimitModels(df)
+  
+  Xpred <- XlimResets(par=singleFits$Xlim$par, data=df)
+  
+  
+  # x <- seq(pi/4, 5 * pi, length.out = 100)
+  # y <- seq(pi/4, 5 * pi, length.out = 100)
+  # r <- as.vector(sqrt(outer(x^2, y^2, "+")))
+  # grid <- expand.grid(x=x, y=y)
+  # grid$z <- cos(r^2) * exp(-r/(pi^3))
+
+  plot(-1000,-1000,main='X-limit', xlab='', ylab='',
+       xlim=c(-2,6),ylim=c(-.5,14),
+       bty='n',ax=FALSE)
+  
+  title(xlab='x coordinate', line=2.5)
+  title(ylab='y coordinate', line=2.5)
+  
+  x = seq(-2,6,0.1)
+  y = seq(0,13.5,0.1)
+  grid <- expand.grid(x=x, y=y)
+  X.sigma <- sd(df$X)
+  X.mu <- mean(df$X)
[…]

### 2021-06-22 Manuscript.Rmd
working on manuscript

--- 2021-03-11 Manuscript.Rmd
@@
 Now we look at the distribution of reset points, and whether they can be explained by a temporal or spatial limit.
-In Figure 4 (middle), we can see that the reset points vary more in when they occur, than how far they are from the real path of the stimulus.
-That is, predicting reset points from a single Y coordinate (time limit) would leave larger residuals than predicting reset points from a single X coordinate (or spatial limit).
-This already indicates that a spatial limit explains more of the variance in reset points than a temporal limit.
+In Figure 4 (middle), we can see that the reset points vary more along the y-axis, than along the x-axis.
+That is, it looks like predicting reset points from a single X coordinate (or spatial limit) than predicing reset points from a given time.
 
-```{r fig5, echo=FALSE, message=FALSE, fig.width=8, fig.height=3, fig.cap="\\label{fig:fig5}**Figure 5:** Reset Models.
-**Left:** Distance of reset from true path predicted from a single temporal limit.
-**Middle:** Time of reset predicted from a single spatial limit.
-**Right:** Both spatial offset and time of resets predicted from a temporal and a spatial limit."}
+```{r fig5, echo=FALSE, message=FALSE, fig.width=7, fig.height=5.5, fig.cap="\\label{fig:fig5}**Figure 5:** Reset Models.
+Model limit shown here are given by weighted bootstrapping from all participants, lines are median parameter values, shaded areas the 95% confidence interval of the limits.
+Gray dots are the reset points from all participants.
+**Left:** Distance of reset from true path as a single spatial limit.
+**Right:** Reset coordinates predicted from a single temporal limit (at two speeds)."}
 #plotResetPoints()
-plotModels()
+#plotModels()
+plotBootstrappedFits()
 ```
 
-We test this with three simple models, based on straighforward trigonometry.
-First, we calculate the slope of a line ($s$, time/illusory offset) through the origin and a sample halfway to the reset point, indicating the strength of the illusion.
-If the perceived path of the gabor is approximated by a straight line (up to a reset), and reset points occur at a set limit, then a temporal limit ($L_y$) would predict spatial offset from the physical path at the time of the reset, $x$:
+We test this with a series of simple models, based on straighforward trigonometry.
+First, we calculate the angle of a line through the origin and a sample on the trajectory halfway to the reset point (with coordinates $X$ and $Y$), as the strength of the illusion ($a = atan2(X,Y)$ expressing the angular deviation the line $x=0$ which is the real path of the gabor).
+This can then be used to calculate a slope ($s = \frac{sin(a)}{cos(a)}$).
 
-[2]  &nbsp;&nbsp;&nbsp;&nbsp;&nbsp;  $x = L_y / s$
+If the perceived path of the gabor is approximated by a straight line (up to a reset), and reset points occur at a set limit, then a spatial limit ($L_x$) would predict the y-coordinate of the first reset, $y$:
 
-And a spatial limit ($L_x$) would predict time until the first reset, $y$:
+[2]  &nbsp;&nbsp;&nbsp;&nbsp;&nbsp;  $y = L_x \cdot s$
 
-[3]  &nbsp;&nbsp;&nbsp;&nbsp;&nbsp;  $y = L_x \cdot s$
+And a temporal limit ($L_t$) would predict both coordinates of the reset, $x$ and $y$:
 
-While resets relying on both limits would use these two equations:
+[3a]  &nbsp;&nbsp;&nbsp;&nbsp;&nbsp;  $x = L_t \cdot v \cdot cos(a)$
+[3b]  &nbsp;&nbsp;&nbsp;&nbsp;&nbsp;  $y = L_t \cdot v \cdot sin(a)$
 
-[4a]  &nbsp;&nbsp;&nbsp;&nbsp;&nbsp;  $x = L_x + L_y / s$
+Where $L_t$ and $L_x$ are a temporal and spatial limit (in seconds and cm respectively) that are free parameters, $a$ and $s$ corresponds to the calculated angle and slope expressing the strength of the illusion, and speed ($v$ in _cm/s_) is given by the stimulus.
+If we fit the two simple models for each participant separately, by minimizing the mean square error (error as absolute distance in centimeters between predicted and actual reset points), most participants are best fit with a spatial limit (MSE: 6/9, AIC: 7/9) and the remaining participants are best fit with a temporal limit.
+We can also combine limits to see if they are complementary.
+Here, we see that for 4/9 participants (same for MSE and AIC) the best fit boils down to a spatial limit as the other limit is set to 0.
+Of the remaining participants 4 are best fit with a combined spatial and temporal limit (4/9) and one participant is best fit combining a vertical offset with a temporal limit.
+This indicates that in this data, a spatial limit explains the reset points best.
+However even the spatial limit does not explain the data very well.
 
-[4b]  &nbsp;&nbsp;&nbsp;&nbsp;&nbsp;  $y = L_x \cdot s + L_y$
-
-Where $L_y$ and $L_x$ are a temporal and spatial limit that are free parameters, and $s$ corresponds to the calculated slope from the start of the trajectory to the halfway point.
-If we fit these simple models by minimizing the mean square error (errors are the absolute distance in centimeters between predicted and actual reset points), the optimal parameters for the single limit models indicate a temporal limit of ~1.82 _s_ (Eq.
-2; AIC=-20.3) and a spatial limit of ~2.46 _cm_ (Eq.
-3; AIC=-28.4), as illustrated in Figure 6 (left and middle).
-This is close to, but not the same as, the mean values of the respective coordinates of reset points (Fig.
-5, middle).
-The model with two limits uses a temporal limit of 0.55 _s_ and a spatial limit of 1.89 _cm_ (Eq.
-4a,b; AIC=-28.8).
-The model with a single spatial limit fits the data better than the model with a temporal limit ( _p_=.016, relative log-likelihood based on AICs) but is not better than the model with combined limits ( _p_=.87).
-This indicates that a spatial limit explains most of the data.
+To obtain population level values for the spatial and temporal limits, we resample reset points from each participant, using the smallest number of detected reset points that fall in the first quadrant, in the group (44), to ensure that each participant contributes the same number of reset points.
+Both models based on a spatial limit (eq.
+2) and on a temporal limit (eq.
+3) as well as a model combining these two were fit 1000 times.
+In the models with two limits there is a strong negative correlation between fitted spatial and temporal limits (r(998)=-0.985, p>.001) indicating parameter redundancy in the combined limit models.
+That means that the two limits probably explain the same parts of the data, making a two-limit model very hard to interpret.
+Hence, only the bootstrapped confidence intervals for the two single limit models are shown in Figure 5.
+Further, since temporal limits might be fit better with an exponential distribution (and minimizing errors amounts to using a normal distribution), we repeated fitting single limit models based on normal, exponential and gamma likelihood distributions for all participants, but found that of the best temporal models, most used a normal distribution (6/9) and the rest an exponential distribution (3/9).
+All model fits combined show that while the reset point data reported here do somewhat favour a spatial limit, they can not conclusively rule out other limits - most notably a temporal limit.
 

### 2021-08-04 R/tracking_bounded.R
slightly more informative legend

@@ -409,7 +409,7 @@ plotBoundedTracking <- function(target='inline') {
   
   
   legend(x=0, y=45, 
-         legend=c('Cavanagh & Tse (2019)', 'participants', 'average'), 
+         legend=c('Cavanagh & Tse (2019)', 'participants (N=4)', 'average'), 
          col=c('black', colors$lightblue$s, colors$yorkred$s), 
          pch=c(NA,1,1), lty=c(1,0,0), 
          bty='n', cex=1)

### 2021-08-04 R/tryout.R

@@ -144,7 +144,6 @@ TlikelihoodTry <- function() {
 
 XTlikelihoodTry <- function() {
   
-  
   source('R/models_likelihood.R')
   source('R/retrace_onePass_V4.R')
   

### 2021-08-04 R/retrace_onePass_V4.R
consolidated & simplified figures for re-tracing experiment

@@ -618,16 +618,15 @@ getCVRdata <- function() {
 
 # Figures -----
 
-
-
-plotData <- function(target='inline') {
+plotResetPoints <- function(target='inline') {
   
+  source('R/models.R')
   
   if (target == 'pdf') {
-    cairo_pdf(filename='doc/Fig4_trajectories_resetpoints_illusionstrength.pdf',onefile=TRUE,width=8,height=4)
+    cairo_pdf(filename='doc/Fig4_trajectories_resetpoints.pdf',onefile=TRUE,width=7.5,height=4)
   }
   if (target == 'svg') {
-    svglite(file='doc/Fig4_trajectories_resetpoints_illusionstrength.svg',width=8,height=4)
+    svglite(file='doc/Fig4_trajectories_resetpoints.svg',width=7.5,height=4)
   }
   
   par(mar=c(3.5, 3.5, 2.5, 0.5))
@@ -639,15 +638,144 @@ plotData <- function(target='inline') {
   # 2: 2D overview of average reset points
   # 3: Cavangh & Tse comparison
   
-  layout(matrix(c(1,2,3), nrow = 1, ncol = 3, byrow = T), widths=c(1.2,1,1.7))
+  layout(matrix(c(1,2,3), nrow = 1, ncol = 3, byrow = T))
   
-  # **********************************
-  # EXAMPLE PARTICIPANT PLOT:
+  # # # # # # # # # # # # # # # # # # 
+  # 
+  # Reset point detection / initial direction
+  # 
+  # # # # # # # # # # # # # # # # # # 
   
   plot(-1000,-1000,
-       main='participant 4, 3 cps, 4 s',xlab='',ylab='',
-       xlim=c(-.25,.5), ylim=c(-.1,.9),
+       main='reset detection',xlab='',ylab='',
+       xlim=c(-1.5,7.5)/13.5, ylim=c(-1,13.5)/13.5,
        bty='n',ax=F, asp=1)
+  
+  title(xlab='cm', line=2.5)
+  title(ylab='cm', line=2.5)
+  
+  df <- read.csv(sprintf('data/onePass_V4/onepass_V4_p%02d.csv',1))
+  tasktrials <- unique(df$trial[df$taskname == 're-trace'])
+  taskdf <- df[which(df$trial_no %in% tasktrials),]
+  
+  #str(taskdf)
[…]

### 2021-08-04 R/models.R
finished function to plot bootstrapped model fits (not used anymore)

@@ -1241,8 +1241,17 @@ bootStrapFits <- function(bootstraps=1000) {
   
 }
 
-plotBootstrappedFits <- function() {
+plotBootstrappedFits <- function(target='inline') {
  
+  
+  if (target == 'pdf') {
+    cairo_pdf(filename='doc/FigY_modeled_limits.pdf',onefile=TRUE,width=5,height=4)
+  }
+  if (target == 'svg') {
+    svglite(file='doc/FigY_modeled_limits.svg',width=5,height=4)
+  }
+  
+  
   colors=getColors()
   layout(matrix(c(1,2),nrow=1,ncol=2,byrow=TRUE))
   
@@ -1310,5 +1319,9 @@ plotBootstrappedFits <- function() {
   
   axis(side=1, at=c(0,4,8))
   axis(side=2, at=seq(0,13.5,length.out = 4))
+  
+  if (target %in% c('pdf','svg')) {
+    dev.off()
+  }
   
 }

### 2021-08-04 Manuscript.Rmd
lots of edits to the manuscript

@@
 Informally, some observers have reported that the gabor’s path may saturate so that it remains parallel to, but offset from the true path.
-Others report that the position resets back toward its physical location whereupon accumulation begins again.
+Others report that the position moves back toward its physical location, either slowly or suddenly, whereupon accumulation begins again.
 The ability to record trajectories of the perceived position will then allow us to study spontaneous reset outcomes.
@@
 However, if $K$ is 1, then the internal motion has equal contribution with the external motion in determining the perceived direction.
-For comparison with other work, we will find the value of $K$ that best describes the strength of the illusion in both of our two experiments.
+For comparison with other work, we will find the value of $K$ that best describes the strength of the illusion in each of our two experiments.
 
@@
 **Bottom right:** The strength of the illusion in degrees angle deviation from the physical path.
-In red are the average illusions strengths for 0 _cps_, ±1 _cps_ and ±3 _cps_, and in light blue the data for the 4 individual participants.
+In red are the average illusion strengths for 0 _cps_, ±1 _cps_ and ±3 _cps_, and in light blue the data for the 4 individual participants.
 Gray dashed line: prediction of the vector combination model of Equation 1 with $K=1$.
@@
 The first (if any) of the remaining direction changes, are then taken as reset points.
-By first averaging the x and y coordinates (in _cm_) of reset points in all 9 participants and 6 stimulus conditions, we obtain 54 average reset points for further analyses and modelling.
-Both the horizontal and vertical (straight ahead) distance of the reset point from the origin of the trajectory are converted to centimeters, and trajectories are scaled in both dimensions so that vertical distance from the origin is 13.5 cm.
-The vertical component can then be converted to time as 1 _cm_ = 3.375 _s_ in the 4 _s_ condition and 1 _cm_ = 4.5 _s_ in the 3 _s_ condition.
+To keep the horizontal and vertical (straight ahead) coordinates of the reset points isometric they are given in centimeters.
+The distance from origin can then be converted to time as 1 _cm_ = 3.375 _s_ in the 4 _s_ condition and 1 _cm_ = 4.5 _s_ in the 3 _s_ condition.
+Participants' ability to retrace perceived trajectories might vary, but normalizing trajectories so that the final Y coordinate is at 13.5 _cm_ does not noticably change the results.
 
@@
 
-```{r fig4, echo=FALSE, message=FALSE, fig.width=8, fig.height=3.5, fig.cap="\\label{fig:fig4}**Figure 4:** Re-tracing Data.
-**Left:** One participant's trajectories with very clear 'return' resets at the purple dots.
+```{r fig4, echo=FALSE, message=FALSE, fig.width=7.5, fig.height=4, fig.cap="\\label{fig:fig4}**Figure 4:** Re-tracing Data.
+**Left:** Reset points are taken at the first sample in a splined-smoothed re-traced trajectory that goes in the direction opposite that of the expected illusion.
+Orange and gray lines: example trajectory.
+Orange dot: reset point.
+Red dot: point at half the distance to the reset point.
+Angle a: measure of illusion strength.
+**Middle:** One participant's trajectories with very clear 'return' resets at the purple dots.
 (Purple lines: trajectories up to resets, gray lines: trajectories after resets, blue lines: trajectories that had no resets in the re-tracing.) Other participants' data were not this clear.
-**Middle:** Average reset point locations for all 9 participants in all 6 conditions (54 points) with standard deviations.
-(Gray arrow: real trajectory, blue line: spatial average, red line: temporal average.) **Right:** Illusion strength as in Figure 2 (bottom right); plotted against predicted strength from Eq.1: the arctangent of the ratio of internal to external velocities.
-Blue dots from 3 s passes, red from 4 s passes, open circles are group averages.
-Predictions from Eq.
-1: Gray dashed line for $K=1$, black line; $K=0.81$, purple line; best fit here at $K=0.51$."}
+**Right:** Distribution of all 859 detected reset points."}
 #plotExampleData()
-plotData()
+plotResetPoints()
 ```
 
-On average, the algorithm detects resets in ~75% of trials.
+On average, this method detects resets in ~75% of trials.
 For three participants, only about half the trials showed resets (42%, 44% and 58%).
@@
 Therefore, we kept all participants and resets in our analysis (reset points in Fig.
-4, middle).
+4, right).
 
-We compare the strength of the illusion as measured with half-way points (see Fig 4, right), with an earlier study [@Cavanagh2019].
+As before, we compare the strength of the illusion (see Fig 5, left), with an earlier study [@Cavanagh2019].
 A line through the origin with slope 0.51 ($K$ from Eq.
 1) predicts illusion strength fairly well in these data, which is weaker than the $K$=0.81 found by Cavanagh & Tse [-@Cavanagh2019] and lower than the $K$=0.72 found in Experiment 1 with online tracking.
-We can also see that illusion strength measured with re-tracing is more variable.
-While the illusion is less strong and more variable, it is nevertheless there and depends on the relationship between internal and external speed as in previous studies.
+While the illusion is less strong, it is nevertheless there and depends on the relationship between internal and external speed as in previous studies.
 
@@
 That is, it looks like predicting reset points from a single X coordinate (or spatial limit) than predicing reset points from a given time.
+However, there is also considerable variance, so that fitting any line through the data, will necessarily leave errors.
 
-```{r fig5, echo=FALSE, message=FALSE, fig.width=7, fig.height=5.5, fig.cap="\\label{fig:fig5}**Figure 5:** Reset Models.
-Model limit shown here are given by weighted bootstrapping from all participants, lines are median parameter values, shaded areas the 95% confidence interval of the limits.
+```{r fig5, echo=FALSE, message=FALSE, fig.width=7.5, fig.height=4, fig.cap="\\label{fig:fig5}**Figure 5:** Illusion strength and reset models.
+**Left:** Illusion strength as in Figure 2 (bottom right); plotted against predicted strength from Eq.1: the arctangent of the ratio of internal to external velocities.
+Purple dots are the group means from 3 s passes, red from 4 s passes (2, 3 and 4 _cps_ internal speeds).
+Predictions from Eq.
+1: Gray dashed line for $K=1$, black line; $K=0.81$, purple line; best fit here at $K=0.51$ and the shaded area shows a 95% confidence interval.
+**Right:** Model limits shown here come from a weighted bootstrapping across participants, lines are median parameter values, shaded areas the 95% confidence intervals of the limits.
 Gray dots are the reset points from all participants.
-**Left:** Distance of reset from true path as a single spatial limit.
-**Right:** Reset coordinates predicted from a single temporal limit (at two speeds)."}
+While a spatial limit provides the best fit, neither a spatial nor a temporal limit explains the data well."}
 #plotResetPoints()
 #plotModels()
-plotBootstrappedFits()
+#plotBootstrappedFits()
+plotIllusionStrength()
 ```
 
-We test this with a series of simple models, based on straighforward trigonometry.
-First, we calculate the angle of a line through the origin and a sample on the trajectory halfway to the reset point (with coordinates $X$ and $Y$), as the strength of the illusion ($a = atan2(X,Y)$ expressing the angular deviation the line $x=0$ which is the real path of the gabor).
-This can then be used to calculate a slope ($s = \frac{sin(a)}{cos(a)}$).
+We test this with a series of simple models.
+First, for every reset point (with coordinates $X$ and $Y$) we use the illusion strength, expressed as either an angle (Fig 4, left: $a = atan2(X,Y)$) or a slope ($s = \frac{sin(a)}{cos(a)}$) to predict reset points.
 
@@
 [3a]  &nbsp;&nbsp;&nbsp;&nbsp;&nbsp;  $x = L_t \cdot v \cdot cos(a)$
+
 [3b]  &nbsp;&nbsp;&nbsp;&nbsp;&nbsp;  $y = L_t \cdot v \cdot sin(a)$
@@
 Where $L_t$ and $L_x$ are a temporal and spatial limit (in seconds and cm respectively) that are free parameters, $a$ and $s$ corresponds to the calculated angle and slope expressing the strength of the illusion, and speed ($v$ in _cm/s_) is given by the stimulus.
-If we fit the two simple models for each participant separately, by minimizing the mean square error (error as absolute distance in centimeters between predicted and actual reset points), most participants are best fit with a spatial limit (MSE: 6/9, AIC: 7/9) and the remaining participants are best fit with a temporal limit.
-We can also combine limits to see if they are complementary.
-Here, we see that for 4/9 participants (same for MSE and AIC) the best fit boils down to a spatial limit as the other limit is set to 0.
+If we fit the two simple models for each participant separately, by minimizing the mean square error (in isometric space, with absolute distance in centimeters between predicted and actual reset points), most participants are best fit with a spatial limit (MSE: 6/9, AIC: 7/9) and the remaining participants are best fit with a temporal limit.
+We can also combine limits to see if they are complementary, including.
+Here, we see that for 4/9 participants (MSE and AIC) the best fit equals a spatial limit as the other limit is set to 0.
 Of the remaining participants 4 are best fit with a combined spatial and temporal limit (4/9) and one participant is best fit combining a vertical offset with a temporal limit.
@@
 Hence, only the bootstrapped confidence intervals for the two single limit models are shown in Figure 5.
-Further, since temporal limits might be fit better with an exponential distribution (and minimizing errors amounts to using a normal distribution), we repeated fitting single limit models based on normal, exponential and gamma likelihood distributions for all participants, but found that of the best temporal models, most used a normal distribution (6/9) and the rest an exponential distribution (3/9).
+Further, since temporal limits might be fit better with an exponential distribution (and minimizing errors amounts to using a normal distribution), we repeated fitting single limit models based on normal, exponential and gamma likelihood distributions for all participants, but found that of the best temporal models, only a few used an exponential distribution (3/9) and the rest a normal distribution (6/9).
 All model fits combined show that while the reset point data reported here do somewhat favour a spatial limit, they can not conclusively rule out other limits - most notably a temporal limit.
@@
 We then set out to test if these spontaneous resets are triggered by a limiting distance of drift away from the physical location, or by a specific limit of time.
-The results showed that the resets are better explained by the distance from the true position, not by elapsed time.
+The results showed that the resets are slightly better explained by the distance from the true position, but a temporal limit is not ruled out.
 
@@
 4, middle) but much more variable in terms of time from the start of the gabor's motion.
-In any case, the temporal mean of these reset points was not 200 _ms_ but 1.5 _s_ and ranged from 0.5 _s_ to almost 4 _s_ (see Fig.
+In any case, the temporal limit of these reset points was not 200 _ms_ but 1.1 _s_ and fell into a wide range from 0.5 _s_ to almost 4 _s_ across trials (see Fig.
 4).
@@
 In our experiment, we did not control for eye-movements, so we can't directly investigate whether eye movements accompanied the “return” resets that participants reported.
-However, “return” resets occurred at a fairly consistent location for each participant, which suggests that they are triggered by the stimulus dynamics and not by attentional distraction or eye movements.
+**However, “return” resets occurred at a fairly consistent location for each participant, which suggests that they are triggered by the stimulus dynamics and not by attentional distraction or eye movements.**
 
@@
 Few if any resets were detected during the online tracking, either because of the additional attentional requirements or because of the short duration (2 s), Participants did report spontaneous resets in delayed, single-pass re-tracing of Experiment 2.
-The resets were best explained by a limiting spatial offset from the true stimulus location.
-This finding suggested that resets resolve a conflict between the physical and illusory position of the stimulus and that both are available to the visual system.
-This confirms the notion that we gauge the position of objects in the world by combining both retinal position signals and when the retinal information has high uncertainty, motion-based predictions.
+The resets were best explained by a limiting spatial offset from the true stimulus location, but do not exclude a temporal limit.
+These findings may either indicate there are multiple causes for spontaneous resets or that a re-tracing method is not suitable to resolve what causes this phenomenon.
+Either way, both a random time from stimulus onset until reset and too large a conflict between retinal and perceived position leading to resets remain possible.
 

### 2021-08-05 R/retrace_onePass_V4.R
changed illusion-strength plot to compare with Heller et al 2021, not Cavanagh and Tse, 2019

@@ -1068,7 +1068,7 @@ plotIllusionStrength <- function(target='inline') {
   
   angles <- c(0,5*(pi/12))
   lines(angles,(angles/pi)*180,col='gray',lty=2)
-  lines(angles,0.81*((angles/pi)*180),col='black',lty=1)
+  lines(angles,0.74*((angles/pi)*180),col='black',lty=1)
   
   #xcoords <- atan(internalspeed / externalspeed)
   
@@ -1085,7 +1085,7 @@ plotIllusionStrength <- function(target='inline') {
   points(avg_xcoords[avg_idxE4], 90 - ((avg_df$angle[avg_idxE4]/pi)*180), col=colors$orange$s, pch=16)
   
   legend(x=0, y=45, 
-         legend = c('group means (3 s)', 'group means (4 s)', 'K=1', 'K=0.81', sprintf('K=%0.2f',slope)), 
+         legend = c('group means (3 s)', 'group means (4 s)', 'K=1', 'K=0.74', sprintf('K=%0.2f',slope)), 
          pch=c(16,16,NA,NA,NA), col=c(colors$purple$s, colors$orange$s, 'gray', 'black', colors$lightblue$s), 
          lty = c(0,0,2,1,1),
          bty='n', cex=0.8)

### 2021-08-05 Manuscript.Rmd
edits suggested by Patrick

@@
     reference_docx: template.docx
+  pdf_document: default
   html_document:
     df_print: paged
-  pdf_document: default
 header-includes:
@@
 # And finally, code to compare this data with Sirui's PSEs
-source('R/time.R')
+# source('R/time.R') # no longer used
 
@@
 OSFdata()
-# this only downloads the data for the paper (well...)
 
@@
 preProcessOnePass_V4()
+# the rest of the paper should now compile without complaint
 ```
@@
 1, top left) generates extreme misjudgments of the moving target’s location which may deviate by as much as several degrees of visual angle from its true location.
-Surprisingly, this extraordinary perceptual illusion does not affect eye movements to the gabor: saccades to these targets are determined by their physical not their perceived location [@Lisi2015], and although immediate eye-movements go towards the true position, delayed eye movements and pointing go to the perceived position [@Lisi2017; @Massendari2018].
+Surprisingly, this extraordinary perceptual illusion does not affect eye movements to the gabor: immediate saccades to these targets are determined by their physical not their perceived location [@Lisi2015].
+In contrast, delayed eye movements and pointing go to the perceived position [@Lisi2017; @Massendari2018].
 These findings suggest that both the true and illusory positions are available in the brain.
@@
 If the gabor itself is not moving [@DeValois1991], the shift saturates quickly [40 _ms_: @Chung2007; 100 _ms_: @Arnold2007], perhaps because the accumulating evidence of actual location of the stationary gabor patch is sufficient to anchor its position.
-However, when the gabor itself also moves as in the double-drift stimulus, there is no stable location information to anchor the position estimate and the build-up continues well beyond 100 ms.
-The perceived location conintues to move away from its physical position by up to several times the size of the gabor.
-The strength of the illusion, the deviation of perceived path of the gabor from its true path, depends on the speed of both the internal and external motion [@Tse2006; @Shapiro2010; @Kwon2015; @Cavanagh2019].
+However, when the gabor itself also moves, as in the double-drift stimulus, there is no stable location information to anchor the position estimate and the build-up continues well beyond 100 ms.
+The perceived location continues to move away from its physical position by up to several times the size of the gabor.
+The strength of the illusion, the deviation of perceived path of the gabor from its true path, depends on the speed of both the internal and external motion [@Tse2006; @Shapiro2010; @Kwon2015; @Cavanagh2019; @Heller2021].
 In our first experiment, participants track their percept with a stylus as the gabor moves back and forth on its trajectory (Fig.
@@
 The stimuli were observed via an upward facing mirror placed exactly between the tablet and monitor, so that the stimuli appear to be in the same horizontal plane as the hand, and the tip of the stylus.
-Experiments were run in Python 2.7 with PsychoPy [@Peirce2019] and analyses were done in R 3.6.1 [@Rproject].
+Experiments were run in Python 2.7 with PsychoPy [@Peirce2019].
 
@@
 
-We will compare the illusory direction of the heading against the predictions of a simple vector combination model of perceived direction from Tse and Hsieh [-@Tse2006], Cavanagh and Tse [-@Cavanagh2019] and Heller, Patel, Faustin, Cavanagh, and Tse [-@Heller2021].
+We compare the illusory direction of the heading against the predictions of a simple vector combination model of perceived direction from Tse and Hsieh [-@Tse2006], Cavanagh and Tse [-@Cavanagh2019] and Heller et al.
+[-@Heller2021].
 In this model, the perceived direction is a simple vector combination of the external ($V_e$) and internal ($V_i$) motions and a constant $K$.
@@
 For comparison with other work, we will find the value of $K$ that best describes the strength of the illusion in each of our two experiments.
+
+All analyses were done in R 3.6.1 [@Rproject]. All data, as well as scripts for the experiments and analyses are available on OSF: [https://osf.io/72ndu/](https://osf.io/72ndu/).
 
@@
 **Left:** Individual trajectories.
-Note that there are only a few obvious discontinuities (highlighted in red).
 In general, the traced paths showed rough linearity.
@@
 Gray dashed line: prediction of the vector combination model of Equation 1 with $K=1$.
-Black line: fit with $K=0.81$ as in Cavanagh & Tse (2019), $K=0.81$.
+Black line: $K=0.74$ as in Heller et al.
+(2021).
 Purple: best fit to our data, $K=0.72$."}
@@
 The data clearly show the illusion, but does tracking decrease the strength of the illusion compared to simply observing it?
-In Figure 2 (bottom right), we compare mean illusion strength to that reported previously [@Cavanagh2019].
-As can be seen the averages from our data here (red circles, purple line) coincide fairly well with the illusion strength predicted by the model (Eq.
+In Figure 2 (bottom right), we compare mean illusion strength to that reported previously [@Heller2021].
+As can be seen the averages from our data here (red circles, purple line) coincide fairly well with the illusion strength predicted by the model fitted to all their data (Eq.
 1, in black).
@@
 
-As before, we compare the strength of the illusion (see Fig 5, left), with an earlier study [@Cavanagh2019].
+As before, we compare the strength of the illusion (see Fig 5, left), with an earlier study [@Heller2021].
 A line through the origin with slope 0.51 ($K$ from Eq.
-1) predicts illusion strength fairly well in these data, which is weaker than the $K$=0.81 found by Cavanagh & Tse [-@Cavanagh2019] and lower than the $K$=0.72 found in Experiment 1 with online tracking.
+1) predicts illusion strength fairly well in these data, which is weaker than the $K=0.74$ found by Heller et al.
+[-@Heller2021] and lower than the $K=0.72$ found in Experiment 1 with online tracking.
 While the illusion is less strong, it is nevertheless there and depends on the relationship between internal and external speed as in previous studies.
@@
 Predictions from Eq.
-1: Gray dashed line for $K=1$, black line; $K=0.81$, purple line; best fit here at $K=0.51$ and the shaded area shows a 95% confidence interval.
+1: Gray dashed line for $K=1$, black line; $K=0.74$, light blue line; best fit here at $K=0.51$ and the shaded area shows a 95% confidence interval.
 **Right:** Model limits shown here come from a weighted bootstrapping across participants, lines are median parameter values, shaded areas the 95% confidence intervals of the limits.
 Gray dots are the reset points from all participants.
-While a spatial limit provides the best fit, neither a spatial nor a temporal limit explains the data well."}
+While a spatial limit provides the better fit, neither a spatial nor a temporal limit explains the data well."}
 #plotResetPoints()
@@
 We test this with a series of simple models.
-First, for every reset point (with coordinates $X$ and $Y$) we use the illusion strength, expressed as either an angle (Fig 4, left: $a = atan2(X,Y)$) or a slope ($s = \frac{sin(a)}{cos(a)}$) to predict reset points.
+First, for every reset point (with coordinates $X$ and $Y$) we use the illusion strength based on a point earlier in the trajectory, expressed as either an angle (Fig 4, left: $a = \arctan{(X/Y)}$) or a slope ($s = sin(a)/cos(a)$) to predict reset points.
 
-If the perceived path of the gabor is approximated by a straight line (up to a reset), and reset points occur at a set limit, then a spatial limit ($L_x$) would predict the y-coordinate of the first reset, $y$:
+If the perceived path of the gabor is approximated by a straight line (up to a reset), and reset points occur at a set limit, then a spatial limit ($L_x$) would equal the x-coordinate and predict the y-coordinate of the first reset, $y$:
 
@@
 If we fit the two simple models for each participant separately, by minimizing the mean square error (in isometric space, with absolute distance in centimeters between predicted and actual reset points), most participants are best fit with a spatial limit (MSE: 6/9, AIC: 7/9) and the remaining participants are best fit with a temporal limit.
-We can also combine limits to see if they are complementary, including.
-Here, we see that for 4/9 participants (MSE and AIC) the best fit equals a spatial limit as the other limit is set to 0.
-Of the remaining participants 4 are best fit with a combined spatial and temporal limit (4/9) and one participant is best fit combining a vertical offset with a temporal limit.
-This indicates that in this data, a spatial limit explains the reset points best.
-However even the spatial limit does not explain the data very well.
+This indicates that in this data, a spatial limit explains the reset points best, but neither limit explains the data very well.
 
@@
 2) and on a temporal limit (eq.
-3) as well as a model combining these two were fit 1000 times.
-In the models with two limits there is a strong negative correlation between fitted spatial and temporal limits (r(998)=-0.985, p>.001) indicating parameter redundancy in the combined limit models.
-That means that the two limits probably explain the same parts of the data, making a two-limit model very hard to interpret.
-Hence, only the bootstrapped confidence intervals for the two single limit models are shown in Figure 5.
-Further, since temporal limits might be fit better with an exponential distribution (and minimizing errors amounts to using a normal distribution), we repeated fitting single limit models based on normal, exponential and gamma likelihood distributions for all participants, but found that of the best temporal models, only a few used an exponential distribution (3/9) and the rest a normal distribution (6/9).
-All model fits combined show that while the reset point data reported here do somewhat favour a spatial limit, they can not conclusively rule out other limits - most notably a temporal limit.
+3)  were fit 1000 times.
+We explored other modeling approaches (e.g.
+based on various distributions of likelihoods or combining limits) but these show the same pattern of results: spatial limits are somewhat favoured, but temporal limits can not be excluded (see OSF repository: [https://osf.io/72ndu/](https://osf.io/72ndu/)).
 
@@
 In our experiment, we did not control for eye-movements, so we can't directly investigate whether eye movements accompanied the “return” resets that participants reported.
-**However, “return” resets occurred at a fairly consistent location for each participant, which suggests that they are triggered by the stimulus dynamics and not by attentional distraction or eye movements.**
+However, the modestly better performance of the spatial limit model here suggests that spontaneous resets are not randomly distributed - as they might be if eye-movements or shifts of attention are the only way to elicit resets.
 
 We find that manual tracking of the double-drift stimulus during its motion did show the expected illusion.
-Few if any resets were detected during the online tracking, either because of the additional attentional requirements or because of the short duration (2 s), Participants did report spontaneous resets in delayed, single-pass re-tracing of Experiment 2.
+Few if any resets were detected during the online tracking, either because of the additional attentional requirements or because of the short duration (2 s).
+Participants did report spontaneous resets in delayed, single-pass re-tracing of Experiment 2.
 The resets were best explained by a limiting spatial offset from the true stimulus location, but do not exclude a temporal limit.
 These findings may either indicate there are multiple causes for spontaneous resets or that a re-tracing method is not suitable to resolve what causes this phenomenon.
-Either way, both a random time from stimulus onset until reset and too large a conflict between retinal and perceived position leading to resets remain possible.
+Either way, both some amount of time from stimulus onset until reset and too large a conflict between retinal and perceived position leading to resets remain possible.
 

### 2021-08-06 Manuscript.Rmd
removed K=1 lines from figures (2 and 5)

@@
 In red are the average illusion strengths for 0 _cps_, ±1 _cps_ and ±3 _cps_, and in light blue the data for the 4 individual participants.
-Gray dashed line: prediction of the vector combination model of Equation 1 with $K=1$.
-Black line: $K=0.74$ as in Heller et al.
+Black line: prediction of the vector combination model of Eq.
+1 with $K=0.74$ as in Heller et al.
 (2021).
@@
 **Left:** Illusion strength as in Figure 2 (bottom right); plotted against predicted strength from Eq.1: the arctangent of the ratio of internal to external velocities.
-Purple dots are the group means from 3 s passes, red from 4 s passes (2, 3 and 4 _cps_ internal speeds).
-Predictions from Eq.
-1: Gray dashed line for $K=1$, black line; $K=0.74$, light blue line; best fit here at $K=0.51$ and the shaded area shows a 95% confidence interval.
+Purple dots are the group means from 3 s passes, orange from 4 s passes (2, 3 and 4 _cps_ internal speeds).
+Predictions from the vector combination model in Eq.
+1: black line; $K=0.74$ as in Heller et al.
+(2021), light blue line; best fit here at $K=0.51$ and the shaded area shows a 95% confidence interval.
 **Right:** Model limits shown here come from a weighted bootstrapping across participants, lines are median parameter values, shaded areas the 95% confidence intervals of the limits.
@@
 Where $L_t$ and $L_x$ are a temporal and spatial limit (in seconds and cm respectively) that are free parameters, $a$ and $s$ corresponds to the calculated angle and slope expressing the strength of the illusion, and speed ($v$ in _cm/s_) is given by the stimulus.
-If we fit the two simple models for each participant separately, by minimizing the mean square error (in isometric space, with absolute distance in centimeters between predicted and actual reset points), most participants are best fit with a spatial limit (MSE: 6/9, AIC: 7/9) and the remaining participants are best fit with a temporal limit.
+If we fit the two simple models for each participant separately, by minimizing the mean square error (in isometric space, with absolute distance in centimeters between predicted and actual reset points), most participants are best fit with a spatial limit (MSE: 6/9, AIC: 7/9).
 This indicates that in this data, a spatial limit explains the reset points best, but neither limit explains the data very well.
@@
 2) and on a temporal limit (eq.
-3)  were fit 1000 times.
+3)  were fit 1000 times (see Fig.
+5, right).
+The median fits use a spatial limit of 3.5 _cm_ or ~3.6 _dva_ and a temporal limit of 1.1 _s_.
 We explored other modeling approaches (e.g.

### 2021-08-06 R/retrace_onePass_V4.R

@@ -1067,7 +1067,7 @@ plotIllusionStrength <- function(target='inline') {
           border=NA)
   
   angles <- c(0,5*(pi/12))
-  lines(angles,(angles/pi)*180,col='gray',lty=2)
+  #lines(angles,(angles/pi)*180,col='gray',lty=2)
   lines(angles,0.74*((angles/pi)*180),col='black',lty=1)
   
   #xcoords <- atan(internalspeed / externalspeed)
@@ -1085,9 +1085,9 @@ plotIllusionStrength <- function(target='inline') {
   points(avg_xcoords[avg_idxE4], 90 - ((avg_df$angle[avg_idxE4]/pi)*180), col=colors$orange$s, pch=16)
   
   legend(x=0, y=45, 
-         legend = c('group means (3 s)', 'group means (4 s)', 'K=1', 'K=0.74', sprintf('K=%0.2f',slope)), 
-         pch=c(16,16,NA,NA,NA), col=c(colors$purple$s, colors$orange$s, 'gray', 'black', colors$lightblue$s), 
-         lty = c(0,0,2,1,1),
+         legend = c('group means (3 s)', 'group means (4 s)', 'Heller et al. (2021)', sprintf('K=%0.2f',slope)), 
+         pch=c(16,16,NA,NA), col=c(colors$purple$s, colors$orange$s, 'black', colors$lightblue$s), 
+         lty = c(0,0,1,1),
          bty='n', cex=0.8)
   
   #axis(side=1,at=seq(0,3*(pi/8),pi/8),labels=c('0',expression(pi/8),expression(pi/4),expression(3*pi/8)))

### 2021-08-06 R/tracking_bounded.R

@@ -392,7 +392,7 @@ plotBoundedTracking <- function(target='inline', discontinuous=FALSE) {
        bty='n',ax=F)
   
   angles <- c(0,pi/4)
-  lines(angles,(angles/pi)*180,col='gray',lty=2)
+  #lines(angles,(angles/pi)*180,col='gray',lty=2)
   lines(angles,0.74*((angles/pi)*180),col='black',lty=1)
   
   

### 2021-08-06 Manuscript.Rmd
updated ethics statement, refer to figures in standardized format

@@
 
+The research was support by funding from NSERC Canada (PC and DYPH), as well as by the Department of Psychological and Brain Science, Dartmouth College (PC).
+
 # Abstract
@@
 This double-drift stimulus (Fig.
-1, top left) generates extreme misjudgments of the moving target’s location which may deviate by as much as several degrees of visual angle from its true location.
+1, left) generates extreme misjudgments of the moving target’s location which may deviate by as much as several degrees of visual angle from its true location.
 Surprisingly, this extraordinary perceptual illusion does not affect eye movements to the gabor: immediate saccades to these targets are determined by their physical not their perceived location [@Lisi2015].
@@
 All participants reported being right handed and had either normal or corrected to normal vision.
-Procedures were approved by York’s Human Participants Review Committee, and all participants provided prior, written, informed consent.
+Procedures were in accordance with the Declaration of Helsinki and were approved by York’s Human Participants Review Committee.
+All participants provided prior, written, informed consent.
 
@@
 
-Participants used a stylus on a horizontal drawing tablet (Wacom Intuos Pro) to indicate where they perceived the location of a gabor (see Fig 1, bottom right), as well as a keypad for additional responses.
+Participants used a stylus on a horizontal drawing tablet (Wacom Intuos Pro) to indicate where they perceived the location of a gabor (see Fig.
+1, right), as well as a keypad for additional responses.
 Stimuli were displayed on a downward facing LED monitor (30 Hz, 20”, 1680x1050), parallel with the drawing tablet.
@@
 As can be seen the averages from our data here (red circles, purple line) coincide fairly well with the illusion strength predicted by the model fitted to all their data (Eq.
-1, in black).
+1, black line).
 That is; tracking the illusion while observing it does not change the strength of the illusion.
@@
 library(magick)
-Fig4b <- image_read_svg('doc/methods_fig_2.svg')
+Fig3 <- image_read_svg('doc/methods_fig_2.svg')
 par(mai=c(0,0,0,0))
-plot(Fig4b)
+plot(Fig3)
 ```
@@
 We use the recorded trajectories to assess points where resets occurred, whether they were a discontinuity in orientation reflecting a saturation (hit the wall reset, Fig.
-4) or a discontinuity in position reflecting a jump back to the real location.
+3) or a discontinuity in position reflecting a jump back to the real location.
 These reset points can then be used to determine if resets are time-limited or space-limited.
@@
 All participants reported to be right handed and had either normal or corrected to normal vision.
-Procedures were approved by York’s Human Participants Review Committee, and all participants provided prior, written, informed consent.
+Procedures were in accordance with the Declaration of Helsinki and were approved by York’s Human Participants Review Committee.
+All participants provided prior, written, informed consent.
 
@@
 
-The setup was the same as used in the online tracking experiment, except that we now also used a small keypad, for responses (see Fig 1, right).
+The setup was the same as used in the online tracking experiment, except that we now also used a small keypad, for responses (Fig.
+1, right).
 Stimuli were identical to those used in the online tracking experiment, but now always started at the near end of the workspace and moved away from the participants.
@@
 
-In Figure 4 (left), we show the trajectories for participant 4 in the condition with 3 _cps_ internal motion and 4 _s_ passes.
+In Figure 4 (middle), we show the trajectories for participant 4 in the condition with 3 _cps_ internal motion and 4 _s_ passes.
 This particular participant's re-traced paths show clear 'return' resets, all of which are detected by our simple algorithm.
-Other participants are not that clear cut.
+Other participants' data are less clear.
 
@@
 We test this with a series of simple models.
-First, for every reset point (with coordinates $X$ and $Y$) we use the illusion strength based on a point earlier in the trajectory, expressed as either an angle (Fig 4, left: $a = \arctan{(X/Y)}$) or a slope ($s = sin(a)/cos(a)$) to predict reset points.
+First, for every reset point (with coordinates $X$ and $Y$) we use the illusion strength based on a point earlier in the trajectory, expressed as either an angle (Fig.
+4, left: $a = \arctan{(X/Y)}$) or a slope ($s = sin(a)/cos(a)$) to predict reset points.
 
-If the perceived path of the gabor is approximated by a straight line (up to a reset), and reset points occur at a set limit, then a spatial limit ($L_x$) would equal the x-coordinate and predict the y-coordinate of the first reset, $y$:
+If the perceived path of the gabor is approximated by a straight line (up to a reset), and reset points occur at a set limit, then a spatial limit ($L_x$) would equal the x-coordinate and predict the y-coordinate, $y$, of the first reset:
 
@@
 
-Where $L_t$ and $L_x$ are a temporal and spatial limit (in seconds and cm respectively) that are free parameters, $a$ and $s$ corresponds to the calculated angle and slope expressing the strength of the illusion, and speed ($v$ in _cm/s_) is given by the stimulus.
+Where $L_t$ and $L_x$ are a temporal and spatial limit (in seconds and cm respectively) that are free parameters, $a$ and $s$ correspond to the calculated angle and slope expressing the strength of the illusion, and speed ($v$ in _cm/s_) is given by the stimulus.
 If we fit the two simple models for each participant separately, by minimizing the mean square error (in isometric space, with absolute distance in centimeters between predicted and actual reset points), most participants are best fit with a spatial limit (MSE: 6/9, AIC: 7/9).
@@
 2) and on a temporal limit (eq.
-3)  were fit 1000 times (see Fig.
+3)  were fit 1000 times (Fig.
 5, right).

### 2021-08-06 Manuscript.Rmd
added references (Ueda et al., 2018, Cormack, 2019) and changed an incorrect ref to a figure (4 right, not 4 middle)

@@
 Surprisingly, this extraordinary perceptual illusion does not affect eye movements to the gabor: immediate saccades to these targets are determined by their physical not their perceived location [@Lisi2015].
-In contrast, delayed eye movements and pointing go to the perceived position [@Lisi2017; @Massendari2018].
+In contrast, delayed eye movements and pointing go to the perceived position [@Lisi2017; @Massendari2018; @Ueda2018].
 These findings suggest that both the true and illusory positions are available in the brain.
@@
 In this experiment, participants were asked to use a stylus on a drawing tablet to track a moving gabor with some amount of internal motion that elicited the double-drift illusion.
+A variant on online tracking has been used before [@Cormack2019] with a continuous task where the stimulus depended on the response.
 We test if tracking the illusion continuously is a good measure of illusion strength, or if it prevents or decreases the illusion.
@@
 Now we look at the distribution of reset points, and whether they can be explained by a temporal or spatial limit.
-In Figure 4 (middle), we can see that the reset points vary more along the y-axis, than along the x-axis.
+In Figure 4 (right), we can see that the reset points vary more along the y-axis, than along the x-axis.
 That is, it looks like predicting reset points from a single X coordinate (or spatial limit) than predicing reset points from a given time.
@@
 
-We first tested whether manual tracking of a double-drift stimulus is susceptible to the illusion as reported by perceptual judgments, pointing [@Lisi2017], and memory saccades [@Massendari2018].
+We first tested whether manual tracking of a double-drift stimulus is susceptible to the illusion as reported by perceptual judgments, pointing [@Lisi2017], and memory saccades [@Massendari2018; @Ueda2018].
 The results of this online tracking experiment show a close agreement between the angle of the manual tracking and the illusion strength expected based on the model from earlier perceptual measures [@Cavanagh2019].

### 2021-08-07 Manuscript.Rmd
updated equations with sin and cos in latex to \sin and \cos

@@
     reference_docx: template.docx
-  pdf_document: default
   html_document:
     df_print: paged
+  pdf_document: default
 header-includes:
@@
 First, for every reset point (with coordinates $X$ and $Y$) we use the illusion strength based on a point earlier in the trajectory, expressed as either an angle (Fig.
-4, left: $a = \arctan{(X/Y)}$) or a slope ($s = sin(a)/cos(a)$) to predict reset points.
+4, left: $a = \arctan{(X/Y)}$) or a slope ($s = \sin{(a)}/\cos{(a)}$) to predict reset points.
 
@@
 
-[3a]  &nbsp;&nbsp;&nbsp;&nbsp;&nbsp;  $x = L_t \cdot v \cdot cos(a)$
+[3a]  &nbsp;&nbsp;&nbsp;&nbsp;&nbsp;  $x = L_t \cdot v \cdot \cos{(a)}$
 
-[3b]  &nbsp;&nbsp;&nbsp;&nbsp;&nbsp;  $y = L_t \cdot v \cdot sin(a)$
+[3b]  &nbsp;&nbsp;&nbsp;&nbsp;&nbsp;  $y = L_t \cdot v \cdot \sin{(a)}$
 

### 2021-11-09 exp/luminance.py
psychopy "experiment" to measure luminance of the monitor

@@ -1,0 +1,169 @@
+# do imports:
+
+import numpy as np
+#import scipy as sp
+#import pandas as pd
+
+#import pygame
+#from pygame.locals import *
+
+import psychopy
+from psychopy import visual, core, data, event, logging, sound, gui
+
+#import time
+
+#import random
+#from random import seed
+
+
+
+
+def runLuminanceTest():
+    
+    # setup configuration object:
+    cfg = {}
+  
+    # add a window for visual stimuli:
+    cfg['fullscr']      = True
+    cfg['monitorIndex'] = 1
+    cfg['flip']         = True
+    cfg['flip']         = False
+    #resolution=[1680, 1050]
+    cfg = createWindow(cfg, resolution=[1920, 1080])
+    
+    cfg = createStimuli(cfg)
+    
+    cfg = createTestColors(cfg, ncolors=18, channels='all')
+    #cfg = createTestColors(cfg, ncolors=6, channels=['white'])
+
+    # run the actual experiment, catch errors to cleanly exit:
+    try:
+
+        cfg = LuminanceTest(cfg)
+
+    except Exception as err:
+        
+        # what went wrong?
+        print(err)
+        #print('some error?')
+
+    finally:
+
+        cfg = cleanlyExit(cfg)
+
+
+
+def createWindow(cfg, resolution=[1920, 1080]):
+  
+    viewScale = [1, 1]
+    if cfg['flip']:
+        viewScale = [1, -1]
+
+    win = visual.Window(resolution, fullscr=cfg['fullscr'], viewScale=viewScale, waitBlanking=True)
+  
+  
+    cfg['win'] = win
+    winsize = win.size
+    cfg['width'] = winsize[0]
+    cfg['height'] = winsize[1]
+    cfg['winpos'] = cfg['win'].pos # might be necessary to subtract this from collected mouse coordinates?
+  
+    return(cfg)
+
+
+def createStimuli(cfg):
+    
+    instruction_pos = [0,0.5]
+    if (cfg['flip']):
+        instruction_pos = [0,-0.5]
+    # instruction text to show the RGB value used
+    cfg['instruction'] = visual.TextStim(win=cfg['win'], text='', pos=instruction_pos, colorSpace='rgb', color=[1,1,1], flipVert=cfg['flip'])
+    
+    # rectangle?
+    #w = cfg['width']
+    #h = cfg['height']
+    #if (w > h):
+    #    rectsize = [w/h, 1]
+    #else:
+    #    rectsize = [1, h/w]
+    
[…]

### 2021-11-09 Manuscript.Rmd
some revisions

@@
 source('R/common.R')
-# The functions in here are used on the data of the bounded tracking experiments:
+# The functions in here are used on the data of the bounded tracking experiment:
 source('R/tracking_bounded.R')
@@
 source('R/compare.R')
-# And then some code for the re-tracing experiment...
+# And then some code for the re-tracing experiment:
 source('R/retrace_onePass_V4.R')
@@
 # we pre-process the data:
-getBoundedTrackingStandardizedSegments()
-getSegmentDirections()
+#getBoundedTrackingStandardizedSegments()
+segmentDirections()
 preProcessOnePass_V4()
@@
 To report the perceived trajectory of the gabor, participants drew it after the gabor was removed from the monitor.
-Resets were detected and analyzed and they suggest that there is a spatial limit beyond which the illusion saturates or resets.
+Resets were detected and analyzed but their distribution neither matches a strict temporal or spatial limit on perception that could trigger resets.
 
@@
 The strength of the illusion, the deviation of perceived path of the gabor from its true path, depends on the speed of both the internal and external motion [@Tse2006; @Shapiro2010; @Kwon2015; @Cavanagh2019; @Heller2021].
-In our first experiment, participants track their percept with a stylus as the gabor moves back and forth on its trajectory (Fig.
-1) with 5 speeds of the internal motion.
-We analyze the angle of the tracking to determine if the manual response is susceptible to the illusion.
+Rather than use point estimates of the perceived location of the stimulus, we want to test how well a continuous trajectory can measure the strength of the illusion, or if perception is changed by this response modality such as in other illusions
+
+CITE SOME PREVIOUS WOKR HERE.
+
+We assess this in our first experiment, using hand movement trajectores recorded during stimulus presentation (Fig.
+1).
+We compare trajectories across 5 different internal motions of the gabor and compare the illusion strength with previous work based on a straightforward vector combination model [@Heller2021].
+As we expected [@mhjh], online tracking provides a similar measure of illusion strength as pointing.
 
@@
 
-In this experiment, participants used their unseen hand and stylus to do continuous online tracking of the double-drift stimulus that moved back and forth along the Y-axis of the horizontal screen for 12 _s_.
+In this experiment, participants used their unseen hand and stylus to do continuous online tracking of the double-drift stimulus while that moved back and forth along the Y-axis of the horizontal screen for 12 _s_.
 Before trial onset, the participant moved the stylus to the middle of the screen, and were given 1.5 _s_ to fixate a point on the left or right of the screen.
@@
 Because there is no set reference point to gauge movement direction against after each direction change, we use instantaneous heading along the trajectory instead.
-We first segmented the tracking trajectories according to stimulus passes (in between direction changes of the gabor), and removed the noisy tracking during the first and last half second, leaving 1 _s_ trajectory segments.
-We then calculated the instantaneous heading (direction, disregarding velocity or distance) of a smooth spline interpolated version of each trajectory segment (using 101 spline interpolated position samples, to get 100 heading samples).
+We first segmented the tracking trajectories according to full stimulus passes (in between direction changes of the gabor), and removed the noisy tracking during the first and last half second, leaving 1 _s_ trajectory segments.
+We then calculated the instantaneous heading (direction, disregarding velocity or distance) between all 32 raw samples for 31 heading samples per segment.
 This heading measure should depend on the illusion strength.
@@
 
+Here we calculate $V_i$ and $V_e$ in _cm/s_; _cps_ * 0.58 _cm_ for $V_i$ and 13.5/2 _cm/s_ for $V_e$.
 If $K$ is 0, then there is no illusion and the perceived direction matches the external motion direction.
@@
 
-In Figure 2 (top right), we plot the average instantaneous heading during the middle second of each 2 _s_ pass for each of the 5 internal motion speeds.
+All raw trajectories are show in Figure 2A.
+In Figure 2C, we plot the average instantaneous heading during the middle second of each 2 _s_ pass for each of the 5 internal motion speeds.
 The average angle of the tracked path is given in degrees deviation from straight forward.
-It seems that continuous tracking is sensitive to the strength of the illusion, and this is confirmed by a repeated-measures F-test on the average angles, using internal motion as a within-subjects factor ( _F_(4,12)=102.6, _p_<.001, $\eta^2$=0.97 ).
-With FDR-corrected, paired t-tests, we find a difference in heading between internal motion speeds of 0 _cps_ and both 1 _cps_ ( _p_=.022) and -1 _cps_ ( _p_=.025) as well as between -3 and -1 _cps_ and 3 and 1 _cps_ (both _p_ <.001).
+It seems that continuous tracking is sensitive to the strength of the illusion, and this is confirmed by a repeated-measures ANOVA on the average angles, using internal motion as a within-subjects factor ( _F_(4,12)=102.6, _p_<.001, $\eta^2$=0.97 ).
+With 4 FDR-corrected [@benjamini1995], paired t-tests, we find a difference in heading between internal motion speeds of 0 _cps_ and both 1 _cps_ ( _p_=.022) and -1 _cps_ ( _p_=.025) as well as between -3 and -1 _cps_ and 3 and 1 _cps_ (both _p_ <.001).
 This makes it clear that continuous tracking does not prevent the illusion from occurring.
@@
 
-```{r fig2, echo=FALSE, message=FALSE, fig.width=8, fig.height=5, fig.cap="\\label{fig:fig2}**Figure 2:** Online Tracking Data.
-**Left:** Individual trajectories.
-In general, the traced paths showed rough linearity.
-**Top right:** Average heading of the middle 1 _s_ of each pass.
+```{r fig2, echo=FALSE, message=FALSE, fig.width=6, fig.height=5, fig.cap="\\label{fig:fig2}**Figure 2:** Online Tracking Data.
+**A:** Raw individual trajectories.
+In general, the traced paths approximate linearity, with the direction depending on the internal motion.
+**B:** Four trials with a possible reset.
+**C:** Average heading of the middle 1 _s_ of each pass.
 Distance from origin is time, angle is the heading angle.
-Dark blue lines the average heading, in light blue a polar heatmap with the distribution of heading.
-**Bottom right:** The strength of the illusion in degrees angle deviation from the physical path.
+Dark blue lines show the average heading.
+In light blue is the average area (across participants) containing 90% of the heading data.
+**D:** The strength of the illusion in degrees angle deviation from the physical path.
 In red are the average illusion strengths for 0 _cps_, ±1 _cps_ and ±3 _cps_, and in light blue the data for the 4 individual participants.
@@
 1 with $K=0.74$ as in Heller et al.
-(2021).
-Purple: best fit to our data, $K=0.72$."}
+(2021), which is indistinguishable from the best fit to our data: $K=0.738$."}
 plotBoundedTracking()
@@
 The data clearly show the illusion, but does tracking decrease the strength of the illusion compared to simply observing it?
-In Figure 2 (bottom right), we compare mean illusion strength to that reported previously [@Heller2021].
+In Figure 2D, we compare mean illusion strength to that reported previously [@Heller2021].
 As can be seen the averages from our data here (red circles, purple line) coincide fairly well with the illusion strength predicted by the model fitted to all their data (Eq.
@@
 In the previous experiment with online tracking, the duration of each pass in the online tracking experiment was 2 seconds and the manual traces (Fig.
-2, left) mostly showed a linear trajectory without saturation or reset of the types shown in Figure 3 (left).
+2A,B) mostly showed a linear trajectory without saturation or reset of the types shown in Figure 3 (left).
 The continuation of the accumulation may be a result of the active tracking – either the extra attention required to track increased the extent of the accumulation or possibly the temporal demands of the tracking caused the participants to average their traces over any breaks.
+Alternatively, post-dictive inference may prevent resets from being perceived.
 To address this, we switched to a delayed, offline recording of the perceived path.
 
-```{r fig3, echo=F, fig.height=4.55, fig.width=8, message=FALSE, error=FALSE, warning=FALSE, fig.cap="\\label{fig:fig3}**Figure 3:** Double-Drift Resets.
+```{r fig3, echo=F, fig.height=3.4125, fig.width=6, message=FALSE, error=FALSE, warning=FALSE, fig.cap="\\label{fig:fig3}**Figure 3:** Double-Drift Resets.
 **Left:** At some point, the perceived position may stop moving further away from its actual path.
 For some people this takes the form of a ‘return’ reset, back toward the true position, for others it means the illusory position remains at a fixed offset moving parallel to the real path, or some combination of the two.
-**Top right:** if these resets are due to a spatial limit on the size of the illusory position shift away from the true vertical path, the locations of the resets should have a constant X coordinatee, the spatial offset from the vertical path.
-**Bottom right:** if they occur instead after some period of time, resets should have a constant Y coordinate, the time since the gabor’s motion began."}
+**Top right:** if these resets are due to a spatial limit on the size of the illusory position shift away from the true vertical path, the locations of the resets should have a constant X coordinate, the spatial offset from the vertical path.
+**Bottom right:** if they occur instead after some period of time, resets should have a constant radius (R); the time since the gabor’s motion began."}
 library(magick)
@@
 We were interested, in particular, in whether the limitation of the accumulation would be set by space or time.
-Since both the true and the illusory position appear to be available in the visual system (the true position drives saccades whereas the illusory position drives perception, Lisi & Cavanagh, -@Lisi2015, it is possible that resets occur once the distance between the real and perceived positions exceeds some limit.
+Since both the true and the illusory position appear to be available in the visual system (the true position drives saccades whereas the illusory position drives perception, [@Lisi2015], it is possible that resets occur once the distance between the real and perceived positions exceeds some limit.
 We can speculate that this spatial limit would be related to the positional uncertainty of the gabor.
@@
 All participants reported to be right handed and had either normal or corrected to normal vision.
-Procedures were in accordance with the Declaration of Helsinki and were approved by York’s Human Participants Review Committee.
+Procedures were in accordance with the Declaration of Helsinki (2013) and were approved by York’s Human Participants Review Committee.
 All participants provided prior, written, informed consent.
@@
 
-The setup was the same as used in the online tracking experiment, except that we now also used a small keypad, for responses (Fig.
+The setup was the same as used in the online tracking experiment, except that we now also used a small keypad for responses (Fig.
 1, right).
 Stimuli were identical to those used in the online tracking experiment, but now always started at the near end of the workspace and moved away from the participants.
-The double drift stimuli would use either 3 or 4 seconds to move 13.5 _cm_, corresponding to speeds of ~4.5 and ~3.75 _cm/s_ or ~4.7 or ~3.5 _dva/s_, and had an internal drift of 2, 3 or 4 _cps_ .
+The double drift stimuli would use either 3 or 4 seconds to move 13.5 _cm_, corresponding to speeds of ~4.5 and ~3.75 _cm/s_ or ~4.7 and ~3.5 _dva/s_, and had an internal drift of 2, 3 or 4 _cps_ .
 
@@
 Using a smooth spline interpolated version of the trajectory we detect right-left direction changes, based on sign changes.
-Since the initial movement is often jittery, we only take direction changes further than ~0.39 cm ( ~0.23 _dva_ ) away from the starting point.
+First, we remove any short segments on the trajectory at the start and end that go toward the participant.
+Then, we downsample the trajectories such that each successive sample is at least 9 _mm_ from the previous, and then applied a median filter with a half-width of 5 samples.
+In these summary trajectories we detect the first inflection point using the X coordinates.
+We then use the raw trajectory again, first locating the corresponding sample for the inflection point, and then find the first previous sample at 97.5% of the X coordinate at the inflection point sample in the raw trajectory.
+Downsampling removes the small periodic signal in the trajectories recorded from participant XYZXYZXYZ that seem to capture the phase shifts of the gabor.
+
+################################
+################################
+################################
+################################
+################################
+################################
+################################
+################################
+################################
+################################
+
 The first (if any) of the remaining direction changes, are then taken as reset points.
@@
 1) predicts illusion strength fairly well in these data, which is weaker than the $K=0.74$ found by Heller et al.
-[-@Heller2021] and lower than the $K=0.72$ found in Experiment 1 with online tracking.
+[-@Heller2021] and in Experiment 1 with online tracking.
 While the illusion is less strong, it is nevertheless there and depends on the relationship between internal and external speed as in previous studies.

### 2021-11-09 R/common.R
move segmentation code to bounded tracking file, add function for panel labels

@@ -143,127 +143,127 @@ rotateCoordinates <- function(df,angle,origin=c(0,0)) {
   
 }
 
-segmentMultiPassTrial <- function(df, velocityCutoff = NA, type='gaborzerocrossing') {
-  
-  # type: gaborreversal OR gaborzerocrossing 
-  
-  # first get the part where the gabor is actually moving (or present on the screen):
-  gabormoves <- which(df$gabory_pix != 0 & !is.na(df$gabory_pix)) # second condition not technically necessary
-  gaboronset <- min(gabormoves) - 1
-  gabormoves <- c(min(gabormoves):max(gabormoves)) # in case a sample accidentally puts the gabor at exactly 0
-  
-  X <- df$handx_pix[gabormoves]
-  Y <- df$handy_pix[gabormoves]
-  
-  if (type == 'gaborzerocrossing') {
-    
-    # now get the zero-crossings of the gabors (the middle of each segment will lag only briefly behind that):
-    crossings <- zerocrossings(df$gabory_pix[gabormoves], bounces = FALSE)
-    #print(crossings)
-    segments <- list()
-    
-    for (cross.idx in c(1:length(crossings))) {
-      
-      #cat(sprintf('\ncrossing: %d\n\n', cross.idx))
-      
-      crossing <- crossings[cross.idx]
-      
-      # beginning of segment:
-      seg.in <- crossing - which(abs(diff(sign(diff(rev(Y[1:crossing]))))) > 0)[1] + 1
-      
-      # end of segment:
-      seg.out <- crossing + which(abs(diff(sign(diff(Y[crossing:length(Y)])))) > 0)[1] - 1
-      
-      #print(c(seg.in,seg.out))
-      
-      # velocity threshold?
-      if (is.numeric(velocityCutoff)) {
-        
-        t <- df$time_ms[gabormoves]
-        V <- getSplinedVelocity(X[c(seg.in:seg.out)], Y[c(seg.in:seg.out)], t[c(seg.in:seg.out)])$velocity
-        
-        #print(V)
-        
-        velCrit <- max(V) * velocityCutoff
-        #print(velCrit)
-        #print(which(V > velCrit))
-        
-        seg.in  <- seg.in  + (which(V      > velCrit)[1] - 1)
-        seg.out <- seg.out - (which(rev(V) > velCrit)[1] - 1) 
-        
-      }
-      
-      seg.in  <- seg.in  + gaboronset
-      seg.out <- seg.out + gaboronset
-      
[…]

### 2021-11-09 R/models.R
finished gaussian and gamma distribution models, started plotting code

@@ -790,8 +790,8 @@ getDataTrials <- function(illusionMinimum=5, illusionMaximum=85) {
   
   
   # add slopes conveniently:
-  df$angle <- ((df$initialdirection)/180)*pi
-  df$angle <- df$initialdirection
+  #df$angle <- ((df$initialdirection)/180)*pi
+  df$angle <- (pi/2) - df$initialdirection
   df$sin.a <- sin(df$angle)
   df$cos.a <- cos(df$angle)
   df$slope <- df$sin.a / df$cos.a
@@ -807,8 +807,19 @@ getDataTrials <- function(illusionMinimum=5, illusionMaximum=85) {
   df$Vi <- df$internalspeed
   df$Ve <- round(0.5 / df$externalspeed)
   
+  
+  df$unMu <- df$slope * df$Y
+  
+  # this seems incorrect:
+  #df$unRT <- pmax(0,df$Y-(df$X/df$slope)) + pmin(df$Y*df$slope,df$X/df$slope)
+  
+  # also wrong:
+  #df$unRT <- sqrt(df$X^2 + (df$X*df$slope)^2) + pmax(0,df$Y-(df$X/df$slope))
+  
+  df$unRT <- ( df$Y + pmax(0, (sqrt(df$X^2 +(df$X*df$slope)^2) - (df$X*df$slope) ) ) ) / df$speed
+  
   # select only useful columns:
-  df <- df[,c('participant','X', 'Y', 'RT', 'speed', 'slope', 'angle', 'sin.a', 'cos.a', 'Vi', 'Ve')]
+  df <- df[,c('participant','X', 'Y', 'RT', 'speed', 'slope', 'angle', 'sin.a', 'cos.a', 'Vi', 'Ve', 'unMu', 'unRT')]
   
   return(df)
   
@@ -1360,9 +1371,9 @@ XoffsetGaussianLikelihood <- function(par,data) {
   
 }
 
-# ************************************************
-# offset gamma probability density functions -----
-# ************************************************
+# *****************************************
+# gamma probability density functions -----
+# *****************************************
 
 # gamma distributions have a 'shape' parameter that puts them
 # on a continuum that spans exponential and normal distributions  
@@ -1387,13 +1398,43 @@ ToffsetGammaLikelihood <- function(par, data) {
   rT <- par['rT']
   oT <- par['oT']
   
-  L <- dgamma( sqrt(data$X^2 + data$Y^2) - (oT * data$speed), shape=sT, rate=rT  )
[…]

### 2021-11-09 R/retrace_onePass_V4.R
finalized new reset detection algorithm

@@ -480,7 +480,8 @@ oldPreProcessOnePass_V4 <- function(participants = c(2,3,4,5,6,8,9,10,11), overw
 
 preProcessOnePass_V4 <- function(participants = c(2,3,4,5,6,8,9,10,11), 
                                  overwrite = TRUE,
-                                 RPFUN=get97.5percMaxX1_resetpoint,
+#                                 RPFUN=get97.5percMaxX1_resetpoint,
+                                 RPFUN=getResampledFilteredInflectResetPoint,
                                  IDFUN=getDirectionHalfway) {
   
   
@@ -535,27 +536,16 @@ preProcessOnePass_V4 <- function(participants = c(2,3,4,5,6,8,9,10,11),
       externalspeed <- c(externalspeed, trialdf$externalMovement[1])
       fixationside <- c(fixationside, trialdf$fixationside[1])
       
-      # only step 99 was retracing (the preceding ones are for stimulus presentation)
-      step <- 99
-      step.idx <- which(trialdf$step == step)
+      trace <- cmTrace(trialdf)
       
-      # get the relevant samples X and Y coordinates and scale to centimeters:
-      x <- ((trialdf$handx_pix[step.idx] / (524)) + 0.0) * 13.5
-      y <- ((trialdf$handy_pix[step.idx] / (524)) + 0.5) * 13.5
+      x <- trace$x
+      y <- trace$y
+      t <- trace$t
       
-      # start at origin:
-      x <- x - x[1]
-      y <- y - y[1]
-      
-      # get time as well (for filters, ended up not using it):
-      t <- trialdf$time_ms[step.idx]
-      
-      # flip to normalize internal motion direction
-      if (trialdf$internalMovement[1] < 0) x <- -x
       
       # GET THE RESET POINT
       # which function is used to get the reset point is a function argument:
-      resetPoint <- RPFUN(x,y,t,verbosity=verbosity) 
+      resetPoint <- RPFUN(x,y,t,verbosity=verbosity, internalspeed=trialdf$internalMovement[1], externalspeed=trialdf$externalMovement[1])
       
       # store the reset point in the output data vectors:
[…]

### 2021-11-09 R/tracking_bounded.R
finished new figure 2

@@ -5,66 +5,256 @@ source('R/common.R')
 
 # Data handling -----
 
-getBoundedTrackingStandardizedSegments <- function(segmentpoints=101, version=1) {
+# getBoundedTrackingStandardizedSegments <- function(segmentpoints=101, version=1) {
+#   
+#   # use these participants:
+#   participants <- c(1,3,5,6,7,8)
+#   participants <- c(3,6,7,8) # 1 and 5 are authors
+#   folderstr <- 'data/bounded_tracking/'
+#   filestr <- 'bounded_tracking_'
+#   
+#   if (version == 2) {
+#     participants <- c(1,2,9,10,11,12,13)
+#     participants <- c(2,9,10,11,12,13) # 1 is author
+#     folderstr <- 'data/bounded_tracking_V2/'
+#     filestr <- 'bounded_tracking_V2_'
+#   }
+#   
+#   if (file.exists(sprintf('%sstandardized_segments.csv',folderstr))) {
+#     return()
+#   }
+#   
+#   # read all the data into one data frame:
+#   # BTdata <- NA
+#   BTSdata <- NA
+#   
+#   for (ppno in participants) {
+#     
+#     ppdf <- read.csv(sprintf('%s%sp%02d.csv',folderstr,filestr,ppno), stringsAsFactors = F)
+#     ppdf$participant <- ppno
+#     
+#     # # get all raw data in one frame?
+#     # if (is.data.frame(BTdata)) {
+#     #   BTdata <- rbind(BTdata, ppdf)
+#     # } else {
+#     #   BTdata <- ppdf
+#     # }
+#     
+#     # segments:
+#     trials <- unique(ppdf$trial_no)
+#     #print(trials)
+#     for (trialnum in trials) {
+#       #print(length(trialnum))
+#       trialdf <- ppdf[which(ppdf$trial_no == trialnum),]
+#       #str(trialdf)
+#       # normalize for external direction:
+#       trialdf$handx_pix <- trialdf$handx_pix * trialdf$externalDirection[1]
+#       
+#       # split into segments:
+#       tracksegments <- segmentMultiPassTrial(trialdf, velocityCutoff = NA, type='gaborreversal')
+#       #print(tracksegments)
+#       for (seg.no in c(1:length(tracksegments))) {
+#         seg <- tracksegments[[seg.no]]
+#         segmentdf <- trialdf[c(seg[1]:seg[2]),]
+#         #segmentdf$segment_no <- seg.no
+#         
+#         if (dim(segmentdf)[1] < 32) {
[…]

### 2021-11-13 R/retrace_onePass_V4.R
moved unused source files to R/old/ folder, and renamed the source files that were left

@@ -1,3236 +1,0 @@
-library('svglite')
-library('ez')
-library('segmented')
-
-source('R/common.R')
-#source('R/models.R')
-
-# Data handling -----
-
-oldPreProcessOnePass_V4 <- function(participants = c(2,3,4,5,6,8,9,10,11), overwrite=FALSE) {
-  
-  if (file.exists('data/onepass_V4/onePass_V4_re-trace.csv') & !overwrite) {
-    return()
-  }
-  
-  # angle_deg <- seq(5,50,0.5)
-  # angle_rad <- (angle_deg / 180) * pi
-  # cos.a <- cos(angle_rad)
-  # sin.a <- sin(angle_rad)
-  # #slope <- sin.a / cos.a
-  # slope <- cos.a / sin.a
-  # 
-  # modeldf <- data.frame(angle_deg, angle_rad, cos.a, sin.a, slope)
-  # 
-  # Lx_MSE <- c()
-  # Ly_MSE <- c()
-  # LxLy_MSE <- c()
-  
-  internalMovements <- c(2,3,4)
-  externalMovements <- rev(c(.125, .167))
-  
-  #for (task in c('arrow','re-trace')) {
-  for (task in c('re-trace')) {
-    
-    # we will generate data files for both tasks, with these columns:
-    # except that there is no bound in the arrow task
-    participant <- c()
-    trial <- c()
-    internalspeed <- c()
-    internaldirection <- c()
-    externalspeed <- c()
-    fixationside <- c()
-    initialdirection <- c()
-    illusionstrength <- c()
-    boundX <-c()
-    boundY <-c()
-    boundXraw <- c()
-    boundYraw <- c()
-    
-    # ***************************
-    # 
-    
-    #pdf(file='doc/individual_participants_trials.pdf',width=12,height=8)
-    
-    #layout(matrix(c(1,2,3,4,5,6),byrow=TRUE,ncol=3,nrow=2))
-    
-    for (participant.idx in c(1:length(participants))) {
-      
-      # plot(-1000,-1000,
-      #      main=sprintf('participant %d',participants[participant.idx]),xlab='',ylab='',
-      #      xlim=c(-4.5,10),ylim=c(-.5,14),
-      #      bty='n',ax=F)
-      
-      # lines(x=c(0,0),y=c(0,13.5),col='blue')
-      
-      ppno <- participants[participant.idx]
-      
[…]

### 2021-11-13 R/exp2_retracing.R
renamed source file for tracing experiment, wrote a function to check some trajectories for participant 2

@@ -1,0 +1,3444 @@
+library('svglite')
+library('ez')
+library('segmented')
+
+source('R/common.R')
+#source('R/models.R')
+
+# Data handling -----
+
+oldPreProcessOnePass_V4 <- function(participants = c(2,3,4,5,6,8,9,10,11), overwrite=FALSE) {
+  
+  if (file.exists('data/onepass_V4/onePass_V4_re-trace.csv') & !overwrite) {
+    return()
+  }
+  
+  # angle_deg <- seq(5,50,0.5)
+  # angle_rad <- (angle_deg / 180) * pi
+  # cos.a <- cos(angle_rad)
+  # sin.a <- sin(angle_rad)
+  # #slope <- sin.a / cos.a
+  # slope <- cos.a / sin.a
+  # 
+  # modeldf <- data.frame(angle_deg, angle_rad, cos.a, sin.a, slope)
+  # 
+  # Lx_MSE <- c()
+  # Ly_MSE <- c()
+  # LxLy_MSE <- c()
+  
+  internalMovements <- c(2,3,4)
+  externalMovements <- rev(c(.125, .167))
+  
+  #for (task in c('arrow','re-trace')) {
+  for (task in c('re-trace')) {
+    
+    # we will generate data files for both tasks, with these columns:
+    # except that there is no bound in the arrow task
+    participant <- c()
+    trial <- c()
+    internalspeed <- c()
+    internaldirection <- c()
+    externalspeed <- c()
+    fixationside <- c()
+    initialdirection <- c()
+    illusionstrength <- c()
+    boundX <-c()
+    boundY <-c()
+    boundXraw <- c()
+    boundYraw <- c()
+    
+    # ***************************
+    # 
+    
+    #pdf(file='doc/individual_participants_trials.pdf',width=12,height=8)
+    
+    #layout(matrix(c(1,2,3,4,5,6),byrow=TRUE,ncol=3,nrow=2))
+    
+    for (participant.idx in c(1:length(participants))) {
+      
+      # plot(-1000,-1000,
+      #      main=sprintf('participant %d',participants[participant.idx]),xlab='',ylab='',
+      #      xlim=c(-4.5,10),ylim=c(-.5,14),
+      #      bty='n',ax=F)
+      
+      # lines(x=c(0,0),y=c(0,13.5),col='blue')
+      
+      ppno <- participants[participant.idx]
+      
[…]

### 2021-11-13 R/common.R
reorganized functions in sections for different kinds of tasks

@@ -1,125 +1,12 @@
 
 
+# useful extra operator:
 `%notin%` <- Negate(`%in%`)
 
 
-
-# 
-# # pixels per cm
-# 
-# # resolution is 1680 x 1050 pixels
-# # which spans ~43.5 x ~27 cm
-# # for width that is roughly 38.799 pixels per cm
-# # for depth that is roughly 38.888 pixels per cm
-# # on average that would be: 38.755 pixels per cm
-# 
-# PPC <- 38.754789272
-# 
-
-
-getPPCdva <- function(lo=3) {
-  
-  props <- list()
-  props['PPC'] <- 38.754789272 # pixels per cm
-  
-  # tan(A) = a/b
-  
-  dva <- expand.grid('eye.dist'  = seq(15, 25, length.out = lo), 
-                     'gabor.pos' = seq(-6.75, 6.75, length.out = lo))
-  
-  b <- 27 # distance between mirror and monitor
-  a <- dva$eye.dist + dva$gabor.pos # horizontal distance (along monitor) 
-  
-  # https://en.wikipedia.org/wiki/Trigonometry
-  A <- atan(a/b) # angle at centre of gabor
-  
-  a1 <- tan(A+((0.5/180)*pi))*b
-  a2 <- tan(A-((0.5/180)*pi))*b
-  
-  dva$cmpd <- a1 - a2
-  
-  ghw <- (100/3) / props[['PPC']]
-  
-  props[['gaborcm']] <- ghw
-  
-  dva$gabor.dva <- ghw / dva$cmpd
-  
-  props[['dva']] <- dva
-  
-  # visual size of the "workspace" (13.5 cm track)
-  dva.track <- expand.grid('eye.dist'  = seq(15, 25, length.out = lo))
-  
-  b <- 27
-  a <- dva.track$eye.dist
-  
-  A1 <- atan(a/(b-6.75)) # the angle of the close end at each head position
-  A2 <- atan(a/(b+6.75)) # the angle of the far end at each head position
-  
-  dva.track$dva <- ((A1-A2)/pi)*180
-  
-  props[['track']] <- dva.track
-  
-  return(props)
-  
-}
-
-
-getColors <- function() {
-  
-  colors <- list()
-  
-  colors[['blue']]      <- list('s'='#005de4ff', 't'='#005de42f')
-  
-  colors[['lightblue']] <- list('s'='#0fd2e2ff', 't'='#0fd2e22f')
-  
[…]

### 2021-11-13 exp/frameGrab.py
generate gabors at 24 phases with 10 pixel border around the 100x100 pixel gabor

@@ -12,15 +12,15 @@ def createExampleGabors():
   cfg['fullscr'] = False
   
   # create fullscreen, upside up window:
-  cfg = createWindow(cfg, resolution=[100,100])
+  cfg = createWindow(cfg, resolution=[120,120])
   cfg = createStimuli(cfg)
   
   # change some settings to increase the resolution:
   cfg['gabor'].size = 100
   cfg['gabor'].sf = 1.5/100
   
-  for phase in np.arange(0,1,45/360):
-    cfg['gabor'].phase = (phase, 0.0)
+  for phase in np.arange(0,350,15):
+    cfg['gabor'].phase = (phase/360., 0.0)
   
     # put on the screen:
     cfg['gabor'].draw()
@@ -28,7 +28,7 @@ def createExampleGabors():
   
     # grab the current frame, and save it to a file:
     cfg['win'].getMovieFrame()
-    cfg['win'].saveMovieFrames('gabor_%d.png'%(int(phase*360)))
+    cfg['win'].saveMovieFrames('background_gabor_%d.png'%(phase))
   
   # release window:
   cfg['win'].close()

### 2021-11-13 Manuscript.Rmd
use reorganized code, some revisions implemented

@@
 knitr::opts_chunk$set(echo = FALSE)
-# The functions here allow downloading the collected data from the OSF project:
+# First we load all custom scripts:
+
+# The function in this file downloads the data from the OSF project:
 source('R/data.R')
@@
 # The functions in here are used on the data of the bounded tracking experiment:
-source('R/tracking_bounded.R')
-# Here we compare the tracking data with an earlier paper:
-source('R/compare.R')
+source('R/exp1_tracking.R')
 # And then some code for the re-tracing experiment:
-source('R/retrace_onePass_V4.R')
+source('R/exp2_retracing.R')
 # As well as models fit to that data:
-source('R/models.R')
-# And finally, code to compare this data with Sirui's PSEs
-# source('R/time.R') # no longer used
+source('R/exp2_models.R')
 
@@
 # we pre-process the data:
-#getBoundedTrackingStandardizedSegments()
+
+# for exp 1, this segments the tracking trajectories
+# to get the middle 1 second for each full pass
 segmentDirections()
+
+# for exp 2, this extracts the reset points:
 preProcessOnePass_V4()
-# the rest of the paper should now compile without complaint
+
+# the rest of the paper should now run without complaint
 ```
@@
 1, right), as well as a keypad for additional responses.
-Stimuli were displayed on a downward facing LED monitor (30 Hz, 20”, 1680x1050), parallel with the drawing tablet.
+Stimuli were displayed on a downward facing LCD monitor (30 Hz, 20”, 1680x1050), parallel with the drawing tablet.
 The stimuli were observed via an upward facing mirror placed exactly between the tablet and monitor, so that the stimuli appear to be in the same horizontal plane as the hand, and the tip of the stylus.
@@
 
-The stimuli were phase-changing gabor patches with a gaussian mask with a sigma (sd) of 0.43 _cm_ or ~0.59 _dva_, spatial frequency of 0.58 _c/cm_ or ~0.42 _c/dva_, vertical orientation, and contrast of 100%.
+The stimuli were phase-changing gabor patches with a sigma (sd) of 0.43 _cm_ or ~0.59 _dva_, spatial frequency of 0.58 _c/cm_ or ~0.42 _c/dva_, vertical orientation, and Michelson contrast of 98.7%.
+The background luminance was 18.46 $cd/m^2$ and the mean luminance of the gabor deviated from the background by less than 1%.
 By simultaneously moving the gabor envelope and drifting its internal sine wave, it becomes a ‘double-drift stimulus’ where the perceived path deviates from the real path.
@@
 In Figure 2D, we compare mean illusion strength to that reported previously [@Heller2021].
-As can be seen the averages from our data here (red circles, purple line) coincide fairly well with the illusion strength predicted by the model fitted to all their data (Eq.
+As can be seen the averages from our data here (red circles) coincide fairly well with the illusion strength predicted by the model fitted to all their data (Eq.
 1, black line).
+If we fit the vector combination model, we get $K=0.738$ which is very close to $K=0.74$ reported by Heller et al.
+[-@Heller2021].
 That is; tracking the illusion while observing it does not change the strength of the illusion.
@@
 3).
-These resets may also occur spontaneously once the accumulation has gone on too long or too far.
+Other percepts of resets have been reported as well, but as far as we know these fall in between these two extremes.
+Resets may also occur spontaneously once the accumulation has gone on too long or too far, depending on what mechanism causes resets.
+Where resets occur would then both depend on the mechanism that causes resets, as well as on the strength of the illusion (purple arrows in Fig.
+3) which in turn depends on the stimulus properties as well as variation in perception within and between participants.
 The purpose of this experiment is to capture this spontaneous resetting and determine its source.
@@
 For some people this takes the form of a ‘return’ reset, back toward the true position, for others it means the illusory position remains at a fixed offset moving parallel to the real path, or some combination of the two.
-**Top right:** if these resets are due to a spatial limit on the size of the illusory position shift away from the true vertical path, the locations of the resets should have a constant X coordinate, the spatial offset from the vertical path.
+**Right:** Given varying strength of the illusion (denoted by the angle of the purple arrows), either due to stimulus properties or within or between subject differences, resets may occur at different points in space.
+**Top right:** if these resets can be explained by a spatial limit on the size of the illusory position shift away from the true vertical path, the locations of the resets should have a constant X coordinate; the spatial offset from the vertical path.
 **Bottom right:** if they occur instead after some period of time, resets should have a constant radius (R); the time since the gabor’s motion began."}
@@
 We were interested, in particular, in whether the limitation of the accumulation would be set by space or time.
-Since both the true and the illusory position appear to be available in the visual system (the true position drives saccades whereas the illusory position drives perception, [@Lisi2015], it is possible that resets occur once the distance between the real and perceived positions exceeds some limit.
+Since both the true and the illusory position appear to be available in the visual system (the true position drives saccades whereas the illusory position drives perception, @Lisi2015), it is possible that resets occur once the distance between the real and perceived positions exceeds some limit.
 We can speculate that this spatial limit would be related to the positional uncertainty of the gabor.
@@
 When the positional uncertainty is high, in the periphery, with the gabor’s mean luminance matched to the background [@Gurnsey2012; @Cavanagh2019], the spatial offset could be quite large before exceeding the range of positional uncertainty around the gabor’s true position.
-Under this hypothesis, the timing of spontaneous resets will depend on the spatial offset from the true path which will be directly proportional to the internal speed (for a constant external motion).
+Under this hypothesis, the occurrence of spontaneous resets will depend on the spatial offset from the true path which will be directly proportional to the internal speed (for a constant external motion).
 On the other hand, if spontaneous resets are the result of the temporal limitations of the integration process, resets would occur after a certain amount of time, independently of the internal speed and the spatial offset it creates.
@@
 Stimuli were identical to those used in the online tracking experiment, but now always started at the near end of the workspace and moved away from the participants.
-The double drift stimuli would use either 3 or 4 seconds to move 13.5 _cm_, corresponding to speeds of ~4.5 and ~3.75 _cm/s_ or ~4.7 and ~3.5 _dva/s_, and had an internal drift of 2, 3 or 4 _cps_ .
+The double drift stimuli would use either 3 or 4 seconds to move 13.5 _cm_, corresponding to speeds of ~4.5 and ~3.75 _cm/s_ or ~4.7 and ~3.5 _dva/s_,
+
+THOSE NUMBERS DON'T MAKE SENSE...
+
+and had an internal drift of 2, 3 or 4 _cps_ .
 
@@
 Each block used the 6 combinations of two external speeds ( corresponding to 3 or 4 _s_ presentation time ), and three internal speeds ( 2, 3 or 4 _cps_ ), 6 times, for a total of 36 trials per block.
-Each of the combinations of internal and external speed was presented 24 times in total.
+Each of the combinations of internal and external speed was presented 24 times in total, so that we have 144 trials in total.
 
@@
 Using a smooth spline interpolated version of the trajectory we detect right-left direction changes, based on sign changes.
-First, we remove any short segments on the trajectory at the start and end that go toward the participant.
-Then, we downsample the trajectories such that each successive sample is at least 9 _mm_ from the previous, and then applied a median filter with a half-width of 5 samples.
+First, we remove the first and last 4 mm of the trajectory, as well as any short segments on the trajectory at the start and end that go toward the participant, to remove any unintended jitters caused by starting and ending the movement.
+Then, we apply a bi-directional low-pass Butterworth filter with a cut-off frequency of 1.5 Hz.
+However, since participants draw at very different speeds, we first linearly interpolate the trajectories such that the sampling rate is normalized to 30 Hz for all trajectories, based on the stimulus duration.
 In these summary trajectories we detect the first inflection point using the X coordinates.
-We then use the raw trajectory again, first locating the corresponding sample for the inflection point, and then find the first previous sample at 97.5% of the X coordinate at the inflection point sample in the raw trajectory.
-Downsampling removes the small periodic signal in the trajectories recorded from participant XYZXYZXYZ that seem to capture the phase shifts of the gabor.
+This could put the reset point too late for hit-the-wall resets, so that we move back to get the previous sample at 95% of the X coordinate, and finally locate the sample in the raw trajectory that is closest to this point.
+Not all trajectories result in a reset point, so we may miss some genuine hit-the-wall resets where the trajectory does not have an inflection point.
+However, there is no simple way to segregate these from trials without a reset.
 
-################################
-################################
-################################
-################################
-################################
-################################
-################################
-################################
-################################
-################################
-
-The first (if any) of the remaining direction changes, are then taken as reset points.
-To keep the horizontal and vertical (straight ahead) coordinates of the reset points isometric they are given in centimeters.
-The distance from origin can then be converted to time as 1 _cm_ = 3.375 _s_ in the 4 _s_ condition and 1 _cm_ = 4.5 _s_ in the 3 _s_ condition.
-Participants' ability to retrace perceived trajectories might vary, but normalizing trajectories so that the final Y coordinate is at 13.5 _cm_ does not noticably change the results.
+To facilitate modeling, the horizontal and vertical (straight ahead) coordinates of the reset points are kept isometric by giving them in centimeters.
+The distance from origin can then be converted to time as 1 _cm_ = 4/13.5 _s_ in the 4 _s_ condition and 1 _cm_ = 3/13.5 _s_ in the 3 _s_ condition.
 
@@
 We take a trajectory sample halfway between the starting point and the detected reset point.
-We then use the angular difference between the gabor's real trajectory and a straight line drawn through that point and the start of the trajectory as a measure of illusion strength.
+We then use the angular difference between the gabor's real trajectory and a straight line drawn through that point and the start of the trajectory as a measure of illusion strength (see Fig.
+4).
 
@@
 (Purple lines: trajectories up to resets, gray lines: trajectories after resets, blue lines: trajectories that had no resets in the re-tracing.) Other participants' data were not this clear.
-**Right:** Distribution of all 859 detected reset points."}
+**Right:** Distribution of all 858 used reset points from 8/9 participants."}
 #plotExampleData()
@@
 
-On average, this method detects resets in ~75% of trials.
-For three participants, only about half the trials showed resets (42%, 44% and 58%).
-For the remaining six participants the algorithm detected resets in 72%-99% of trials, with an average of 88%.
-The minimum number of resets in each of the six conditions for any participant is 7 (out of 24 trials), with an average of ~17.
+On average, this method detects resets in ~68% of trials.
+For two participants, less than half the trials showed resets (9% and 22%).
+For the remaining seven participants the algorithm detected resets in 65%-99% of trials, with an average of 82%.
 Since participants experiencing a "hit-the-wall reset" (Fig.
-3) as opposed to a "jump" reset, might not always display a detectable direction change in their paths, 50% of trials with resets seems reasonable.
-Either type of perceived resets appears valid, and the average reset positions still indicates some limit to the illusion.
-Therefore, we kept all participants and resets in our analysis (reset points in Fig.
+3) as opposed to a "jump" reset, might not always display a detectable direction change in their paths, most of these percentages of trials with resets seems reasonable.
+However, the participant with 9% resets will not be used for figures and statistics, but we have verified that excluding their data does not change the results.
+This means we analyse a total of 858 reset points (Fig.
 4, right).
@@
 As before, we compare the strength of the illusion (see Fig 5, left), with an earlier study [@Heller2021].
-A line through the origin with slope 0.51 ($K$ from Eq.
-1) predicts illusion strength fairly well in these data, which is weaker than the $K=0.74$ found by Heller et al.
-[-@Heller2021] and in Experiment 1 with online tracking.
+A line through the origin with $K=0.54$ predicts illusion strength fairly well in these data, which is weaker than the $K=0.74$ found by Heller et al.
+[-@Heller2021].
+Illusion strength is also lower than in Experiment 1 with online tracking and it has a fair amount of between-subject variation in this task (blue shaded area is 95% confidence interval across participants).
 While the illusion is less strong, it is nevertheless there and depends on the relationship between internal and external speed as in previous studies.

### 2021-11-19 R/common.R
moved partial eta squared function to common.R

@@ -325,9 +325,9 @@ getGaborLuminanceValues <- function(mask=NULL) {
   
 }
 
-# # # # # # # # # # # # # #
-# statistics function -----
-# # # # # # # # # # # # # #
+# # # # # # # # # # # # # # #
+# statistics functions  -----
+# # # # # # # # # # # # # # #
 
 getConfidenceInterval <- function(data, variance = var(data), conf.level = 0.95, method='t-distr', resamples=1000, FUN=mean) {
   
@@ -426,6 +426,28 @@ confidenceEllipse <- function(x, y=NA, interval=.95, vectors=100) {
   
 }
 
+#' add partial eta squared to ezANOVA output
+#'
+#'
+#' @param ezANOVAResult ezANOVA result
+#' @return ezANOVA with partial eta-squared
+#' @details ezANOVA must be run with parameter 'detailed=T'.
+#' @author Frank Papenmeier
+#' @export
+#'
+ezANOVA.pes <- function (ezANOVAResult)
+{
+  if (is.null(ezANOVAResult$ANOVA$SSn) | is.null(ezANOVAResult$ANOVA$SSd))
+  {
+    stop("ezANOVA must be run with parameter 'detailed=T'.")
+  }
+  
+  ezANOVAResult$ANOVA$pes <- ezANOVAResult$ANOVA$SSn/(ezANOVAResult$ANOVA$SSn+ezANOVAResult$ANOVA$SSd)
+  
+  return(ezANOVAResult)
+}
+
+
 # # # # # # # # # # # # # # # # # # # # # 
 # extended plotting functionality -----
 # # # # # # # # # # # # # # # # # # # # # 
@@ -440,7 +462,7 @@ getColors <- function(transparency=81.5) {
   colors$blue[['t']] <- t_col(colors$blue$s,percent=transparency)
   
   colors[['lightblue']] <- list('s'='#0fd2e2ff')
-  colors$lightblue[['t']] <- t_col(colors$blue$s,percent=transparency)
+  colors$lightblue[['t']] <- t_col(colors$lightblue$s,percent=transparency)
   
   colors[['yorkred']]   <- list('s'='#e51636ff')
   colors$yorkred[['t']] <- t_col(colors$yorkred$s,percent=transparency)

### 2021-11-19 R/exp1_tracking.R
added confidence intervals of the mean to the heading plot

@@ -589,11 +589,13 @@ plotBoundedTracking <- function(target='inline', discontinuous=FALSE) {
   title(xlab='internal speed [cps]', line=2.4)
   title(main='C: heading', font.main=1, cex.main=1.5, adj=0, line=0.5)
   
+  headingCIs <- aggregate(heading ~ internalSpeed, data=heading[['statistics']], FUN=getConfidenceInterval, conf.level=0.95)
+  
   for (internalSpeed.idx in c(1:length(internalSpeeds))) {
     
     internalSpeed <- internalSpeeds[internalSpeed.idx]
 
-    # 2/3 of heading samples falls in this interval:
+    # 90% of ALL heading samples falls in this interval:
     interval <- unlist(heading[['interval']][which(heading[['interval']]$internalSpeed == internalSpeed),])
     lo <- interval[2]
     hi <- interval[3]
@@ -601,6 +603,13 @@ plotBoundedTracking <- function(target='inline', discontinuous=FALSE) {
     pX <- c( cos(arc) * 0.5, cos(rev(arc)) * 1.75 ) + internalSpeed.idx
     pY <- c( sin(arc) * 0.5, sin(rev(arc)) * 1.75 ) + 0.75
     polygon(pX,pY,border=NA,col=colors$blue$t)
+    # 95% CI across participants:
+    ci95 <- as.vector(unlist(headingCIs[which(headingCIs$internalSpeed == internalSpeed),c(2)]))
+    arc <- seq(ci95[1],ci95[2],length.out=25)
+    pX <- c( cos(arc) * 0.5, cos(rev(arc)) * 1.75 ) + internalSpeed.idx
+    pY <- c( sin(arc) * 0.5, sin(rev(arc)) * 1.75 ) + 0.75
+    polygon(pX,pY,border=NA,col=colors$blue$t)
+    
     
     # now the average trajectory:
     avgDir <- heading[['average']]$heading[which(heading[['average']]$internalSpeed == internalSpeed)] 
@@ -648,6 +657,7 @@ plotBoundedTracking <- function(target='inline', discontinuous=FALSE) {
   angles <- c(0,pi/4)
   # Heller et al. (2021):
   lines(angles,0.74*((angles/pi)*180),col='black',lty=1)
+  lines(angles,0.60*((angles/pi)*180),col='black',lty=2)
   
   # individual participants
[…]

### 2021-11-19 R/exp2_models.R
standardized all parameters in likelihood-based models, added code to plot the simplest models

@@ -730,7 +730,7 @@ compareAllModels <- function(N=9,AICfun=AIC,verbosity=1) {
 
 #source('R/retrace_onePass_V4.R')
 
-getData54 <- function(illusionMinimum=5) {
+getData54 <- function(illusionMinimum=0) {
   
   df <- summarizeTraceBoundsV4()
   
@@ -769,7 +769,7 @@ getData54 <- function(illusionMinimum=5) {
   
 }
 
-getDataTrials <- function(illusionMinimum=5, illusionMaximum=85) {
+getDataTrials <- function(illusionMinimum=0, illusionMaximum=90) {
   
   # load re-trace resets and illusion strengths
   df <- read.csv('data/onePass_V4/onePass_V4_re-trace.csv', stringsAsFactors = F)
@@ -1349,8 +1349,8 @@ ToffsetGaussianLikelihood <- function(par,data) {
   
   # time limit
   
-  mT <- par['mT']
-  sT <- par['sT']
+  mT <- par['mTn']
+  sT <- par['sTn']
   
   L <- (1 / (sT * sqrt(2 * pi)) ) * exp( -0.5 * ( ( data$RT - mT) / sT )^2 )
   
@@ -1362,8 +1362,8 @@ XoffsetGaussianLikelihood <- function(par,data) {
   
   # space limit
   
-  mX <- par['mX']
-  sX <- par['sX']
+  mX <- par['mXn']
+  sX <- par['sXn']
   
   L <- (1 / (sX * sqrt(2 * pi)) ) * exp( -0.5 * ( (data$X-mX) / sX )^2 )
   
@@ -1380,9 +1380,9 @@ XoffsetGaussianLikelihood <- function(par,data) {
 
 XoffsetGammaLikelihood <- function(par, data) {
   
-  sX <- par['sX']
-  rX <- par['rX']
-  oX <- par['oX']
+  sX <- par['sXg']
+  rX <- par['rXg']
+  oX <- par['oXg']
   
   L <- dgamma( data$X - oX, shape=sX, rate=rX )
   # log(0) will be -Inf, not a good probability to maximize
@@ -1394,9 +1394,9 @@ XoffsetGammaLikelihood <- function(par, data) {
 
 ToffsetGammaLikelihood <- function(par, data) {
   
-  sT <- par['sT']
-  rT <- par['rT']
-  oT <- par['oT']
+  sT <- par['sTg']
+  rT <- par['rTg']
+  oT <- par['oTg']
   
   # hmmm... this is actually done in centimeters ?
[…]

### 2021-11-19 R/exp2_retracing.R
added angular cutoff / tolerance to main reset point detection function, moved stats function out, create new figure to illustrate reset points and illusion strength

@@ -483,7 +483,8 @@ preProcessOnePass_V4 <- function(participants = c(2,3,4,5,6,8,9,10,11),
 #                                 RPFUN=get97.5percMaxX1_resetpoint,
                                  RPFUN=getResampledFilteredInflectResetPoint,
                                  IDFUN=getDirectionHalfway,
-                                 verbosity=0) {
+                                 verbosity=0,
+                                 angularcutoff=0.0) {
   
   
 
@@ -546,12 +547,18 @@ preProcessOnePass_V4 <- function(participants = c(2,3,4,5,6,8,9,10,11),
       
       # GET THE RESET POINT
       # which function is used to get the reset point is a function argument:
-      resetPoint <- RPFUN(x,y,t,verbosity=verbosity, internalspeed=trialdf$internalMovement[1], externalspeed=trialdf$externalMovement[1])
+      resetPoint <- RPFUN(x,
+                          y,
+                          t,
+                          verbosity=verbosity, 
+                          internalspeed=trialdf$internalMovement[1], 
+                          externalspeed=trialdf$externalMovement[1], 
+                          minimum=c('x'=0.5,'y'=0.5),
+                          angularcutoff=angularcutoff)
       
       # store the reset point in the output data vectors:
       resetX <-c(resetX, resetPoint$X)
       resetY <-c(resetY, resetPoint$Y)
-      
       
       # GET THE INITIAL DIRECTION
       # the function is also an argument to this function:
@@ -612,7 +619,7 @@ cmTrace <- function(trialdf) {
 }
 
 
-get97.5percMaxX1_resetpoint <- function(x,y,t,verbosity=0,internalspeed=NA,externalspeed=NA,returnTrajectory=FALSE) {
+get97.5percMaxX1_resetpoint <- function(x,y,t,verbosity=0,internalspeed=NA,externalspeed=NA,returnTrajectory=FALSE,minimum=c('x'=NULL,'y'=NULL),angularcutoff=NULL) {
   
   # x <- pdf$x
   # y <- pdf$y
[…]

### 2021-11-19 Manuscript.Rmd
work on revisions

@@
 To report the perceived trajectory of the gabor, participants drew it after the gabor was removed from the monitor.
-Resets were detected and analyzed but their distribution neither matches a strict temporal or spatial limit on perception that could trigger resets.
+Resets were detected and analyzed but their distribution neither matches a temporal nor a spatial limit on perception that could trigger resets.
 
@@
 The perceived location continues to move away from its physical position by up to several times the size of the gabor.
-The strength of the illusion, the deviation of perceived path of the gabor from its true path, depends on the speed of both the internal and external motion [@Tse2006; @Shapiro2010; @Kwon2015; @Cavanagh2019; @Heller2021].
-Rather than use point estimates of the perceived location of the stimulus, we want to test how well a continuous trajectory can measure the strength of the illusion, or if perception is changed by this response modality such as in other illusions
 
-CITE SOME PREVIOUS WOKR HERE.
+The strength of the illusion; the deviation of perceived path of the gabor from its true path, depends on the speed of both the internal and external motion [@Tse2006; @Shapiro2010; @Kwon2015; @Cavanagh2019; @Heller2021].
+Rather than use point estimates of the perceived location of the stimulus, we want to test how well a continuous trajectory can measure the strength of the illusion.
+Eye and hand movement are differentially sensitive to many illusions, such as the Ponzo illusion [@Gamble2017], Ebbinghaus illusion [@Knol2017], flash-lag effect [@Blohm2003], the Duncker illusion [@Soechting2001] and many others.
+Since it has been shown that the double-drift illusion affects pointing movements, but not saccades [@Lisi2017] we expect that hand trajectories will also be sensitive to the illusion.
 
@@
 We compare trajectories across 5 different internal motions of the gabor and compare the illusion strength with previous work based on a straightforward vector combination model [@Heller2021].
-As we expected [@mhjh], online tracking provides a similar measure of illusion strength as pointing.
+As we expected, online tracking provides a similar measure of illusion strength as pointing.
 
@@
 At some point, the illusion stops accumulating.
-The accumulation can be reset by introducing a temporal break [@Lisi2015] or by distraction attention [@Nakayama2020], but it may also reset spontaneously once the accumulation has gone on too long or too far.
+The accumulation can be reset by introducing a temporal break [@Lisi2015] or by distracting attention [@Nakayama2020], but it may also reset spontaneously once the accumulation has gone on too long or too far.
 Informally, some observers have reported that the gabor’s path may saturate so that it remains parallel to, but offset from the true path.
@@
 
+In particular, we compare the illusion strengths we find with those reported by Heller et al.
+[-@Heller2021].
+They show a double-drift stimulus with 36 combinations of internal and external motion for 500 ms and then have participants indicate on a ring where they perceived the path of the stimulus to intersect with this ring.
+While they hypothesize that illusion strength decreases with faster external motion (perhaps because of resets or limits on perception), here we will mostly use their overall estimate of illusion strength, given by $K=0.74$.
+
 All analyses were done in R 3.6.1 [@Rproject]. All data, as well as scripts for the experiments and analyses are available on OSF: [https://osf.io/72ndu/](https://osf.io/72ndu/).
@@
 **Left:** At some point, the perceived position may stop moving further away from its actual path.
-For some people this takes the form of a ‘return’ reset, back toward the true position, for others it means the illusory position remains at a fixed offset moving parallel to the real path, or some combination of the two.
-**Right:** Given varying strength of the illusion (denoted by the angle of the purple arrows), either due to stimulus properties or within or between subject differences, resets may occur at different points in space.
+For some people this takes the form of a ‘return’ reset, back toward the true position either suddenly or slowly, for others it means the illusory position remains at a fixed offset moving parallel to the real path.
+**Right:** Given varying strength of the illusion (denoted by the angle of the purple arrows), e.g.
+due to stimulus properties or within or between subject differences, resets may occur at different points in space.
 **Top right:** if these resets can be explained by a spatial limit on the size of the illusory position shift away from the true vertical path, the locations of the resets should have a constant X coordinate; the spatial offset from the vertical path.
@@
 All participants reported to be right handed and had either normal or corrected to normal vision.
-Procedures were in accordance with the Declaration of Helsinki (2013) and were approved by York’s Human Participants Review Committee.
+Procedures were in accordance with the Declaration of Helsinki (2003) and were approved by York’s Human Participants Review Committee.
 All participants provided prior, written, informed consent.
@@
 Each block used the 6 combinations of two external speeds ( corresponding to 3 or 4 _s_ presentation time ), and three internal speeds ( 2, 3 or 4 _cps_ ), 6 times, for a total of 36 trials per block.
-Each of the combinations of internal and external speed was presented 24 times in total, so that we have 144 trials in total.
+Each of the combinations of internal and external speed was presented 24 times in total, so that we have 144 trials for each participant, and 1296 trials in total.
 
@@
 This could put the reset point too late for hit-the-wall resets, so that we move back to get the previous sample at 95% of the X coordinate, and finally locate the sample in the raw trajectory that is closest to this point.
-Not all trajectories result in a reset point, so we may miss some genuine hit-the-wall resets where the trajectory does not have an inflection point.
-However, there is no simple way to segregate these from trials without a reset.
+Finally, we remove a few reset points that are detected very close to the veridical path without any build-up of the illusion by only keeping reset points that fell 5 mm within the quadrant where we expect them.
+This heuristic will result in some misses and some false alarms, and while changing the algorithm may improve performance, there is no ground truth to evaluate this.
+All raw trajectories and the detected reset points can be found on the OSF repository (XXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXX add link?).
 

### 2021-11-19 DoubleDriftResetLimits.Rmd
tried out new data download function in notebook

@@
 # The functions in here are used on the data of the bounded tracking experiments:
-source('R/tracking_bounded.R')
+source('R/exp1_tracking.R')
 # And finally some code for the re-tracing experiment...
-source('R/retrace_onePass_V4.R')
+source('R/exp2_retracing.R')
 # As well as models fit to that data:
-source('R/models.R')
-# Comparison with Sirui Liu's data:
-source('R/time.R')
+source('R/exp2_models.R')
 ```
@@
 # from R/data.R
-downloadRawData(overwrite=FALSE, experiments=c('bounded_tracking', 'bounded_tracking_V2', 'onePass_V4'))
+OSFdata()
 ```
@@
 retrace$value <- 1
-retrace$value[which(is.na(retrace$boundX))] <- 0
+retrace$value[which(is.na(retrace$resetX))] <- 0
 resetcounts <- aggregate(value ~ participant, data=retrace, FUN=sum)
@@
 print(resetcounts)
+#print(mean(retrace$value))
 ```
@@
 ```{r}
-resetcountsCOND <- aggregate(value ~ participant + internalspeed + externalspeed, data=retrace, FUN=sum)
+resetcountsCOND <- aggregate(value ~ participant + abs(internalspeed) + externalspeed, data=retrace, FUN=sum)
 print(resetcountsCOND)

### 2021-11-20 R/exp2_models.R
make fitting and plotting functions standalone and ready for paper

@@ -2369,7 +2369,7 @@ plotUncoupledModels <- function(df, target='inline') {
 # which is what the orthogonal and normal models compare (it is the same)
 # and then there is an offset gamma distribution for comparison
 
-fitJointModel <- function(df) {
+fitSomeModels <- function(df, jointModel=FALSE, Xnormal=TRUE, Xgamma=TRUE, Tgamma=TRUE) {
   
   outputlist <- list()
   
@@ -2438,98 +2438,149 @@ fitJointModel <- function(df) {
   
   
   # **********************************
-  # the X-Normal distributions
+  # the X-Normal distribution
   # **********************************
   
-  # create search "grids":
-  mX = seq(0, 8, length.out = 16)
-  sX = seq(0, 8, length.out = 16)
-  
-  # make them into data frames:
-  searchgridXlim <- expand.grid('mXn'=mX, 'sXn'=sX)
-  
-  # get Likelihoods for points in search grid:
-  expXlimLLs <- apply(searchgridXlim,FUN=resetLogLikelihood,MARGIN=c(1),data=df,fitFUN=XoffsetGaussianLikelihood)
-  
-  # get 5 best points in the grid:
-  topgridXlim <- searchgridXlim[order(expXlimLLs, decreasing = TRUE)[c(1,5,9)],]
-  
-  control <- list( 'maximize' = TRUE )
-  
-  # print(topgridXlim)
-  
-  # do the actual fitting:
-  allXlimFits <- do.call("rbind",
-                         apply( topgridXlim,
-                                MARGIN=c(1),
-                                FUN=optimx,
-                                fn=resetLogLikelihood,
-                                method=c('nlminb'),
-                                lower=c( 0.0, 0.0 ),
-                                upper=c(13.5, 13.5),
-                                control=control,
-                                data=df,
-                                fitFUN=XoffsetGaussianLikelihood,) )
-  
-  
-  
-  # pick the best fit:
-  winXlimFit <- allXlimFits[order(allXlimFits$value, decreasing = TRUE)[1],]
-  
-  winXparN <- unlist(winXlimFit[c('mXn','sXn')])
-  
-  winXvalN <- as.numeric(winXlimFit$value)
-  names(winXvalN) <- c('logL')
-  
-  
-  outputlist[['XdistNormal']]=list('par'=winXparN,'logL'=winXvalN)
-  
+  if (Xnormal) {
+    
+    # create search "grids":
[…]

### 2021-11-20 R/common.R
function to create a linear colormap betweenn two colors in RGB space

@@ -571,7 +571,28 @@ t_col <- function(color, percent = 50, name = NULL) {
 }
 ## END
 
-
+linPal <- function(from='#FFFFFF', to='#E51636', n=256, alpha=NULL) {
+  
+  from <- col2rgb(from,alpha=TRUE)/255
+  to   <- col2rgb(to,alpha=TRUE)/255
+  
+  R <- seq(from[1],to[1],length.out = n)
+  G <- seq(from[2],to[2],length.out = n)
+  B <- seq(from[3],to[3],length.out = n)
+  A <- seq(from[4],to[4],length.out = n)
+  
+  if (!is.null(alpha)) {
+    if (length(alpha) == 1) {
+      A <- rep(alpha,n)
+    }
+    if (length(alpha) == 2) {
+      A <- seq(alpha[1],alpha[2],length.out = n)
+    }
+  }
+  
+  return(rgb(red=R, green=G, blue=B, alpha=A))
+  
+}
 
 
 

### 2021-11-20 Manuscript.Rmd
fixed reference

@@
 It seems that continuous tracking is sensitive to the strength of the illusion, and this is confirmed by a repeated-measures ANOVA on the average angles, using internal motion as a within-subjects factor ( _F_(4,12)=102.6, _p_<.001, $\eta^2$=0.97 ).
-With 4 FDR-corrected [@benjamini1995], paired t-tests, we find a difference in heading between internal motion speeds of 0 _cps_ and both 1 _cps_ ( _p_=.022) and -1 _cps_ ( _p_=.025) as well as between -3 and -1 _cps_ and 3 and 1 _cps_ (both _p_ <.001).
+With 4 FDR-corrected [@Benjamini1995], paired t-tests, we find a difference in heading between internal motion speeds of 0 _cps_ and both 1 _cps_ ( _p_=.022) and -1 _cps_ ( _p_=.025) as well as between -3 and -1 _cps_ and 3 and 1 _cps_ (both _p_ <.001).
 This makes it clear that continuous tracking does not prevent the illusion from occurring.
@@
 #plotResetPoints()
-#plotModels()
+plotModels()
 #plotBootstrappedFits()
-plotIllusionStrength()
+#plotIllusionStrength()
 ```

### 2021-11-24 R/exp3_gaborjumps.R
add code to analyse 3rd experiment

@@ -1,0 +1,201 @@
+
+# do the psychophysics -----
+
+getPSEs <- function(groups=c('exp','ctrl'), overwrite=FALSE) {
+  
+  for (group in groups) {
+    
+    # check if the file is already there, skip if it is:
+    if (file.exists(sprintf('data/gaborjump/%s_PSEs.csv', group))) {
+      next
+    }
+    
+    # extracted data:
+    df <- read.csv(sprintf('data/gaborjump/gaborjump_%s_data.csv',group), stringsAsFactors = FALSE)
+    
+    # invert leftward trials:
+    ext1 <- which(df$exmotion == 1)
+    df$gapshift[ext1] <- -1 * df$gapshift[ext1]
+    df$response[ext1] <- -1 * df$response[ext1]
+    
+    # we want to scale between 0 and 1:
+    df$response[which(df$response == -1)] <- 0
+    
+    # get mean proportion of right-ward responses per condition:
+    agg_df <- aggregate(response ~ participant + path_length + stimdur + exmotion + gapshift, data=df, FUN=mean)
+    
+    conditions <- expand.grid('path_length'=unique(agg_df$path_length),
+                              #'exmotion'=unique(agg_df$exmotion),
+                              'stimdur'=unique(agg_df$stimdur)
+    )
+    
+    
+    participant <- c()
+    path_length <- c()
+    stimdur     <- c()
+    #exmotion    <- c()
+    PSE         <- c()
+    std         <- c()
+    
+    
+    # need to know the participants, so we can analyze their data separately:
+    participants <- unique(agg_df$participant)
+    
+    # 12 PSEs per participant:
+    #layout(mat=matrix(c(1:12),ncol = 4,nrow = 3, byrow = TRUE))
+    
+    gapshifts_df <- data.frame('gapshift'=seq(-4,4,length.out = 161))
+    
+    for (ppno in participants) {
+      
+      p_df <- agg_df[which(agg_df$participant == ppno),]
+      
+      for (c.row in c(1:dim(conditions)[1])) {
+        c_df <- df[which(df$participant == ppno &
+                           df$path_length == conditions$path_length[c.row] &
+                           #df$exmotion    == conditions$exmotion[c.row] &
+                           df$stimdur     == conditions$stimdur[c.row]),]
+        #print(c_df)
+        #model <- glm(response ~ gapshift, data=c_df, family = binomial(link="probit"))
[…]

### 2021-11-24 R/data.R
download data for 3rd experiment

@@ -14,7 +14,7 @@ OSFdata <- function(overwrite=FALSE, removezip=FALSE, extradata=FALSE) {
   
   datadir <- osfr::osf_ls_files(dirs[idx,])
   
-  zipfiles <- c('bounded_tracking.zip','onepass_V4.zip','luminance.zip')
+  zipfiles <- c('bounded_tracking.zip','onepass_V4.zip','luminance.zip','gaborjump.zip')
   if (extradata) {
     zipfiles <- c(zipfiles, 'infinite_tracking.zip', 'pilots2.zip', 'CVRdemo.zip')
   }

### 2021-11-24 R/exp2_models.R
check X/T limit and gamma models for each participant

@@ -2902,6 +2902,82 @@ plotModels <- function(target='inline') {
   
 }
 
+AIClL <- function(lL, k) {
+  return((2*k) - (2*lL))
+}
+
+fitParticipantModels <- function() {
+  
+  df <- getDataTrials()
+  
+  participant <- c()
+  
+  XlimMSE <- c()
+  XlimAIC <- c()
+  XlimP   <- c()
+  TlimMSE <- c()
+  TlimAIC <- c()
+  TlimP   <- c()
+  XgamLik <- c()
+  XgamAIC <- c()
+  XgamP   <- c()
+  TgamLik <- c()
+  TgamAIC <- c()
+  TgamP   <- c()
+  
+  for (ppno in c(unique(df$participant),-1)) {
+    
+    if (ppno > 0) {
+      participant <- c(participant, sprintf('%d',ppno))
+      pdf <- df[which(df$participant==ppno),]
+    } else {
+      participant <- c(participant, 'all')
+      pdf <- df
+    }
+    
+    p_fits <- fitSomeModels(df=pdf, Xnormal = FALSE )
+    
+    X_MSE <- p_fits$XlimOrth$MSE
+    T_MSE <- p_fits$TlimOrth$MSE
+    
+    AICs <- AICc(MSE=c(X_MSE, T_MSE), k=c(2,2), N=6)
+    limLL <- relativeLikelihood(AICs)
+    
+    # print(AICs)
+    # print(limLL)
+    
+    XlimMSE <- c(XlimMSE, X_MSE)
+    XlimAIC <- c(XlimAIC, AICs['Lx'])
+    TlimMSE <- c(TlimMSE, T_MSE)
+    TlimAIC <- c(TlimAIC, AICs['Lt'])
+    XlimP <- c(XlimP, limLL['Lx'])
+    TlimP <- c(TlimP, limLL['Lt'])
+    
+    X_lL <- as.numeric(p_fits$XdistGamma$logL)
+    T_lL <- as.numeric(p_fits$TdistGamma$logL)
+    
+    logL <- c('Xll'=X_lL, 'Tll'=T_lL)
+    AICs <- AIClL(logL, k=c(2,2))
+    gamLL <- relativeLikelihood(AICs)
+    
+    #print(AICs)
+    #print(limLL)
+    
+    XgamLik <- c(XgamLik, X_lL)
[…]

### 2021-11-24 R/exp2_retracing.R
change figure dimensions to work with new figure

@@ -1194,10 +1194,10 @@ plotResetPoints <- function(target='inline') {
   #source('R/models.R')
   
   if (target == 'pdf') {
-    cairo_pdf(filename='doc/Fig4_trajectories_reset_points.pdf',onefile=TRUE,width=6,height=5)
+    cairo_pdf(filename='doc/Fig4_trajectories_reset_points.pdf',onefile=TRUE,width=6,height=6)
   }
   if (target == 'svg') {
-    svglite(file='doc/Fig4_trajectories_reset_points.svg',width=6,height=5)
+    svglite(file='doc/Fig4_trajectories_reset_points.svg',width=6,height=6)
   }
   
 

### 2021-11-24 Manuscript.Rmd
revisions

@@
 - Bernard Marius 't Hart 1,*
+- Sirui Liu 2
+- Peter Ulrich Tse 2
 - Denise Y.P. Henriques 1
@@
 
+# new addition with more classic psychophysics on perceived offsets:
+source('R/exp3_gaborjumps.R')
+
 # we download the data:
@@
 
+# for exp 3, we fit psychometric functions to get PSEs:
+getPSEs()
+
 # the rest of the paper should now run without complaint
@@
 _1: Centre for Vision Research, York University, Toronto, ON M3J1P3, Canada_\
-_2: Department of Psychology, Glendon College, Toronto, ON M4N3M6, Canada_\
-_3: Department of Psychological and Brain Sciences, Dartmouth College, Hanover, NJ 03755, USA_\
+_2: Department of Psychological and Brain Sciences, Dartmouth College, Hanover, NJ 03755, USA_\
+_3: Department of Psychology, Glendon College, Toronto, ON M4N3M6, Canada_\
 
@@
 We test if tracking the illusion continuously is a good measure of illusion strength, or if it prevents or decreases the illusion.
+
+Potential extra citation: [@PatricioDecima2022].
 
@@
 Distance from origin is time, angle is the heading angle.
-Dark blue lines show the average heading.
-In light blue is the average area (across participants) containing 90% of the heading data.
+Dark blue lines show the average heading over time.
+The light blue depicts the average area (across participants) containing 90% of the heading data.
+The darker blue areas depict the 95% confidence interval of the mean heading.
 **D:** The strength of the illusion in degrees angle deviation from the physical path.
@@
 Stimuli were identical to those used in the online tracking experiment, but now always started at the near end of the workspace and moved away from the participants.
-The double drift stimuli would use either 3 or 4 seconds to move 13.5 _cm_, corresponding to speeds of ~4.5 and ~3.75 _cm/s_ or ~4.7 and ~3.5 _dva/s_,
-
-THOSE NUMBERS DON'T MAKE SENSE...
-
-and had an internal drift of 2, 3 or 4 _cps_ .
+The double drift stimuli would use either 3 or 4 seconds to move 13.5 _cm_, corresponding to speeds of ~4.5 and ~3.75 _cm/s_ or ~4.7 and ~3.5 _dva/s_, and had an internal drift of 2, 3 or 4 _cps_ .
 
@@
 
-```{r fig4, echo=FALSE, message=FALSE, fig.width=7.5, fig.height=4, fig.cap="\\label{fig:fig4}**Figure 4:** Re-tracing Data.
-**Left:** Reset points are taken at the first sample in a splined-smoothed re-traced trajectory that goes in the direction opposite that of the expected illusion.
-Orange and gray lines: example trajectory.
-Orange dot: reset point.
-Red dot: point at half the distance to the reset point.
-Angle a: measure of illusion strength.
-**Middle:** One participant's trajectories with very clear 'return' resets at the purple dots.
-(Purple lines: trajectories up to resets, gray lines: trajectories after resets, blue lines: trajectories that had no resets in the re-tracing.) Other participants' data were not this clear.
-**Right:** Distribution of all 858 used reset points from 8/9 participants."}
+```{r fig4, echo=FALSE, message=FALSE, fig.width=6, fig.height=6, fig.cap="\\label{fig:fig4}**Figure 4:** Re-tracing Data.
+**A-D:** Example trajectories and reset points.
+All trajectory samples in transparent light blue (denser samples result in darker color).
+Dark red lines are, pruned, interpolated and filtered trajectories (see Methods).
+Purple circles indicate reset points.
+**A:** Trajectory with several sudden resets.
+**B:** 'Hit-the-wall' reset.
+**C:** Slow return reset.
+**D:** Continued accumulation of the illusion.
+**E:** Distribution of all 827 detected reset points.
+**F:** Using the angle to a point along the trajectory at half the distance to the reset point as a measure of illusion strength, we can compare this data to other data.
+The illusion is weaker ($K=0.60$, blue line) and its strength varies between participants, but it is nevertheless clearly there."}
 #plotExampleData()
@@
 Now we look at the distribution of reset points, and whether they can be explained by a temporal or spatial limit.
-In Figure 4 (right), we can see that the reset points vary more along the y-axis, than along the x-axis.
-That is, it looks like predicting reset points from a single X coordinate (or spatial limit) than predicing reset points from a given time.
+In Figure 4E, we can see that the reset points vary more along the y-axis, than along the x-axis.
+This would be consistent with a spatial limit to the illusion resulting in resets.
 However, there is also considerable variance, so that fitting any line through the data, will necessarily leave errors.
 
-```{r fig5, echo=FALSE, message=FALSE, fig.width=7.5, fig.height=4, fig.cap="\\label{fig:fig5}**Figure 5:** Illusion strength and reset models.
-**Left:** Illusion strength as in Figure 2 (bottom right); plotted against predicted strength from Eq.1: the arctangent of the ratio of internal to external velocities.
-Purple dots are the group means from 3 s passes, orange from 4 s passes (2, 3 and 4 _cps_ internal speeds).
-Predictions from the vector combination model in Eq.
-1: black line; $K=0.74$ as in Heller et al.
-(2021), light blue line; best fit here at $K=0.51$ and the shaded area shows a 95% confidence interval.
-**Right:** Model limits shown here come from a weighted bootstrapping across participants, lines are median parameter values, shaded areas the 95% confidence intervals of the limits.
-Gray dots are the reset points from all participants.
-While a spatial limit provides the better fit, neither a spatial nor a temporal limit explains the data well."}
+```{r fig5, echo=FALSE, message=FALSE, fig.width=6, fig.height=5.333333, fig.cap="\\label{fig:fig5}**Figure 5:** Reset models.
+**A:** Model limits relative to reset point locations.
+Reset points neither fall along a spatial limit (blue line) or a temporal limit (red lines).
+Marginal distributions show that reset points' X coordinates and time since stimulus approximately fall in a gamma distribution.
+**B:** Gamma distribution on time (log likelihood: -766.6).
+**C:** Gamma distribution on spatial offset (log likelihood: -1319.3)."}
 #plotResetPoints()
-plotModels()
+invisible(capture.output(plotModels()))
 #plotBootstrappedFits()

### 2021-11-29 R/exp3_gaborjumps.R
plot function for PSE data, Sirui's simplest time model

@@ -6,7 +6,7 @@ getPSEs <- function(groups=c('exp','ctrl'), overwrite=FALSE) {
   for (group in groups) {
     
     # check if the file is already there, skip if it is:
-    if (file.exists(sprintf('data/gaborjump/%s_PSEs.csv', group))) {
+    if (file.exists(sprintf('data/gaborjump/%s_PSEs.csv', group)) & !overwrite) {
       next
     }
     
@@ -199,3 +199,554 @@ ANOVAonPSEs <- function() {
   }
   
 }
+
+# figures -----
+
+plotJumpPSEs <- function(target='inline', individualDots=FALSE) {
+  
+  colors <- getColors()
+  
+  if (target == 'pdf') {
+    pdf( file = 'doc/fig_6_gaborjump_exp_alt.pdf', width=6, height=4 )
+  }
+  
+  layout(mat = matrix(c(1,2), 
+                      ncol  = 2,
+                      nrow  = 1,
+                      byrow = TRUE),
+         widths = c(1.125,1.1))
+  
+  methodsFig <- magick::image_read_pdf('doc/methods_fig_3.pdf')
+  
+  # extract width and height properties:
+  width <- magick::image_info(methodsFig)[['width']]
+  height <- magick::image_info(methodsFig)[['height']]
+  
+  par(mar=c(1,1,2.2,1))
+  
+  if (individualDots) {
+    pos <- 0.15 # Plot OffSet
+  } else {
+    pos <- 0
+  }
+  
+  plot(methodsFig,xlim=c(0,width),ylim=c(0,height),asp=1)
+  title('A: methods', font.main=1, cex.main=1.5, adj=0, line=0.5)
+  
+  
+  PSEdf <- read.csv('data/gaborjump/exp_PSEs.csv', stringsAsFactors = FALSE)
+  groupPSEs <- PSEdf[which(PSEdf$participant == 'group'),]
+  participantPSEs <- PSEdf[which(PSEdf$participant != 'group'),]
+  
+  par(mar = c(2.8,2.8,2.2,0.1),
+      mgp = c(2, 0.55, 0) )
+  
+  plot(-1000,-1000,main='',xlab='',ylab='',
+       xlim=c(0,8),ylim=c(0,3.5),
+       bty='n',ax=F)
+  
+  title(main='B: results', font.main=1, cex.main=1.5, adj=0, line=0.5)
[…]

### 2021-11-29 Manuscript.Rmd
removed extra authors again: can't be done if they don't respond also: more revisions

@@
 author: 
-- Bernard Marius 't Hart 1,*
-- Sirui Liu 2
-- Peter Ulrich Tse 2
-- Denise Y.P. Henriques 1
-- Patrick Cavanagh 1,2,3
+- Bernard Marius 't Hart $^{1,*}$
+- Denise Y.P. Henriques $^1$
+- Patrick Cavanagh $^{1,2,3}$
 output:
+  html_document:
+    df_print: paged
+  pdf_document: default
   word_document:
     reference_docx: template.docx
-  pdf_document: default
-  html_document:
-    df_print: paged
 header-includes:
@@
 
-# for exp 3, we fit psychometric functions to get PSEs:
-getPSEs()
+# for exp 3, we make sure the psychometric PSEs are there:
+getPSEs(overwrite = FALSE)
+# set overwrite to TRUE to recalculate the PSEs
 
@@
 _1: Centre for Vision Research, York University, Toronto, ON M3J1P3, Canada_\
-_2: Department of Psychological and Brain Sciences, Dartmouth College, Hanover, NJ 03755, USA_\
-_3: Department of Psychology, Glendon College, Toronto, ON M4N3M6, Canada_\
+_2: Department of Psychology, Glendon College, Toronto, ON M4N3M6, Canada_\
+_3: Department of Psychological and Brain Sciences, Dartmouth College, Hanover, NJ 03755, USA_\
 
@@
 To report the perceived trajectory of the gabor, participants drew it after the gabor was removed from the monitor.
-Resets were detected and analyzed but their distribution neither matches a temporal nor a spatial limit on perception that could trigger resets.
+Resets were detected and analyzed and their distribution neither matches a temporal nor a spatial limit, but rather a gamma distribution over time.
 
@@
 The ability to record trajectories of the perceived position will then allow us to study spontaneous reset outcomes.
+We find that there is no hard temporal or spatial limit leading to resets, but it seems that resets occur randomly over time, based on a gamma distribution.
 
@@
 
+We test this with a series of simple models.
+If there is a spatial offset limit that triggers resets, the resets points should fall on, or be close to a line with a constant X coordinate.
+If there is a temporal limit, all resets should have the same or a similar reset time (distance from origin / speed).
+If we assume errors are normally distributed, we can then fit a normal distribution (with 2 parameters: mean and standard deviation) on both the X coordinates and the time points of resets.
+The distributions of neither the X coordinates nor the time points of resets appear normally distributed though, but seem better approximated by a gamma distribution (see Fig 5A).
+So we additionally fit gamma distributions (with 2 parameters: rate and shape) on the X coordinates and time points of resets as well.
+We then calculate the likelihoods of these four simple distribution models both using all resets and within individual participants and use this to calculate an AIC to determine a relative probability that compares models.
+
 ```{r fig5, echo=FALSE, message=FALSE, fig.width=6, fig.height=5.333333, fig.cap="\\label{fig:fig5}**Figure 5:** Reset models.
@@
 
-We test this with a series of simple models.
-First, for every reset point (with coordinates $X$ and $Y$) we use the illusion strength based on a point earlier in the trajectory, expressed as either an angle (Fig.
-4, left: $a = \arctan{(X/Y)}$) or a slope ($s = \sin{(a)}/\cos{(a)}$) to predict reset points.
-
-If the perceived path of the gabor is approximated by a straight line (up to a reset), and reset points occur at a set limit, then a spatial limit ($L_x$) would equal the x-coordinate and predict the y-coordinate, $y$, of the first reset:
-
-[2]  &nbsp;&nbsp;&nbsp;&nbsp;&nbsp;  $y = L_x \cdot s$
-
-And a temporal limit ($L_t$) would predict both coordinates of the reset, $x$ and $y$:
-
-[3a]  &nbsp;&nbsp;&nbsp;&nbsp;&nbsp;  $x = L_t \cdot v \cdot \cos{(a)}$
-
-[3b]  &nbsp;&nbsp;&nbsp;&nbsp;&nbsp;  $y = L_t \cdot v \cdot \sin{(a)}$
-
-Where $L_t$ and $L_x$ are a temporal and spatial limit (in seconds and cm respectively) that are free parameters, $a$ and $s$ correspond to the calculated angle and slope expressing the strength of the illusion, and speed ($v$ in _cm/s_) is given by the stimulus.
-If we fit the two simple models for each participant separately, by minimizing the mean square error (in isometric space, with absolute distance in centimeters between predicted and actual reset points), most participants are best fit with a spatial limit (MSE: 6/9, AIC: 7/9).
-This indicates that in this data, a spatial limit explains the reset points best, but neither limit explains the data very well.
-
-To obtain population level values for the spatial and temporal limits, we resample reset points from each participant, using the smallest number of detected reset points that fall in the first quadrant, in the group (44), to ensure that each participant contributes the same number of reset points.
-Both models based on a spatial limit (eq.
-2) and on a temporal limit (eq.
-3)  were fit 1000 times (Fig.
-5, right).
-The median fits use a spatial limit of 3.5 _cm_ or ~3.6 _dva_ and a temporal limit of 1.1 _s_.
-We explored other modeling approaches (e.g.
-based on various distributions of likelihoods or combining limits) but these show the same pattern of results: spatial limits are somewhat favoured, but temporal limits can not be excluded (see OSF repository: [https://osf.io/72ndu/](https://osf.io/72ndu/)).
+Within the normal distribution models of reset points (means in Fig 5A, blue: spatial, red: time), the time-based distribution fits better for the group data ( _p_<.001 ) and all participants, and the same holds within gamma distribution models of reset models (Fig 5B,C).
+Furthermore, the gamma-based distribution on time of resets performed better than the normal distribution in the pooled data ( _p_<.001 ), and this holds for most participants (6/9) as well.
+This means that time-based distributions explain our reset points better than offset- or space based models.
 
@@
 We first tested whether manual tracking of a double-drift stimulus is susceptible to the illusion as reported by perceptual judgments, pointing [@Lisi2017], and memory saccades [@Massendari2018; @Ueda2018].
-The results of this online tracking experiment show a close agreement between the angle of the manual tracking and the illusion strength expected based on the model from earlier perceptual measures [@Cavanagh2019].
+The results of this online tracking experiment show a close agreement between the angle of the manual tracking and the illusion strength expected based on the model from earlier perceptual measures [@Heller2021].
 However, since no unambiguous resets were captured in the traces of this experiment, we followed up with a second experiment that did record evidence of resets.
@@
 We then set out to test if these spontaneous resets are triggered by a limiting distance of drift away from the physical location, or by a specific limit of time.
-The results showed that the resets are slightly better explained by the distance from the true position, but a temporal limit is not ruled out.
+A gamma distribution of reset time seems to explain this data best.
+
+However, it does not seem like there is a specific temporal limit on time, i.e.
+resets do not occur at a given time after stimulus onset, but are more or less randomly pulled from a wide distribution.
+This relies on the hand trajectories we recorded here to provide an accurate measure of reset points.
+Unpublished work [poster: @Liu2019] used a psychophysics-based approach to probe how spontaneous resets might affect the accumulated illusion after several amounts of time (1, 2 or 3 _s_) while keeping the actual path (2 or 4 _dva_) and maximum illusion (1.4 or 2.8 _dva_) constant (see Fig 6A).
+After the main stimulus presentation, there is a 350 _ms_ gap, where there is no stimulus, and this is know to evoke a reset, after which the illusion resumes accumulating from the position where the gabor appears after the gap [@Lisi2015].
+The data from 10 participants shows that as a stimulus is presented for longer, the illusory offset at the end is smaller than would be expected based on the individually calibrated illusion strengths.
+
+```{r fig6, echo=FALSE, message=FALSE, fig.width=6, fig.height=4, fig.cap="\\label{fig:fig6}**Figure 6:** Spontaneous resets without hand trajectories.
+Experiment and Results based on data from Liu et al.
+[-@Liu2019] **A:** By ensuring the perceived path of a double-drift stimulus is straight down, any gap-induced reset, should be equal to the accumulated illusion if there are no spontaneous resets, but smaller if there are.
+Gaps were presented after 1, 2 or 3 seconds, and 2 or 4 dva real paths, which allows psychophysics to quantify the amount of illusion left after several periods of abserving a double-dirft stimulus.
+**B:** Points of subject equivalence show that the size of the illusion decreases with time even as the real path, and maximum illusion is kept constant.
+These data are in line with resets occurring randomly over time."}
+plotJumpPSEs(target='inline', individualDots = FALSE)
+```
 

### 2021-11-29 R/exp2_models.R
fit gamma and normal models on all participants to get likelihoods -> AICs -> relative probabilities

@@ -2369,7 +2369,7 @@ plotUncoupledModels <- function(df, target='inline') {
 # which is what the orthogonal and normal models compare (it is the same)
 # and then there is an offset gamma distribution for comparison
 
-fitSomeModels <- function(df, jointModel=FALSE, Xnormal=TRUE, Xgamma=TRUE, Tgamma=TRUE) {
+fitSomeModels <- function(df, jointModel=FALSE, Xnormal=TRUE, Tnormal=TRUE, Xgamma=TRUE, Tgamma=TRUE, Orthogonal=FALSE) {
   
   outputlist <- list()
   
@@ -2377,64 +2377,65 @@ fitSomeModels <- function(df, jointModel=FALSE, Xnormal=TRUE, Xgamma=TRUE, Tgamm
   # ortogonal models, just because
   # ********************************
   
-  
-  Lx=seq(0, 8, length.out = 41)
-  Lt=seq(0, 4, length.out = 41)
-  
-  # make them into data frames:
-  searchgridXlim <- expand.grid('Lx'=Lx)
-  searchgridTlim <- expand.grid('Lt'=Lt)
-  
-  # get MSE for points in search grid:
-  XlimMSE <- apply(searchgridXlim,FUN=XorthogonalMSE,MARGIN=c(1),data=df,fitFUN=XlimResets)
-  TlimMSE <- apply(searchgridTlim,FUN=TorthogonalMSE,MARGIN=c(1),data=df,fitFUN=TlimResets)
-  
-  # get 4 best points in the grid:
-  topgridXlim <- data.frame('Lx'=searchgridXlim[order(XlimMSE)[c(1,3,5)],])
-  topgridTlim <- data.frame('Lt'=searchgridTlim[order(TlimMSE)[c(1,3,5)],])
-  
-  # do the actual fitting:
-  allXlimFits <- do.call("rbind",
-                         apply( topgridXlim,
-                                MARGIN=c(1),
-                                FUN=optimx,
-                                fn=XorthogonalMSE,
-                                method='L-BFGS-B',
-                                lower=c(0),
-                                upper=c(13.5),
-                                data=df,
-                                fitFUN=XlimResets) )
-  
-  allTlimFits <- do.call("rbind",
[…]

### 2021-11-29 R/exp2_retracing.R
changed colors of reset distribution and the K figure

@@ -1295,7 +1295,7 @@ plotResetPoints <- function(target='inline') {
   #print(dim(df))
   
   plot(df$resetX,df$resetY,
-       pch=16,col=t_col(colors$blue$s, percent=97.5),cex=3,
+       pch=16,col=t_col(colors$purple$s, percent=97.5),cex=3,
        main='',xlab='',ylab='',
        xlim=c(-1,9),ylim=c(0,16),
        bty='n',ax=F,asp=1)
@@ -1384,15 +1384,16 @@ plotResetPoints <- function(target='inline') {
   for (speed in c(3,4)) {
     
     idx <- which(avg_df_g$Ve == 13.5/speed)
-    speedcol <- c(colors$blue$s,colors$yorkred$s)[speed-2]
-    points(avg_df_g$vector_prediction[idx],avg_df_g$angle[idx],col=speedcol,pch=1,cex=2)
+    speedcol <- c(colors$orange$s,colors$orange$s)[speed-2]
+    speedpch <- c(1,2)[speed-2]
+    points(avg_df_g$vector_prediction[idx],avg_df_g$angle[idx],col=speedcol,pch=speedpch,cex=2)
     
   }
   
   legend(x=0, y=45, 
          legend = c('group means (3 s)', 'group means (4 s)', 'Heller et al. (2021)', sprintf('K=%0.2f',group_k)),
-         col=c(colors$blue$s, colors$yorkred$s, 'black', colors$lightblue$s), 
-         pch=c(1,1,NA,NA), lty=c(0,0,1,1), 
+         col=c(colors$orange$s, colors$orange$s, 'black', colors$lightblue$s), 
+         pch=c(1,2,NA,NA), lty=c(0,0,1,1), 
          bty='n', cex=1)
   
   axis(side=1,at=seq(0,75,15))

### 2021-11-29 R/old/time.R
reduced number of runs

@@ -357,7 +357,7 @@ fitResetProbability <- function() {
   fit <- optim(par          = par,
                fn           = timeLimitMSE, 
                conditions   = conditions,
-               runs         = 10000,
+               runs         = 5000,
                PSEs         = PSEs,
                method       = 'Brent',
                lower        = c(0),

### 2021-12-03 Manuscript.Rmd
revisions

@@
 ---
-title: "Measuring the double-drift illusion with hand tracking"
+title: "Measuring the double-drift illusion with hand trajectories"
 author: 
-- Bernard Marius 't Hart $^{1,*}$
-- Denise Y.P. Henriques $^1$
-- Patrick Cavanagh $^{1,2,3}$
+- Bernard Marius 't Hart 1,*
+- Denise Y.P. Henriques 1
+- Patrick Cavanagh 1,2,3
 output:
@@
     df_print: paged
-  pdf_document: default
   word_document:
     reference_docx: template.docx
+  pdf_document: default
 header-includes:
@@
 To report the perceived trajectory of the gabor, participants drew it after the gabor was removed from the monitor.
-Resets were detected and analyzed and their distribution neither matches a temporal nor a spatial limit, but rather a gamma distribution over time.
+Resets were detected and analyzed and their distribution matches neither a temporal nor a spatial limit, but rather a wide gamma distribution over time.
 
@@
 the flash lag [@Nijhawan1994]; or the flash grab effect [@Cavanagh2013].
-These cases of position extrapolation may be functional, helping the motor system overcome inevitable neural delays in targeting a moving object [@Hogendoorn2020; @Nijhawan1994; @Duhamel1992].
+Although there are other explanations for the flash-lag effect [@Eagleman2000; @Cai2001], these cases of position extrapolation may be functional, helping the motor system overcome inevitable neural delays in targeting a moving object [@Hogendoorn2020; @Nijhawan1994; @Duhamel1992].
 A stronger and very different motion-induced position shift arises when a moving gabor has internal motion orthogonal to its path and is viewed in the periphery [@Tse2006; @Shapiro2010; @Gurnsey2012; @Kwon2015; @Lisi2015].
@@
 library(magick)
-Fig1 <- image_read_svg('doc/methods_fig_1.svg')
+Fig1 <- image_read_pdf('doc/methods_fig_1.pdf')
 par(mai=c(0,0,0,0))
@@
 However, when the gabor itself also moves, as in the double-drift stimulus, there is no stable location information to anchor the position estimate and the build-up continues well beyond 100 ms.
-The perceived location continues to move away from its physical position by up to several times the size of the gabor.
+The perceived location may move from the physical position by up to several times the size of the gabor.
 
-The strength of the illusion; the deviation of perceived path of the gabor from its true path, depends on the speed of both the internal and external motion [@Tse2006; @Shapiro2010; @Kwon2015; @Cavanagh2019; @Heller2021].
-Rather than use point estimates of the perceived location of the stimulus, we want to test how well a continuous trajectory can measure the strength of the illusion.
+At some point, the illusion does stop accumulating.
+The accumulation can be reset by introducing a temporal break [@Lisi2015] or by distracting attention [@Nakayama2020], but it may also reset spontaneously, perhaps when the accumulation has gone on too long or too far.
+Informally, some observers have reported that the gabor’s path may saturate so that it remains parallel to, but offset from the true path.
+Others report that the position moves back toward its physical location, either slowly or suddenly, whereupon accumulation begins again.
+The point where the accumulation of illusory offset stops could be informative about how the brain combines motion and position information.
+
+Here we use hand movements to investigate where or when the double-drift illusion resets.
 Eye and hand movement are differentially sensitive to many illusions, such as the Ponzo illusion [@Gamble2017], Ebbinghaus illusion [@Knol2017], flash-lag effect [@Blohm2003], the Duncker illusion [@Soechting2001] and many others.
-Since it has been shown that the double-drift illusion affects pointing movements, but not saccades [@Lisi2017] we expect that hand trajectories will also be sensitive to the illusion.
+Similarly, it has been shown that the double-drift illusion affects pointing movements, but not saccades [@Lisi2017], and hand movements have been used to test other perceptual phenomena [@PatricioDecima2022].
+Hence, we expect that hand trajectories will be an effective method to capture the full illusory trajectory, rather than a point estimate along the perceived path, and this will allow capturing resets of the double-drift illusion.
 
-We assess this in our first experiment, using hand movement trajectores recorded during stimulus presentation (Fig.
+The strength of the illusory percept; the deviation of the perceived path of the gabor from its true path, depends on the speed of both the internal and external motion [@Tse2006; @Shapiro2010; @Kwon2015; @Cavanagh2019; @Heller2021].
+We want to test how well a continuous trajectory reflects the percept of the illusion.
+We assess this in our first experiment, using hand movement trajectories recorded during stimulus presentation (Fig.
 1).
@@
 As we expected, online tracking provides a similar measure of illusion strength as pointing.
+However, only few possible resets are apparent in the data (>4% trials), and no limit to the illusion is clear.
 
-In our second experiment, we look at the spatial and temporal limits of the illusion.
-At some point, the illusion stops accumulating.
-The accumulation can be reset by introducing a temporal break [@Lisi2015] or by distracting attention [@Nakayama2020], but it may also reset spontaneously once the accumulation has gone on too long or too far.
-Informally, some observers have reported that the gabor’s path may saturate so that it remains parallel to, but offset from the true path.
-Others report that the position moves back toward its physical location, either slowly or suddenly, whereupon accumulation begins again.
-The ability to record trajectories of the perceived position will then allow us to study spontaneous reset outcomes.
-We find that there is no hard temporal or spatial limit leading to resets, but it seems that resets occur randomly over time, based on a gamma distribution.
+In our second experiment, we therefore slow down the external stimulus motion, and have participants reproduce the perceived trajectory after stimulus presentation.
+Here we do observe spontaneous resets, which allows us to investigate any spatial or temporal limits of the illusion.
+However, we find that there is no hard temporal or spatial limit leading to resets, but it seems that resets occur randomly over time, based on a gamma distribution.
 
@@
 We test if tracking the illusion continuously is a good measure of illusion strength, or if it prevents or decreases the illusion.
-
-Potential extra citation: [@PatricioDecima2022].
+Additionally, we use a relatively long trajectory to allow the double-drift illusion to exceed any perceptual limits it may have, so that we can observe what these are.
 
@@
 
-For this experiment, 4 participants were recruited from the lab (3 female; ages 21 – 31, mean: 25.75).
+For this experiment, 4 participants were recruited from the lab (3 female; ages 21 – 31, mean: 25.75) and they were naive to the purpose of the study.
 All participants reported being right handed and had either normal or corrected to normal vision.
-Procedures were in accordance with the Declaration of Helsinki and were approved by York’s Human Participants Review Committee.
+Procedures were in accordance with the Declaration of Helsinki (2003) and were approved by York’s Human Participants Review Committee.
 All participants provided prior, written, informed consent.
@@
 1, right), as well as a keypad for additional responses.
-Stimuli were displayed on a downward facing LCD monitor (30 Hz, 20”, 1680x1050), parallel with the drawing tablet.
+Stimuli were displayed on a downward facing LCD monitor (30 Hz, 20”, 1680x1050, Dell E2009Wt), parallel with the drawing tablet.
 The stimuli were observed via an upward facing mirror placed exactly between the tablet and monitor, so that the stimuli appear to be in the same horizontal plane as the hand, and the tip of the stylus.
@@
 In this experiment, participants used their unseen hand and stylus to do continuous online tracking of the double-drift stimulus while that moved back and forth along the Y-axis of the horizontal screen for 12 _s_.
-Before trial onset, the participant moved the stylus to the middle of the screen, and were given 1.5 _s_ to fixate a point on the left or right of the screen.
+Before trial onset, the participant moved the stylus to the middle of the screen, and were given 1.5 _s_ to fixate a point on 16.5 _cm_ (~22.5 _dva_) to the left or right of the centre of the screen.
 The gabor would then appear at the middle of the screen and start moving.
-Internal motion could be 3, 1, 0, -1 or -3 _cps_.
+Internal motion could be 3, 1, 0, -1 or -3 _cps_ (each repeated 8 times).
 Since each moving gabor was shown for 12 _s_, and it moved from one end to another in _2_ s, it completed 5 full and 2 half passes of the path.
@@
 They show a double-drift stimulus with 36 combinations of internal and external motion for 500 ms and then have participants indicate on a ring where they perceived the path of the stimulus to intersect with this ring.
-While they hypothesize that illusion strength decreases with faster external motion (perhaps because of resets or limits on perception), here we will mostly use their overall estimate of illusion strength, given by $K=0.74$.
+While they hypothesize that illusion strength decreases with faster external motion (perhaps because of resets or limits on perception), here we will use their overall estimate of illusion strength, given by $K=0.74$.
 
@@
 
-All raw trajectories are show in Figure 2A.
+All raw trajectories are shown in Figure 2A.
 In Figure 2C, we plot the average instantaneous heading during the middle second of each 2 _s_ pass for each of the 5 internal motion speeds.
@@
 This makes it clear that continuous tracking does not prevent the illusion from occurring.
-We will use the heading angle as a measure of the strength of the illusion.
+We will use heading angle as a measure of the strength of the illusion.
 
@@
 Other percepts of resets have been reported as well, but as far as we know these fall in between these two extremes.
-Resets may also occur spontaneously once the accumulation has gone on too long or too far, depending on what mechanism causes resets.
+Such resets may occur spontaneously once the accumulation has gone on too long or too far, depending on what mechanism causes resets.
 Where resets occur would then both depend on the mechanism that causes resets, as well as on the strength of the illusion (purple arrows in Fig.
@@
 library(magick)
-Fig3 <- image_read_svg('doc/methods_fig_2.svg')
+Fig3 <- image_read_pdf('doc/methods_fig_2.pdf')
 par(mai=c(0,0,0,0))
@@
 
-In experiment 2, participants first watched a double-drift stimulus and then either reproduced the perceived path of the stimulus by retracing it on the tablet, or indicated the initial movement direction by changing the orientation of a line originating in the same position as the doube-drift stimulus (data not used, but corresponds to illusion strength determined from trajectories).
+Before the experiment started, the illusion and its resets were explained to particpants, and they were instructed to make sure to replicate any changes in direction of the perceived path as accurately as they could manage.
+In experiment 2, participants first peripherally observed a double-drift stimulus, while fixating to the left or right of the stimulus.
+Then they either reproduced the perceived path of the stimulus by retracing it on the tablet, or indicated the initial movement direction by changing the orientation of a line originating in the same position as the doube-drift stimulus (data not used, but corresponds to illusion strength determined from trajectories).
 More details are given below.
@@
 Block types were alternated, and the order was counterbalanced across participants.
-Each block used the 6 combinations of two external speeds ( corresponding to 3 or 4 _s_ presentation time ), and three internal speeds ( 2, 3 or 4 _cps_ ), 6 times, for a total of 36 trials per block.
+Each block used all 6 combinations of two external speeds ( corresponding to 3 or 4 _s_ presentation time ), and three internal speeds ( 2, 3 or 4 _cps_ ), 6 times, for a total of 36 trials per block.
 Each of the combinations of internal and external speed was presented 24 times in total, so that we have 144 trials for each participant, and 1296 trials in total.
@@
 The individual trajectories reveal some points that are potential spontaneous resets that we try to extract.
-Using a smooth spline interpolated version of the trajectory we detect right-left direction changes, based on sign changes.
-First, we remove the first and last 4 mm of the trajectory, as well as any short segments on the trajectory at the start and end that go toward the participant, to remove any unintended jitters caused by starting and ending the movement.
+We created a heuristic to extract likely reset points based on right-left inflections in the trajectories.
+First, we remove the first and last 4 mm of the trajectory, as well as any short segments on the trajectory at the start and end that go toward the participant (i.e.
+more than 180 degrees away from the true external motion), to remove any unintended jitters caused by starting and ending the movement.
 Then, we apply a bi-directional low-pass Butterworth filter with a cut-off frequency of 1.5 Hz.
@@
 In these summary trajectories we detect the first inflection point using the X coordinates.
-This could put the reset point too late for hit-the-wall resets, so that we move back to get the previous sample at 95% of the X coordinate, and finally locate the sample in the raw trajectory that is closest to this point.
+This could put the reset point too late for hit-the-wall resets, so that we move back to get the previous sample at 95% of the X coordinate at the inflection point.
+Then, we locate the sample in the raw trajectory that is closest to this point.
 Finally, we remove a few reset points that are detected very close to the veridical path without any build-up of the illusion by only keeping reset points that fell 5 mm within the quadrant where we expect them.
 This heuristic will result in some misses and some false alarms, and while changing the algorithm may improve performance, there is no ground truth to evaluate this.
-All raw trajectories and the detected reset points can be found on the OSF repository (XXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXX add link?).
+All raw trajectories and the detected reset points can be found on the OSF repository.
 
@@
 
-In Figure 4 (middle), we show the trajectories for participant 4 in the condition with 3 _cps_ internal motion and 4 _s_ passes.
-This particular participant's re-traced paths show clear 'return' resets, all of which are detected by our simple algorithm.
-Other participants' data are less clear.
+Four example trajectories and the location of resets determined by our heuristic are shown in Figure 4.
+On average, this method detects resets in ~64% of trials.
+For two participants, less than half the trials showed resets (9% and 22%).
+For the remaining seven participants the algorithm detected resets in 65%-99% of trials, with an average of 82%.
+Since participants experiencing a "hit-the-wall reset" (Fig.
+3) as opposed to a "jump" reset, might not always display a detectable direction change in their paths, most of these percentages of trials with resets seems reasonable.
+However, the participant with 9% resets will not be used for calculating illusion strength as they have no resets in several conditions.
+This means we analyse a total of 827 reset points (Fig.
+4, right).
 
@@
 **F:** Using the angle to a point along the trajectory at half the distance to the reset point as a measure of illusion strength, we can compare this data to other data.
+For this graph, we exclude one participant who produced very few resets.
 The illusion is weaker ($K=0.60$, blue line) and its strength varies between participants, but it is nevertheless clearly there."}
@@
 
-On average, this method detects resets in ~68% of trials.
-For two participants, less than half the trials showed resets (9% and 22%).
-For the remaining seven participants the algorithm detected resets in 65%-99% of trials, with an average of 82%.
-Since participants experiencing a "hit-the-wall reset" (Fig.
-3) as opposed to a "jump" reset, might not always display a detectable direction change in their paths, most of these percentages of trials with resets seems reasonable.
-However, the participant with 9% resets will not be used for figures and statistics, but we have verified that excluding their data does not change the results.
-This means we analyse a total of 858 reset points (Fig.
-4, right).
-
-As before, we compare the strength of the illusion (see Fig 5, left), with an earlier study [@Heller2021].
-A line through the origin with $K=0.54$ predicts illusion strength fairly well in these data, which is weaker than the $K=0.74$ found by Heller et al.
+As before, we compare the strength of the illusion (Fig.
+4F), with an earlier study [@Heller2021].
+A line through the origin with $K=0.60$ predicts illusion strength fairly well in these data, which is weaker than the $K=0.74$ found by Heller et al.
 [-@Heller2021].
-Illusion strength is also lower than in Experiment 1 with online tracking and it has a fair amount of between-subject variation in this task (blue shaded area is 95% confidence interval across participants).
+Illusion strength is also lower than in Experiment 1 with online tracking and it has a fair amount of between-subject variation in this task (blue shaded area is 95% confidence interval on K across participants, and organge lines are 95% confidence intervals across participants for illusion strength in each condition).
 While the illusion is less strong, it is nevertheless there and depends on the relationship between internal and external speed as in previous studies.
@@
 We test this with a series of simple models.
-If there is a spatial offset limit that triggers resets, the resets points should fall on, or be close to a line with a constant X coordinate.
+If there is a spatial offset limit that triggers resets, the reset points should fall on, or be close to a line with a constant X coordinate.
 If there is a temporal limit, all resets should have the same or a similar reset time (distance from origin / speed).
@@
 **A:** Model limits relative to reset point locations.
-Reset points neither fall along a spatial limit (blue line) or a temporal limit (red lines).
-Marginal distributions show that reset points' X coordinates and time since stimulus approximately fall in a gamma distribution.
+Reset points neither fall along a spatial limit (blue line) or a temporal limit (red lines: continuous for 4 _s_ passes, dashed for 3 _s_ passes).
+Marginal distributions show that reset points time and spatial offset approximate a gamma distribution.
 **B:** Gamma distribution on time (log likelihood: -766.6).
@@
 
-Within the normal distribution models of reset points (means in Fig 5A, blue: spatial, red: time), the time-based distribution fits better for the group data ( _p_<.001 ) and all participants, and the same holds within gamma distribution models of reset models (Fig 5B,C).
-Furthermore, the gamma-based distribution on time of resets performed better than the normal distribution in the pooled data ( _p_<.001 ), and this holds for most participants (6/9) as well.
-This means that time-based distributions explain our reset points better than offset- or space based models.
+Within the normal distribution models of reset points (means in Fig 5A, blue: spatial, red: time), the time-based distribution fits better for the group data ( _p_<.001 ) as well as all individual participants, and the same holds within gamma distribution models of reset models (Fig 5B,C).
+Furthermore, the gamma-based distribution on time of resets performed better than the normal distribution of reset time in the pooled data ( _p_<.001 ), and this holds for most participants (6/9) as well.
+This means that time-based distributions explain our reset points better than models assuming a spatial offset.
 
@@
 The results of this online tracking experiment show a close agreement between the angle of the manual tracking and the illusion strength expected based on the model from earlier perceptual measures [@Heller2021].
-However, since no unambiguous resets were captured in the traces of this experiment, we followed up with a second experiment that did record evidence of resets.
+However, we were also interested in  spontaneous resets and only ~4% of trials show a deviation in the trajectory that might reflect a reset.
+In the second experiment, 62% of trials showed a reset, and 36% of those occurred within 1 _s_ and 84% within 2 _s_.
+Based on these numbers one might expect spontaneous resets in at least 22.5% of each of the two half passes in the online tracking experiment and 52.5% of each of the full 2 _s_ passes - if each pass is independent.
+This would mean ~86% of trials would show at least 1 spontaneous reset which is much larger than the ~4% of trials we observe ( _p_<.001, exact binomial test).
+The question remains as to why so few spontaneous resets occurred in this experiment.
+Simple explanations could be that duration of stimulation presentation requires relatively fast hand movements, leaving little time for drawing spontaneous resets or reduced attention for the double-drift stimulus.
+The one feature that sets this task apart from others is that the tracking is done while perceiving the stimulus, such that action could affect perception.
+One possible mechanism is that predicted sensory consequences of the hand movement inform perception ("I will move my hand in this direction, so I expect to see motion in that direction as well").
+And another is post-dictive inference ("my hand has moved in this direction so I must have seen the visual stimulus move in the same direction").
+But this is a question for future research.
 
-To measure resets, we asked participants to draw the perceived path of the stimulus after a single pass.
-The orientation of the drawn traces reflect the strength of the illusion, and in most trials there was a clear reset point, for either a "return" reset or a "hit-the-wall" reset.
+In our second experiment, we asked participants to draw the perceived path of the stimulus after observing a single pass.
+The orientation of the drawn traces reflect the strength of the illusion, and in a majority of trials (62%) there was a clearly indentifiable reset point, for either a "return" reset or a "hit-the-wall" reset.
 We then set out to test if these spontaneous resets are triggered by a limiting distance of drift away from the physical location, or by a specific limit of time.
@@
 However, it does not seem like there is a specific temporal limit on time, i.e.
-resets do not occur at a given time after stimulus onset, but are more or less randomly pulled from a wide distribution.
-This relies on the hand trajectories we recorded here to provide an accurate measure of reset points.
-Unpublished work [poster: @Liu2019] used a psychophysics-based approach to probe how spontaneous resets might affect the accumulated illusion after several amounts of time (1, 2 or 3 _s_) while keeping the actual path (2 or 4 _dva_) and maximum illusion (1.4 or 2.8 _dva_) constant (see Fig 6A).
-After the main stimulus presentation, there is a 350 _ms_ gap, where there is no stimulus, and this is know to evoke a reset, after which the illusion resumes accumulating from the position where the gabor appears after the gap [@Lisi2015].
-The data from 10 participants shows that as a stimulus is presented for longer, the illusory offset at the end is smaller than would be expected based on the individually calibrated illusion strengths.
+resets do not occur at a given time after stimulus onset, but appear randomly sampled from a wide distribution.
+In part, this may rely on the hand trajectories we recorded here to provide an accurate measure of reset points, as they may contain an unknown amount of noise.
+Therefore, we compare our results with those from an as of yet unpublished work [pre-print: @Liu2021].
+This study used a psychophysics-based approach to probe how spontaneous resets might affect the accumulated illusion throughout time (1, 2 or 3 _s_) while keeping the actual path (2 or 4 _dva_) and maximum illusion (1.4 or 2.8 _dva_, respectively) constant.
+Spontaneous resets should decrease the accumulated illusion (see Fig 6A).
+After the main stimulus presentation, there is a 350 _ms_ gap, during which the double-drift stimulus is not shown, and this is know to evoke a reset.
+After this gap-evoked reset the illusion resumes accumulating from the veridical position where the gabor appears after the gap [@Lisi2015].
+By changing the location where the double-drift reappears and asking participants if it jumped left or right, the point of subjective equivalence that expresses the size of the gap-evoked jump reset can be determined.
+
+The data from 10 participants shows that as a stimulus is presented for longer, the illusory offset at the end gets smaller (Fig 6B, round dots).
+This can be explained by (among others) resets randomly occuring throughout time.
+Liu et al.
+[-@Liu2021] present a Poisson process model with a constant chance of reset occurring every 10 ms of ~0.007 or ~0.7 / _s_ (Fig 6B, light diamonds).
+If we compare this to the gamma distribution fit, with a rate of $2.81$ events per second, and reset occurring at the $3.74^{th}$ event (on average), this amounts to a ~0.75 (2.81/3.74) chance of a reset per second.
+If we simply take the parameters of the gamma distribution fit to reset times in the re-tracing experiment (without optimization), we can also predict PSEs (Fig 6B, dark diamonds).
+We see that the two findings, based on very different approaches, are mostly in agreement with each other.
 
@@
 Experiment and Results based on data from Liu et al.
-[-@Liu2019] **A:** By ensuring the perceived path of a double-drift stimulus is straight down, any gap-induced reset, should be equal to the accumulated illusion if there are no spontaneous resets, but smaller if there are.
+[-@Liu2021], re-analyzed here.
+**A:** By ensuring the perceived path of a double-drift stimulus is straight down (teal arrows), any gap-induced reset, should be equal to the accumulated illusion if there are no spontaneous resets (black arrow), but smaller if there are (light blue arrow).
 Gaps were presented after 1, 2 or 3 seconds, and 2 or 4 dva real paths, which allows psychophysics to quantify the amount of illusion left after several periods of abserving a double-dirft stimulus.
 **B:** Points of subject equivalence show that the size of the illusion decreases with time even as the real path, and maximum illusion is kept constant.
-These data are in line with resets occurring randomly over time."}
-plotJumpPSEs(target='inline', individualDots = FALSE)
+These data are in line with resets occurring randomly over time.
+The simplest model presented by Liu et al.
+(2021) keeps the illusion strength constant and fits a chance of reset per time unit as ~0.007 / 10 ms to match the gap-evoked reset (diamond symbols)."}
+plotJumpPSEs(target = 'inline', individualDots = FALSE, modelDots = TRUE, gammaDots=TRUE)
 ```
 
-Recently, Kwon, Tadin, and Knill [-@Kwon2015] proposed a Bayesian object tracking model to explain the double-drift illusion.
+Kwon, Tadin, and Knill [-@Kwon2015] proposed a Bayesian object tracking model to explain the double-drift illusion and other phenomena.
 In their model, when the position signals are weak, the motion signals bias perceived position away from the true location.
-In their case, this deviation saturated after about 200 _ms_ and the perceived path then continued parallel to the physical path, as in the “hit the wall” resets we described above [@Kwon2015, Fig.
+Their model predicts that this deviation saturates after about 200 _ms_ and the perceived path then continues parallel to the physical path, as in the “hit the wall” resets we describe above [@Kwon2015, Fig.
 3].
-In contrast to their model, our participants in Experiment 2 showed quite variable timing for the reset point where the accumulation of offset ended.
-The resets were more stable in terms of spatial offset (Fig.
-4, middle) but much more variable in terms of time from the start of the gabor's motion.
-In any case, the temporal limit of these reset points was not 200 _ms_ but 1.1 _s_ and fell into a wide range from 0.5 _s_ to almost 4 _s_ across trials (see Fig.
-4).
-Moreover, 4/9 of our participants regularly reported “return” resets that the Kwon et al.
-[-@Kwon2015] model does not predict.
-The “return” resets suggest that both the physical and perceived locations are available to the visual system and that when the deviation exceeds some limit, the perceived location returns toward the physical location.
-At that point, the build-up of the illusion may start again.
+In contrast to their model, the reset in our re-tracing task show quite variable timing, and do not show a clear spatial offset (Fig.
+5A) but seem to fall on a fairly wide gamma distribution on reset time (Fig.
+5B), which is at odds with a set limit (either spatial or temporal) where resets occur.
+This seems to argue against a perceptual mechanism underlying resets, such as on going Bayesian inference until the discrepancy between retinal and perceived position signals is too large.
+In any case, the best fitting temporal limit of these reset points was not 200 _ms_ but 1.3 _s_ and 99% of detected reset times are above 200 _ms_ (see Fig.
+5A).
+Moreover, 4/9 of our participants clearly reported “return” resets that the Kwon et al.
+[-@Kwon2015] model does not predict, and only 1/9 participants showed clear "hit-the-wall" resets.
+The “return” resets suggest that both the physical and perceived locations are available to the visual system and that when spontaneous resets occur, the perceived location returns toward the physical location.
+While our results speak against a particular, stimulus or perception induced mechanism leading to resets, the illusion itself may still have limits, and the way resets are perceived may relate to stimulus properties as well as to illusory perception.
+Regardless, future models that aim to predict perception in double-drift stimuli need to also predict resets.
 
@@
 The participants in this experiment also retraced the perceived path and reported “return” resets.
-Nakayama and Holcombe also suggest that planned eye movements can trigger a “return” reset back to the physical path.
+Nakayama and Holcombe [-@Nakayama2020] suggest that planned eye movements can trigger a “return” reset back to the physical path.
 In our experiment, we did not control for eye-movements, so we can't directly investigate whether eye movements accompanied the “return” resets that participants reported.
-However, the modestly better performance of the spatial limit model here suggests that spontaneous resets are not randomly distributed - as they might be if eye-movements or shifts of attention are the only way to elicit resets.
+However, the random distribution of resets throughout time are in line with a mechanism based on distracted attention, in some cases leading to eye-movement planning.
 
-We find that manual tracking of the double-drift stimulus during its motion did show the expected illusion.
-Few if any resets were detected during the online tracking, either because of the additional attentional requirements or because of the short duration (2 s).
+We find that manual tracking of the double-drift stimulus during its motion did show the expected illusion, although not always as strong.
+Few if any resets were detected during online tracking, perhaps because of the additional attentional requirements or the short duration (2 s), although interactions with the simultaneous hand movements could explain this too.
 Participants did report spontaneous resets in delayed, single-pass re-tracing of Experiment 2.
-The resets were best explained by a limiting spatial offset from the true stimulus location, but do not exclude a temporal limit.
-These findings may either indicate there are multiple causes for spontaneous resets or that a re-tracing method is not suitable to resolve what causes this phenomenon.
-Either way, both some amount of time from stimulus onset until reset and too large a conflict between retinal and perceived position leading to resets remain possible.
+The resets were best explained by a gamma distribution on reset time.
+This finding seems to indicate that spontaneous resets occur independently from the stimulus or how it is perceived, and spontaneous reset may even be random.
+If there is a mechanism causing resets we have not uncovered it here, and if resets are independent from perception, the illusory percept itself may still be limited, and give rise to how resets are perceived.
 

### 2021-12-03 R/common.R
added color-mixing function

@@ -594,6 +594,21 @@ linPal <- function(from='#FFFFFF', to='#E51636', n=256, alpha=NULL) {
   
 }
 
+mixCol <- function(a='#FFFFFF', b='#E51636', balance=c(1,1)) {
+  
+  a <- col2rgb(a,alpha=TRUE)/255
+  b <- col2rgb(b,alpha=TRUE)/255
+  
+  w <- balance / sum(balance)
+  
+  R <- (a[1]*w[1]) + (b[1]*w[2])
+  G <- (a[2]*w[1]) + (b[2]*w[2])
+  B <- (a[3]*w[1]) + (b[3]*w[2])
+  A <- (a[4]*w[1]) + (b[4]*w[2])
+  
+  return(rgb(red=R, green=G, blue=B, alpha=A))
+  
+}
 
 
 

### 2021-12-03 R/exp2_models.R
fixed model plotting code to actually fit/use the correct models

@@ -2691,7 +2691,7 @@ plotModels <- function(target='inline') {
   
   df <- getDataTrials()
   
-  modelfits <- fitSomeModels(df, Xnormal=FALSE)
+  modelfits <- fitSomeModels(df)
   
   colors <- getColors()
   
@@ -2719,7 +2719,8 @@ plotModels <- function(target='inline') {
   title(xlab='reset X [cm]', line=2.4)
   title(ylab='reset Y [cm]', line=2.4)
   
-  Lx <- modelfits$XlimOrth$par['Lx']
+  Lx <- modelfits$XdistNormal$par['mXn']
+  #print(Lx)
   lines(x   = rep(Lx,2),
         y   = c(0,13.5), 
         col = colors$blue$s,
@@ -2729,7 +2730,7 @@ plotModels <- function(target='inline') {
           col = colors$blue$s,
           border=NA)
   
-  Lt <- as.numeric(modelfits$TlimOrth$par['Lt'])
+  Lt <- as.numeric(modelfits$TdistNormal$par['mTn'])
   for (speed in c(3,4)) {
     lines(x   = sin(seq(0,pi/2,length.out = 50)) * Lt * speed,
           y   = cos(seq(0,pi/2,length.out = 50)) * Lt * speed,

### 2021-12-03 R/exp2_retracing.R
add confidence intervals to the 6 re-tracing conditions for the illusion-strength plot

@@ -1381,19 +1381,29 @@ plotResetPoints <- function(target='inline') {
           border=NA)
   lines(c(0,75),group_k*c(0,75),col=colors$lightblue$s,lty=1)
   
+  condCI <- aggregate(angle ~ Vi + Ve, data=avg_df_p, FUN=getConfidenceInterval)
+  condCI$vector_prediction <- (atan( condCI$Vi / condCI$Ve ) / pi ) * 180
+  
   for (speed in c(3,4)) {
     
-    idx <- which(avg_df_g$Ve == 13.5/speed)
+    
     speedcol <- c(colors$orange$s,colors$orange$s)[speed-2]
-    speedpch <- c(1,2)[speed-2]
-    points(avg_df_g$vector_prediction[idx],avg_df_g$angle[idx],col=speedcol,pch=speedpch,cex=2)
+    
+    for (Vi in unique(condCI$Vi)) {
+      idx <- which(condCI$Vi == Vi & condCI$Ve == 13.5/speed)
+      lines(rep(condCI$vector_prediction[idx],2),condCI$angle[idx,c(1,2)],col=t_col(speedcol,percent=81.5),lw=6,lend=1)
+    }
+    
+    idx <- which(avg_df_g$Ve == 13.5/speed)
+    speedpch <- c(16,17)[speed-2]
+    points(avg_df_g$vector_prediction[idx],avg_df_g$angle[idx],col=speedcol,pch=speedpch,cex=1.5)
     
   }
   
   legend(x=0, y=45, 
          legend = c('group means (3 s)', 'group means (4 s)', 'Heller et al. (2021)', sprintf('K=%0.2f',group_k)),
          col=c(colors$orange$s, colors$orange$s, 'black', colors$lightblue$s), 
-         pch=c(1,2,NA,NA), lty=c(0,0,1,1), 
+         pch=c(16,17,NA,NA), lty=c(0,0,1,1), 
          bty='n', cex=1)
   
   axis(side=1,at=seq(0,75,15))

### 2021-12-03 R/exp3_gaborjumps.R
add gamma distribution model to PSE plot (but it doesn't work as intended, unless the poisson PSEs are also plotted)

@@ -202,12 +202,12 @@ ANOVAonPSEs <- function() {
 
 # figures -----
 
-plotJumpPSEs <- function(target='inline', individualDots=FALSE) {
+plotJumpPSEs <- function(target='inline', individualDots=FALSE, modelDots=FALSE, gammaDots=FALSE) {
   
   colors <- getColors()
   
   if (target == 'pdf') {
-    pdf( file = 'doc/fig_6_gaborjump_exp_alt.pdf', width=6, height=4 )
+    pdf( file = 'doc/fig_6_gaborjump_exp.pdf', width=6, height=4 )
   }
   
   layout(mat = matrix(c(1,2), 
@@ -224,19 +224,27 @@ plotJumpPSEs <- function(target='inline', individualDots=FALSE) {
   
   par(mar=c(1,1,2.2,1))
   
-  if (individualDots) {
-    pos <- 0.15 # Plot OffSet
-  } else {
-    pos <- 0
+  # 1 data, 2 individual dots, 3 model fits
+  pos <- c(0,0,0)
+  
+  if (individualDots & !modelDots) {
+    pos <- c(0.2, -0.2, 0) # Plot OffSet
+  }
+  if (!individualDots & modelDots) {
+    pos <- c(-0.2, 0, 0.2)
+  }
+  if (individualDots & modelDots) {
+    pos <- c(0, -0.3, 0.3)
   }
   
   plot(methodsFig,xlim=c(0,width),ylim=c(0,height),asp=1)
   title('A: methods', font.main=1, cex.main=1.5, adj=0, line=0.5)
   
-  
   PSEdf <- read.csv('data/gaborjump/exp_PSEs.csv', stringsAsFactors = FALSE)
   groupPSEs <- PSEdf[which(PSEdf$participant == 'group'),]
   participantPSEs <- PSEdf[which(PSEdf$participant != 'group'),]
+  
+  modelPSEs <- read.csv('data/gaborjump/modelfitPSEs.csv', stringsAsFactors = FALSE)
   
   par(mar = c(2.8,2.8,2.2,0.1),
       mgp = c(2, 0.55, 0) )
@@ -277,10 +285,21 @@ plotJumpPSEs <- function(target='inline', individualDots=FALSE) {
       pPSEs <- participantPSEs$PSE[which(participantPSEs$path_length == pathlength & participantPSEs$stimdur == stimdur)]
       
[…]

### 2021-12-09 Manuscript.Rmd
include Patrick's feedback in manuscript

@@
     df_print: paged
+  pdf_document: default
   word_document:
     reference_docx: template.docx
-  pdf_document: default
 header-includes:
@@
 To report the perceived trajectory of the gabor, participants drew it after the gabor was removed from the monitor.
-Resets were detected and analyzed and their distribution matches neither a temporal nor a spatial limit, but rather a wide gamma distribution over time.
+Resets were detected and analyzed and their distribution matches neither a temporal nor a spatial limit, but rather a broad gamma distribution over time.
+This suggests that resets are triggered randomly, about once per 1.3 _s_, possible by extraneous distractions or eye movements.
 
@@
 
-Here we use hand movements to investigate where or when the double-drift illusion resets.
+Here we first investigate wether hand-tracking of the gabor demonstrates the illusion.
 Eye and hand movement are differentially sensitive to many illusions, such as the Ponzo illusion [@Gamble2017], Ebbinghaus illusion [@Knol2017], flash-lag effect [@Blohm2003], the Duncker illusion [@Soechting2001] and many others.
 Similarly, it has been shown that the double-drift illusion affects pointing movements, but not saccades [@Lisi2017], and hand movements have been used to test other perceptual phenomena [@PatricioDecima2022].
-Hence, we expect that hand trajectories will be an effective method to capture the full illusory trajectory, rather than a point estimate along the perceived path, and this will allow capturing resets of the double-drift illusion.
+Hence, we expect that hand trajectories will be an effective method to capture the full illusory trajectory, rather than a point estimate of its endpoint [@Hui2020] or its overall angle [@Lisi2015].
 
-The strength of the illusory percept; the deviation of the perceived path of the gabor from its true path, depends on the speed of both the internal and external motion [@Tse2006; @Shapiro2010; @Kwon2015; @Cavanagh2019; @Heller2021].
-We want to test how well a continuous trajectory reflects the percept of the illusion.
-We assess this in our first experiment, using hand movement trajectories recorded during stimulus presentation (Fig.
+The strength of the illusory percept — the deviation of the perceived path of the gabor from its true path — depends on the speed of both the internal and external motion [@Tse2006; @Shapiro2010; @Kwon2015; @Cavanagh2019; @Heller2021].
+In this first experiment, we test how well manual tracking reflects the percept of the illusion.
+To do so, we use hand movement trajectories recorded during stimulus presentation (Fig.
 1).
 We compare trajectories across 5 different internal motions of the gabor and compare the illusion strength with previous work based on a straightforward vector combination model [@Heller2021].
-As we expected, online tracking provides a similar measure of illusion strength as pointing.
-However, only few possible resets are apparent in the data (>4% trials), and no limit to the illusion is clear.
+As we expected, online tracking provides a measure of illusion strength that is similar to that found with pointing [@Lisi2017].
+However, the traces reveal only a few possible resets (>4% trials), and no limit to the illusion is clear.
 
-In our second experiment, we therefore slow down the external stimulus motion, and have participants reproduce the perceived trajectory after stimulus presentation.
-Here we do observe spontaneous resets, which allows us to investigate any spatial or temporal limits of the illusion.
-However, we find that there is no hard temporal or spatial limit leading to resets, but it seems that resets occur randomly over time, based on a gamma distribution.
+To study resets, we therefore slow down the external stimulus motion in our second experiment, and have participants reproduce the perceived trajectory after stimulus presentation.
+Now we do observe spontaneous resets (64% of trials) that allow us to investigate the spatial and temporal limits of the illusion.
+We find that there is no hard temporal or spatial limit leading to resets.
+Instead, resets appear to occur randomly over time, based on a gamma distribution.
 
@@
 A variant on online tracking has been used before [@Cormack2019] with a continuous task where the stimulus depended on the response.
-We test if tracking the illusion continuously is a good measure of illusion strength, or if it prevents or decreases the illusion.
+Here, we test if tracking the illusion continuously is a good measure of illusion strength, or if it prevents or decreases the illusion.
 Additionally, we use a relatively long trajectory to allow the double-drift illusion to exceed any perceptual limits it may have, so that we can observe what these are.
@@
 In general, the traced paths approximate linearity, with the direction depending on the internal motion.
-**B:** Four trials with a possible reset.
+**B:** Four sample trials with anomalous deviations that may be resets.
 **C:** Average heading of the middle 1 _s_ of each pass.
@@
 Other percepts of resets have been reported as well, but as far as we know these fall in between these two extremes.
-Such resets may occur spontaneously once the accumulation has gone on too long or too far, depending on what mechanism causes resets.
-Where resets occur would then both depend on the mechanism that causes resets, as well as on the strength of the illusion (purple arrows in Fig.
-3) which in turn depends on the stimulus properties as well as variation in perception within and between participants.
+Such resets may occur spontaneously once the accumulation has gone on too long or too far, depending on what mechanism causes the resets, and the strength of the illusion (purple arrows in Fig.
+3).
+The illusion strength in turn in turn depends on the stimulus properties (speeds, gabor size, carrier frequency, eccentricity; @Gurnsey2012) as well as variations within and between participants.
 The purpose of this experiment is to capture this spontaneous resetting and determine its source.
@@
 2A,B) mostly showed a linear trajectory without saturation or reset of the types shown in Figure 3 (left).
-The continuation of the accumulation may be a result of the active tracking – either the extra attention required to track increased the extent of the accumulation or possibly the temporal demands of the tracking caused the participants to average their traces over any breaks.
-Alternatively, post-dictive inference may prevent resets from being perceived.
-To address this, we switched to a delayed, offline recording of the perceived path.
+The continuation of the accumulation without resets may be a result of the active tracking.
+We do not know precisely why this occurred but we speculate it was either because 1) the extra attention required to track decreased the chance of attentional distraction introducing resets [@Nakayama2020] or 2) the temporal demands of the drawing the trace in real time caused the participants to miss the resets that did occur, or 3) the resolution of online tracing at these speeds is poor and participants may have just averaged their traces over any breaks.
+In addition, post-dictive inference may prevent resets from being perceived.
+To address the possibility that it was the active tracking that suppressed the resets, we switched to a delayed, offline recording of the perceived path and the participants' trajectory now showed frequent resets.
 
@@
 We were interested, in particular, in whether the limitation of the accumulation would be set by space or time.
-Since both the true and the illusory position appear to be available in the visual system (the true position drives saccades whereas the illusory position drives perception, @Lisi2015), it is possible that resets occur once the distance between the real and perceived positions exceeds some limit.
+Since both the true and the illusory position appear to be available in the visual system (the true position drives saccades whereas the illusory position drives perception, @Lisi2015 ), it is possible that resets occur once the distance between the real and perceived positions exceeds some limit.
 We can speculate that this spatial limit would be related to the positional uncertainty of the gabor.
@@
 For this experiment, 9 participants were recruited from the lab (6 female; ages 19 - 27, mean: 22.8).
-All participants reported to be right handed and had either normal or corrected to normal vision.
+All participants reported themsleves as right handed and had either normal or corrected-to-normal vision.
 Procedures were in accordance with the Declaration of Helsinki (2003) and were approved by York’s Human Participants Review Committee.
@@
 Each block used all 6 combinations of two external speeds ( corresponding to 3 or 4 _s_ presentation time ), and three internal speeds ( 2, 3 or 4 _cps_ ), 6 times, for a total of 36 trials per block.
-Each of the combinations of internal and external speed was presented 24 times in total, so that we have 144 trials for each participant, and 1296 trials in total.
+Each of the combinations of internal and external speed was presented 24 times in total, so that we had 144 trials for each participant, and 1296 trials in total.
 
@@
 
-The individual trajectories reveal some points that are potential spontaneous resets that we try to extract.
-We created a heuristic to extract likely reset points based on right-left inflections in the trajectories.
-First, we remove the first and last 4 mm of the trajectory, as well as any short segments on the trajectory at the start and end that go toward the participant (i.e.
+The individual trajectories revealed inflections of direction that were potential spontaneous resets and we created a heuristic to localize them.
+First, we removed the first and last 4 mm of the trajectory, as well as any short segments on the trajectory at the start and end that went toward the participant (i.e.
 more than 180 degrees away from the true external motion), to remove any unintended jitters caused by starting and ending the movement.
-Then, we apply a bi-directional low-pass Butterworth filter with a cut-off frequency of 1.5 Hz.
-However, since participants draw at very different speeds, we first linearly interpolate the trajectories such that the sampling rate is normalized to 30 Hz for all trajectories, based on the stimulus duration.
-In these summary trajectories we detect the first inflection point using the X coordinates.
-This could put the reset point too late for hit-the-wall resets, so that we move back to get the previous sample at 95% of the X coordinate at the inflection point.
-Then, we locate the sample in the raw trajectory that is closest to this point.
-Finally, we remove a few reset points that are detected very close to the veridical path without any build-up of the illusion by only keeping reset points that fell 5 mm within the quadrant where we expect them.
-This heuristic will result in some misses and some false alarms, and while changing the algorithm may improve performance, there is no ground truth to evaluate this.
-All raw trajectories and the detected reset points can be found on the OSF repository.
+Then, we applied a bi-directional low-pass Butterworth filter with a cut-off frequency of 1.5 Hz.
+However, since participants draw at very different speeds, we resample them at 30 Hz using linear interpolation, giving 90 samples for the 3 _s_ trajectories and 120 samples for the 4 _s_ trajectories.
+After filtering, we detected the first inflection point in these summary trajectories using only the x coordinates (local maximum of x).
+This could put the reset point too late for saturation or “hit-the-wall” resets, so that we moved back to get the previous sample at 95% of the x coordinate at the inflection point.
+Then, we located the sample in the raw trajectory that was closest to this point.
+We then excluded a few reset points that were less than 5 mm, within the expected quadrant.
+Those that were very close to y<5 _mm_ are unlikely resets because they either have extremely high illusion strength, or not enough room or time for actual accumulation of illusory drift.
+Those that were very close to the veridical path (x<5 _mm_) likely don't have any illusory drift, and were probably random wiggles in an attempt to make a vertical tracing.
+Out of 1296 trials, the heuristic identified reset points in 827 (~64%).
+This heuristic likely yielded some misses and some false alarms, and while changing the algorithm may improve performance, there is no ground truth to evaluate this.
+All raw trajectories and the detected reset points could be found on the OSF repository.
 
@@
 
-Apart from reset points, we need a measure of illusion strength.
-We take a trajectory sample halfway between the starting point and the detected reset point.
-We then use the angular difference between the gabor's real trajectory and a straight line drawn through that point and the start of the trajectory as a measure of illusion strength (see Fig.
-4).
+Apart from reset points, we also extracted a measure of illusion strength.
+We took a trajectory sample at half the distance between the starting point and the detected reset point.
+We then used the angular difference between the gabor's real trajectory and a straight line drawn through that point and the start of the trajectory as a measure of illusion strength.
 
@@
 
-Four example trajectories and the location of resets determined by our heuristic are shown in Figure 4.
-On average, this method detects resets in ~64% of trials.
+Four example trajectories and the location of the first resets determined by our heuristic are shown in Figure 4 (A-D).
+The x-axis here indicates illusory drift, since the gabor moved with a constant speed along the y-axis.
+Distance to origin can be converted to the time of reset by dividing by speed.
+On average, our heuristic detected resets in ~64% of trials.
 For two participants, less than half the trials showed resets (9% and 22%).
@@
 3) as opposed to a "jump" reset, might not always display a detectable direction change in their paths, most of these percentages of trials with resets seems reasonable.
-However, the participant with 9% resets will not be used for calculating illusion strength as they have no resets in several conditions.
-This means we analyse a total of 827 reset points (Fig.
-4, right).
+However, the participant with only 9% of trials showing a reset was excluded for calculating illusion strength as they had no resets in several conditions.
+This means we analysed a total of 827 reset points (Fig.
+4E).
 
@@
 **C:** Slow return reset.
-**D:** Continued accumulation of the illusion.
+**D:** Continued accumulation of the illusion without any reset.
 **E:** Distribution of all 827 detected reset points.
-**F:** Using the angle to a point along the trajectory at half the distance to the reset point as a measure of illusion strength, we can compare this data to other data.
+**F:** Using the angle from the origin of the trace to a point along the trajectory at half the distance between origin and reset point as a measure of illusion strength, we can compare this data to other data.
 For this graph, we exclude one participant who produced very few resets.
-The illusion is weaker ($K=0.60$, blue line) and its strength varies between participants, but it is nevertheless clearly there."}
+The illusion is weaker than in Experiment 1 ($K=0.60$, blue line) and its strength varies between participants, but it is nevertheless clearly there."}
 #plotExampleData()
@@
 This would be consistent with a spatial limit to the illusion resulting in resets.
-However, there is also considerable variance, so that fitting any line through the data, will necessarily leave errors.
+However, there is also considerable variance along the x-axis, so that fitting any vertical line through the data, (a spatial limit, the blue line in Fig 5A) provides a poor explanation of the pattern of resets.
 
-We test this with a series of simple models.
-If there is a spatial offset limit that triggers resets, the reset points should fall on, or be close to a line with a constant X coordinate.
-If there is a temporal limit, all resets should have the same or a similar reset time (distance from origin / speed).
-If we assume errors are normally distributed, we can then fit a normal distribution (with 2 parameters: mean and standard deviation) on both the X coordinates and the time points of resets.
-The distributions of neither the X coordinates nor the time points of resets appear normally distributed though, but seem better approximated by a gamma distribution (see Fig 5A).
-So we additionally fit gamma distributions (with 2 parameters: rate and shape) on the X coordinates and time points of resets as well.
-We then calculate the likelihoods of these four simple distribution models both using all resets and within individual participants and use this to calculate an AIC to determine a relative probability that compares models.
+We examined 4 simple models.
+If there is a spatial offset limit that triggers resets, the reset points should fall on, or be close to a line with a constant x coordinate (blue line in Fig.
+5A).
+If there is a temporal limit, all resets should have the same or a similar reset time (distance from origin / speed, the red arcs in Fig.
+5A).
+If we assume thaat these spatial or temporal limits have some random variation from trial to trial, we can fit a normal distribution (with 2 parameters: mean and standard deviation) on both the x coordinates and the time points of resets.
+However, neither the x coordinates, nor the time points of resets were normally distributed (both _p_<.001, Kolmogorov-Smirnov tests of data against optimal normal distribution) suggesting that there is no hard space or time limit that triggers resets.
+We then used gamma distributions to capture any processes where the resets happen at random time in a trial or a random illusory offset.
+So we additionally fit gamma distributions (with 2 parameters: rate and shape) on the X coordinates and time points of resets as well (see Fig 5A, marginal distributions; Fig 5B,C).
+We then calculate the likelihoods of these four simple distribution models both using all resets and within individual participants and use this to calculate an AIC to determine a relative probability to compare the models.
 
@@
 Marginal distributions show that reset points time and spatial offset approximate a gamma distribution.
+**B,C:** Color illustrates the gamma probability distributions, where white is zero probability, and the darkest red or blue is the maximum probability density.
+This is a relatively scale-free unit, where the total surface of the full probability density function is 1.
+The data and likelihood distributions match fairly well.
 **B:** Gamma distribution on time (log likelihood: -766.6).
 **C:** Gamma distribution on spatial offset (log likelihood: -1319.3)."}
-#plotResetPoints()
 invisible(capture.output(plotModels()))
-#plotBootstrappedFits()
-#plotIllusionStrength()
 ```
 
-Within the normal distribution models of reset points (means in Fig 5A, blue: spatial, red: time), the time-based distribution fits better for the group data ( _p_<.001 ) as well as all individual participants, and the same holds within gamma distribution models of reset models (Fig 5B,C).
+For the space and time limit with normal distributions (means in Fig 5A, blue: spatial, red: time), the time-based distribution fits better for the group data (normal space AIC=2,834.83, normal time _AIC_=1,677.98; _p_<.001) as well as all individual participants, and the same holds within gamma distribution models of reset models (gamma space AIC=2,642.69, gamma time AIC=1,537.10; _p_<.001).
 Furthermore, the gamma-based distribution on time of resets performed better than the normal distribution of reset time in the pooled data ( _p_<.001 ), and this holds for most participants (6/9) as well.
-This means that time-based distributions explain our reset points better than models assuming a spatial offset.
+This means that time-based distributions explain our reset points better than models assuming a spatial offset, so that the time-based gamma distribution was the most successful of all 4.
 
@@
 We first tested whether manual tracking of a double-drift stimulus is susceptible to the illusion as reported by perceptual judgments, pointing [@Lisi2017], and memory saccades [@Massendari2018; @Ueda2018].
-The results of this online tracking experiment show a close agreement between the angle of the manual tracking and the illusion strength expected based on the model from earlier perceptual measures [@Heller2021].
-However, we were also interested in  spontaneous resets and only ~4% of trials show a deviation in the trajectory that might reflect a reset.
-In the second experiment, 62% of trials showed a reset, and 36% of those occurred within 1 _s_ and 84% within 2 _s_.
-Based on these numbers one might expect spontaneous resets in at least 22.5% of each of the two half passes in the online tracking experiment and 52.5% of each of the full 2 _s_ passes - if each pass is independent.
-This would mean ~86% of trials would show at least 1 spontaneous reset which is much larger than the ~4% of trials we observe ( _p_<.001, exact binomial test).
-The question remains as to why so few spontaneous resets occurred in this experiment.
-Simple explanations could be that duration of stimulation presentation requires relatively fast hand movements, leaving little time for drawing spontaneous resets or reduced attention for the double-drift stimulus.
+The results of this online tracking experiment showed a close agreement between the angle of the manual tracking and the illusion strength expected based on the model from earlier perceptual measures [@Heller2021].
+However, we were also interested in spontaneous resets and only ~4% of trials in this first experiment had deviations that could have been a reset.
+In contrast, in the second experiment, 63.8% of trials showed a reset.
+There was a reset within 1 _s_ in 22.3% of trials, and within 2 _s_ in 52.2% of trials.
+We can use these results to determine the rate of spontaneous resets that would have been expected in the first experiment.
+If each pass is independent and the perceived trajectories were equivalent in the two experiments, this would mean ~98.5% of trials of Experiment 1 should have shown at least 1 spontaneous reset which is far larger than the ~4% of trials we observed ( _p_<.001, exact binomial test).
+We are not sure why so few spontaneous resets occurred in the first experiment.
+Possible explanations could be either that online tracking requires relatively fast hand movements, leaving little time for drawing the details of any spontaneous resets or the demands of the online tracking increased the attention to the double-drift stimulus and with less chance of attentional distraction, the likelihood of resets was reduced [@Nakayama2020].
 The one feature that sets this task apart from others is that the tracking is done while perceiving the stimulus, such that action could affect perception.
-One possible mechanism is that predicted sensory consequences of the hand movement inform perception ("I will move my hand in this direction, so I expect to see motion in that direction as well").
+One possible mechanism is that predicted sensory consequences of the hand movement influence perception ("I will move my hand in this direction, so I expect to see motion in that direction as well").
 And another is post-dictive inference ("my hand has moved in this direction so I must have seen the visual stimulus move in the same direction").
@@
 In our second experiment, we asked participants to draw the perceived path of the stimulus after observing a single pass.
-The orientation of the drawn traces reflect the strength of the illusion, and in a majority of trials (62%) there was a clearly indentifiable reset point, for either a "return" reset or a "hit-the-wall" reset.
-We then set out to test if these spontaneous resets are triggered by a limiting distance of drift away from the physical location, or by a specific limit of time.
-A gamma distribution of reset time seems to explain this data best.
+The orientation of the drawn traces reflect the strength of the illusion, and in a majority of trials (63.8%) there was a clearly indentifiable reset point, of either a "return" or a "hit-the-wall" type.
+We then set out to test if these spontaneous resets were triggered by a limiting distance of drift away from the physical location, or by a specific limit of time.
+Neither of these provided a viable explanation of the pattern of resets.
+Resets points are broadly distributed throughout the workspace, suggesting that resets are not linked to illusion strength or stimulus onset, but may caused by a more "random" process.
 
-However, it does not seem like there is a specific temporal limit on time, i.e.
-resets do not occur at a given time after stimulus onset, but appear randomly sampled from a wide distribution.
-In part, this may rely on the hand trajectories we recorded here to provide an accurate measure of reset points, as they may contain an unknown amount of noise.
+We observe that reset times appear randomly sampled from a broad gamma distribution.
+In part, this may rely on how (un)suitable hand trajectories are as an accurate measure of reset points, as they may contain an unknown amount of noise.
 Therefore, we compare our results with those from an as of yet unpublished work [pre-print: @Liu2021].
-This study used a psychophysics-based approach to probe how spontaneous resets might affect the accumulated illusion throughout time (1, 2 or 3 _s_) while keeping the actual path (2 or 4 _dva_) and maximum illusion (1.4 or 2.8 _dva_, respectively) constant.
+This study probed how spontaneous resets might affect the accumulated illusion throughout time (1, 2 or 3 _s_) while keeping the actual path (2 or 4 _dva_) and maximum illusion constant (45° deviation, so 1.4 or 2.8 _dva_ offsets, respectively).
 Spontaneous resets should decrease the accumulated illusion (see Fig 6A).
@@
 
-The data from 10 participants shows that as a stimulus is presented for longer, the illusory offset at the end gets smaller (Fig 6B, round dots).
+The data from 10 participants showed that for a fixed path and illusion strength, longer durations produced smaller illusory offsets at the end (Fig 6B, round dots).
 This can be explained by (among others) resets randomly occuring throughout time.
@@
 5A) but seem to fall on a fairly wide gamma distribution on reset time (Fig.
-5B), which is at odds with a set limit (either spatial or temporal) where resets occur.
+5B), which is at odds with a set limit (either spatial or temporal) where saturation or resets occur.
 This seems to argue against a perceptual mechanism underlying resets, such as on going Bayesian inference until the discrepancy between retinal and perceived position signals is too large.

### 2021-12-09 R/exp1_tracking.R
remove K=0.60 line from exp 1 figure

@@ -657,7 +657,7 @@ plotBoundedTracking <- function(target='inline', discontinuous=FALSE) {
   angles <- c(0,pi/4)
   # Heller et al. (2021):
   lines(angles,0.74*((angles/pi)*180),col='black',lty=1)
-  lines(angles,0.60*((angles/pi)*180),col='black',lty=2)
+  #lines(angles,0.60*((angles/pi)*180),col='black',lty=2)
   
   # individual participants
   points(xcoords, (df$heading/pi)*180, col=colors$lightblue$s, pch=1, cex=1.0)
@@ -682,9 +682,9 @@ plotBoundedTracking <- function(target='inline', discontinuous=FALSE) {
   
   
   legend(x=0, y=45, 
-         legend=c('Heller et al. (2021), K=0.74', 'K=0.60', 'participants (N=4)', 'average'), 
-         col=c('black','black', colors$lightblue$s, colors$yorkred$s), 
-         pch=c(NA,NA,1,1), lty=c(1,2,0,0), 
+         legend=c('Heller et al. (2021), K=0.74', 'participants (N=4)', 'average'), 
+         col=c('black', colors$lightblue$s, colors$yorkred$s), 
+         pch=c(NA,1,1), lty=c(1,0,0), 
          bty='n', cex=1)
   
   #axis(side=1,at=seq(0,pi/4,pi/8),labels=c('0',expression(pi/8),expression(pi/4)))

### 2021-12-11 Manuscript.Rmd
integrated Patrick's changes and feedback on Discussion

@@
 5A).
-If we assume thaat these spatial or temporal limits have some random variation from trial to trial, we can fit a normal distribution (with 2 parameters: mean and standard deviation) on both the x coordinates and the time points of resets.
-However, neither the x coordinates, nor the time points of resets were normally distributed (both _p_<.001, Kolmogorov-Smirnov tests of data against optimal normal distribution) suggesting that there is no hard space or time limit that triggers resets.
-We then used gamma distributions to capture any processes where the resets happen at random time in a trial or a random illusory offset.
-So we additionally fit gamma distributions (with 2 parameters: rate and shape) on the X coordinates and time points of resets as well (see Fig 5A, marginal distributions; Fig 5B,C).
-We then calculate the likelihoods of these four simple distribution models both using all resets and within individual participants and use this to calculate an AIC to determine a relative probability to compare the models.
+If we assume that these spatial or temporal limits do allow a little random variation from trial to trial, we can fit a normal distribution (with 2 parameters: mean and standard deviation) on both the x coordinates and the time points of resets.
+However, neither the x coordinates, nor the time points of resets were normally distributed (both _p_<.001, Kolmogorov-Smirnov tests of data against optimal normal distribution).
+Additionally, if resets are triggered by either a spatial or temporal limit the data would all fall relatively close to it, but this does not appear to be the case.
+
+These first two models assumed that resets were triggered at fixed locations in space or time, but the next tro propose that the resets are random processes that occur with some distribution across space or time.
+We used gamma distributions to capture these processes (with 2 parameters: rate and shape) on the X coordinates and time points of resets as well (see Fig 5A, marginal distributions; Fig 5B,C).
+To compare across the four models, we then calculated the likelihoods of their predictions (based on the underlying distributions' probability density functions) using all resets together and also within individual participants.
+From these likelihood values, we calculated an AIC to determine a relative probability for each model.
 
@@
 **A:** Model limits relative to reset point locations.
-Reset points neither fall along a spatial limit (blue line) or a temporal limit (red lines: continuous for 4 _s_ passes, dashed for 3 _s_ passes).
+Reset points neither fall along a spatial limit (blue line; log likelihood: -1415.4) or a temporal limit (red lines: continuous for 4 _s_ passes, dashed for 3 _s_ passes; log likelihood: -837.0).
 Marginal distributions show that reset points time and spatial offset approximate a gamma distribution.
@@
 There was a reset within 1 _s_ in 22.3% of trials, and within 2 _s_ in 52.2% of trials.
-We can use these results to determine the rate of spontaneous resets that would have been expected in the first experiment.
-If each pass is independent and the perceived trajectories were equivalent in the two experiments, this would mean ~98.5% of trials of Experiment 1 should have shown at least 1 spontaneous reset which is far larger than the ~4% of trials we observed ( _p_<.001, exact binomial test).
+We can use these results to determine the rate of spontaneous resets that would have been expected in the first experiment if it had the same reset dynamics as the second experiment.
+If each pass was independent and the perceived trajectories were equivalent in the two experiments, this would mean ~98.5% of trials of Experiment 1 should have shown at least 1 spontaneous reset which is far larger than the ~4% of trials we observed ( _p_<.001, exact binomial test).
 We are not sure why so few spontaneous resets occurred in the first experiment.
-Possible explanations could be either that online tracking requires relatively fast hand movements, leaving little time for drawing the details of any spontaneous resets or the demands of the online tracking increased the attention to the double-drift stimulus and with less chance of attentional distraction, the likelihood of resets was reduced [@Nakayama2020].
-The one feature that sets this task apart from others is that the tracking is done while perceiving the stimulus, such that action could affect perception.
-One possible mechanism is that predicted sensory consequences of the hand movement influence perception ("I will move my hand in this direction, so I expect to see motion in that direction as well").
-And another is post-dictive inference ("my hand has moved in this direction so I must have seen the visual stimulus move in the same direction").
+We first suggest two simple explanations for the difference: 1) Online tracking required relatively fast hand movements, and therefore left little time for drawing the details of any spontaneous resets.
+2) The demands of the online tracking increased the attention to the double-drift stimulus and with less chance of attentional distraction, the likelihood of resets was reduced [@Nakayama2020].
+However, this task, unlike previous tests of the double-drift stimulus, may involve an interaction between action and perception.
+For example, the predicted sensory consequences of the hand movement may influence perception ("I will move my hand in this direction, so I expect to see motion in that direction as well").
+Or vice versa, as in post-dictive inference ("my hand has moved in this direction so I must have seen the visual stimulus move in the same direction").
 But this is a question for future research.
@@
 Neither of these provided a viable explanation of the pattern of resets.
-Resets points are broadly distributed throughout the workspace, suggesting that resets are not linked to illusion strength or stimulus onset, but may caused by a more "random" process.
+Resets points are broadly distributed, suggesting that resets are not linked to illusion strength or stimulus onset, but may be caused by a more "random" process.
 
-We observe that reset times appear randomly sampled from a broad gamma distribution.
-In part, this may rely on how (un)suitable hand trajectories are as an accurate measure of reset points, as they may contain an unknown amount of noise.
-Therefore, we compare our results with those from an as of yet unpublished work [pre-print: @Liu2021].
-This study probed how spontaneous resets might affect the accumulated illusion throughout time (1, 2 or 3 _s_) while keeping the actual path (2 or 4 _dva_) and maximum illusion constant (45° deviation, so 1.4 or 2.8 _dva_ offsets, respectively).
-Spontaneous resets should decrease the accumulated illusion (see Fig 6A).
-After the main stimulus presentation, there is a 350 _ms_ gap, during which the double-drift stimulus is not shown, and this is know to evoke a reset.
+Recently, it has been suggested that resets can be triggered by a distraction of attention [@Nakayama2020].
+The participants in their experiment also retraced the perceived path and reported “return” resets.
+Nakayama and Holcombe [-@Nakayama2020] claim that planned eye movements can trigger a “return” reset back to the physical path.
+In our experiment, we did not control for eye-movements, so we can't directly investigate whether eye movements accompanied the “return” resets that participants reported.
+However, the random distribution of resets throughout time are in line with a mechanism based on distracted attention, which may involve eye-movement planning and execution.
+The rate of resets that we found here, ~0.75 / _s_, are too infrequent for microsaccades (~3 / _s_) so the distractions may be more significant than these small eye movements.
+On the other hand, random occurrance of distractions, a fraction of which triggers a reset, is in line with the kind of processes underlying gamma distributions [e.g.
+@McGill1965; @Wolfe2002].
+
+We observe that reset times were well modeled by broad gamma distribution, suggesting that resets were best explained as occurring randomly in time with an approximate rate of 1 every 1.3 _s_.
+However, we have to qualify our conclusions given the noisy nature of hand-drawn trajectories as a measure of the perceived trajectory and the reset points.
+Therefore, we compare our results with those from s similar study that analyzed trajectory endpoints [pre-print: @Liu2021].
+In this study, the internal speed of the Gabor texture was adjusted across conditions and participants so that the illusion always had a local direction of 45° compared to its actual vertical direction.
+This then generated a highly predictable offset of the end of the trajectory that would be $\sqrt{2}$ of the path length if there were no spontaneous resets.
+Anything less than that revealed the presence and size of spontaneous resets.
+Liu et al.
+tested 3 different durations (1, 2 and 3 s) and two path lengths (2 and 4 dva) for the first segment of the Gabor’s motion (see Fig.
+6A).
+They then presented a 350 _ms_ gap during which the double-drift stimulus is not shown, and this is known to evoke a full reset to the physical location [@Lisi2015].
 After this gap-evoked reset the illusion resumes accumulating from the veridical position where the gabor appears after the gap [@Lisi2015].
-By changing the location where the double-drift reappears and asking participants if it jumped left or right, the point of subjective equivalence that expresses the size of the gap-evoked jump reset can be determined.
+By changing the location where the double-drift reappeared and asking participants if it shifted left or right, the endpoint of the first segment was determined.
+The difference between that point and the expected value for no spontaneous resets indicated the presence and size of any that had occurred.
 
@@
 Liu et al.
-[-@Liu2021] present a Poisson process model with a constant chance of reset occurring every 10 ms of ~0.007 or ~0.7 / _s_ (Fig 6B, light diamonds).
-If we compare this to the gamma distribution fit, with a rate of $2.81$ events per second, and reset occurring at the $3.74^{th}$ event (on average), this amounts to a ~0.75 (2.81/3.74) chance of a reset per second.
-If we simply take the parameters of the gamma distribution fit to reset times in the re-tracing experiment (without optimization), we can also predict PSEs (Fig 6B, dark diamonds).
+[-@Liu2021] present a Poisson process model with a constant chance of reset of ~0.7 / _s_ (Fig 6B, light diamonds).
+In comparison, with the gamma distribution fit to our data here (rate=$2.81$, shape=$3.74$), we found a ~0.75 / _s_ chance of a reset.
+Moreover, if we simply take the parameters of the gamma distribution fit to reset times in the re-tracing experiment (without further optimization), we can also generate the PSEs that should be found in Liu et al.'s experiment (Fig 6B, dark diamonds).
 We see that the two findings, based on very different approaches, are mostly in agreement with each other.
@@
 **A:** By ensuring the perceived path of a double-drift stimulus is straight down (teal arrows), any gap-induced reset, should be equal to the accumulated illusion if there are no spontaneous resets (black arrow), but smaller if there are (light blue arrow).
-Gaps were presented after 1, 2 or 3 seconds, and 2 or 4 dva real paths, which allows psychophysics to quantify the amount of illusion left after several periods of abserving a double-dirft stimulus.
+Gaps were presented after 1, 2 or 3 seconds, and 2 or 4 dva real paths, which allows psychophysics to quantify the amount of illusion left after several periods of observing a double-drift stimulus.
 **B:** Points of subject equivalence show that the size of the illusion decreases with time even as the real path, and maximum illusion is kept constant.
@@
 Kwon, Tadin, and Knill [-@Kwon2015] proposed a Bayesian object tracking model to explain the double-drift illusion and other phenomena.
-In their model, when the position signals are weak, the motion signals bias perceived position away from the true location.
-Their model predicts that this deviation saturates after about 200 _ms_ and the perceived path then continues parallel to the physical path, as in the “hit the wall” resets we describe above [@Kwon2015, Fig.
+In their model, when the position signals are weak they have less weight and the motion signals make a prediction that moves the perceived position away from the true location.
+Their model predicts that after about 200 _ms_ the motion prediction saturates and the perceived path then continues parallel to the physical path, as in the “hit the wall” resets we describe above [@Kwon2015, Fig.
 3].
-In contrast to their model, the reset in our re-tracing task show quite variable timing, and do not show a clear spatial offset (Fig.
-5A) but seem to fall on a fairly wide gamma distribution on reset time (Fig.
-5B), which is at odds with a set limit (either spatial or temporal) where saturation or resets occur.
-This seems to argue against a perceptual mechanism underlying resets, such as on going Bayesian inference until the discrepancy between retinal and perceived position signals is too large.
-In any case, the best fitting temporal limit of these reset points was not 200 _ms_ but 1.3 _s_ and 99% of detected reset times are above 200 _ms_ (see Fig.
+In contrast to their model, the resets in our re-tracing task show quite variable timing (Fig.
+5A) and are better described by a fairly broad gamma distribution (Fig.
+5B).
+This argues against a  mechanism that is based on comparing the current location to the physical location and saturating or triggering a reset when the discrepancy is too large.
+In any case, the best fitting temporal limit of the reset points in our data was not 200 _ms_ but 1.3 _s_ and 99% of detected reset times are above 200 _ms_ (see Fig.
 5A).
@@
 The “return” resets suggest that both the physical and perceived locations are available to the visual system and that when spontaneous resets occur, the perceived location returns toward the physical location.
-While our results speak against a particular, stimulus or perception induced mechanism leading to resets, the illusion itself may still have limits, and the way resets are perceived may relate to stimulus properties as well as to illusory perception.
-Regardless, future models that aim to predict perception in double-drift stimuli need to also predict resets.
+Our results suggest that the resets, whether saturation or returns, happen randomly over time, rather than at a particular deviation in spatial offset or after a fixed duration.
+Future models of position perception for the double-drift stimulus need to include a realistic model of the locations and type of resets.
 
-Recently, it has been suggested that resets can be triggered by a distraction of attention [@Nakayama2020].
-The participants in this experiment also retraced the perceived path and reported “return” resets.
-Nakayama and Holcombe [-@Nakayama2020] suggest that planned eye movements can trigger a “return” reset back to the physical path.
-In our experiment, we did not control for eye-movements, so we can't directly investigate whether eye movements accompanied the “return” resets that participants reported.
-However, the random distribution of resets throughout time are in line with a mechanism based on distracted attention, in some cases leading to eye-movement planning.
-
-We find that manual tracking of the double-drift stimulus during its motion did show the expected illusion, although not always as strong.
-Few if any resets were detected during online tracking, perhaps because of the additional attentional requirements or the short duration (2 s), although interactions with the simultaneous hand movements could explain this too.
-Participants did report spontaneous resets in delayed, single-pass re-tracing of Experiment 2.
+In summary, we find that manual tracking of the double-drift stimulus during its motion did show the expected illusion at the expected strength (experiment 1).
+Few if any resets were detected during online tracking, perhaps because of the additional attentional requirements or the short duration (2 _s_), although interactions with the simultaneous hand movements could explain this too.
+In contrast, participants did report relatively frequent, spontaneous resets in the delayed, single-pass re-tracing of Experiment 2.
 The resets were best explained by a gamma distribution on reset time.
-This finding seems to indicate that spontaneous resets occur independently from the stimulus or how it is perceived, and spontaneous reset may even be random.
-If there is a mechanism causing resets we have not uncovered it here, and if resets are independent from perception, the illusory percept itself may still be limited, and give rise to how resets are perceived.
+This finding suggests that spontaneous resets occur independently of the stimulus and how or where it is perceived.
+Instead, they may occur randomly over time, perhaps due to unrelated attentional distractions.
 

### 2021-12-11 Manuscript.Rmd
updated fig 6, according to Patrick's feedback

@@
 output:
+  pdf_document: default
   html_document:
@@
     reference_docx: template.docx
-  pdf_document: default
 header-includes:
@@
 (2021) keeps the illusion strength constant and fits a chance of reset per time unit as ~0.007 / 10 ms to match the gap-evoked reset (diamond symbols)."}
-plotJumpPSEs(target = 'inline', individualDots = FALSE, modelDots = TRUE, gammaDots=TRUE)
+plotJumpPSEs(target = 'inline', individualDots=FALSE, modelDots=FALSE, gammaDots=FALSE, modelLine=TRUE, gammaLine=TRUE)
 ```

### 2021-12-11 R/exp3_gaborjumps.R

@@ -202,7 +202,7 @@ ANOVAonPSEs <- function() {
 
 # figures -----
 
-plotJumpPSEs <- function(target='inline', individualDots=FALSE, modelDots=FALSE, gammaDots=FALSE) {
+plotJumpPSEs <- function(target='inline', individualDots=FALSE, modelDots=FALSE, gammaDots=FALSE, modelLine=FALSE, gammaLine=FALSE) {
   
   colors <- getColors()
   
@@ -279,7 +279,9 @@ plotJumpPSEs <- function(target='inline', individualDots=FALSE, modelDots=FALSE,
       mu <- groupPSEs$PSE[idx]
       sigma <- groupPSEs$std[idx]
       
-      color <- colors[[ c('orange', 'yorkred', 'purple')[stimdur] ]]$s
+      #color <- colors[[ c('orange', 'yorkred', 'purple')[stimdur] ]]$s
+      #color <- colors[[ c('orange', 'orange', 'orange')[stimdur] ]]$s
+      color <- colors$orange$s
       lc <- c(lc, color)
       
       pPSEs <- participantPSEs$PSE[which(participantPSEs$path_length == pathlength & participantPSEs$stimdur == stimdur)]
@@ -292,8 +294,8 @@ plotJumpPSEs <- function(target='inline', individualDots=FALSE, modelDots=FALSE,
       }
       
       # modelPSEs
+      idx <- which(modelPSEs$path_length == pathlength & modelPSEs$duration == stimdur)
       if (modelDots) {
-        idx <- which(modelPSEs$path_length == pathlength & modelPSEs$duration == stimdur)
         fitPSE <- modelPSEs$fitPSE[idx]
         points(stimdur+pos[3]+xoff,fitPSE,col=mixCol('#FFFFFF', color, balance=c(2,1)),pch=18,cex=1.5)
       }
@@ -310,11 +312,22 @@ plotJumpPSEs <- function(target='inline', individualDots=FALSE, modelDots=FALSE,
       
     }
     
+    if (modelLine) {
+      idx <- which(modelPSEs$path_length == pathlength)
+      lines(c(1,2,3)+pos[1]+xoff,modelPSEs$fitPSE[idx],col=colors$yorkred$s, lty=3)
+    }
+    if (gammaLine) {
+      idx <- which(modelPSEs$path_length == pathlength)
+      lines(c(1,2,3)+pos[1]+xoff,modelPSEs$gamma[idx],col=colors$purple$s, lty=3)
+    }
+    
+    
+    
[…]

### 2021-12-11 Manuscript.Rmd
made fig 6 use same language as some paragraphs in the discussion

--- 2021-12-11 Manuscript.Rmd
@@
 output:
+  pdf_document: default
+  word_document:
+    reference_docx: template.docx
   html_document:
     df_print: paged
-  word_document:
-    reference_docx: template.docx
-  pdf_document: default
 header-includes:
@@
 The simplest model presented by Liu et al.
-(2021) keeps the illusion strength constant and fits a chance of reset per time unit as ~0.007 / 10 ms to match the gap-evoked reset (diamond symbols)."}
-plotJumpPSEs(target = 'inline', individualDots = FALSE, modelDots = TRUE, gammaDots=TRUE)
+[-@Liu2021] keeps the illusion strength constant and fits a chance of reset per time unit as ~0.7 / _s_ to match the gap-evoked reset (diamond symbols)."}
+plotJumpPSEs(target = 'inline', individualDots=FALSE, modelDots=FALSE, gammaDots=FALSE, modelLine=TRUE, gammaLine=TRUE)
 ```

### 2021-12-12 R/exp3_gaborjumps.R
changing fig 6

@@ -202,7 +202,7 @@ ANOVAonPSEs <- function() {
 
 # figures -----
 
-plotJumpPSEs <- function(target='inline', individualDots=FALSE, modelDots=FALSE, gammaDots=FALSE, modelLine=FALSE, gammaLine=FALSE) {
+plotJumpPSEs <- function(target='inline', individualDots=FALSE, modelDots=FALSE, gammaDots=FALSE, modelLine=FALSE, gammaLine=FALSE, propExpect=FALSE) {
   
   colors <- getColors()
   
@@ -249,27 +249,38 @@ plotJumpPSEs <- function(target='inline', individualDots=FALSE, modelDots=FALSE,
   par(mar = c(2.8,2.8,2.2,0.1),
       mgp = c(2, 0.55, 0) )
   
+  ylim <- c(0,3.5)
+  if (propExpect) {ylim <- c(0,1.1)}
   plot(-1000,-1000,main='',xlab='',ylab='',
-       xlim=c(0,8),ylim=c(0,3.5),
+       xlim=c(0,8),ylim=ylim,
        bty='n',ax=F)
   
   title(main='B: results', font.main=1, cex.main=1.5, adj=0, line=0.5)
   title(xlab='stimulus duration [s]', line=1.6, cex.lab=0.7)
-  title(ylab='gap-evoked reset [dva]', line=1.6, cex.lab=0.7)
+  if (propExpect) {
+    title(ylab='gap-evoked reset [% expected]', line=1.6, cex.lab=0.7)
+  } else {
+    title(ylab='gap-evoked reset [dva]', line=1.6, cex.lab=0.7)
+  }
+  
+  if (propExpect) {
+    lines(c(0.5,7.5)+xoff,c(1,1),col='#000000',lty=2)
+  }
   
   for (pathlength in c(2,4)) {
     
     xoff <- (pathlength-2)*2
     
-    lines(c(0.5,3.5)+xoff,rep(pathlength/sqrt(2),2),col='#000000',lty=2)
-    
-    text( x=3.5+xoff, 
-          y=(pathlength/sqrt(2))+0.14, 
-          sprintf('%0.1f dva', pathlength/sqrt(2)), 
-          adj=c(1,1),
-          cex=0.7, 
-          col='#000000'  )
-    
+    if (!propExpect) {
+      lines(c(0.5,3.5)+xoff,rep(pathlength/sqrt(2),2),col='#000000',lty=2)
+      text( x=3.5+xoff, 
[…]

### 2021-12-12 Manuscript.Rmd
suggestions from Patrick

--- 2021-12-11 Manuscript.Rmd
@@
 Illusion strength is also lower than in Experiment 1 with online tracking and it has a fair amount of between-subject variation in this task (blue shaded area is 95% confidence interval on K across participants, and organge lines are 95% confidence intervals across participants for illusion strength in each condition).
-While the illusion is less strong, it is nevertheless there and depends on the relationship between internal and external speed as in previous studies.
+While the illusion, as captured by retracing, is less strong, it is nevertheless there and depends on the relationship between internal and external speed as in previous studies.
 
@@
 
-These first two models assumed that resets were triggered at fixed locations in space or time, but the next tro propose that the resets are random processes that occur with some distribution across space or time.
-We used gamma distributions to capture these processes (with 2 parameters: rate and shape) on the X coordinates and time points of resets as well (see Fig 5A, marginal distributions; Fig 5B,C).
+These first two models assumed that resets were triggered at fixed locations in space or time, but the next two propose that the resets are random processes that occur with some distribution across space or time.
+We used gamma distributions to capture these processes (also with 2 parameters: rate; the chance of an event per second, and shape; how many events are needed, in this case, until a reset is triggered) on the X coordinates and time points of resets as well (see Fig 5A, marginal distributions; Fig 5B,C).
 To compare across the four models, we then calculated the likelihoods of their predictions (based on the underlying distributions' probability density functions) using all resets together and also within individual participants.
@@
 2) The demands of the online tracking increased the attention to the double-drift stimulus and with less chance of attentional distraction, the likelihood of resets was reduced [@Nakayama2020].
-However, this task, unlike previous tests of the double-drift stimulus, may involve an interaction between action and perception.
+However, this task, unlike previous tests of the double-drift stimulus, may also involve an interaction between action and perception [e.g.
+@Beets2010].
 For example, the predicted sensory consequences of the hand movement may influence perception ("I will move my hand in this direction, so I expect to see motion in that direction as well").
@@
 Neither of these provided a viable explanation of the pattern of resets.
-Resets points are broadly distributed, suggesting that resets are not linked to illusion strength or stimulus onset, but may be caused by a more "random" process.
+Resets points are broadly distributed, suggesting that resets are not linked to the degree of illusory offset or time from stimulus onset, but instead may be caused by a more "random" process.
 
@@
 The rate of resets that we found here, ~0.75 / _s_, are too infrequent for microsaccades (~3 / _s_) so the distractions may be more significant than these small eye movements.
-On the other hand, random occurrance of distractions, a fraction of which triggers a reset, is in line with the kind of processes underlying gamma distributions [e.g.
-@McGill1965; @Wolfe2002].
+On the other hand, gamma distributions [e.g.
+@McGill1965; @Wolfe2002] do model a random occurence of events (rate parameter) where only some fraction of these trigger an action (shape parameter).
+This could be consistent with the more frequent microsaccades happening randomly at a faster rate where only some of there would trigger a reset.
+In other words, the gamma distribution is not only a good fit to our reset data but it may reveal the processes underlying the resets.
 
-We observe that reset times were well modeled by broad gamma distribution, suggesting that resets were best explained as occurring randomly in time with an approximate rate of 1 every 1.3 _s_.
+The fit showed that reset times were well modeled by a broad gamma distribution, with resets occurring randomly in time (set by the rate and shape parameter) with an average rate of 1 reset every 1.3 _s_ over the 3 to 4 _s_ duration of the trials.
 However, we have to qualify our conclusions given the noisy nature of hand-drawn trajectories as a measure of the perceived trajectory and the reset points.
-Therefore, we compare our results with those from s similar study that analyzed trajectory endpoints [pre-print: @Liu2021].
-In this study, the internal speed of the Gabor texture was adjusted across conditions and participants so that the illusion always had a local direction of 45° compared to its actual vertical direction.
+To gauge the generality of our results, we compared them to those from a similar study that analyzed trajectory endpoints [pre-print: @Liu2021] to determine the presence and size of spontaneous resets.
+In their study, the internal speed of the Gabor texture was adjusted across conditions and participants so that the illusion always had a local direction of 45° compared to its actual vertical direction.
 This then generated a highly predictable offset of the end of the trajectory that would be $\sqrt{2}$ of the path length if there were no spontaneous resets.
-Anything less than that revealed the presence and size of spontaneous resets.
+Any offset less than that revealed the presence and size of spontaneous resets.
 Liu et al.
-tested 3 different durations (1, 2 and 3 s) and two path lengths (2 and 4 dva) for the first segment of the Gabor’s motion (see Fig.
+[-@Liu2021] tested different durations and path lengths for the first segment of the Gabor’s motion (see Fig.
 6A).
-They then presented a 350 _ms_ gap during which the double-drift stimulus is not shown, and this is known to evoke a full reset to the physical location [@Lisi2015].
-After this gap-evoked reset the illusion resumes accumulating from the veridical position where the gabor appears after the gap [@Lisi2015].
-By changing the location where the double-drift reappeared and asking participants if it shifted left or right, the endpoint of the first segment was determined.
-The difference between that point and the expected value for no spontaneous resets indicated the presence and size of any that had occurred.
+Then a 350 _ms_ gap in stimulus presentation is introduced, which is known to evoke a full reset to the physical location [@Lisi2015].
+The difference between the perceived points where the Gabor disappeared (illusion minus spontaneous resets) and reappeared (veridical), can be used to assess spontaneous resets.
 
 The data from 10 participants showed that for a fixed path and illusion strength, longer durations produced smaller illusory offsets at the end (Fig 6B, round dots).
-This can be explained by (among others) resets randomly occuring throughout time.
+This can be explained if resets occurred randomly over time, so that longer durations would have more resets and a smaller final offset at the endpoint.
 Liu et al.
-[-@Liu2021] present a Poisson process model with a constant chance of reset of ~0.7 / _s_ (Fig 6B, light diamonds).
-In comparison, with the gamma distribution fit to our data here (rate=$2.81$, shape=$3.74$), we found a ~0.75 / _s_ chance of a reset.
-Moreover, if we simply take the parameters of the gamma distribution fit to reset times in the re-tracing experiment (without further optimization), we can also generate the PSEs that should be found in Liu et al.'s experiment (Fig 6B, dark diamonds).
+[-@Liu2021] modeled this with a Poisson process having a constant chance of reset of ~0.7 / _s_ (Fig 6B, light diamonds).
+In comparison, the gamma distribution fit to our data here (rate=$2.81$, shape=$3.74$), gave a ~0.75 / _s_ chance of a reset.
+In addition, we simply used the parameters of the gamma distribution fit to our data to generate the PSEs that should be found in Liu et al.'s experiment (Fig 6B, dark diamonds).
 We see that the two findings, based on very different approaches, are mostly in agreement with each other.
@@
 Gaps were presented after 1, 2 or 3 seconds, and 2 or 4 dva real paths, which allows psychophysics to quantify the amount of illusion left after several periods of observing a double-drift stimulus.
-**B:** Points of subject equivalence show that the size of the illusion decreases with time even as the real path, and maximum illusion is kept constant.
-These data are in line with resets occurring randomly over time.
-The simplest model presented by Liu et al.
-(2021) keeps the illusion strength constant and fits a chance of reset per time unit as ~0.007 / 10 ms to match the gap-evoked reset (diamond symbols)."}
-plotJumpPSEs(target = 'inline', individualDots = FALSE, modelDots = TRUE, gammaDots=TRUE)
+**B:** The endpoints of the first segments indicate the size of the illusion and the difference between these values and the expected value (1.4 or 2.8 _dva_) indicates the size of any intervening spontaneous resets.
+This shows that the size of the illusion decreases over time when path length and local illusion angle are kept constant.
+These data are in line with resets occurring randomly over time at ~0.7 / _s_."}
+plotJumpPSEs(target = 'inline', individualDots=FALSE, modelDots=FALSE, gammaDots=FALSE, modelLine=TRUE, gammaLine=TRUE)
 ```
@@
 Kwon, Tadin, and Knill [-@Kwon2015] proposed a Bayesian object tracking model to explain the double-drift illusion and other phenomena.
-In their model, when the position signals are weak they have less weight and the motion signals make a prediction that moves the perceived position away from the true location.
+In their model, when the position signals are weak they have less weight and the predictions based on motion signals get more weight, moving the perceived position away from the true location.
 Their model predicts that after about 200 _ms_ the motion prediction saturates and the perceived path then continues parallel to the physical path, as in the “hit the wall” resets we describe above [@Kwon2015, Fig.
@@
 The “return” resets suggest that both the physical and perceived locations are available to the visual system and that when spontaneous resets occur, the perceived location returns toward the physical location.
-Our results suggest that the resets, whether saturation or returns, happen randomly over time, rather than at a particular deviation in spatial offset or after a fixed duration.
+Our results suggest that the resets, whether saturation or return resets, happen randomly over time, rather than at a particular deviation in spatial offset or after a fixed duration.
 Future models of position perception for the double-drift stimulus need to include a realistic model of the locations and type of resets.
@@
 In summary, we find that manual tracking of the double-drift stimulus during its motion did show the expected illusion at the expected strength (experiment 1).
-Few if any resets were detected during online tracking, perhaps because of the additional attentional requirements or the short duration (2 _s_), although interactions with the simultaneous hand movements could explain this too.
-In contrast, participants did report relatively frequent, spontaneous resets in the delayed, single-pass re-tracing of Experiment 2.
+Few if any resets were detected during online tracking, however, perhaps because of the additional attentional requirements or the short duration (2 _s_), although interactions with the simultaneous hand movements may contribute as well.
+In contrast, participants did report relatively frequent, spontaneous resets in the delayed, single-pass re-tracing of experiment 2.
 The resets were best explained by a gamma distribution on reset time.

### 2021-12-15 Manuscript.Rmd
edits from Patrick

@@
 **E:** Distribution of all 827 detected reset points.
-**F:** Using the angle from the origin of the trace to a point along the trajectory at half the distance between origin and reset point as a measure of illusion strength, we can compare this data to other data.
+**F:** Using the angle from the origin of the trace to a point along the trajectory at half the distance between origin and reset point as a measure of illusion strength, we can compare these data to other data.
 For this graph, we exclude one participant who produced very few resets.
@@
 These first two models assumed that resets were triggered at fixed locations in space or time, but the next two propose that the resets are random processes that occur with some distribution across space or time.
-We used gamma distributions to capture these processes (also with 2 parameters: rate; the chance of an event per second, and shape; how many events are needed, in this case, until a reset is triggered) on the X coordinates and time points of resets as well (see Fig 5A, marginal distributions; Fig 5B,C).
+We used gamma distributions to capture these processes (these have 2 parameters: rate, the chance of an event per second; and shape, how many events are needed, in this case, until a reset is triggered) on the X coordinates and time points of resets as well (see Fig 5A, marginal distributions; Fig 5B,C).
 To compare across the four models, we then calculated the likelihoods of their predictions (based on the underlying distributions' probability density functions) using all resets together and also within individual participants.
@@
 Neither of these provided a viable explanation of the pattern of resets.
-Resets points are broadly distributed, suggesting that resets are not linked to the degree of illusory offset or time from stimulus onset, but instead may be caused by a more "random" process.
+Resets points are broadly distributed, suggesting that resets are not linked to the degree of illusory offset or to the time from stimulus onset, but instead may be caused by a more "random" process.
 
@@
 However, the random distribution of resets throughout time are in line with a mechanism based on distracted attention, which may involve eye-movement planning and execution.
-The rate of resets that we found here, ~0.75 / _s_, are too infrequent for microsaccades (~3 / _s_) so the distractions may be more significant than these small eye movements.
+The rate of resets that we found here, ~0.75 / _s_, is too infrequent for microsaccades (~3 / _s_) so the distractions may be more significant than these small eye movements.
 On the other hand, gamma distributions [e.g.
 @McGill1965; @Wolfe2002] do model a random occurence of events (rate parameter) where only some fraction of these trigger an action (shape parameter).
-This could be consistent with the more frequent microsaccades happening randomly at a faster rate where only some of there would trigger a reset.
+This could be consistent with the more frequent microsaccades happening randomly at a faster rate where only some of these would trigger a reset.
 In other words, the gamma distribution is not only a good fit to our reset data but it may reveal the processes underlying the resets.
@@
 To gauge the generality of our results, we compared them to those from a similar study that analyzed trajectory endpoints [pre-print: @Liu2021] to determine the presence and size of spontaneous resets.
-In their study, the internal speed of the Gabor texture was adjusted across conditions and participants so that the illusion always had a local direction of 45° compared to its actual vertical direction.
-This then generated a highly predictable offset of the end of the trajectory that would be $\sqrt{2}$ of the path length if there were no spontaneous resets.
-Any offset less than that revealed the presence and size of spontaneous resets.
+In their study, the internal speed of the Gabor texture was adjusted across conditions and participants so that the illusion always appeared locally downward (vertical, ignoring any resets) compared to its actual 45° direction.
+This generated a highly predictable offset of path length / $\sqrt{2}$ at the end of the trajectory if there were no spontaneous resets.
+Any measured offset less than the expected value (path length / $\sqrt{2}$ ) then revealed the presence and size of spontaneous resets.
 Liu et al.
-[-@Liu2021] tested different durations and path lengths for the first segment of the Gabor’s motion (see Fig.
+[-@Liu2021] used this technique to measure spontaneous resets for  3 different durations and 2 path lengths (see Fig.
 6A).
-Then a 350 _ms_ gap in stimulus presentation is introduced, which is known to evoke a full reset to the physical location [@Lisi2015].
-The difference between the perceived points where the Gabor disappeared (illusion minus spontaneous resets) and reappeared (veridical), can be used to assess spontaneous resets.
 
-The data from 10 participants showed that for a fixed path and illusion strength, longer durations produced smaller illusory offsets at the end (Fig 6B, round dots).
-This can be explained if resets occurred randomly over time, so that longer durations would have more resets and a smaller final offset at the endpoint.
+The data from 10 participants showed that for a fixed path and illusion strength, longer durations produced smaller illusory offsets at the end (Fig 6B, light blue dots).
+This can be explained if spontaneous resets occurred randomly over time, so that longer durations would have more resets and a smaller final offset at the endpoint.
 Liu et al.
-[-@Liu2021] modeled this with a Poisson process having a constant chance of reset of ~0.7 / _s_ (Fig 6B, light diamonds).
-In comparison, the gamma distribution fit to our data here (rate=$2.81$, shape=$3.74$), gave a ~0.75 / _s_ chance of a reset.
-In addition, we simply used the parameters of the gamma distribution fit to our data to generate the PSEs that should be found in Liu et al.'s experiment (Fig 6B, dark diamonds).
+[-@Liu2021] modeled this with a Poisson process having a constant chance of reset of ~0.7 / _s_ (Fig 6B, orange line).
+In comparison, in our experiment, the gamma distribution fit to the data of explicit resets (rate=$2.81$, shape=$3.74$), gave a ~0.75 / _s_ chance of a reset, a remarkaly similar estimate.
+That's why, we also used the parameters of our gamma distribution fit from data to generate the PSEs that should be found in Liu et al.'s experiment with no additional fitting (Fig 6B, dashed purple line).
 We see that the two findings, based on very different approaches, are mostly in agreement with each other.
 
-```{r fig6, echo=FALSE, message=FALSE, fig.width=6, fig.height=4, fig.cap="\\label{fig:fig6}**Figure 6:** Spontaneous resets without hand trajectories.
-Experiment and Results based on data from Liu et al.
+```{r fig6, echo=FALSE, message=FALSE, fig.width=6, fig.height=4, fig.cap="\\label{fig:fig6}**Figure 6:** Spontaneous resets without hand trajectories from Liu et al.
 [-@Liu2021], re-analyzed here.
-**A:** By ensuring the perceived path of a double-drift stimulus is straight down (teal arrows), any gap-induced reset, should be equal to the accumulated illusion if there are no spontaneous resets (black arrow), but smaller if there are (light blue arrow).
-Gaps were presented after 1, 2 or 3 seconds, and 2 or 4 dva real paths, which allows psychophysics to quantify the amount of illusion left after several periods of observing a double-drift stimulus.
-**B:** The endpoints of the first segments indicate the size of the illusion and the difference between these values and the expected value (1.4 or 2.8 _dva_) indicates the size of any intervening spontaneous resets.
-This shows that the size of the illusion decreases over time when path length and local illusion angle are kept constant.
-These data are in line with resets occurring randomly over time at ~0.7 / _s_."}
-plotJumpPSEs(target = 'inline', individualDots=FALSE, modelDots=FALSE, gammaDots=FALSE, modelLine=TRUE, gammaLine=TRUE)
+**A:** The Gabor’s internal motion was adjusted for each condition and participant so that the perceived path was always locally straight down (yellow arrows).
+The 350 ms temporal gap then ensured a full reset [@Lisi2015] to the physical location when the Gabor continued on its path following the gap (blue arrow).
+Gaps were presented after 1, 2 or 3 seconds, and 2 or 4 _dva_ real paths and the offset measured between the pre- and post-gap paths (the vertical green lines) indicated the amount of illusion left at the end of  the pre-gap path (measured PSE).
+**B:** Percent reduction of the illusion (difference between measured offset and the offset expected with no intervening spontaneous resets  as a percentage of expected offset) as a function of path length and duration.
+The size of the illusion decreased over time for a fixed path length and local illusion angle.
+These data are consistent with resets occurring randomly over time at ~0.7 / _s_ modeled as a Poisson process."}
+plotJumpPSEs(target='inline', individualDots=F, modelDots=F, gammaDots=F, modelLine=T, gammaLine=T, propExpect=T)
 ```

### 2021-12-15 R/common.R
indentation improvements

@@ -458,20 +458,23 @@ getColors <- function(transparency=81.5) {
   
   colors <- list()
   
-  colors[['blue']]      <- list('s'='#005de4ff')
-  colors$blue[['t']] <- t_col(colors$blue$s,percent=transparency)
+  colors[['blue']]        <- list('s'='#005de4ff')
+  colors$blue[['t']]      <- t_col(colors$blue$s,percent=transparency)
   
-  colors[['lightblue']] <- list('s'='#0fd2e2ff')
+  colors[['lightblue']]   <- list('s'='#0fd2e2ff')
   colors$lightblue[['t']] <- t_col(colors$lightblue$s,percent=transparency)
   
-  colors[['yorkred']]   <- list('s'='#e51636ff')
-  colors$yorkred[['t']] <- t_col(colors$yorkred$s,percent=transparency)
+  colors[['yorkred']]     <- list('s'='#e51636ff')
+  colors$yorkred[['t']]   <- t_col(colors$yorkred$s,percent=transparency)
   
-  colors[['orange']]    <- list('s'='#ff8200ff')
-  colors$orange[['t']] <- t_col(colors$orange$s,percent=transparency)
+  colors[['orange']]      <- list('s'='#ff8200ff')
+  colors$orange[['t']]    <- t_col(colors$orange$s,percent=transparency)
   
-  colors[['purple']]    <- list('s'='#b400e4ff')
-  colors$purple[['t']] <- t_col(colors$purple$s,percent=transparency)
+  colors[['purple']]      <- list('s'='#b400e4ff')
+  colors$purple[['t']]    <- t_col(colors$purple$s,percent=transparency)
+  
+  colors[['darkgreen']]   <- list('s'='#408040ff')
+  colors$darkgreen[['t']] <- t_col(colors$purple$s,percent=transparency)
   
   return(colors)
   

### 2021-12-15 R/exp3_gaborjumps.R
clearer figure for Sirui's data & model

@@ -182,6 +182,8 @@ ANOVAonPSEs <- function() {
     
     df <- df[which(df$participant != 'group'),]
     
+    df$normPSE <- df$PSE / (df$path_length / sqrt(2))
+    
     df$participant <- as.factor(df$participant)
     df$path_length <- as.factor(df$path_length)
     df$stim_dur <- as.factor(df$stimdur)
@@ -250,7 +252,7 @@ plotJumpPSEs <- function(target='inline', individualDots=FALSE, modelDots=FALSE,
       mgp = c(2, 0.55, 0) )
   
   ylim <- c(0,3.5)
-  if (propExpect) {ylim <- c(0,1.1)}
+  if (propExpect) {ylim <- c(-.1,1.3)}
   plot(-1000,-1000,main='',xlab='',ylab='',
        xlim=c(0,8),ylim=ylim,
        bty='n',ax=F)
@@ -258,21 +260,24 @@ plotJumpPSEs <- function(target='inline', individualDots=FALSE, modelDots=FALSE,
   title(main='B: results', font.main=1, cex.main=1.5, adj=0, line=0.5)
   title(xlab='stimulus duration [s]', line=1.6, cex.lab=0.7)
   if (propExpect) {
-    title(ylab='gap-evoked reset [% expected]', line=1.6, cex.lab=0.7)
+    title(ylab='% reduction of expected illusion', line=1.6, cex.lab=0.7, adj=0.5/1.3)
   } else {
     title(ylab='gap-evoked reset [dva]', line=1.6, cex.lab=0.7)
   }
   
   if (propExpect) {
-    lines(c(0.5,7.5)+xoff,c(1,1),col='#000000',lty=2)
+    lines(c(0.25,7.75),c(1,1),col='#000000',lty=1)
+    #text(4,1.03,'expected illusion: no spontaneous resets', cex=0.7)
   }
   
   for (pathlength in c(2,4)) {
     
     xoff <- (pathlength-2)*2
     
+    expected <- pathlength/sqrt(2)
+    
     if (!propExpect) {
-      lines(c(0.5,3.5)+xoff,rep(pathlength/sqrt(2),2),col='#000000',lty=2)
+      lines(c(0.5,3.5)+xoff,rep(expected,2),col='#000000',lty=2)
       text( x=3.5+xoff, 
             y=(pathlength/sqrt(2))+0.14, 
             sprintf('%0.1f dva', pathlength/sqrt(2)), 
[…]

### 2021-12-16 Manuscript.Rmd
added comparison to illusion strengths used in Heller et al, and mentioned range of K across participants in exp 2, in the caption to fig 4

@@
 output:
+  word_document:
+    reference_docx: template.docx
+  pdf_document: default
   html_document:
     df_print: paged
-  pdf_document: default
-  word_document:
-    reference_docx: template.docx
 header-includes:
@@
 While they hypothesize that illusion strength decreases with faster external motion (perhaps because of resets or limits on perception), here we will use their overall estimate of illusion strength, given by $K=0.74$.
+While in _dva/s_ our stimuli have relatively high external and relatively low internal motion, compared to Heller et al.
+[-@Heller2021], they are within the range of tested stimuli as is the predicted illusion strength.
 
@@
 For this graph, we exclude one participant who produced very few resets.
-The illusion is weaker than in Experiment 1 ($K=0.60$, blue line) and its strength varies between participants, but it is nevertheless clearly there."}
+The illusion is weaker than in Experiment 1 ($K=0.60$, blue line) and its strength varies between participants (from $K=0.40$ to $K=0.81$), but it is nevertheless clearly there."}
 #plotExampleData()

### 2021-12-16 R/common.R
added function to convert figures to pdf format

@@ -810,4 +810,35 @@ density2D <- function(x, y, bw=1, weights=NULL, n=100, from=NULL, to=NULL, cut=3
   
 }
 
+library('rsvg')
+library(magick)
 
+replotMethodsFigures <- function(figno=1, target='inline') {
+  
+  if (figno == 1) {
+    width <- 6
+    height <- 2.4
+    sourceno <- 1
+    outputfile <- 'doc/Fig1_methods1'
+  } 
+  
+  if (figno == 3) {
+    width <- 6
+    height <- 3.4125
+    sourceno <- 2
+    outputfile <- 'doc/Fig3_methods2'
+  }
+  
+  if (target == 'pdf') {
+    pdf(file=sprintf('%s.pdf',outputfile), width=width, height=height)
+  }
+  
+  Fig <- image_read_pdf(sprintf('doc/methods_fig_%d.pdf',sourceno))
+  par(mai=c(0,0,0,0))
+  plot(Fig)
+  
+  if (target %in% c('pdf')) {
+    dev.off()
+  }
+  
+}

### 2021-12-16 R/exp1_tracking.R
added options for some plotting functions to produce pdf files

@@ -457,6 +457,9 @@ plotBoundedTracking <- function(target='inline', discontinuous=FALSE) {
   if (target=='svg') {
     svglite(file='doc/Fig2-new.svg',width=6,height=5,fix_text_size = FALSE)
   }
+  if (target=='pdf') {
+    pdf(file='doc/Fig2_tracking.pdf', width=6,height=5)
+  }
   
   layout( matrix(c(1,1,2,3,4,4), nrow = 3, ncol = 2, byrow = FALSE), 
           heights=c(1.1,1.0,0.9), widths=c(0.9,1.1))
@@ -692,7 +695,7 @@ plotBoundedTracking <- function(target='inline', discontinuous=FALSE) {
   axis(side=2,at=seq(0,45,15))
   
   
-  if (target %in% c('svg')) {
+  if (target %in% c('svg','pdf')) {
     dev.off()
   }
   

### 2021-12-16 R/exp3_gaborjumps.R

@@ -209,7 +209,7 @@ plotJumpPSEs <- function(target='inline', individualDots=FALSE, modelDots=FALSE,
   colors <- getColors()
   
   if (target == 'pdf') {
-    pdf( file = 'doc/fig_6_gaborjump_exp.pdf', width=6, height=4 )
+    pdf( file = 'doc/Fig6_gaborjump_exp.pdf', width=6, height=4 )
   }
   
   layout(mat = matrix(c(1,2), 

### 2021-12-16 DoubleDriftResetLimits.Rmd
deleted everything from the notebook that is no longer relevant (most of it) will have re-write this document later

@@
 We want to see if we can find reset points in tajectories people draw while or after seeing a double-drift stimulus, with the goal to see if we can figure out if those resets occur after a certain time interval (like perception in other illusions) or if it occurs when the percept reaches certain spatial difference with the true position.
-
-![Stimuli and Setup.
-*Top left*: a gabor with internal motion (phase shifts) and external motion, that is viewed peripherally, appears to follow a path deviating from the true path.
-Eye-movements immediately following disappearance of the gabor land at the true position, but pointing movements and delayed eye-movements go to the illusory position.
-*Bottom left*: at some point, the perceived position will "reset" towards the true position.
-For some people this takes the form of a "jump" back to the true position, for others it means the illusory position doesn't deviate more from the true position, and there are some in between version.
-*Top right*: if these resets are due to a spatial limit on the size of the illusory position shift, the points of the resets should have a constant X coordinate.
-*Middle right*: if they occur after some period of time, they should have a constant Y coordinate.
-*Bottom right*: we have people track or re-trace their percept and try to isolate reset points from the drawn trajectories to figure out where or when resets occur.
-](doc/stimulus_reset.png){width=100%}
-
-# Setting up the R session
-
-Before starting any analyses, we need to make sure that we have all the code and data necessary, this will be set up by the chunks below
-
-## Libraries
-
-We've minimized depencies on other packages, but there are three that help. Of course, you can skip the chunks that use them, or just install them by uncommenting the second line of this chunk:
-
-```{r}
-# install the dependencies using this command:
-# install.packages(c('svglite','ez'))
-# they seem to rely on other packages: 'data.table', 'ggplot2', 'dplyr', 'lme4', and 'car'
-
-library('svglite')
-library('ez')
-
-# - svglite generates SVG versions of the figures, but you can see the inline R figures in the notebook:
-# - ez allows one F-test to be run, but it's not central to the paper
-```
 
@@
 source('R/exp2_models.R')
+# And to make comparisons with the Liu et al. (preprint) data:
+source('R/exp3_gaborjumps.R')
 ```
@@
 
-In this experiment, 6 participants were instructed to look at a fixation point at the left or right side of the screen, while moving a tablet controlled cursor to the center of the screen.
+In this experiment, 4 participants were instructed to look at a fixation point at the left or right side of the screen, while moving a tablet controlled cursor to the center of the screen.
 Subsequently, a double-drift stimulus would appear and then move on a straight, 13.5 cm path along the Y axis of the screen (not sideways, but forward/backward, or upward/downward if you like), starting in one of two directions at random.
@@
 The internal motion could have one of 5 values -3, -1, 0, 1 or 3 cycles per second (the number of complete phases the gabor would cycle through in 1 second).
-Each of these 5 internal motion values was repeated 8 times (with random initial direction and fixation side) for a total of 40 trials.
+Each of these 5 internal motion values was repeated 8 times (with random initial direction and fixation side) for a total of 40 trials per participant.
 
@@
 ```{r}
-# from R/tracking_bounded.R:
-getBoundedTrackingStandardizedSegments(segmentpoints=101, version=1)
-getSegmentDirections(version=1)
+# is there a function for this?
 ```
 
-This will create pre-processed data files in the directory: `data/bounded_tracking`.
+This will (re?) create pre-processed data files in the directory: `data/bounded_tracking`.
 
@@
 
-```{r fig.width=7, fig.height=6}
+```{r fig.width=6, fig.height=6}
 # from R/tracking_bounded.R:
@@
 ```
-
-The top row shows average instantaneous heading (as actual direction) over time (distance from origin), with a polar heatmap of all the data in the background.
-The first and last 0.5 second of each pass is stripped, so that the inversions at the far and close end of the workspace are excluded from the average heading data.
-In this case the middle 1 second of each pass of the gabor is used for the heading plots.
-Since heading doesn't change throughout these segments (stats?) we take the average heading across all segments for every participant and every internal speed to do statistical analyses.
 
@@
 
-First we want to see if internal motion affects illusion strength as measured by instantaneous heading direction.
-A repeated measures ANOVA (or F-test) on a model using internal speed to predict heading direction shows that internal motion affects heading direction in tracking data (F(4,12)=102.7, p<.001, ges=0.97, Greenhouse-Geisser corrected).
-This effect is not caused by one internal speed giving different directions, but by each internal speed resulting in different heading directions from it's direct neighbour (4 FDR-corrected, pairwise, one-sided t-tests: all p<.025).
-That means heading direction is a good measure of illusion strength.
+A lot of stuff is missing right now. Will fill in later.
 
-A few of the tracking paths might be reflecting a reset (four examples are plotted in red in Fig 2).
-However, there are only a few and it's not clear they represent resets, or tracking errors made when changing direction at the far and close end of the workspace.
-The heading distributions don't seem to show any systematic (time-locked) deflections from the average heading.
+## Expected resets in experiment 1, based on numbers in experiment 2
 
-# Experiment 1b: continuous tracking with 4 second segments
+If we expect resets at the same rate as they occur in experiment 2, and each pass in experiment 1 is independent. This becomes a fairly simply calculation of chance.
 
-Therefore, we did another experiment that mimicked the first, but with two changes, all designed to increase the occurrence of resets.
-First, we only used -3 and 3 cycles per second for the internal motion as this would increase the strength of the illusion and perhaps lead to a higher chance of reset.
-Second, the external motion was halved as earlier work showed that resets often occur after around 2 seconds [@Lisi2015 ??], and 2 seconds was how long one pass (from one end of the screen to the other) took in the first experiment.
-So now, the double-drift stimulus would do a single pass in 4 seconds instead of 2 seconds.
-This allows more opportunity for resets, but might also make it easier to track the gabor's position, so that errors at the direction changes in the far and close end of the workspace are less likely.
-Since the passes are longer, we use only 3 full passes in each trial, but run 32 trials in total (instead of the 8 per condition).
+```{r}
+# data:
+df <- getDataTrials()
 
-Before calculating heading, we've normalized the data, so that the direction of all segments _should_ be the same.
+# percentage of trials with resets before 1 second (0.2229938)
+sp <- sum(df$RT < 1) / 1296
+# percentage of trials with resets before 2 seconds (0.5223765)
+lp <- sum(df$RT < 2) / 1296
 
-First, here is a figure with the heading distributions across time for all participants as well as the group average:
+exp1trial <- 1 - ( (1-sp)^2 * (1-lp)^5 )
+exp1trial
 
-```{r fig.width=8, fig.height=4}
-plotBoundedTracking_V2(target='inline')
 ```
 
-Again, we see no time-dependent decrease of heading towards zero, except perhaps at the start of the data for participant 12.
-Since these are preceded by very large headings, and then settle into the normal range, this participant might have had trouble responding to the direction reversals of the gabor rather then show resets, but it's somewhat unclear.
-The group average reflects this tendency of participant 12 somewhat, but overall the average heading is very close to what was seen in the first trakcing experiment: 28.9 compared to 27.4 or 26.7 degrees.
-So depsite the slower external movement, the illusion was about equally strong, but did not lead to consistent resets.
-
-# Experiment 2: one-pass tracking and psychophysics
-
-Since it seemed that no clear spontaneous resets occurred in the continuous tracking paradigm, we used a re-tracing approach (c.f.
-@Nakayama2019).
-In this next experiment, participants first were presented with a double-drift stimulus starting close, and moving away from them, for a single pass across the workspace (while fixating to the left or right, at random).
-In two kinds of blocks participants had to either indicate the initial direction by adjusting the orientation of a line originating at the sarting point of the gabor, or they had to retrace the perceived path of the gabor.
-When retracing the path, the trajectory so far was shown on the screen, and it could be deleted to start over.
-The idea was that if the simultaneous tracking from the first tasks either interfered with the percept or made it hard to quickly report resets, this paradigm would overcome either problem, or both.
-
-In this experiment we did not control the strength of the illusion.
-In order to sample diverse illusion strengths, the external motion of 13.5 cm took either 3 or 4 seconds (~5.2 or ~3.4 cm/s), and the internal motion was set to either 2, 3 or 4 cps (corresponding to about 3.4, 5.2 or 6.9 cm/s).
-That means there are 6 different stimuli.
-Fixation was to the left or right (at random) and the internal motion could either be to the left or right as well (counterbalanced, and normalized to rightward motion in pre-processing).
-The resulting trajectories in the re-tracing condition seemed to show a reset in the majority of trials, for a few examples, see Fig XXXX.
-
-In order to determine the point of reset automatically, we took the first sign change of the X coordinate.
-Since the start of the tracing often was somewhat "shaky" we disregarded sign changes in the first 75 px or XXXX cm.
-Since the trajectories are measured in discrete pixel values, sign changes were clearer after slight smoothing of the trajectories, using splines.
+This would mean that 98.5% of trials in exp 1 should have shown at least 1 reset. Let's see if this is statistically different from the 6/160 (~4%) that we observed:
 
 ```{r}
-retrace <- read.csv('data/onePass_V4/onePass_V4_re-trace.csv', stringsAsFactors = F)
-retrace$value <- 1
-retrace$value[which(is.na(retrace$resetX))] <- 0
-resetcounts <- aggregate(value ~ participant, data=retrace, FUN=sum)
-resetcounts$value <- resetcounts$value / (6*6*4)
-print(resetcounts)
-#print(mean(retrace$value))
+binom.test(6, 160, p=exp1trial)
 ```
 
-We can also look at this, further split by condition:
+Of course it is.
+
+## Figures for Journal of Vision
+
+From the website it is not directly clear which formats are and are not accepted, but it seems they want vector-based formats like pdf or eps. Let's go with pdf for now.
 
 ```{r}
-resetcountsCOND <- aggregate(value ~ participant + abs(internalspeed) + externalspeed, data=retrace, FUN=sum)
-print(resetcountsCOND)
-hist(resetcountsCOND$value)
+replotMethodsFigures(target='pdf', figno=1)
+plotBoundedTracking(target='pdf')
+replotMethodsFigures(target='pdf', figno=3)
+plotResetPoints(target='pdf')
+plotModels(target='pdf')
+plotJumpPSEs(target='pdf', individualDots=F, modelDots=F, gammaDots=F, modelLine=T, gammaLine=T, propExpect=T)
 ```
-
-This distribution appears to be bimodal.
-Since resets here depend on direction inversion, and some people may experience more "hit-the-wall resets", it seems that the lower peak could be from people with hit-the-wall resets, while the other participants perceive some return to the true position.
-
-We validate this measure of detecting resets as follows.
-The angle of a straight line to the reset point should have approximately the same angle as the oriented lines indicating the initial direction of the gabor.
-We take the average initial directions and the average X and Y coordinates of detected reset points, for every participant in each of the 6 conditions.
-First, we test if the initial perceived direction as participants inidicated with the oriented line, responds to the different internal and external motion speeds.
-Second, we perform a regression of the angle to the reset point on the angle of the initial direction responses.
-
-```{r fig.width=8, fig.height=4}
-plotValidity(target='inline')
-```
-
-It looks like the initial percept angle responds to internal and external motion (speed) somewhat, and like there is some relationship between both measures, although not as strong as it might be.
-We verify both observations.
-
-```{r}
-testValidity()
-```
-
-First, we perform a repeated measures ANOVA on a model predicting (averaged) line angle, as a measure of the initial percept of the direction of motion from the internal and external speed of the stimulus.
-There is an effect of internal speed (F(2,16)=12.7, p<.001, ges=0.049) as well as of external speed (F(1,8)=6.89, p=.030, ges=0.031) on the line angles, and no interaction.
-This means that stimulus speeds affected perceived paths as expected and that the line angle responses captured this to some extent.
-
-Now, we want to see if the re-tracing paths capture the strength of the illusion, similarly to how the line angles capture illusion strength - at least up to the reset point.
-We do a linear regression of average reset point angle on average line angle in the same condition for every participant (then remove two outliers and redo the regression), **but this is not significant (R2(50)=.051, p=.058)**.
-(This is not the regression shown in the figure.) When looking at the raw data in the figure, it seems like the error in each variable is larger for increasing values of the other variable.
-If true, this would mean two things.
-First, given the apparently increasing variability of measurement (measurement error?) of each variable on the magnitude of the other variable, the errors of the fit for each point need to be weighted (e.g.
-by the distance to the origin, or something similar).
-Second, simple linear regression strictly speaking doesn't allow for (much) measurement error in the independent variable: orthogonal distance regression (or maybe total least squares) would be better suited for such data.
-
-Let's first solve the weighting problem, as that is straightforward: we add `1 / (distance to origin)` as the vector of weights to the linear model.
-The line angle responses in this weighted linear model _do_ predict reset point angle (R2(50)=0.15, p=.003).
-Orthogonal distance regression with weights should be possible, but there is no ready-made implementation (at least not that I could find), and it may also be completely redundant.
-That is: I will skip this for now.
-
-We're reasonably sure that the method of distilling reset points from trajectories works, so let's see what data there is.
-First we look at the data, and plot the reset coordinates in a time-spatial representation: X coordinates in centimeters, and Y coordinates converted to seconds (or scaled normalized centimeters that account for the different durations of the 3 s and 4 s pass conditions).
-We can also plot the X and Y coordinates over the strength of the illusion, expressed as the angle of the line from the start point of the gabor to the reset point relative to straight forward.
-
-```{r fig.width=7, fig.height=5}
-plotResetPoints()
-```
-
-In the figure with the reset point coordinates, we show the true path of the gabor as a gray arrow, and add potential limits by plotting the median X and Y coordinates as dashed lines.
-These are put at ~1.6 cm for the spatial limit and ~1.0 s for the temporal limit.
-The median X coordinate seems to describe the reset point distribution somewhat better.
-
-We can also fit models that use one or the other limit, using a least-squared error algorithm (using a constrained, quasi-Newton method).
-In these single-limit models, the errors are the distance of the fitted reset point from the actual reset point in normalized centimeters.
-Perhaps unsurprisingly, the limits found by these two models are qualitatively not very different from the medians of the coordinates.
-Of course, using a single, set value for the Y coordinate should result in a worse fit.
-And this is confirmed by AIC-based relative likelihoods:
-
-```{r}
-fitSeparateModelsOnData(verbosity = 1)
-```
-
-The mean, squared error and AIC is larger for a temporal limit, than for a spatial limit, and the relative likelihood of a temporal limit is a fraction of that of a spatial limit (p<.001).
-It does seem that a spatial limit would fit the data better.
-
-Nevertheless, we now try to fit both coordinates with a single model.
-In a first model, we simply combine both, i.e.
-there is both a spatial and a temporal limit, and we fit a weight parameter that balances the two components.
-The data with the fitted model looks like this:
-
-```{r, fig.width=7, fig.height=5}
-plotInterdependentModel()
-```
-
-Of course the MSE for this model is lower than either of the single-limit models: 2.267313. But the AIC might not be:
-
-```{r}
-# MSEs <- c('Lx'=2.603413, 'Ly'=4.027378, 'wLx+wLy'=2.267313)
-# 
-# N <- (9*6)-1 # number of observations (participants? or participants times conditions?)
-# # this is then used for C:
-# C <- N*(log(2*pi)+1)
-# 
-# AICs <- (2 * c(1,1,3)) + N*log(MSEs) + C
-# cat('model AICs:\n')
-# print(AICs)
-#   
-# relativeLikelihoods <- exp((min(AICs)-AICs)/2)
-# cat('\nrelative log-likelihoods:\n')
-# print(relativeLikelihoods)
-```
-
-If we use the number of participants as the number of observations, the single limit on X is still the favourable model.
-If we use the rows of the data frame, or the number of points in the plot, it is the double limit model.
-
-The main advantage of this model is that it includes a weight that decides how much of the result depends on a spatial limit and also how much depends on a time limit.
-This weight is fitted at around 80%, which gives a rather different picture then the relative likelihoods of the single limit models: there is substantial room for an effect of time.
-
-It is possible that after crossing a spatial limit, some small amount of time has to pass before perception is reset towards the true, retinal position.
-Therefore, we fit another model that expresses this idea.
-
-```{r}
-plotSequentialModel()
-```
-
-The fit seems the same as that of the weighted model, and indeed the MSE is identical.
-That is: the two models lead to exactly the same predicted reset points.
-Let's see what the AICs and relative log-likelihoods are.
-
-```{r}
-MSEs <- c('Lx'=2.603413, 'Ly'=4.027378, 'wLx+wLy'=2.267313, 'Lx->Ly'=2.267313)
-
-N <- (9*6)-1 # number of observations (participants? or participants times conditions?)
-# this is then used for C:
-C <- N*(log(2*pi)+1)
-
-AICs <- (2 * c(1,1,3,2)) + N*log(MSEs) + C
-cat('model AICs:\n')
-print(AICs)
-
-relativeLikelihoods <- exp((min(AICs)-AICs)/2)
-cat('\nrelative log-likelihoods:\n')
-print(relativeLikelihoods)
-```
-
-Since the sequential model has two parameters instead of three, it is slightly better, but not significantly so.
-While the sequential model allows estimating the actual spatial and temporal limits, the weighted model allows estimating the relative importance of the spatial and temporal limits.
-
-# Alternative Time Model
-
-This model is based on the work by Sirui Liu and colleagues who found that when probing the location of the stimulus at a few fixed temporal offsets from stimulus onset it deviated less and less after more time.
-This is consistent with a simple model with a fixed, fairly low chance of a reset occurring at every time point.
-That is: as time goes by the likelihood of reset having occurred is larger, and as a result the average percept is closer to the true position over time.
-This model fits very well and is completely at odds with the findings here.
-So which is true?
-
-One way both may be true at the same time, is if after a reset, the illusion builds up less or not at all.
-In the current experiment people report the full path of the stimulus, so we can also look at the average deviation of the perceived position from the true position over time, continuously.
-This allows to see if it is in line with the data reported by Liu, Tse & Cavanagh [].
-
-Here is a figure that splits participants by whether or not they have an average percept that is consistent with the data Lui et al found, and that is inconsistent with it.
-
-```{r}
-plotSplitDeviations()
-```
-
-Let's also have a look at the distribution of first reset points on Y-axis (~time).
-Sirui's model predicts a certain distribution of first reset points, so we can compare wat is measured here, with the model.
-
-```{r}
-plotResetYdistribution()
-```

### 2021-12-16 R/exp2_models.R
are colours better when saved to svg, then converted to pdf by inkscape?

@@ -2698,6 +2698,9 @@ plotModels <- function(target='inline') {
   if (target == 'pdf') {
     pdf(file = 'doc/Fig5_marginals_distributions.pdf', width=6, height=4/0.75, bg='white')
   }
+  if (target == 'svg') {
+    svglite(filename='doc/Fig5_marginals_distributions.svg', width=6, height=4/0.75, fix_text_size = FALSE)
+  }
   
   # layout(mat=matrix(c(2,1,4,2,1,5,7,3,6),byrow=TRUE,ncol=3,nrow=3),
   #        widths = c(0.75,2,1.5), heights = c(1,1,1))

### 2021-12-16 Manuscript.Rmd
last few changes demanded by reviewers

@@
 output:
-  word_document:
-    reference_docx: template.docx
   pdf_document: default
@@
     df_print: paged
+  word_document:
+    reference_docx: template.docx
 header-includes:
@@
 [-@Heller2021].
+A more sophisticated model allows predicting perceived speed as well [@Johnston2013], but we are interested in illusion strength only, so we use the simple vector combination model.
 In this model, the perceived direction is a simple vector combination of the external ($V_e$) and internal ($V_i$) motions and a constant $K$.
@@
 While in _dva/s_ our stimuli have relatively high external and relatively low internal motion, compared to Heller et al.
-[-@Heller2021], they are within the range of tested stimuli as is the predicted illusion strength.
+[-@Heller2021], they are within the range of tested stimuli as are the predicted illusion strengths.
 
@@
 **Top right:** if these resets can be explained by a spatial limit on the size of the illusory position shift away from the true vertical path, the locations of the resets should have a constant X coordinate; the spatial offset from the vertical path.
-**Bottom right:** if they occur instead after some period of time, resets should have a constant radius (R); the time since the gabor’s motion began."}
+**Bottom right:** if they occur instead after some period of time, resets should have a constant radius (T); the time since the gabor’s motion began."}
 library(magick)
@@
 For this experiment, 9 participants were recruited from the lab (6 female; ages 19 - 27, mean: 22.8).
-All participants reported themsleves as right handed and had either normal or corrected-to-normal vision.
+All participants reported themselves as right handed and had either normal or corrected-to-normal vision.
 Procedures were in accordance with the Declaration of Helsinki (2003) and were approved by York’s Human Participants Review Committee.
@@
 Resets points are broadly distributed, suggesting that resets are not linked to the degree of illusory offset or to the time from stimulus onset, but instead may be caused by a more "random" process.
+Overall, both the spatial offsets and time of resets seem to follow a gamma distribution about equally well (Fig 5A, marginal distributions).
+However, average spatial offsets seems to be lower with lower Y coordinates, such that the gamma distribution on spatial offsets doesn’t work as well throughout the data set (Fig 5C).
+In contrast, a gamma distribution on reset time works equally well, irrespective of angle (Fig 5B), and this is reflected in the better fit of this model.
+Hence it seems most likely that reset points are randomly distributed over time, according to a gamma distribution.
 
@@
 The 350 ms temporal gap then ensured a full reset [@Lisi2015] to the physical location when the Gabor continued on its path following the gap (blue arrow).
-Gaps were presented after 1, 2 or 3 seconds, and 2 or 4 _dva_ real paths and the offset measured between the pre- and post-gap paths (the vertical green lines) indicated the amount of illusion left at the end of  the pre-gap path (measured PSE).
-**B:** Percent reduction of the illusion (difference between measured offset and the offset expected with no intervening spontaneous resets  as a percentage of expected offset) as a function of path length and duration.
+Gaps were presented after 1, 2 or 3 seconds, and 2 or 4 _dva_ real paths and the offset measured between the pre- and post-gap paths (the vertical green lines) indicated the amount of illusion left at the end of the pre-gap path (measured illusion), which can be contrasted with the expected illusion if there were no spontaneous resets.
+**B:** Percent reduction of the illusion (difference between measured illusion and the expected illusion with no intervening spontaneous resets as a percentage of expected illusion) as a function of path length and duration.
 The size of the illusion decreased over time for a fixed path length and local illusion angle.

### 2021-12-17 Manuscript.Rmd
fixing issues with renv, dependencies

@@
 output:
-  pdf_document: default
+  word_document:
+    reference_docx: template.docx
   html_document:
     df_print: paged
-  word_document:
-    reference_docx: template.docx
+  pdf_document: default
 header-includes:

### 2021-12-17 R/common.R

@@ -1,4 +1,8 @@
 
+# some dependencies that are not automatically picked up by renv:
+
+library('pdftools')
+
 
 # useful extra operator:
 `%notin%` <- Negate(`%in%`)

### 2021-12-17 R/exp2_retracing.R

@@ -1,6 +1,6 @@
 library('svglite')
 library('ez')
-library('segmented')
+#library('segmented')
 
 source('R/common.R')
 #source('R/models.R')
@@ -516,7 +516,7 @@ preProcessOnePass_V4 <- function(participants = c(2,3,4,5,6,8,9,10,11),
     ppno <- participants[participant.idx]
     
     # load participants data file:
-    df <- read.csv(sprintf('data/onePass_V4/onepass_V4_p%02d.csv',ppno))
+    df <- read.csv(sprintf('data/onepass_V4/onepass_V4_p%02d.csv',ppno))
     
     # get the numbers of trials that are in the re-trace condition
     trials <- unique(df$trial[df$taskname == task])
@@ -587,7 +587,7 @@ preProcessOnePass_V4 <- function(participants = c(2,3,4,5,6,8,9,10,11),
                    illusionstrength)
   
   # write data frame to disk:
-  write.csv(df, file=sprintf('data/onePass_V4/onePass_V4_%s.csv', task), quote=F, row.names=F)
+  write.csv(df, file=sprintf('data/onepass_V4/onePass_V4_%s.csv', task), quote=F, row.names=F)
   
 }
 
@@ -748,37 +748,37 @@ getSplineReversalResetPoint <- function(x,y,t,internalspeed=NA,externalspeed=NA,
   
 }
 
-getMultilinearResetPoint <- function(x,y,t,internalspeed=NA,externalspeed=NA,returnTrajectory=FALSE) {
-  
-  # THIS FUNCTION WAS NOT FINISHED AS IT WAS GETTING BAROQUE
-  
-  # make data frame and switch coordinates:
-  df <- data.frame('x'=y,
-                   'y'=x,
-                   't'=t)  # don't really need t, right?
-  
-  # starting point for segmented linear regression:
-  start.lm <- lm(y ~ x, data=df)
-  
-  predict.lm( object = start.lm )
-  
-  breakpoints <- 1
-  
-  seg.lm <- segmented::segmented( start.lm, 
-                                  seg.Z   = ~x, 
-                                  npsi    = breakpoints, # 1 break point
-                                  control = seg.control( display = FALSE ) )
-  
-  predict.segmented()
-  
-  # this test indicates that (when p<.05) there MAY be an additional breakpoint
-  # davies.test(start.lm)
-  
-  # here put a while-loop:
[…]

### 2021-12-17 R/exp2_retracing.R
update directory, skip loading arrow data

@@ -65,7 +65,7 @@ oldPreProcessOnePass_V4 <- function(participants = c(2,3,4,5,6,8,9,10,11), overw
       
       ppno <- participants[participant.idx]
       
-      df <- read.csv(sprintf('data/onePass_V4/onepass_V4_p%02d.csv',ppno))
+      df <- read.csv(sprintf('data/onepass_V4/onepass_V4_p%02d.csv',ppno))
       
       tasktrials <- unique(df$trial[df$taskname == task])
       
@@ -456,7 +456,7 @@ oldPreProcessOnePass_V4 <- function(participants = c(2,3,4,5,6,8,9,10,11), overw
     # boundY <-c()
     
     df <- data.frame(participant, trial, internalspeed, internaldirection, externalspeed, fixationside, initialdirection, illusionstrength, boundX, boundY)
-    write.csv(df, file=sprintf('data/onePass_V4/onePass_V4_%s.csv', task), quote=F, row.names=F)
+    write.csv(df, file=sprintf('data/onepass_V4/onePass_V4_%s.csv', task), quote=F, row.names=F)
     
   }
   
@@ -1041,7 +1041,7 @@ summarizeTraceBoundsV4 <- function() {
   
   df$internalspeed <- abs(df$internalspeed)
   
-  df2 <- read.csv('data/onePass_V4/onePass_V4_arrow.csv', stringsAsFactors = F)
+  #df2 <- read.csv('data/onepass_V4/onePass_V4_arrow.csv', stringsAsFactors = F)
   
   # participants <- unique(df$participant)
   # 
@@ -1674,7 +1674,7 @@ getTrajectoryExamples <- function(participants = c(2,3,4,5,6,8,9,10,11), nsample
   for (participant in participants) {
     
     # load participant's data:
-    df <- read.csv(sprintf('data/onePass_V4/onepass_V4_p%02d.csv',participant))
+    df <- read.csv(sprintf('data/onepass_V4/onepass_V4_p%02d.csv',participant))
     
     # select only re-tracing trials:
     tasktrials <- unique(df$trial[df$taskname == 're-trace'])

### 2021-12-17 Manuscript.Rmd
restore renv in first big chunk

@@
 knitr::opts_chunk$set(echo = FALSE)
+
+# this may take a long time, but sets up the R environemt
+# such that (I hope) all packages are compatible with my code:
+renv::restore()
+
 # First we load all custom scripts:

### 2021-12-19 Manuscript.Rmd
found errors in caption to Fig 6

@@
 
-The data from 10 participants showed that for a fixed path and illusion strength, longer durations produced smaller illusory offsets at the end (Fig 6B, light blue dots).
+The data from 10 participants showed that for a fixed path and illusion strength, longer durations produced smaller illusory offsets at the end (Fig 6B, green dots).
 This can be explained if spontaneous resets occurred randomly over time, so that longer durations would have more resets and a smaller final offset at the endpoint.
@@
 In comparison, in our experiment, the gamma distribution fit to the data of explicit resets (rate=$2.81$, shape=$3.74$), gave a ~0.75 / _s_ chance of a reset, a remarkaly similar estimate.
-That's why, we also used the parameters of our gamma distribution fit from data to generate the PSEs that should be found in Liu et al.'s experiment with no additional fitting (Fig 6B, dashed purple line).
+That's why, we also used the parameters of our gamma distribution fit from data to generate the PSEs that should be found in Liu et al.'s experiment with no additional fitting (Fig 6B, dashed orange line).
 We see that the two findings, based on very different approaches, are mostly in agreement with each other.
@@
 **A:** The Gabor’s internal motion was adjusted for each condition and participant so that the perceived path was always locally straight down (yellow arrows).
-The 350 ms temporal gap then ensured a full reset [@Lisi2015] to the physical location when the Gabor continued on its path following the gap (blue arrow).
-Gaps were presented after 1, 2 or 3 seconds, and 2 or 4 _dva_ real paths and the offset measured between the pre- and post-gap paths (the vertical green lines) indicated the amount of illusion left at the end of the pre-gap path (measured illusion), which can be contrasted with the expected illusion if there were no spontaneous resets.
+The 350 ms temporal gap (white lines) then ensured a full reset [@Lisi2015] to the physical location when the Gabor continued on its path following the gap (blue arrow).
+Gaps were presented after 1, 2 or 3 seconds, and 2 or 4 _dva_ real paths and the offset measured between the pre- and post-gap paths (vertical green line) indicated the amount of illusion left at the end of the pre-gap path (measured illusion), which can be contrasted with the expected illusion if there were no spontaneous resets (black line).
 **B:** Percent reduction of the illusion (difference between measured illusion and the expected illusion with no intervening spontaneous resets as a percentage of expected illusion) as a function of path length and duration.
-The size of the illusion decreased over time for a fixed path length and local illusion angle.
-These data are consistent with resets occurring randomly over time at ~0.7 / _s_ modeled as a Poisson process."}
+The size of the illusion decreased over time for a fixed path length and local illusion angle (green dots).
+These data are consistent with resets occurring randomly over time at ~0.7 / _s_ modeled as a Poisson process (continuous orange line), and largely with the gamma distribution fit to the reset times found here (dashed orange line)."}
 plotJumpPSEs(target='inline', individualDots=F, modelDots=F, gammaDots=F, modelLine=T, gammaLine=T, propExpect=T)

### 2022-01-03 Manuscript.Rmd
fixed typo (occurence) and unclear line in caption to Fig 5

@@
 Reset points neither fall along a spatial limit (blue line; log likelihood: -1415.4) or a temporal limit (red lines: continuous for 4 _s_ passes, dashed for 3 _s_ passes; log likelihood: -837.0).
-Marginal distributions show that reset points time and spatial offset approximate a gamma distribution.
+Marginal distributions show that both the time and spatial offset of reset points approximate a gamma distribution.
 **B,C:** Color illustrates the gamma probability distributions, where white is zero probability, and the darkest red or blue is the maximum probability density.
@@
 On the other hand, gamma distributions [e.g.
-@McGill1965; @Wolfe2002] do model a random occurence of events (rate parameter) where only some fraction of these trigger an action (shape parameter).
+@McGill1965; @Wolfe2002] do model a random occurrence of events (rate parameter) where only some fraction of these trigger an action (shape parameter).
 This could be consistent with the more frequent microsaccades happening randomly at a faster rate where only some of these would trigger a reset.

### 2022-01-03 R/exp2_models.R
normal models need to be fit for lines in 5A, made likelihood distribution a weighted average for the two speeds in 5B

@@ -2415,7 +2415,7 @@ plotUncoupledModels <- function(df, target='inline') {
 # which is what the orthogonal and normal models compare (it is the same)
 # and then there is an offset gamma distribution for comparison
 
-fitSomeModels <- function(df, jointModel=FALSE, Xnormal=FALSE, Tnormal=FALSE, Xgamma=TRUE, Tgamma=TRUE, Ygamma=FALSE, Orthogonal=FALSE, TexGaussian=FALSE, XexGaussian=FALSE) {
+fitSomeModels <- function(df, jointModel=FALSE, Xnormal=TRUE, Tnormal=TRUE, Xgamma=TRUE, Tgamma=TRUE, Ygamma=FALSE, Orthogonal=FALSE, TexGaussian=FALSE, XexGaussian=FALSE) {
   
   outputlist <- list()
   
@@ -3073,7 +3073,17 @@ plotModels <- function(target='inline') {
   
   
   par <- modelfits$TdistGamma$par
-  likelihoods <- TGammaLikelihood(par,data)
+  
+  W3 <- mean(df$speed == 4.5)
+  W4 <- mean(df$speed == 3.375)
+  
+  data$speed <- 4.5
+  likelihoods3 <- TGammaLikelihood(par,data)
+  
+  data$speed <- 3.375
+  likelihoods4 <- TGammaLikelihood(par,data)
+  
+  likelihoods <- (W3 * likelihoods3) + (W4 * likelihoods4) 
   
   Z <- matrix(likelihoods$L,
               ncol=length(data_y),

### 2022-01-05 Manuscript.Rmd
manuscript edits

@@
 [-@Heller2021].
-A more sophisticated model allows predicting perceived speed as well [@Johnston2013], but we are interested in illusion strength only, so we use the simple vector combination model.
-In this model, the perceived direction is a simple vector combination of the external ($V_e$) and internal ($V_i$) motions and a constant $K$.
+A more sophisticated model allows predicting perceived speed as well [@Johnston2013], but this model predicts perceived direction and speed for a mixture of first order motions, and it is unclear if this works for a combination of first and second order motion (what we call internal and external speed here).
+In the vector average model we use here, the perceived direction is a simple vector combination of the external ($V_e$) and internal ($V_i$) motions and a constant $K$.
 That is, the deviation from the physical direction is given by:
@@
 3).
-The illusion strength in turn in turn depends on the stimulus properties (speeds, gabor size, carrier frequency, eccentricity; @Gurnsey2012) as well as variations within and between participants.
+The illusion strength in turn depends on the stimulus properties (speeds, gabor size, carrier frequency, eccentricity; @Gurnsey2012) as well as variations within and between participants.
 The purpose of this experiment is to capture this spontaneous resetting and determine its source.
@@
 
+The X coordinates of the detected reset points are used as the spatial offset of the reset.
+We also need the time of the reset.
+Using the Y coordinates of the drawn trajectory (divided by speed) would assume that the real and perceived vertical position of the gabor are the same, which would imply that for very strong illusions the perceived speed can be many times the real speed.
+This seems unrealistic.
+As described above, it is unclear if the overall perceived speed of a mixture of many first order motions [@Johnston2013] applies to the double-drift illusion and if so, whether or not that can be used to calculate perceived time from path length.
+Instead, we chose a simplification by using the radius (R; distance from origin to reset, divided by speed) as the reset time.
+However, while useful for the current project, it remains to be tested how perceived speed and perceived path length are related to each other and to stimulus properties in the double-drift illusion.
+
 To facilitate modeling, the horizontal and vertical (straight ahead) coordinates of the reset points are kept isometric by giving them in centimeters.
 The distance from origin can then be converted to time as 1 _cm_ = 4/13.5 _s_ in the 4 _s_ condition and 1 _cm_ = 3/13.5 _s_ in the 3 _s_ condition.
+For each model using a probability density function to estimate model likelihood, the variable that is used is divided by its median value such that each of their probability densities are on a comparable scale.
 
@@
 **A:** Model limits relative to reset point locations.
-Reset points neither fall along a spatial limit (blue line; log likelihood: -1415.4) or a temporal limit (red lines: continuous for 4 _s_ passes, dashed for 3 _s_ passes; log likelihood: -837.0).
+Reset points neither fall close to a spatial limit (blue line) or a temporal limit (red line: 4 _s_ passes, orange line: 3 _s_ passes).
 Marginal distributions show that both the time and spatial offset of reset points approximate a gamma distribution.
 **B,C:** Color illustrates the gamma probability distributions, where white is zero probability, and the darkest red or blue is the maximum probability density.
-This is a relatively scale-free unit, where the total surface of the full probability density function is 1.
-The data and likelihood distributions match fairly well.
-**B:** Gamma distribution on time (log likelihood: -766.6).
-**C:** Gamma distribution on spatial offset (log likelihood: -1319.3)."}
+This is a relatively scale-free unit, where the total surface of the full probability density function (along a line orthogonal to the predictor variable) is 1.
+**B:** Gamma distribution on time (log likelihood: -605.6).
+**C:** Gamma distribution on spatial offset (log likelihood: -632.6)."}
 invisible(capture.output(plotModels()))
 ```
+
+When applied to the full dataset, the gamma distributions fit better than the normal distributions (normal space AIC=XXX, gamma space _AIC_=XXX; _p_<.XXX; normal time AIC=XXX, gamma time _AIC_=XXX; _p_<.XXX), but within neither of these there is a difference between the spatial and temporal distributions (normal _p_=.XXX; gamma _p_=.XXX).
+Therefore, we test how well each gamma distribution generalizes across sections of the data.
+For the spatial offset, we divide the data by Y coordinate in 5 equally sized bins.
+For the temporal offset, we divide the data by angle in 5 equally sized bins.
+We fit the gamma distribution to each bin.
+The fits within each bin are shown in Fig 5B&C.
+We then get likelihood estimates for the data in the 4 other bins, and sum these for a total estimate of how well a single distribution describes the whole data set.
+We find that the model on reset time generalizes better than the model on spatial reset offset (STATS).
 

### 2022-01-05 R/exp2_models.R
adding cross-bin gamma models

@@ -547,6 +547,8 @@ relativeLikelihood <- function(crit) {
 # % explained variance -----
 # # # # # # # # # # # # # #
 
+# 2D variance is NOT as straightforward as I thought
+# DO NOT USE:
 
 D2var <- function(coords, ddof=0) {
   
@@ -825,7 +827,7 @@ getDataTrials <- function(illusionMinimum=0, illusionMaximum=90, bin=5) {
   if (is.numeric(bin)) {
     # add bin-indices
     indices <- round( seq(1, dim(df)[1]+1, length.out=bin+1) )
-    for (variable in c('X','Y','RT')) {
+    for (variable in c('angle','Y','X')) {
       bvn <- sprintf('%s_bin',variable)
       df[[bvn]] <- NA
       vo <- order(df[,variable])
@@ -2421,9 +2423,9 @@ plotUncoupledModels <- function(df, target='inline') {
   
 }
 
-# ****************************
-# fit / plot joint model -----
-# ****************************
+# ***********************************
+# fit / plot distribution model -----
+# ***********************************
 
 # this first function fits models that fit single limits/offsets/distributions
 # i.e. on one coordinate only, for now this is the X and T coordinates
@@ -2432,13 +2434,16 @@ plotUncoupledModels <- function(df, target='inline') {
 # which is what the orthogonal and normal models compare (it is the same)
 # and then there is an offset gamma distribution for comparison
 
-fitSomeModels <- function(df, jointModel=FALSE, Xnormal=TRUE, Tnormal=TRUE, Xgamma=TRUE, Tgamma=TRUE, Ygamma=FALSE, Orthogonal=FALSE, TexGaussian=FALSE, XexGaussian=FALSE) {
+fitSomeModels <- function(df, scaleFUN=median, jointModel=FALSE, Xnormal=TRUE, Tnormal=TRUE, Xgamma=TRUE, Tgamma=TRUE, Ygamma=FALSE, Orthogonal=FALSE, TexGaussian=FALSE, XexGaussian=FALSE, XgammaBin=FALSE, TgammaBin=FALSE) {
   
   outputlist <- list()
   
   # ********************************
   # ortogonal models, just because
   # ********************************
+  
+  # these are fitted / evaluated in isometric coordinates (cm,cm)
+  # so no normalization or scaling is needed
   
   if (Orthogonal) {
     
[…]

### 2022-01-07 R/exp2_models.R
adjust colors, test bin models with bins as variable

@@ -3058,13 +3058,15 @@ fitSomeModels <- function(df, scaleFUN=median, jointModel=FALSE, Xnormal=TRUE, T
 
 plotModels <- function(target='inline') {
   
-  bins <- 4
+  bins <- 3
   df <- getDataTrials(bins=bins)
   
   modelfits <- fitSomeModels(df)
   
   colors <- getColors()
   darkorange <- mixCol(colors$yorkred$s, colors$orange$s, balance=c(1,0))
+  lightred  <- mixCol(colors$yorkred$s, '#FFFFFF', balance=c(1,0.10))
+  darkred   <- mixCol(colors$yorkred$s, '#000000', balance=c(1,0.30))
   
   if (target == 'pdf') {
     pdf(file = 'doc/Fig5_marginals_distributions.pdf', width=6, height=4/0.75, bg='white')
@@ -3114,12 +3116,12 @@ plotModels <- function(target='inline') {
   for (speed in c(3,4)) {
     lines(x   = sin(seq(0,pi/2,length.out = 50)) * Lt * speed,
           y   = cos(seq(0,pi/2,length.out = 50)) * Lt * speed,
-          col = c(colors$yorkred$s, darkorange)[speed-2],
+          col = c(darkred, lightred)[speed-2],
           lw  = 2,
           lty = 1)
     polygon(x   = c(-.2,0,0),
             y   = (Lt * speed) + c(0, .2, -.2),
-            col = c(colors$yorkred$s, darkorange)[speed-2],
+            col = c(darkred, lightred)[speed-2],
             border=NA)
   }
   
@@ -3254,7 +3256,7 @@ plotModels <- function(target='inline') {
        pch=16,col=t_col('#666666', percent = 0),cex=0.2,
        bty='n',ax=F,asp=1)
   
-  title(main='B: time model',
+  title(main='B: reset time',
         font.main=1, cex.main=1.5, adj=0, line=0.25)
   title(xlab='reset X [cm]', line=2.4)
   title(ylab='reset Y [cm]', line=2.4)
@@ -3298,7 +3300,7 @@ plotModels <- function(target='inline') {
       lines(x=sin(rads)*mf,
             y=cos(rads)*mf,
             lty=1,
-            col=c(colors$yorkred$s, darkorange)[speedno],
+            col=c(darkred, lightred)[speedno],
             lw=2)
     }
   }
@@ -3318,7 +3320,7 @@ plotModels <- function(target='inline') {
[…]

### 2022-01-07 Manuscript.Rmd
replying to reviewer feedback

@@
 ---
-  title: "Measuring the double-drift illusion with hand trajectories"
+title: "Measuring the double-drift illusion and its resets with hand trajectories"
 author: 
@@
 output:
-  word_document:
-    reference_docx: template.docx
   html_document:
@@
   pdf_document: default
+  word_document:
+    reference_docx: template.docx
 header-includes:
@@
 [-@Heller2021].
-Illusion strength is also lower than in Experiment 1 with online tracking and it has a fair amount of between-subject variation in this task (blue shaded area is 95% confidence interval on K across participants, and organge lines are 95% confidence intervals across participants for illusion strength in each condition).
+Illusion strength is also lower than in Experiment 1 with online tracking and it has a fair amount of between-subject variation in this task (blue shaded area is 95% confidence interval on K across participants, and orange lines are 95% confidence intervals across participants for illusion strength in each condition).
 While the illusion, as captured by retracing, is less strong, it is nevertheless there and depends on the relationship between internal and external speed as in previous studies.
@@
 **A:** Model limits relative to reset point locations.
-Reset points neither fall close to a spatial limit (blue line) or a temporal limit (red line: 4 _s_ passes, orange line: 3 _s_ passes).
+Reset points neither fall close to a spatial limit (blue line) or a temporal limit (dark red line: 4 _s_ passes, light red line: 3 _s_ passes).
 Marginal distributions show that both the time and spatial offset of reset points approximate a gamma distribution.
@@
 **B:** Gamma distribution on time (log likelihood: -605.6).
-**C:** Gamma distribution on spatial offset (log likelihood: -632.6)."}
+Dark and light red lines: denote the median reset time for 4 _s_ and 3 _s_ passes in angular bins with an equal number of resets.
+**C:** Gamma distribution on spatial offset (log likelihood: -632.6).
+Blue lines indicate median reset offsets in Y-coordinate bins with an equal number of resets."}
 invisible(capture.output(plotModels()))
@@
 
-When applied to the full dataset, the gamma distributions fit better than the normal distributions (normal space AIC=XXX, gamma space _AIC_=XXX; _p_<.XXX; normal time AIC=XXX, gamma time _AIC_=XXX; _p_<.XXX), but within neither of these there is a difference between the spatial and temporal distributions (normal _p_=.XXX; gamma _p_=.XXX).
-Therefore, we test how well each gamma distribution generalizes across sections of the data.
-For the spatial offset, we divide the data by Y coordinate in 5 equally sized bins.
-For the temporal offset, we divide the data by angle in 5 equally sized bins.
-We fit the gamma distribution to each bin.
-The fits within each bin are shown in Fig 5B&C.
-We then get likelihood estimates for the data in the 4 other bins, and sum these for a total estimate of how well a single distribution describes the whole data set.
-We find that the model on reset time generalizes better than the model on spatial reset offset (STATS).
-
-For the space and time limit with normal distributions (means in Fig 5A, blue: spatial, red: time), the time-based distribution fits better for the group data (normal space AIC=2,834.83, normal time _AIC_=1,677.98; _p_<.001) as well as all individual participants, and the same holds within gamma distribution models of reset models (gamma space AIC=2,642.69, gamma time AIC=1,537.10; _p_<.001).
+For the space and time limit with normal distributions (means in Fig 5A, blue: spatial, red: time), the time-based distribution fits better for the group data (normal space AIC=2,834.83, normal time _AIC_=1,677.98; _p_<.001) as well as all individual participants, and the same holds within gamma distribution models of reset models (gamma space AIC=1,269.18, gamma time AIC=1,215.12; _p_<.001).
 Furthermore, the gamma-based distribution on time of resets performed better than the normal distribution of reset time in the pooled data ( _p_<.001 ), and this holds for most participants (6/9) as well.
 This means that time-based distributions explain our reset points better than models assuming a spatial offset, so that the time-based gamma distribution was the most successful of all 4.
+
+While the marginal distribution plot in Fig 5A qualitatively appear an equally good fit to the data, the better fit for the distribution on the time of resets could be explained be a better generalization throughout the data set.
+We test this by dividing the data into segments, each with an equal number of resets, based on variables orthogonal to the fitted variable (Y-coordinates for the spatial offset of resets and angle for reset time).
+The median for each segment of data is shown in Fig 5B,C.
+We fit a gamma distribution to each segment, then use the data points outside the segment to get a log likelihood as a measure of generalization and sum them (spatial: XXXXX, temporal: XXXXX).
+We use this to calculate AICs and a relative likelihood (spatial: XXXXX, temporal: XXXXX, _p_<XXXXX) which shows that a gamma distribution fit to reset times in one segment of the data generalizes better to the rest of the data, as compared to a gamma distribution fit to the spatial offset of resets.
 
@@
 In our experiment, we did not control for eye-movements, so we can't directly investigate whether eye movements accompanied the “return” resets that participants reported.
-However, the random distribution of resets throughout time are in line with a mechanism based on distracted attention, which may involve eye-movement planning and execution.
+However, the distribution of resets throughout time are largely in line with a mechanism based on random distractions of attention, which may involve eye-movement planning and execution.
 The rate of resets that we found here, ~0.75 / _s_, is too infrequent for microsaccades (~3 / _s_) so the distractions may be more significant than these small eye movements.
@@
 @McGill1965; @Wolfe2002] do model a random occurrence of events (rate parameter) where only some fraction of these trigger an action (shape parameter).
-This could be consistent with the more frequent microsaccades happening randomly at a faster rate where only some of these would trigger a reset.
-In other words, the gamma distribution is not only a good fit to our reset data but it may reveal the processes underlying the resets.
+In other words, the good fit of a gamma distribution may reveal some properties of the process underlying resets of the double-drift illusion.
+While the process giving rise to resets is likely memoryless, a process giving rise to a gamma distribution (with shape > 1) is not, it does capture two ideas: 1) resets are triggered by random events (such as distractions of attention or microsaccades), and 2) only some of these actually lead to a reset.
+An ex-Gaussian distribution does not fit the data as well as a gamma distribution, but could be a more realistic alternative.
+However, which process actually leads to resets remains to be seen in future work.
 

### 2022-01-18 Manuscript.Rmd
revisions from Patrick

@@
 - Denise Y.P. Henriques 1
-- Patrick Cavanagh 1,2,3
+- Patrick Cavanagh 1,2
 output:
@@
 _2: Department of Psychology, Glendon College, Toronto, ON M4N3M6, Canada_\
-_3: Department of Psychological and Brain Sciences, Dartmouth College, Hanover, NJ 03755, USA_\
 
@@
 
-The research was support by funding from NSERC Canada (PC and DYPH), as well as by the Department of Psychological and Brain Science, Dartmouth College (PC).
+The research was support by funding from NSERC Canada (PC and DYPH).
 
@@
 A more sophisticated model allows predicting perceived speed as well [@Johnston2013], but this model predicts perceived direction and speed for a mixture of first order motions, and it is unclear if this works for a combination of first and second order motion (what we call internal and external speed here).
-In the vector average model we use here, the perceived direction is a simple vector combination of the external ($V_e$) and internal ($V_i$) motions and a constant $K$.
+In the vector average model we use here, the perceived direction
+
+To determine if hand tracking produces an illusion similar to that seen in other studies, we compare our results to those reported by Guernsey and Biard [-@Guernsey2012], Cavanagh and Tse [-@Cavanagh2019] and Heller et al.
+[-@Heller2021] who all used similar displays.
+We also compare the illusory direction of the heading against the predictions of a simple vector combination model [@Tse2006; @Cavanagh2019; @Heller2021].
+There are alternative models that could be considered, one from Kwon et al.
+[-@Kwon2015] and another from Johnston and Scarfe [-@Johnston2013].
+Kwon et al.’s [-@Kwon2015] Kalman filter tracking model assumes a saturation of the position shifts by 200 ms, something we do not observe here.
+Johnson and Scarfe’s [-@Johnston2013] model used a harmonic average to combine vectors, a process that gives more weight to vectors with lower magnitudes.
+This model was used to estimate the perceived direction of an array of stationary Gabors with various internal motions but is less suited to the double-drift stimulus that has only two vectors, one of which is a second-order motion (the motion of the Gabor itself).
+In the simple vector combination model we use here, the perceived direction is a vector combination of the external ($V_e$) and internal ($V_i$) motions weighted by a constant $K$.
 That is, the deviation from the physical direction is given by:
@@
 
-The data clearly show the illusion, but does tracking decrease the strength of the illusion compared to simply observing it?
-In Figure 2D, we compare mean illusion strength to that reported previously [@Heller2021].
-As can be seen the averages from our data here (red circles) coincide fairly well with the illusion strength predicted by the model fitted to all their data (Eq.
-1, black line).
-If we fit the vector combination model, we get $K=0.738$ which is very close to $K=0.74$ reported by Heller et al.
+The illusion size we find with tracking appears to be in the range of that reported in other article for similar speeds and eccentricity.
+For example, Heller et al.
+(2021) reported an illusion of 30° for internal and external speeds of 7.2 and 7.2 dva/s compared to the 27.8° here (average of left and right 3Hz internal motion) with internal and external speeds both 7 dva/s.
+The data clearly show the illusion, and we can compare the mean illusion strengths here to those of Heller et al.
+[-@Heller2021] in more detail using their vector combination model.
+The averages from our data here (Fig 2D: red circles) do coincide fairly well with the illusion strength predicted by the model fitted to their data.
+The best fit value of the vector combination model (Eq.
+1 above; Fig 2D: black line) to our data $K=0.738$ is very close to $K=0.74$ reported by Heller et al.
 [-@Heller2021].
-That is; tracking the illusion while observing it does not change the strength of the illusion.
+We can conclude that tracking the illusion while observing it does not appear to change its strength.
 
@@
 
-The X coordinates of the detected reset points are used as the spatial offset of the reset.
+The X coordinates of the detected reset points give the deviation from the vertical path and are used as the spatial offset of the reset.
 We also need the time of the reset.
-Using the Y coordinates of the drawn trajectory (divided by speed) would assume that the real and perceived vertical position of the gabor are the same, which would imply that for very strong illusions the perceived speed can be many times the real speed.
-This seems unrealistic.
-As described above, it is unclear if the overall perceived speed of a mixture of many first order motions [@Johnston2013] applies to the double-drift illusion and if so, whether or not that can be used to calculate perceived time from path length.
-Instead, we chose a simplification by using the radius (R; distance from origin to reset, divided by speed) as the reset time.
-However, while useful for the current project, it remains to be tested how perceived speed and perceived path length are related to each other and to stimulus properties in the double-drift illusion.
+The gabor moves at a uniform speed along the Y axis so distance along Y denotes the elapsed time for the physical path.
+However, the illusory path of the gabor does not have to stay in alignment with its physical location on the Y axis.
+If it did, its speed and path length would have to increase for larger illusion angles: for example, 41% faster and farther for a 45° illusion, 73% faster and farther for a 60° illusion.
+Observers have not reported these increases in speed or path length compared to the control with no internal motion.
+Moreover, in a recent unpublished study we directly measured the perceived path length of the illusion.
+The path length was always fixed proportion of the physical path (around 70%) for all illusion strengths, Gabor speeds and physical path lengths.
+This means that the perceived speed must be a fixed proportion of the physical speed as well and therefore, in the tracings, it is the distance along the traced path (the radius in our diagrams) that tells us the time since the trial began — not the distance along the physical path (the Y axis).
+For completeness, we also analyze the distributions along the Y axis.
 
-To facilitate modeling, the horizontal and vertical (straight ahead) coordinates of the reset points are kept isometric by giving them in centimeters.
+To facilitate modeling, the horizontal (illusory deviation) and vertical (straight ahead) coordinates of the reset points are kept isometric by giving them in centimeters.
 The distance from origin can then be converted to time as 1 _cm_ = 4/13.5 _s_ in the 4 _s_ condition and 1 _cm_ = 3/13.5 _s_ in the 3 _s_ condition.
-For each model using a probability density function to estimate model likelihood, the variable that is used is divided by its median value such that each of their probability densities are on a comparable scale.
+A probability density function is used to estimate each model’s likelihood and the variable that is used (X or T) is divided by its median value so that each of their probability densities are on a comparable scale.
 
@@
 **B,C:** Color illustrates the gamma probability distributions, where white is zero probability, and the darkest red or blue is the maximum probability density.
-This is a relatively scale-free unit, where the total surface of the full probability density function (along a line orthogonal to the predictor variable) is 1.
-**B:** Gamma distribution on time (log likelihood: -605.6).
+**B:** Gamma distribution on time (_AIC_ = 1215.12).
 Dark and light red lines: denote the median reset time for 4 _s_ and 3 _s_ passes in angular bins with an equal number of resets.
-**C:** Gamma distribution on spatial offset (log likelihood: -632.6).
+**C:** Gamma distribution on spatial offset (_AIC_ = 1269.18).
 Blue lines indicate median reset offsets in Y-coordinate bins with an equal number of resets."}
@@
 
-For the space and time limit with normal distributions (means in Fig 5A, blue: spatial, red: time), the time-based distribution fits better for the group data (normal space AIC=2,834.83, normal time _AIC_=1,677.98; _p_<.001) as well as all individual participants, and the same holds within gamma distribution models of reset models (gamma space AIC=1,269.18, gamma time AIC=1,215.12; _p_<.001).
+For the space and time limit with normal distributions (means in Fig 5A, blue: spatial, red: time), the time-based distribution fits better for the group data (normal space _AIC_=2,834.83, normal time _AIC_=1,677.98; _p_<.001) as well as all individual participants, and the same holds within gamma distribution models of reset models (gamma space _AIC_=1,269.18, gamma time _AIC_=1,215.12; _p_<.001).
 Furthermore, the gamma-based distribution on time of resets performed better than the normal distribution of reset time in the pooled data ( _p_<.001 ), and this holds for most participants (6/9) as well.
@@
 While the process giving rise to resets is likely memoryless, a process giving rise to a gamma distribution (with shape > 1) is not, it does capture two ideas: 1) resets are triggered by random events (such as distractions of attention or microsaccades), and 2) only some of these actually lead to a reset.
-An ex-Gaussian distribution does not fit the data as well as a gamma distribution, but could be a more realistic alternative.
-However, which process actually leads to resets remains to be seen in future work.
+An alternative to the gamma distribution, the ex-Gaussian (ref), could be a more realistic alternative but does not fit the data as well.
+However, the process that actually leads to resets remains to be seen in future work.
 
@@
 [-@Liu2021] modeled this with a Poisson process having a constant chance of reset of ~0.7 / _s_ (Fig 6B, orange line).
-In comparison, in our experiment, the gamma distribution fit to the data of explicit resets (rate=$2.81$, shape=$3.74$), gave a ~0.75 / _s_ chance of a reset, a remarkaly similar estimate.
-That's why, we also used the parameters of our gamma distribution fit from data to generate the PSEs that should be found in Liu et al.'s experiment with no additional fitting (Fig 6B, dashed orange line).
+In comparison, in our experiment, the gamma distribution fit to the data of explicit resets (rate=$2.81$, shape=$3.74$), gave a ~0.75 / _s_ chance of a reset, a remarkably similar estimate.
+We then used the parameters of our gamma distribution fit to our data to generate the PSEs that should be found in Liu et al.'s experiment with no additional fitting (Fig 6B, dashed orange line).
 We see that the two findings, based on very different approaches, are mostly in agreement with each other.

### 2022-01-20 Manuscript.Rmd
a few more edits to the text

@@
 1, left) generates extreme misjudgments of the moving target’s location which may deviate by as much as several degrees of visual angle from its true location.
-Surprisingly, this extraordinary perceptual illusion does not affect eye movements to the gabor: immediate saccades to these targets are determined by their physical not their perceived location [@Lisi2015].
+Surprisingly, this extraordinary perceptual illusion does not affect eye movements to the gabor: immediate saccades to these targets are determined by their physical, not their perceived location [@Lisi2015].
 In contrast, delayed eye movements and pointing go to the perceived position [@Lisi2017; @Massendari2018; @Ueda2018].
@@
 
-The stimuli were phase-changing gabor patches with a sigma (sd) of 0.43 _cm_ or ~0.59 _dva_, spatial frequency of 0.58 _c/cm_ or ~0.42 _c/dva_, vertical orientation, and Michelson contrast of 98.7%.
+The stimuli were phase-changing gabor patches with a sigma ($\sigma$ or sd) of 0.43 _cm_ or ~0.59 _dva_, spatial frequency of 0.58 _c/cm_ or ~0.42 _c/dva_, vertical orientation, and Michelson contrast of 98.7%.
 The background luminance was 18.46 $cd/m^2$ and the mean luminance of the gabor deviated from the background by less than 1%.
@@
 
-To determine if hand tracking produces an illusion similar to that seen in other studies, we compare our results to those reported by Guernsey and Biard [-@Guernsey2012], Cavanagh and Tse [-@Cavanagh2019] and Heller et al.
+To determine if hand tracking produces an illusion similar to that seen in other studies, we compare our results to those reported by Gurnsey and Biard [-@Gurnsey2012], Cavanagh and Tse [-@Cavanagh2019] and Heller et al.
 [-@Heller2021] who all used similar displays.

### 2022-01-20 Manuscript.Rmd
added stats for the binned fits, and references for the ex-Gaussian distribution

@@
 The median for each segment of data is shown in Fig 5B,C.
-We fit a gamma distribution to each segment, then use the data points outside the segment to get a log likelihood as a measure of generalization and sum them (spatial: XXXXX, temporal: XXXXX).
-We use this to calculate AICs and a relative likelihood (spatial: XXXXX, temporal: XXXXX, _p_<XXXXX) which shows that a gamma distribution fit to reset times in one segment of the data generalizes better to the rest of the data, as compared to a gamma distribution fit to the spatial offset of resets.
+We fit a gamma distribution to each segment, then use the data points outside the segment to get a log likelihood as a measure of generalization and sum them (spatial: -3087.9, temporal: -1806.3).
+We use this to calculate AICs and a relative likelihood (spatial: 6191.7, temporal: 3628.6, _p_<.001) which shows that a gamma distribution fit to reset times in one segment of the data generalizes better to the rest of the data, as compared to a gamma distribution fit to the spatial offset of resets.
 
@@
 While the process giving rise to resets is likely memoryless, a process giving rise to a gamma distribution (with shape > 1) is not, it does capture two ideas: 1) resets are triggered by random events (such as distractions of attention or microsaccades), and 2) only some of these actually lead to a reset.
-An alternative to the gamma distribution, the ex-Gaussian (ref), could be a more realistic alternative but does not fit the data as well.
+An alternative to the gamma distribution, the ex-Gaussian [e.g.
+@Matzke2009; @Guy2020], could be a more realistic alternative but does not fit the data as well.
 However, the process that actually leads to resets remains to be seen in future work.

### 2022-01-21 Manuscript.Rmd
final round of manuscript edits before 3rd submission

@@
 
-We compare the illusory direction of the heading against the predictions of a simple vector combination model of perceived direction from Tse and Hsieh [-@Tse2006], Cavanagh and Tse [-@Cavanagh2019] and Heller et al.
-[-@Heller2021].
-A more sophisticated model allows predicting perceived speed as well [@Johnston2013], but this model predicts perceived direction and speed for a mixture of first order motions, and it is unclear if this works for a combination of first and second order motion (what we call internal and external speed here).
-In the vector average model we use here, the perceived direction
-
-To determine if hand tracking produces an illusion similar to that seen in other studies, we compare our results to those reported by Gurnsey and Biard [-@Gurnsey2012], Cavanagh and Tse [-@Cavanagh2019] and Heller et al.
-[-@Heller2021] who all used similar displays.
+To determine if hand tracking produces an illusion similar to that seen in other studies, we first compare our results to those reported by Heller et al.
+[-@Heller2021] who used similar displays.
 We also compare the illusory direction of the heading against the predictions of a simple vector combination model [@Tse2006; @Cavanagh2019; @Heller2021].
@@
 The path length was always fixed proportion of the physical path (around 70%) for all illusion strengths, Gabor speeds and physical path lengths.
-This means that the perceived speed must be a fixed proportion of the physical speed as well and therefore, in the tracings, it is the distance along the traced path (the radius in our diagrams) that tells us the time since the trial began — not the distance along the physical path (the Y axis).
-For completeness, we also analyze the distributions along the Y axis.
+This means that the perceived speed along the illusory path must be a fixed proportion of the physical speed as well and therefore, in the tracings, it is the distance along the traced path (the radius in our diagrams) that tells us the time since the trial began — not the distance along the physical path (the Y axis).
 
@@
 
-While the marginal distribution plot in Fig 5A qualitatively appear an equally good fit to the data, the better fit for the distribution on the time of resets could be explained be a better generalization throughout the data set.
-We test this by dividing the data into segments, each with an equal number of resets, based on variables orthogonal to the fitted variable (Y-coordinates for the spatial offset of resets and angle for reset time).
+While the marginal distribution plot in Fig 5A qualitatively appear an equally good fit to the data, the fit for the distribution on the time of resets maintains a better fit throughout the data set.
+We demonstrate this by dividing the data into segments, each with an equal number of resets, based on variables orthogonal to the fitted variable (Y-coordinates for the spatial offset of resets and angle for reset time).
 The median for each segment of data is shown in Fig 5B,C.
-We fit a gamma distribution to each segment, then use the data points outside the segment to get a log likelihood as a measure of generalization and sum them (spatial: -3087.9, temporal: -1806.3).
+We fit a gamma distribution to each segment, then use the data points outside the segment to get a log likelihood as a measure of generalization and sum them.
 We use this to calculate AICs and a relative likelihood (spatial: 6191.7, temporal: 3628.6, _p_<.001) which shows that a gamma distribution fit to reset times in one segment of the data generalizes better to the rest of the data, as compared to a gamma distribution fit to the spatial offset of resets.
@@
 In our experiment, we did not control for eye-movements, so we can't directly investigate whether eye movements accompanied the “return” resets that participants reported.
-However, the distribution of resets throughout time are largely in line with a mechanism based on random distractions of attention, which may involve eye-movement planning and execution.
+However, the distribution of resets throughout time are roughly in line with a mechanism based on random distractions of attention, which may involve eye-movement planning and execution.
 The rate of resets that we found here, ~0.75 / _s_, is too infrequent for microsaccades (~3 / _s_) so the distractions may be more significant than these small eye movements.
@@
 In other words, the good fit of a gamma distribution may reveal some properties of the process underlying resets of the double-drift illusion.
-While the process giving rise to resets is likely memoryless, a process giving rise to a gamma distribution (with shape > 1) is not, it does capture two ideas: 1) resets are triggered by random events (such as distractions of attention or microsaccades), and 2) only some of these actually lead to a reset.
+While the process giving rise to resets is likely memoryless, a process giving rise to a gamma distribution (with shape > 1) is not.
+Nevertheless, it does capture two ideas: 1) resets are triggered by random events (such as distractions of attention or microsaccades), and 2) only some of these actually lead to a reset.
 An alternative to the gamma distribution, the ex-Gaussian [e.g.
 @Matzke2009; @Guy2020], could be a more realistic alternative but does not fit the data as well.
-However, the process that actually leads to resets remains to be seen in future work.
+In any case, the process that actually leads to resets remains to be seen in future work.
 

### 2022-01-22 Manuscript.Rmd
changed osf repo to registered version of repo

@@
 
-All analyses were done in R 3.6.1 [@Rproject]. All data, as well as scripts for the experiments and analyses are available on OSF: [https://osf.io/72ndu/](https://osf.io/72ndu/).
+All analyses were done in R 3.6.1 [@Rproject]. All data, as well as scripts for the experiments and analyses are available on OSF: [https://osf.io/q65as/](https://osf.io/q65as/).
 

### 2022-02-16 Manuscript.Rmd
changed to OSF link to the (final?) registration of the project
tweaked figure 1

@@
 library(magick)
-Fig1 <- image_read_pdf('doc/methods_fig_1.pdf')
+Fig1 <- image_read_pdf('doc/methods_fig_1_v2.pdf')
 par(mai=c(0,0,0,0))
@@
 
-All analyses were done in R 3.6.1 [@Rproject]. All data, as well as scripts for the experiments and analyses are available on OSF: [https://osf.io/q65as/](https://osf.io/q65as/).
+All analyses were done in R 3.6.1 [@Rproject]. All data, as well as scripts for the experiments and analyses are available on OSF: [https://osf.io/8zugx/](https://osf.io/8zugx/).
 

### 2022-02-16 R/common.R

@@ -820,7 +820,7 @@ density2D <- function(x, y, bw=1, weights=NULL, n=100, from=NULL, to=NULL, cut=3
 library('rsvg')
 library(magick)
 
-replotMethodsFigures <- function(figno=1, target='inline') {
+replotMethodsFigures <- function(figno=1, target='inline', version=NULL) {
   
   if (figno == 1) {
     width <- 6
@@ -840,7 +840,11 @@ replotMethodsFigures <- function(figno=1, target='inline') {
     pdf(file=sprintf('%s.pdf',outputfile), width=width, height=height)
   }
   
-  Fig <- image_read_pdf(sprintf('doc/methods_fig_%d.pdf',sourceno))
+  if (is.null(version)) {
+    Fig <- image_read_pdf(sprintf('doc/methods_fig_%d.pdf',sourceno))
+  } else {
+    Fig <- image_read_pdf(sprintf('doc/methods_fig_%d_v%d.pdf',sourceno,version))
+  }
   par(mai=c(0,0,0,0))
   plot(Fig)
   
